---
title: "Contrast, crossnobis, and linear RSA as one geometry"
description: >
  An executable theorem, independent oracle, estimator claim table, and external parity receipt for Crossform's fixed bilinear geometry.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrast, crossnobis, and linear RSA as one geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)

pair_matrix <- function(q) t(utils::combn(seq_len(q), 2L))

oracle_geometry <- function(blocks, gamma, metric) {
  answer <- matrix(0, nrow(blocks[[1L]]), nrow(blocks[[1L]]))
  for (r in seq_along(blocks)) {
    for (s in seq_along(blocks)) {
      answer <- answer + gamma[r, s] *
        blocks[[r]] %*% metric %*% t(blocks[[s]])
    }
  }
  answer
}

oracle_energy <- function(operator, blocks, gamma, metric) {
  total <- 0
  for (r in seq_along(blocks)) {
    for (s in seq_along(blocks)) {
      for (i in seq_len(nrow(operator))) {
        for (j in seq_len(ncol(operator))) {
          for (a in seq_len(nrow(metric))) {
            for (b in seq_len(ncol(metric))) {
              total <- total + gamma[r, s] * operator[i, j] *
                blocks[[r]][i, a] * metric[a, b] * blocks[[s]][j, b]
            }
          }
        }
      }
    }
  }
  total
}

rdm_from_geometry <- function(geometry) {
  pairs <- pair_matrix(nrow(geometry))
  vapply(seq_len(nrow(pairs)), function(k) {
    u <- numeric(nrow(geometry))
    u[pairs[k, ]] <- c(1, -1)
    drop(crossprod(u, geometry %*% u))
  }, numeric(1))
}

rdm_matrix <- function(values, labels) {
  pairs <- pair_matrix(length(labels))
  answer <- matrix(0, length(labels), length(labels),
                   dimnames = list(labels, labels))
  answer[cbind(pairs[, 1L], pairs[, 2L])] <- values
  answer + t(answer)
}
```

This article makes one bounded claim: fixed contrast energy, pairwise
crossvalidated squared distance, crossnobis under a declared precision, and a
fixed linear RSA coefficient are linear queries of the same geometry. It does
not say that every method called RSA, decoding, or multivariate inference is
implemented by one formula.

The normative derivation and exact independent implementation are
[`design/common-geometry-equivalence.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/common-geometry-equivalence.md)
and
[`design/oracles/common-geometry-equivalence.R`](https://github.com/bbuchsbaum/crossform/blob/main/design/oracles/common-geometry-equivalence.R).
The calculations below reproduce their central identities without sourcing
package internals or the oracle file.

## The theorem

Let (B_r) be the effects-by-features matrix estimated in partition (r),
(H) a fixed effect-side operator, (K) a fixed neural metric, and
(\Gamma) the declared partition-pair weights. Define

\[
G_K = \sum_{r,s} \Gamma_{rs} B_r K B_s^\top,
\qquad
\mathcal E(H,K) = \langle H,G_K\rangle_F.
\]

Linearity of trace gives the equivalent direct expression

\[
\mathcal E(H,K)
= \sum_{r,s}\Gamma_{rs}
  \operatorname{tr}(H^\top B_r K B_s^\top).
\]

Three substitutions give the familiar estimators.

- A contrast (c) uses (H=cc^\top), so the output is (c^\top G_Kc).
- An RDM entry for effects (i,j) uses (u_{ij}=e_i-e_j) and
  (H=u_{ij}u_{ij}^\top). With a declared noise precision for (K), this is
  crossnobis.
- A fixed OLS RSA coefficient uses the row (a_\ell^\top) of
  ((X^\top X)^{-1}X^\top) and the RDM adjoint
  (H=\mathcal D^*(a_\ell)=\sum_{i<j}a_{\ell,ij}u_{ij}u_{ij}^\top).

The equalities concern the estimand. Crossnobis additionally needs independent
evaluated partitions and an honestly fixed or cross-fitted precision. RSA
additionally needs a fixed, full-rank design and a linear objective.

## A hand-sized exact case

The fixture has two partitions, three effects, two features, off-diagonal
partition weights, and a non-identity metric. Every displayed value is
computed in this article.

```{r hand-fixture}
effect_labels <- c("a", "b", "c")
blocks <- list(
  run1 = matrix(c(1, 0, 0, 1, 1, 1), 3L, 2L, byrow = TRUE,
                dimnames = list(effect_labels, NULL)),
  run2 = matrix(c(2, 0, 0, 2, 1, 3), 3L, 2L, byrow = TRUE,
                dimnames = list(effect_labels, NULL))
)
gamma <- matrix(c(0, 0.5, 0.5, 0), 2L, 2L,
                dimnames = list(names(blocks), names(blocks)))
metric <- diag(c(2, 1))
geometry <- oracle_geometry(blocks, gamma, metric)
knitr::kable(geometry, digits = 3, caption = "Generated common geometry")
```

The pairwise squared distances are generated from (u^\top G_Ku). The direct
court below independently expands every partition, effect, and feature index.

```{r hand-rdm}
pairs <- pair_matrix(length(effect_labels))
pair_labels <- apply(pairs, 1L, function(x) {
  paste(effect_labels[x], collapse = " - ")
})
rdm_values <- rdm_from_geometry(geometry)
direct_values <- vapply(seq_len(nrow(pairs)), function(k) {
  u <- numeric(length(effect_labels))
  u[pairs[k, ]] <- c(1, -1)
  oracle_energy(tcrossprod(u), blocks, gamma, metric)
}, numeric(1))
hand_rdm <- data.frame(
  pair = pair_labels,
  from_geometry = rdm_values,
  direct_index_oracle = direct_values,
  absolute_error = abs(rdm_values - direct_values)
)
knitr::kable(hand_rdm, digits = 15)
```

Now fit a fixed linear RSA model to the generated RDM. The same coefficients
are recovered by compiling each coefficient row back into an (H) operator.

```{r hand-rsa}
rsa_design <- cbind(`(Intercept)` = 1, category = c(1, 0, 1))
coefficient_map <- solve(crossprod(rsa_design), t(rsa_design))
rsa_from_rdm <- drop(coefficient_map %*% rdm_values)
rsa_from_adjoint <- vapply(seq_len(nrow(coefficient_map)), function(term) {
  operator <- matrix(0, length(effect_labels), length(effect_labels))
  for (k in seq_len(nrow(pairs))) {
    u <- numeric(length(effect_labels))
    u[pairs[k, ]] <- c(1, -1)
    operator <- operator + coefficient_map[term, k] * tcrossprod(u)
  }
  oracle_energy(operator, blocks, gamma, metric)
}, numeric(1))
hand_rsa <- data.frame(
  term = colnames(rsa_design),
  from_materialized_rdm = rsa_from_rdm,
  from_compiled_adjoint = rsa_from_adjoint,
  absolute_error = abs(rsa_from_rdm - rsa_from_adjoint)
)
knitr::kable(hand_rsa, digits = 15)
```

```{r hand-production-check, include=FALSE}
domain <- abstract_domain(ncol(blocks[[1L]]), id = "common-geometry-article")
rel <- relation(blocks, domain = domain)
plan <- plan_geometry(
  rel,
  compile_frame(whole_brain(normalization = "none"), domain),
  cross_partitions(rel, independence = "independent"),
  metric = noise_precision(metric, domain,
    provenance = list(source = "fixed article metric"))
)
contrast <- stats::setNames(c(1, -1, 0), effect_labels)
model_rdm <- rdm_matrix(c(1, 0, 1), effect_labels)
production_contrast <- crossnobis(plan, contrast)$values
production_rdm <- as.numeric(rdm(plan)$values)
production_rsa <- as.numeric(rsa(plan, models = list(category = model_rdm))$coefficients)
stopifnot(
  max(abs(production_contrast - direct_values[[1L]])) < 1e-12,
  max(abs(production_rdm - rdm_values)) < 1e-12,
  max(abs(production_rsa - rsa_from_rdm)) < 1e-12,
  max(hand_rdm$absolute_error) < 1e-12,
  max(hand_rsa$absolute_error) < 1e-12
)
```

The hidden assertions above run the public `crossnobis()`, `rdm()`, and
`rsa()` paths against the direct oracle. A build stops if any mapping differs.

## Randomized production-to-oracle evidence

The hand case is exact but small. This bounded generated court changes blocks,
SPD metrics, contrasts, and RSA models across replayable seeds. The oracle is
the explicit matrix definition above; the candidate is the public package
path.

```{r randomized-court}
run_case <- function(seed) {
  set.seed(seed)
  q <- 4L
  d <- 3L
  runs <- paste0("run", seq_len(3L))
  effects <- paste0("e", seq_len(q))
  generated_blocks <- stats::setNames(lapply(runs, function(run) {
    matrix(stats::rnorm(q * d), q, d, dimnames = list(effects, NULL))
  }), runs)
  factor <- matrix(stats::rnorm(d * d), d, d)
  generated_metric <- crossprod(factor) + diag(0.25, d)
  generated_gamma <- matrix(1 / 6, 3L, 3L,
                            dimnames = list(runs, runs))
  diag(generated_gamma) <- 0
  generated_geometry <- oracle_geometry(
    generated_blocks, generated_gamma, generated_metric
  )
  generated_rdm <- rdm_from_geometry(generated_geometry)
  generated_contrast <- stats::setNames(stats::rnorm(q), effects)
  generated_contrast <- generated_contrast - mean(generated_contrast)

  model_vectors <- cbind(
    model1 = stats::rnorm(length(generated_rdm)),
    model2 = seq_along(generated_rdm)
  )
  design <- cbind(`(Intercept)` = 1, model_vectors)
  expected_rsa <- drop(solve(crossprod(design), t(design)) %*% generated_rdm)
  models <- lapply(seq_len(ncol(model_vectors)), function(j) {
    rdm_matrix(model_vectors[, j], effects)
  })
  names(models) <- colnames(model_vectors)

  generated_domain <- abstract_domain(d, id = paste0("article-seed-", seed))
  generated_relation <- relation(generated_blocks, domain = generated_domain)
  generated_plan <- plan_geometry(
    generated_relation,
    compile_frame(whole_brain("none"), generated_domain),
    cross_partitions(generated_relation, independence = "independent"),
    metric = noise_precision(generated_metric, generated_domain,
      provenance = list(source = "generated fixed metric"))
  )

  expected_contrast <- drop(crossprod(
    generated_contrast,
    generated_geometry %*% generated_contrast
  ))
  observed <- c(
    contrast = as.numeric(crossnobis(
      generated_plan, generated_contrast
    )$values),
    as.numeric(rdm(generated_plan)$values),
    as.numeric(rsa(generated_plan, models = models)$coefficients)
  )
  expected <- c(expected_contrast, generated_rdm, expected_rsa)
  max(abs(observed - expected))
}

court_seeds <- 1201:1206
court_errors <- vapply(court_seeds, run_case, numeric(1))
randomized_summary <- data.frame(
  cases = length(court_seeds),
  first_seed = min(court_seeds),
  last_seed = max(court_seeds),
  worst_absolute_error = max(court_errors),
  declared_tolerance = 5e-10,
  passes = max(court_errors) <= 5e-10
)
knitr::kable(randomized_summary, digits = 15)
stopifnot(all(randomized_summary$passes))
```

The package test court extends this article with rank-deficient PSD metrics,
unequal partition weights, near-singular metrics, nonfinite refusals,
permutation equivariance, scaling laws, and a 50-seed deep mode. See the
[`test-common-geometry-*` files](https://github.com/bbuchsbaum/crossform/tree/main/tests/testthat)
and their independent oracle link above.

## External differential evidence

The external court uses a deterministic six-condition, four-run fixture, a
fixed non-spherical pooled-residual precision, local support normalization,
and pinned Python packages. It compares the same RDM vector and fixed OLS RSA
coefficients to `rsatoolbox`. The table is read from the generated package
certification artifact; no result is typed into this article.

```{r external-parity}
parity_candidates <- c(
  file.path("..", "inst", "extdata", "certification",
            "common-geometry-external-parity.csv"),
  file.path("inst", "extdata", "certification",
            "common-geometry-external-parity.csv"),
  system.file("extdata", "certification",
              "common-geometry-external-parity.csv", package = "crossform")
)
parity_path <- parity_candidates[nzchar(parity_candidates) &
                                   file.exists(parity_candidates)][1L]
if (is.na(parity_path)) stop("External parity certification artifact missing.")
external_parity <- utils::read.csv(parity_path, stringsAsFactors = FALSE)
external_summary <- external_parity[
  external_parity$quantity %in%
    c("crossnobis_rdm", "linear_rsa_coefficients") &
    !duplicated(external_parity$quantity),
  c("rsatoolbox_version", "quantity", "comparator", "n_values",
    "max_abs_diff", "tolerance", "passes")
]
knitr::kable(external_summary, digits = 15)
stopifnot(all(external_parity$passes))
```

The full fixture, reproduction instructions, convention mapping, and
source/artifact manifest live in
[`exemplars/rsatoolbox-parity`](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/rsatoolbox-parity).
The manifest makes an edited producer, claim, environment lock, or result a
test failure until the external evidence is regenerated or explicitly
revalidated.

## Estimator claim table

This generated table separates an algebraic mapping from the extra assumptions
that license a familiar estimator name.

```{r claim-table, results='asis'}
claim_table <- data.frame(
  estimator = c(
    "Fixed bilinear contrast energy",
    "Pairwise crossvalidated squared-distance RDM",
    "Crossnobis RDM",
    "Fixed multiple-regression RSA coefficient",
    "Correlation/cosine RDM or RSA",
    "Spearman/Kendall RSA",
    "Adaptive classifier, kernel, or embedding"
  ),
  H = c(
    "`c c^T`",
    "`u_ij u_ij^T`",
    "`u_ij u_ij^T`",
    "`D*(a_l)`",
    "No fixed `H` in general",
    "No fixed `H`",
    "No fixed `H`"
  ),
  K = c(
    "Declared fixed metric",
    "Declared fixed metric",
    "Fixed or honestly cross-fitted noise precision",
    "Same declared metric as the source RDM",
    "Data-dependent norm",
    "Any, followed by ranks",
    "Learned from evaluated data"
  ),
  Gamma = c(
    "Declared partition weights",
    "Independent cross-partition weights",
    "Independent cross-partition weights",
    "Inherited unchanged from the source geometry",
    "Not sufficient to linearize",
    "Not sufficient to linearize",
    "Data-adaptive"
  ),
  preprocessing_assumptions = c(
    "Fixed contrast; centering only when translation invariance is claimed",
    "Zero-sum pair differences; declared normalization",
    "Same plus fixed/cross-fitted precision provenance",
    "Fixed full-rank OLS design, pair order, intercept, and model scaling",
    "Observed-pattern normalization is nonlinear",
    "Ranking is nonlinear and tie policy matters",
    "Training rule is part of a different estimand"
  ),
  output = c(
    "Scalar contrast energy",
    "One signed squared-distance estimate per pair",
    "Noise-normalized signed squared-distance estimate",
    "One scalar coefficient per fixed model term",
    "Normalized dissimilarity/similarity",
    "Rank association",
    "Prediction or learned representation"
  ),
  evidence = c(
    "Algebra + exact/randomized/adversarial courts",
    "Algebra + exact/randomized/adversarial courts",
    "Same courts + pinned external parity",
    "Adjoint proof + same courts + pinned external parity",
    "Explicit boundary/refusal",
    "Explicit boundary/refusal",
    "Explicit boundary only"
  ),
  status = c(
    rep("Equivalent special case", 4L),
    rep("Outside this theorem", 3L)
  ),
  stringsAsFactors = FALSE
)
knitr::kable(claim_table, escape = FALSE)
```

“Equivalent special case” means unification of the named fixed estimand under
the stated assumptions. It does not mean that Crossform implements every
estimator family, preprocessing convention, uncertainty procedure, null law,
or population generalization target associated with that name.

## Limits and evidence status

The theorem is algebraic. The hand, randomized, adversarial, and external
courts are computational evidence. None supplies population coverage,
calibrates a null distribution, or proves validity after data-adaptive model or
metric selection. External parity validates only the mapped fixed-linear case,
not all of `rsatoolbox` and not all RSA.

For the broader package evidence boundary, see the
[`certification report`](https://github.com/bbuchsbaum/crossform/blob/main/design/certification-report.md).
For the vocabulary separating theorem, numerical verification, statistical
validation, empirical demonstration, and interpretation, see the
[`unification contract`](https://github.com/bbuchsbaum/crossform/blob/main/design/unification-contract.md).
