---
title: "Conservative frames: attribution, not detection"
description: >
  Partition a global contrast budget across space, compare coherent shares across scales,
  and distinguish signed attribution from latent descriptive summaries.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conservative frames: attribution, not detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

How much of a contrast's total evidence is assigned to each territory? A
**conservative frame** answers this question by partitioning one global total
among its spatial measurements. Its map can be added across territories.

This guide assumes the `relation → frame → plan → contrast_energy()` workflow
from [the introduction](introduction.html). It uses the same generated fMRI
example to compare two normalizations:

| Question | Normalization | Meaning of one node |
|---|---|---|
| Where is evidence locally strong? | `"local"`: normalize rows | Mean evidence density in the node's support |
| How is the global total distributed? | `"conservative"`: normalize columns | Evidence assigned to the node from a fixed global budget |

A frame has one row per measurement and one column per neural feature. Column
normalization allocates each feature across the nodes that contain it. Start
with the conservation check and territory ledger; the later sections explain
scale, signed estimates, and metric choices. The outputs show the numerical identities; quiet assertions check their
tolerances during rendering. Their derivations are in the
[geometry contract](https://github.com/bbuchsbaum/crossform/blob/main/design/conservative-geometry-contract.md).
References marked § below point to that contract.

```{r example}
example <- example_fmri_effects()
domain <- example$domain
pairing <- cross_partitions(
  example$fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
energy <- function(frame) {
  contrast_energy(
    plan_geometry(example$fit$relation, frame, pairing), example$contrast
  )
}
c(features = domain$n_features, conditions = length(example$contrast))
```

The generated fixture plants two blocks of 19 voxels at opposite ends of an
8 × 7 × 5 volume, at the same per-voxel amplitude, never overlapping. They
differ only in sign structure: in the **pattern** block the sign alternates
between neighbouring voxels, so a neighbourhood average nearly cancels; in the
**mean** block every voxel moves the same way. Same effect size, two different
effect *geometries*. Section 3 compares their coherent shares across neighbourhood sizes.
These are generated data, not evidence about a brain; the
[Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
(script `07`) is the real-data companion to this article.

## 1. Two instruments from one fit

The same relation, the same 8 mm neighbourhoods, the same six cross-run pairs.
Only the normalization changes.

```{r two-frames}
detection <- compile_frame(searchlights(4, "local"), domain)
attribution <- compile_frame(searchlights(4, "conservative"), domain)
global <- compile_frame(whole_brain("none"), domain)
```

`whole_brain()` defaults to `"local"`, which divides by the feature count.
`"none"` is the unnormalized whole-domain operator, and it is the *only*
correct comparator for a conservation claim (§2, precondition 1). Using the
default here is the usual source of a spurious conservation failure.

```{r conservation}
detection_mass <- frame_conservation(detection)
attribution_mass <- frame_conservation(attribution)
c(
  detection_conserves = detection_mass$conserved,
  attribution_conserves = attribution_mass$conserved
)
c(
  detection_deviation = detection_mass$max_deviation,
  attribution_deviation = attribution_mass$max_deviation
)
```

```{r conservative-assert-1, include=FALSE}
stopifnot(
  !detection_mass$conserved,
  attribution_mass$conserved,
  attribution_mass$max_deviation < 1e-12,
  identical(attribution_mass$component, "total")
)
```

`frame_conservation()` reports on the **total** component only, and it says so
in `$component`. Section 4 below is why that scope is load-bearing rather than
an implementation detail.

Now the two maps, and the identity that separates them.

```{r budget-identity}
detection_map <- energy(detection)
attribution_map <- energy(attribution)
whole_domain <- energy(global)

identity_gap <- sum(attribution_map$total) - whole_domain$total
c(
  ledger_sum = sum(attribution_map$total),
  whole_domain_total = whole_domain$total,
  identity = identity_gap
)
```

```{r conservative-assert-2, include=FALSE}
stopifnot(abs(identity_gap) <= 1e-12 * abs(whole_domain$total))
```

That is claim 2 of the contract: for a column-normalized frame,
`sum_x <H, G_x> = <H, G_Omega>` for every fixed query `H`, exactly. The 280
searchlight values are not 280 measurements of a density; they are 280 pieces
of one number.

The detection map has no such identity, and the way it fails is instructive.

```{r detection-sum}
c(
  detection_sum = sum(detection_map$total),
  whole_domain_total = whole_domain$total,
  ratio = sum(detection_map$total) / whole_domain$total
)
```

```{r conservative-assert-3, include=FALSE}
stopifnot(
  abs(sum(detection_map$total) - whole_domain$total) > 1e-6,
  abs(sum(detection_map$total) / whole_domain$total - 1) < 0.05
)
```

Here the detection sum happens to lie close to the global total. Row
normalization does not guarantee this: a feature's aggregate weight depends on
all the overlapping neighbourhoods that include it. The sum therefore has no
global-budget interpretation. `contribution()` checks this requirement before
aggregating a map:

```{r contribution-refusal}
territory <- ifelse(
  domain$feature_ids %in% example$truth$planted_feature_ids, "pattern block",
  ifelse(domain$feature_ids %in% example$truth$mean_feature_ids, "mean block",
    "elsewhere")
)
refused <- catch_refusal(contribution(detection_map, by = territory))
refused$capability
```

```{r conservative-assert-4, include=FALSE}
stopifnot(identical(refused$capability, "conservative_frame"))
```

Read the two maps for what each is. The detection map's peak
(`r round(max(detection_map$total), 3)`) is a density, comparable across nodes
of different size and correctly reported as a peak. The attribution map's peak
is a *share of a budget*, and a large node holds more of it than a small one
for the trivial reason that it covers more features. Never compare them
node-for-node.

## 2. The ledger the frame reallocates

Before using the attribution map, understand what it is made of — because the
answer bounds what a multiscale panel is allowed to claim.

Under an identity or diagonal metric the node total is
`G_x = sum_v w_xv b_v^L b_v^R'`. The per-voxel outer products form a **fixed
ledger**, and the frame appears only as a nonnegative linear map applied after
the multivariate content has been fixed per voxel. The consequence is strong
and easy to check: the attribution map at any radius is a spatially smoothed
version of one univariate per-voxel map, and carries no cross-voxel information
whatsoever (§3).

`voxelwise()` is that per-voxel ledger — a point frame, trivially conservative.
Contracting it with the searchlight weights should reproduce the searchlight
map exactly.

```{r voxel-ledger}
ledger <- energy(compile_frame(voxelwise(), domain))
smoothed <- as.numeric(attribution$weights %*% ledger$total)
c(
  ledger_sums_to_global = sum(ledger$total) - whole_domain$total,
  smoothed_equals_map = max(abs(smoothed - attribution_map$total))
)
```

```{r conservative-assert-5, include=FALSE}
stopifnot(
  abs(sum(ledger$total) - whole_domain$total) <= 1e-12 * abs(whole_domain$total),
  max(abs(smoothed - attribution_map$total)) < 1e-12
)
```

The searchlight attribution map *is* the voxel ledger, reallocated. Nothing
between the two steps mixes features. This is scoped to metrics for which
`metric_capabilities()$feature_additive` is `TRUE`; section 6 is the case where
the premise fails.

### Per-scale totals are your own weights

Now stack several radii. `searchlights()` takes a vector, which builds an
α-weighted **family**: each member is column-normalized on its own, the weights
sum to one, and the stack is therefore conservative. Every row carries its own
provenance.

```{r family}
radii <- c(3.01, 4.3, 6.1)
family <- compile_frame(searchlights(radii, "conservative"), domain)
head(family$index, 3)
```

```{r conservative-assert-6, include=FALSE}
stopifnot(
  nrow(family$index) == 3L * domain$n_features,
  all(c("family", "scale", "center", "alpha") %in% names(family$index)),
  frame_conservation(family)$conserved
)
```

`coherence_spectrum()` aggregates that family by scale. Look at the `total`
column first.

```{r per-scale-identity}
spectrum <- coherence_spectrum(
  plan_geometry(example$fit$relation, family, pairing), example$contrast
)
as.data.frame(spectrum)[c("scale", "alpha", "total", "coherence_fraction")]

per_scale_gap <- max(abs(
  spectrum$total - as.data.frame(spectrum)$alpha * whole_domain$total
))
per_scale_gap
```

```{r conservative-assert-7, include=FALSE}
stopifnot(per_scale_gap <= 1e-12 * abs(whole_domain$total))
```

The `total` column is the same number three times, and it is exactly
`alpha_s` times the whole-domain total — **whatever the data say** (§3.1). It
is a picture of the `weights` argument you passed, not of spatial scale. Change
the weights and the column changes with them, because you changed it.

The plotting API makes this distinction explicit:

```{r energy-panel-refusal}
panel <- catch_refusal(plot(spectrum, which = "profile"))
panel$capability
```

```{r conservative-assert-8, include=FALSE}
stopifnot(identical(panel$capability, "scale_energy_panel"))
```

Use the decomposition view to see how the data divide each scale's budget.
The next section explains why its coherent fraction can vary with scale even
when total energy is fixed.

## 3. The coherence spectrum is the informative object

What the weights do *not* fix is the split of each scale's fixed budget into a
coherent part — the rank-one component along the node's own weighted common
mode — and a configuration part. That share depends on the data through
cross-voxel products, and it is the only scale-resolved quantity in the table
above that the weights do not determine.

It is stronger than "not proportional to α": the share is **exactly invariant**
to α, because total and coherent are both homogeneous of degree one under a
rescaling of a row, so the ratio cancels α (§3.2). Measure it rather than
believe it: rerun the same three radii under a lopsided weighting.

```{r alpha-invariance}
skewed <- coherence_spectrum(
  plan_geometry(
    example$fit$relation,
    compile_frame(searchlights(radii, "conservative", weights = c(0.6, 0.3, 0.1)), domain),
    pairing
  ),
  example$contrast
)
c(
  energy_moved = max(abs(skewed$total - spectrum$total)),
  share_moved = max(abs(skewed$coherence_fraction - spectrum$coherence_fraction))
)
```

```{r conservative-assert-9, include=FALSE}
stopifnot(
  max(abs(skewed$total - spectrum$total)) > 1,
  max(abs(skewed$coherence_fraction - spectrum$coherence_fraction)) < 1e-12
)
```

Changing α moves total energy while leaving the coherent fraction unchanged
within numerical precision. The fraction still depends on the chosen radii,
frame, and metric, which should accompany its interpretation. The aggregation
record names the weight-determined column as `alpha_fixed` and the invariant
column as `alpha_invariant`.

```{r spectrum-provenance}
spectrum_record <- spectrum$metadata$aggregation
c(
  fixed = spectrum_record$alpha_fixed,
  invariant = spectrum_record$alpha_invariant
)
```

```{r conservative-assert-10, include=FALSE}
stopifnot(
  identical(spectrum_record$alpha_fixed, "total"),
  identical(spectrum_record$alpha_invariant, "coherence_fraction")
)
```

### Two effect geometries, one budget

The share is a function of `(location, scale)`, not a number per scale, so a
location-wise reading is a table rather than a collapse: `by_location = TRUE`
returns one row per centre per radius, and no reduction over scales is offered.
The fixture's two planted blocks are the demonstration — same amplitude, same
voxel count, disjoint, differing only in whether the effect keeps one sign
across neighbouring voxels.

```{r by-location}
by_location <- as.data.frame(
  coherence_spectrum(
    plan_geometry(example$fit$relation, family, pairing), example$contrast,
    by_location = TRUE
  )
)
by_location$block <- ifelse(
  by_location$center %in% as.character(example$truth$planted_feature_ids),
  "pattern block",
  ifelse(by_location$center %in% as.character(example$truth$mean_feature_ids),
    "mean block", "elsewhere")
)
knitr::kable(tapply(
  by_location$coherence_fraction,
  list(by_location$block, by_location$scale),
  median, na.rm = TRUE
), digits = 3, caption = "Median valid coherent fraction by block and radius (mm).")
```

```{r by-location-check, include=FALSE}
shares <- tapply(
  by_location$coherence_fraction,
  list(by_location$block, by_location$scale), median, na.rm = TRUE
)
stopifnot(
  identical(dim(shares), c(3L, 3L)),
  all(shares["mean block", ] > 0.35),
  all(shares["pattern block", ] < 0.15),
  all(shares["mean block", ] > 4 * shares["pattern block", ]),
  shares["elsewhere", 1L] > shares["elsewhere", 3L]
)
```

Among nodes with valid fractions, the mean block's median coherent share is
roughly two fifths at each radius. The pattern block's median is much smaller:
alternating signs largely cancel under neighbourhood averaging. Its evidence
is primarily configuration at these radii. The elsewhere median decreases as
neighbourhoods grow in this fixture; it is a descriptive result for these data,
not a general scale law for noise.

Two disciplines apply to that table. Shares are **masked, never clamped**,
wherever the components are not a nonnegative partition, so `NA` in this table
means "this node's decomposition does not admit a fraction", not zero. A node's fraction divides its coherent component by its own total. A
territory's share of summed coherent mass, introduced next, has a different
denominator. That coherent budget is **frame-relative** (§4): changing the
frame changes the common modes, so their sum is not a conserved global quantity.

```{r masking}
c(
  rows = nrow(by_location),
  valid_shares = sum(!is.na(by_location$coherence_fraction))
)
```

```{r conservative-assert-11, include=FALSE}
stopifnot(
  any(is.na(by_location$coherence_fraction)),
  !all(is.na(by_location$coherence_fraction))
)
```

## 4. Territory ledgers

Because the map is a budget, it adds. `contribution()` is that addition and
nothing else: group the rows, sum them, and report what conserves and what does
not.

Grouping is **by row**. A searchlight belongs to the territory its centre falls
in, because a node is not divisible; splitting an overlapping node's mass
between two territories needs a second partition that can double-count, and it
is deliberately not offered. The measurement labels of a centre-assigned
searchlight frame are the centre feature identifiers, which is what makes the
alignment checkable rather than assumed.

```{r conservative-assert-12, include=FALSE}
stopifnot(identical(
  as.character(attribution$index$measurement), as.character(domain$feature_ids)
))
```

```{r territory}
ledger_by_territory <- contribution(attribution_map, by = territory)
as.data.frame(ledger_by_territory)
```

```{r territory-identity}
territory_gap <- sum(ledger_by_territory$total) - whole_domain$total
territory_gap
```

```{r territory-identity-check, include=FALSE}
stopifnot(
  abs(territory_gap) <= 1e-12 * abs(whole_domain$total),
  sum(ledger_by_territory$index$n_rows) == length(attribution_map$total),
  identical(ledger_by_territory$metadata$aggregation$overlap_split, FALSE)
)
```

The territories re-add to the whole-domain total exactly. Two readings a
detection map cannot give. First, the 19 nodes centred in the mean block — 7 %
of the 280 — hold about 30 % of the whole contrast budget, and so do the 19
centred in the pattern block. Second, those two near-equal budgets are split
completely differently: roughly two thirds of the mean block's is coherent
against about 4 % of the pattern block's. Equal amplitude, equal extent,
comparable share of the budget, opposite geometry.

What the aggregate refuses to give is as important.

```{r ledger-caveats}
ledger_record <- ledger_by_territory$metadata$aggregation
c(budget_exact = ledger_record$budget_exact, masked = ledger_record$masked)
ledger_record$frame_relative
```

```{r conservative-assert-13, include=FALSE}
stopifnot(
  identical(ledger_record$budget_exact, "total"),
  identical(ledger_record$masked, "signed"),
  isTRUE(ledger_record$frame_relative),
  all(is.na(ledger_by_territory$signed))
)
```

`signed` is masked rather than summed. A contrast view's signed marginal is the
local weighted *mean* contrast, already divided by the node's own frame mass —
a density — and adding densities over a territory is exactly the error a
conservative frame exists to avoid. So a territory ledger carries no direction;
read direction at the node level, from the signed map.

`coherent` and `configuration` add as arithmetic but their sums are
frame-relative, and the object says so rather than leaving it to the reader.

## 5. Descriptive spectra from a nonnegative projection

Crossvalidated estimates are **signed**. That is the visible cost of the
cross-partition pairing that removes the noise term, and conservation is a
statement about a signed sum: it does not make the summands nonnegative.

```{r signed-estimates}
c(
  negative_nodes = sum(attribution_map$total < 0),
  nodes = length(attribution_map$total)
)
```

```{r conservative-assert-14, include=FALSE}
stopifnot(any(attribution_map$total < 0))
```

So the arithmetic that treats a node's value as part of a nonnegative whole —
fractions of the total, cumulative-contribution curves, effective counts,
"top-k explains X %" — is invalid on this layer. The implied shares can leave
`[0, 1]`, and clipping to fix that reintroduces the bias the pairing removes
and destroys the conservation identity along with it.

Those functionals are legal on a **declared** nonnegative projection, which is
a separate, labelled object that records the mass it moved (§6).

```{r latent}
geometry <- materialize_geometry(
  plan_geometry(example$fit$relation, attribution, pairing)
)
latent <- latent_geometry(geometry)
latent
```

```{r latent-cost}
peak <- which.max(attribution_map$total)
all(latent$spectrum >= 0)
c(
  measurements = length(latent$n_eff),
  clipped = sum(latent$moved_mass > 0),
  masked = sum(is.na(latent$n_eff)),
  median_moved_share = round(median(latent$moved_share), 3)
)
c(
  peak_n_eff = round(latent$n_eff[[peak]], 3),
  peak_moved_share = round(latent$moved_share[[peak]], 4),
  peak_C1 = round(unname(latent$cumulative[peak, 1L]), 4)
)
```

```{r conservative-assert-15, include=FALSE}
stopifnot(
  all(latent$spectrum >= 0),
  identical(latent$projection$method, "psd_projection"),
  sum(latent$moved_mass > 0) > 0,
  all(is.na(latent$cumulative[is.na(latent$n_eff), ]))
)
```

The projection is not a formality on this fit: every node clipped something,
the median node moved about a third of its absolute mass, and a handful of
nodes had **no** nonnegative mass at all. Those are masked — `n_eff` and the
whole cumulative row are `NA` — rather than reported as zero, on the same
template the coherence fraction uses. At the strongest node the projected
geometry is effectively one-dimensional and the projection cost almost nothing;
that is a property of that node, not of the map.

None of this is inference, and none of it may be read back onto the signed
estimates of section 1. The object says so itself: its `reading:` line prints
*latent descriptive layer; not for inference*. And the projection enters
identity rather than sitting beside it.

```{r latent-discipline}
substring(latent$receipt$task_partition_id,
  nchar(latent$receipt$task_partition_id) - 14L)
query_readout <- catch_refusal(latent_geometry(attribution_map))
query_readout$capability
```

```{r conservative-assert-16, include=FALSE}
stopifnot(
  grepl("psd_projection$", latent$receipt$task_partition_id),
  identical(query_readout$capability, "latent_projection_source")
)
```

A contrast view is refused because its values are already contracted against
fixed weights: there is no spectrum left to truncate, and clamping such a value
at zero would be a *different* projection moving different mass. The layer
takes a named projection from a closed set rather than substituting one
silently.

A rank budget makes a second descriptive choice: retain at most that many
positive modes. The receipt separates negative mass clipped for PSD
admission from positive mass discarded by the rank budget.

```{r latent-rank-budget}
rank_one <- latent_geometry(geometry, rank = 1)
c(
  peak_clipped_negative = rank_one$clipped_negative_mass[[peak]],
  peak_truncated_positive = rank_one$truncated_positive_mass[[peak]],
  peak_moved = rank_one$moved_mass[[peak]]
)
```

```{r conservative-assert-17, include=FALSE}
stopifnot(
  all(rowSums(rank_one$spectrum > 0) <= 1L),
  max(abs(rank_one$moved_mass - rank_one$clipped_negative_mass -
    rank_one$truncated_positive_mass)) < 1e-12,
  identical(rank_one$clipped_negative_mass, latent$clipped_negative_mass),
  max(abs(rank_one$truncated_positive_mass -
    rowSums(latent$spectrum[, -1L, drop = FALSE]))) < 1e-12
)
```

This does not show that one dimension predicts independent data. For that
question, `fit_geometry()` learns a regularized model-supported form and
`score_geometry()` reads it on separate observations. With the same frozen
prediction, coherent and configuration test inner products still add; the
prediction's squared norm is subtracted once. Separately projected or fitted
latent components need not add to a projection of their total. See
[Predictive geometry](predictive-geometry.html)
(`vignette("predictive-geometry")` offline) for an executable example.

## 6. Which metrics preserve the budget?

Everything above assumed the node metric is the identity. A metric composes
with the frame by symmetric congruence in the square-root weights,
`K_x = D(sqrt(w_x)) Q D(sqrt(w_x))`, and whether conservation survives depends
entirely on whether `Q` is diagonal.

A dense metric is a `p × p` object and the point here is algebraic, so this
section drops to a nine-feature line domain.

```{r metric-fixture}
set.seed(20260820L)
p <- 9L
line <- abstract_domain(p, coordinates = cbind(seq_len(p) - 1, 0),
  feature_ids = paste0("v", seq_len(p)), id = "conservative-frames-metric")
block <- function() {
  value <- matrix(stats::rnorm(3 * p), 3, p)
  rownames(value) <- c("face", "house", "tool")
  value
}
line_relation <- relation(list(run1 = block(), run2 = block()), domain = line)
line_pairing <- cross_partitions(line_relation, independence = "independent")
line_contrast <- c(face = 1, house = -1, tool = 0)
line_frame <- compile_frame(searchlights(1.01, "conservative"), line)
line_global <- compile_frame(whole_brain("none"), line)

line_energy <- function(frame, metric = NULL, composition = "native") {
  contrast_energy(
    plan_geometry(line_relation, frame, line_pairing,
      metric = metric, composition = composition),
    line_contrast
  )
}
```

`metric_capabilities()` answers, before anything is read, whether the metric
keeps the per-voxel ledger of section 2 intact.

```{r metric-capabilities}
diagonal <- neural_metric(diag(stats::runif(p, 0.5, 2.5)), line)
raw <- matrix(stats::rnorm(p * p), p, p)
dense <- neural_metric(crossprod(raw) / p + diag(p), line)
c(
  diagonal = metric_capabilities(diagonal)$feature_additive,
  dense = metric_capabilities(dense)$feature_additive
)
```

```{r conservative-assert-18, include=FALSE}
stopifnot(
  metric_capabilities(diagonal)$feature_additive,
  !metric_capabilities(dense)$feature_additive
)
```

A diagonal metric folds into the frame weights: `D(sqrt(w)) D(q) D(sqrt(w))` is
`D(w q)`, still diagonal, still feature-additive, so `sum_x w_xv = 1` still
closes the ledger. A dense metric picks up off-diagonal terms
`S_uv = sum_x sqrt(w_xu w_xv)`, which column normalization forces to one only on
the diagonal. Conservation fails, and it fails as algebra rather than as
tolerance.

```{r metric-conservation}
gap <- function(metric = NULL, composition = "native") {
  local_sum <- sum(line_energy(line_frame, metric, composition)$total)
  reference <- line_energy(line_global, metric, composition)$total
  (local_sum - reference) / abs(reference)
}
relative_gaps <- c(
  identity = gap(),
  diagonal = gap(diagonal),
  dense_native = gap(dense),
  dense_whitened = gap(dense, "whitened")
)
signif(relative_gaps, 3)
```

```{r conservative-assert-19, include=FALSE}
stopifnot(
  abs(relative_gaps[["identity"]]) <= 1e-12,
  abs(relative_gaps[["diagonal"]]) <= 1e-12,
  abs(relative_gaps[["dense_native"]]) > 0.01,
  abs(relative_gaps[["dense_whitened"]]) <= 1e-12
)
```

The dense metric's measured gap belongs to this fixture. Its size and sign
can vary with the data, while the algebra explains why column normalization
does not guarantee conservation. A small gap on one dataset would not establish
a conservation law.

`composition = "whitened"` places the frame in whitened coordinates,
`Q^(1/2) D(w_x) Q^(1/2)`, whose sum over a conservative frame is `Q` exactly for
any SPD `Q`. It conserves — and it is **a different estimand, not a bug fix**.

```{r whitened-estimand}
native_nodes <- line_energy(line_frame, dense)$total
whitened_nodes <- line_energy(line_frame, dense, "whitened")$total
node_gap <- max(abs(native_nodes - whitened_nodes)) / max(abs(native_nodes))
round(node_gap, 3)
```

```{r conservative-assert-20, include=FALSE}
stopifnot(node_gap > 0.01)
```

Under the native composition a node weights *features* and measures them in the
`Q` geometry. Under the whitened composition it weights *whitened coordinates*,
which are spatially delocalized whenever `Q` is dense — the node's support is no
longer its support. Here the two disagree by roughly a quarter of the largest
node value. Both are defensible; they answer different questions, and the
package makes you name which.

That naming is enforced through plan identity, so a whitened analysis is never
mistaken for a native one after the fact.

```{r plan-identity}
native_plan <- plan_geometry(line_relation, line_frame, line_pairing,
  metric = dense)
whitened_plan <- plan_geometry(line_relation, line_frame, line_pairing,
  metric = dense, composition = "whitened")
!identical(native_plan$scientific_plan_id, whitened_plan$scientific_plan_id)
```

```{r conservative-assert-21, include=FALSE}
stopifnot(
  !identical(native_plan$scientific_plan_id, whitened_plan$scientific_plan_id)
)
```

The composition and the root convention — `"whitened"` means the symmetric
positive semidefinite root, since conservation holds for *any* root `R R' = Q`
while the node values do not — both enter the plan's scientific identity. A
conservation certificate is therefore not sufficient evidence that two whitened
analyses computed the same thing. And the switch must never be applied silently
to repair a failed conservation check: that is a change of estimand, and it
belongs in the record.

## 7. Reporting an attribution map

**No inference.** Conservation is a point-estimate law. `sum_x theta_x =
theta_Omega` says nothing about uncertainty: node estimates of an overlapping
frame can be correlated, variances do not generally add, and per-node
standard errors must never be summed to put an error bar on a conserved budget
(§7.6a). The cross-node sampling covariance that would license one does not
exist yet; `sampling_covariance(scope = "cross_measurement")` refuses by
capability rather than returning per-node margins that could be mistaken for a
joint block. Until then a conservative attribution map is a point ledger
reported without inference.

**No cross-voxel content in a total.** Section 2 is a caveat as much as a
demonstration: under a feature-additive metric, nothing in a `total` column at
any radius is multivariate across space. The multivariate content of this
package lives in the coherent/configuration split and in the RDM and RSA views,
not in the ledger.

**Advanced tier.** This is an advanced-tier topic and uses advanced-tier
exports: `frame_conservation()`, `materialize_geometry()`, `latent_geometry()`,
`neural_metric()`, and `metric_capabilities()`. Everything else here is core:
`example_fmri_effects()`, `compile_frame()`, `searchlights()`, `voxelwise()`,
`whole_brain()`, `abstract_domain()`, `relation()`, `cross_partitions()`,
`plan_geometry()`, `contrast_energy()`, `contribution()`,
`coherence_spectrum()`, and `catch_refusal()`.

## See also

- `vignette("interpreting-results")` — reading a detection map, and the three
  interpretive traps that make a coherent share look like a finding when it is
  an artefact of the frame. Read it before reporting any share from this
  article.
- `vignette("introduction")` — the same fixture as a detection-first workflow,
  and where the planted blocks come from.
- [`design/conservative-geometry-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/conservative-geometry-contract.md)
  — the normative document: the two estimands, the conservation theorem, the
  smoothed-ledger claim and its metric precondition, the α-invariance argument,
  and the numerical tolerances used by this guide's checks.
- [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001),
  script `07-conservative-geometry.R` and section 07 of its README — the same
  four instruments on one subject's ventral temporal cortex, with committed
  receipts.
- [Population form](population-form.html) — carry each participant's
  conservative ledger to shared group nodes and estimate a group model. This
  layer is experimental.
