---
title: "Conservative frames: attribution, not detection"
description: >
  The second spatial instrument. A conservative frame partitions one fixed global budget instead of reporting a density per neighbourhood, so its map is a ledger you may add up over territories — and its per-scale totals are the analyst's own weights, which is why the coherence spectrum, not the energy spectrum, is the finding.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conservative frames: attribution, not detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

Almost every searchlight analysis you have read answers one question: *where is
there evidence?* This article is about the other one: *how is the total effect
distributed?* They are different instruments, they are not comparable
node-for-node, and choosing between them is a decision about the question
rather than a preference about scaling.

The difference is one argument. A frame is a nonnegative weight matrix, one row
per measurement, one column per neural feature. `normalization = "local"`
normalizes the **rows**, so a node reports the mean evidence density inside its
own support — a detection map. `normalization = "conservative"` normalizes the
**columns**, so every feature's unit of evidence is *partitioned* among the
nodes that see it, and a node reports its share of one fixed global budget — an
attribution map.

Everything below is asserted in code you can see. If an identity in this
article stopped holding, the article would stop knitting. The governing
document is
[`design/conservative-geometry-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/conservative-geometry-contract.md),
and section numbers below refer to it.

```{r example}
example <- example_fmri_effects()
domain <- example$domain
pairing <- cross_partitions(
  example$fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
energy <- function(frame) {
  contrast_energy(
    plan_geometry(example$fit$relation, frame, pairing), example$contrast
  )
}
c(features = domain$n_features, conditions = length(example$contrast))
```

The generated fixture plants two blocks of 19 voxels at opposite ends of an
8 × 7 × 5 volume, at the same per-voxel amplitude, never overlapping. They
differ only in sign structure: in the **pattern** block the sign alternates
between neighbouring voxels, so a neighbourhood average nearly cancels; in the
**mean** block every voxel moves the same way. Same effect size, two different
effect *geometries*. Section 3 is where that difference becomes the finding.
These are generated data, not evidence about a brain; the
[Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
(script `07`) is the real-data companion to this article.

## 1. Two instruments from one fit

The same relation, the same 8 mm neighbourhoods, the same six cross-run pairs.
Only the normalization changes.

```{r two-frames}
detection <- compile_frame(searchlights(4, "local"), domain)
attribution <- compile_frame(searchlights(4, "conservative"), domain)
global <- compile_frame(whole_brain("none"), domain)
```

`whole_brain()` defaults to `"local"`, which divides by the feature count.
`"none"` is the unnormalized whole-domain operator, and it is the *only*
correct comparator for a conservation claim (§2, precondition 1). Using the
default here is the usual source of a spurious conservation failure.

```{r conservation}
detection_mass <- frame_conservation(detection)
attribution_mass <- frame_conservation(attribution)
c(
  detection_conserves = detection_mass$conserved,
  attribution_conserves = attribution_mass$conserved
)
c(
  detection_deviation = detection_mass$max_deviation,
  attribution_deviation = attribution_mass$max_deviation
)
stopifnot(
  !detection_mass$conserved,
  attribution_mass$conserved,
  attribution_mass$max_deviation < 1e-12,
  identical(attribution_mass$component, "total")
)
```

`frame_conservation()` reports on the **total** component only, and it says so
in `$component`. Section 4 below is why that scope is load-bearing rather than
an implementation detail.

Now the two maps, and the identity that separates them.

```{r budget-identity}
detection_map <- energy(detection)
attribution_map <- energy(attribution)
whole_domain <- energy(global)

identity_gap <- sum(attribution_map$total) - whole_domain$total
c(
  ledger_sum = sum(attribution_map$total),
  whole_domain_total = whole_domain$total,
  identity = identity_gap
)
stopifnot(abs(identity_gap) <= 1e-12 * abs(whole_domain$total))
```

That is claim 2 of the contract: for a column-normalized frame,
`sum_x <H, G_x> = <H, G_Omega>` for every fixed query `H`, exactly. The 280
searchlight values are not 280 measurements of a density; they are 280 pieces
of one number.

The detection map has no such identity, and the way it fails is instructive.

```{r detection-sum}
c(
  detection_sum = sum(detection_map$total),
  whole_domain_total = whole_domain$total,
  ratio = sum(detection_map$total) / whole_domain$total
)
stopifnot(
  abs(sum(detection_map$total) - whole_domain$total) > 1e-6,
  abs(sum(detection_map$total) / whole_domain$total - 1) < 0.05
)
```

Read that pair of assertions together. Summing the detection map lands within
2 % of the right number — and that is the trap, not a reassurance. The sum of a
detection map estimates nothing: overlapping neighbourhoods double-count every
voxel they share, and landing near the budget is an accident of how much these
particular neighbourhoods happen to overlap. A frame with more overlap would
miss by more, in a direction nothing in the analysis records. The package does
not let you take that sum by accident.

```{r contribution-refusal}
territory <- ifelse(
  domain$feature_ids %in% example$truth$planted_feature_ids, "pattern block",
  ifelse(domain$feature_ids %in% example$truth$mean_feature_ids, "mean block",
    "elsewhere")
)
refused <- catch_refusal(contribution(detection_map, by = territory))
refused$capability
stopifnot(identical(refused$capability, "conservative_frame"))
```

Read the two maps for what each is. The detection map's peak
(`r round(max(detection_map$total), 3)`) is a density, comparable across nodes
of different size and correctly reported as a peak. The attribution map's peak
is a *share of a budget*, and a large node holds more of it than a small one
for the trivial reason that it covers more features. Never compare them
node-for-node.

## 2. The ledger the frame reallocates

Before using the attribution map, understand what it is made of — because the
answer bounds what a multiscale panel is allowed to claim.

Under an identity or diagonal metric the node total is
`G_x = sum_v w_xv b_v^L b_v^R'`. The per-voxel outer products form a **fixed
ledger**, and the frame appears only as a nonnegative linear map applied after
the multivariate content has been fixed per voxel. The consequence is strong
and easy to check: the attribution map at any radius is a spatially smoothed
version of one univariate per-voxel map, and carries no cross-voxel information
whatsoever (§3).

`voxelwise()` is that per-voxel ledger — a point frame, trivially conservative.
Contracting it with the searchlight weights should reproduce the searchlight
map exactly.

```{r voxel-ledger}
ledger <- energy(compile_frame(voxelwise(), domain))
smoothed <- as.numeric(attribution$weights %*% ledger$total)
c(
  ledger_sums_to_global = sum(ledger$total) - whole_domain$total,
  smoothed_equals_map = max(abs(smoothed - attribution_map$total))
)
stopifnot(
  abs(sum(ledger$total) - whole_domain$total) <= 1e-12 * abs(whole_domain$total),
  max(abs(smoothed - attribution_map$total)) < 1e-12
)
```

The searchlight attribution map *is* the voxel ledger, reallocated. Nothing
between the two steps mixes features. This is scoped to metrics for which
`metric_capabilities()$feature_additive` is `TRUE`; section 6 is the case where
the premise fails.

### Per-scale totals are your own weights

Now stack several radii. `searchlights()` takes a vector, which builds an
α-weighted **family**: each member is column-normalized on its own, the weights
sum to one, and the stack is therefore conservative. Every row carries its own
provenance.

```{r family}
radii <- c(3.01, 4.3, 6.1)
family <- compile_frame(searchlights(radii, "conservative"), domain)
head(family$index, 3)
stopifnot(
  nrow(family$index) == 3L * domain$n_features,
  all(c("family", "scale", "center", "alpha") %in% names(family$index)),
  frame_conservation(family)$conserved
)
```

`coherence_spectrum()` aggregates that family by scale. Look at the `total`
column first.

```{r per-scale-identity}
spectrum <- coherence_spectrum(
  plan_geometry(example$fit$relation, family, pairing), example$contrast
)
as.data.frame(spectrum)[c("scale", "alpha", "total", "coherence_fraction")]

per_scale_gap <- max(abs(
  spectrum$total - as.data.frame(spectrum)$alpha * whole_domain$total
))
per_scale_gap
stopifnot(per_scale_gap <= 1e-12 * abs(whole_domain$total))
```

The `total` column is the same number three times, and it is exactly
`alpha_s` times the whole-domain total — **whatever the data say** (§3.1). It
is a picture of the `weights` argument you passed, not of spatial scale. Change
the weights and the column changes with them, because you changed it.

This is normative, not advisory, and the package enforces it.

```{r energy-panel-refusal}
panel <- catch_refusal(plot(spectrum, which = "profile"))
panel$capability
stopifnot(identical(panel$capability, "scale_energy_panel"))
```

An unqualified `plot()` on a spectrum draws the decomposition instead, which is
the panel whose content the data can move. **A multiscale panel of total energy
against scale is not a finding and may not be presented as one.**

## 3. The coherence spectrum is the informative object

What the weights do *not* fix is the split of each scale's fixed budget into a
coherent part — the rank-one component along the node's own weighted common
mode — and a configuration part. That share depends on the data through
cross-voxel products, and it is the only scale-resolved quantity in the table
above that the weights do not determine.

It is stronger than "not proportional to α": the share is **exactly invariant**
to α, because total and coherent are both homogeneous of degree one under a
rescaling of a row, so the ratio cancels α (§3.2). Measure it rather than
believe it: rerun the same three radii under a lopsided weighting.

```{r alpha-invariance}
skewed <- coherence_spectrum(
  plan_geometry(
    example$fit$relation,
    compile_frame(searchlights(radii, "conservative", weights = c(0.6, 0.3, 0.1)), domain),
    pairing
  ),
  example$contrast
)
c(
  energy_moved = max(abs(skewed$total - spectrum$total)),
  share_moved = max(abs(skewed$coherence_fraction - spectrum$coherence_fraction))
)
stopifnot(
  max(abs(skewed$total - spectrum$total)) > 1,
  max(abs(skewed$coherence_fraction - spectrum$coherence_fraction)) < 1e-12
)
```

The energy column moves by tens of units; the share moves by less than a part
in `1e15`. That is why a coherence spectrum may be reported without disclosing
α — the exact opposite of the energy panel — and it is what the aggregation
records: `alpha_fixed` names the column the weights determine, `alpha_invariant`
names the one they do not.

```{r spectrum-provenance}
spectrum_record <- spectrum$metadata$aggregation
c(
  fixed = spectrum_record$alpha_fixed,
  invariant = spectrum_record$alpha_invariant
)
stopifnot(
  identical(spectrum_record$alpha_fixed, "total"),
  identical(spectrum_record$alpha_invariant, "coherence_fraction")
)
```

### Two effect geometries, one budget

The share is a function of `(location, scale)`, not a number per scale, so a
location-wise reading is a table rather than a collapse: `by_location = TRUE`
returns one row per centre per radius, and no reduction over scales is offered.
The fixture's two planted blocks are the demonstration — same amplitude, same
voxel count, disjoint, differing only in whether the effect keeps one sign
across neighbouring voxels.

```{r by-location}
by_location <- as.data.frame(
  coherence_spectrum(
    plan_geometry(example$fit$relation, family, pairing), example$contrast,
    by_location = TRUE
  )
)
by_location$block <- ifelse(
  by_location$center %in% as.character(example$truth$planted_feature_ids),
  "pattern block",
  ifelse(by_location$center %in% as.character(example$truth$mean_feature_ids),
    "mean block", "elsewhere")
)
round(tapply(
  by_location$coherence_fraction,
  list(by_location$block, by_location$scale),
  median, na.rm = TRUE
), 3)
```

```{r by-location-check}
shares <- tapply(
  by_location$coherence_fraction,
  list(by_location$block, by_location$scale), median, na.rm = TRUE
)
stopifnot(
  identical(dim(shares), c(3L, 3L)),
  all(shares["mean block", ] > 0.35),
  all(shares["pattern block", ] < 0.15),
  all(shares["mean block", ] > 4 * shares["pattern block", ]),
  shares["elsewhere", 1L] > shares["elsewhere", 3L]
)
```

The mean block keeps roughly two fifths of its budget in the common mode at
every radius. The pattern block never has a common mode to find: an effect that
alternates sign between neighbours has nearly nothing left after a
neighbourhood average, at any radius the frame can resolve, and its budget sits
almost entirely in configuration. Away from both blocks the share falls as the
neighbourhood grows — the ordinary behaviour of noise plus a fixed budget, as a
node takes in more territory over which nothing keeps one sign.

Two disciplines apply to that table. Shares are **masked, never clamped**,
wherever the components are not a nonnegative partition, so `NA` in this table
means "this node's decomposition does not admit a fraction", not zero. And a
coherent budget is **frame-relative** (§4): the coherent parts of a frame's
nodes do not sum to any global quantity, so a coherent share is a share of
*this frame's* coherent mass and two frames give two incomparable denominators.

```{r masking}
c(
  rows = nrow(by_location),
  valid_shares = sum(!is.na(by_location$coherence_fraction))
)
stopifnot(
  any(is.na(by_location$coherence_fraction)),
  !all(is.na(by_location$coherence_fraction))
)
```

## 4. Territory ledgers

Because the map is a budget, it adds. `contribution()` is that addition and
nothing else: group the rows, sum them, and report what conserves and what does
not.

Grouping is **by row**. A searchlight belongs to the territory its centre falls
in, because a node is not divisible; splitting an overlapping node's mass
between two territories needs a second partition that can double-count, and it
is deliberately not offered. The measurement labels of a centre-assigned
searchlight frame are the centre feature identifiers, which is what makes the
alignment checkable rather than assumed.

```{r territory}
stopifnot(identical(
  as.character(attribution$index$measurement), as.character(domain$feature_ids)
))
ledger_by_territory <- contribution(attribution_map, by = territory)
as.data.frame(ledger_by_territory)
```

```{r territory-identity}
territory_gap <- sum(ledger_by_territory$total) - whole_domain$total
territory_gap
stopifnot(
  abs(territory_gap) <= 1e-12 * abs(whole_domain$total),
  sum(ledger_by_territory$index$n_rows) == length(attribution_map$total),
  identical(ledger_by_territory$metadata$aggregation$overlap_split, FALSE)
)
```

The territories re-add to the whole-domain total exactly. Two readings a
detection map cannot give. First, the 19 nodes centred in the mean block — 7 %
of the 280 — hold about 30 % of the whole contrast budget, and so do the 19
centred in the pattern block. Second, those two near-equal budgets are split
completely differently: roughly two thirds of the mean block's is coherent
against about 4 % of the pattern block's. Equal amplitude, equal extent,
comparable share of the budget, opposite geometry.

What the aggregate refuses to give is as important.

```{r ledger-caveats}
ledger_record <- ledger_by_territory$metadata$aggregation
c(budget_exact = ledger_record$budget_exact, masked = ledger_record$masked)
ledger_record$frame_relative
stopifnot(
  identical(ledger_record$budget_exact, "total"),
  identical(ledger_record$masked, "signed"),
  isTRUE(ledger_record$frame_relative),
  all(is.na(ledger_by_territory$signed))
)
```

`signed` is masked rather than summed. A contrast view's signed marginal is the
local weighted *mean* contrast, already divided by the node's own frame mass —
a density — and adding densities over a territory is exactly the error a
conservative frame exists to avoid. So a territory ledger carries no direction;
read direction at the node level, from the signed map.

`coherent` and `configuration` add as arithmetic but their sums are
frame-relative, and the object says so rather than leaving it to the reader.

## 5. The latent layer, and what its projection cost

Crossvalidated estimates are **signed**. That is the visible cost of the
cross-partition pairing that removes the noise term, and conservation is a
statement about a signed sum: it does not make the summands nonnegative.

```{r signed-estimates}
c(
  negative_nodes = sum(attribution_map$total < 0),
  nodes = length(attribution_map$total)
)
stopifnot(any(attribution_map$total < 0))
```

So the arithmetic that treats a node's value as part of a nonnegative whole —
fractions of the total, cumulative-contribution curves, effective counts,
"top-k explains X %" — is invalid on this layer. The implied shares can leave
`[0, 1]`, and clipping to fix that reintroduces the bias the pairing removes
and destroys the conservation identity along with it.

Those functionals are legal on a **declared** nonnegative projection, which is
a separate, labelled object that records the mass it moved (§6).

```{r latent}
geometry <- materialize_geometry(
  plan_geometry(example$fit$relation, attribution, pairing)
)
latent <- latent_geometry(geometry)
latent
```

```{r latent-cost}
peak <- which.max(attribution_map$total)
all(latent$spectrum >= 0)
c(
  measurements = length(latent$n_eff),
  clipped = sum(latent$moved_mass > 0),
  masked = sum(is.na(latent$n_eff)),
  median_moved_share = round(median(latent$moved_share), 3)
)
c(
  peak_n_eff = round(latent$n_eff[[peak]], 3),
  peak_moved_share = round(latent$moved_share[[peak]], 4),
  peak_C1 = round(unname(latent$cumulative[peak, 1L]), 4)
)
stopifnot(
  all(latent$spectrum >= 0),
  identical(latent$projection$method, "psd_projection"),
  sum(latent$moved_mass > 0) > 0,
  all(is.na(latent$cumulative[is.na(latent$n_eff), ]))
)
```

The projection is not a formality on this fit: every node clipped something,
the median node moved about a third of its absolute mass, and a handful of
nodes had **no** nonnegative mass at all. Those are masked — `n_eff` and the
whole cumulative row are `NA` — rather than reported as zero, on the same
template the coherence fraction uses. At the strongest node the projected
geometry is effectively one-dimensional and the projection cost almost nothing;
that is a property of that node, not of the map.

None of this is inference, and none of it may be read back onto the signed
estimates of section 1. The object says so itself: its `reading:` line prints
*latent descriptive layer; not for inference*. And the projection enters
identity rather than sitting beside it.

```{r latent-discipline}
substring(latent$receipt$task_partition_id,
  nchar(latent$receipt$task_partition_id) - 14L)
query_readout <- catch_refusal(latent_geometry(attribution_map))
query_readout$capability
stopifnot(
  grepl("psd_projection$", latent$receipt$task_partition_id),
  identical(query_readout$capability, "latent_projection_source")
)
```

A contrast view is refused because its values are already contracted against
fixed weights: there is no spectrum left to truncate, and clamping such a value
at zero would be a *different* projection moving different mass. The layer
takes a named projection from a closed set rather than substituting one
silently.

## 6. Metrics: what folds, what breaks, what is a different estimand

Everything above assumed the node metric is the identity. A metric composes
with the frame by symmetric congruence in the square-root weights,
`K_x = D(sqrt(w_x)) Q D(sqrt(w_x))`, and whether conservation survives depends
entirely on whether `Q` is diagonal.

A dense metric is a `p × p` object and the point here is algebraic, so this
section drops to a nine-feature line domain.

```{r metric-fixture}
set.seed(20260820L)
p <- 9L
line <- abstract_domain(p, coordinates = cbind(seq_len(p) - 1, 0),
  feature_ids = paste0("v", seq_len(p)), id = "conservative-frames-metric")
block <- function() {
  value <- matrix(stats::rnorm(3 * p), 3, p)
  rownames(value) <- c("face", "house", "tool")
  value
}
line_relation <- relation(list(run1 = block(), run2 = block()), domain = line)
line_pairing <- cross_partitions(line_relation, independence = "independent")
line_contrast <- c(face = 1, house = -1, tool = 0)
line_frame <- compile_frame(searchlights(1.01, "conservative"), line)
line_global <- compile_frame(whole_brain("none"), line)

line_energy <- function(frame, metric = NULL, composition = "native") {
  contrast_energy(
    plan_geometry(line_relation, frame, line_pairing,
      metric = metric, composition = composition),
    line_contrast
  )
}
```

`metric_capabilities()` answers, before anything is read, whether the metric
keeps the per-voxel ledger of section 2 intact.

```{r metric-capabilities}
diagonal <- neural_metric(diag(stats::runif(p, 0.5, 2.5)), line)
raw <- matrix(stats::rnorm(p * p), p, p)
dense <- neural_metric(crossprod(raw) / p + diag(p), line)
c(
  diagonal = metric_capabilities(diagonal)$feature_additive,
  dense = metric_capabilities(dense)$feature_additive
)
stopifnot(
  metric_capabilities(diagonal)$feature_additive,
  !metric_capabilities(dense)$feature_additive
)
```

A diagonal metric folds into the frame weights: `D(sqrt(w)) D(q) D(sqrt(w))` is
`D(w q)`, still diagonal, still feature-additive, so `sum_x w_xv = 1` still
closes the ledger. A dense metric picks up off-diagonal terms
`S_uv = sum_x sqrt(w_xu w_xv)`, which column normalization forces to one only on
the diagonal. Conservation fails, and it fails as algebra rather than as
tolerance.

```{r metric-conservation}
gap <- function(metric = NULL, composition = "native") {
  local_sum <- sum(line_energy(line_frame, metric, composition)$total)
  reference <- line_energy(line_global, metric, composition)$total
  (local_sum - reference) / abs(reference)
}
relative_gaps <- c(
  identity = gap(),
  diagonal = gap(diagonal),
  dense_native = gap(dense),
  dense_whitened = gap(dense, "whitened")
)
signif(relative_gaps, 3)
stopifnot(
  abs(relative_gaps[["identity"]]) <= 1e-12,
  abs(relative_gaps[["diagonal"]]) <= 1e-12,
  abs(relative_gaps[["dense_native"]]) > 0.01,
  abs(relative_gaps[["dense_whitened"]]) <= 1e-12
)
```

The size and sign of the dense failure are fixture-specific — across draws it
ranges over both signs and tens of percent, and a draw can land within 1 % of
zero by chance. The claim is the algebraic law, never a percentage, which is
why the assertion above reads `> 1 %` and not an equality. A small measured
deviation on one dataset is not evidence that a dense metric conserves.

`composition = "whitened"` places the frame in whitened coordinates,
`Q^(1/2) D(w_x) Q^(1/2)`, whose sum over a conservative frame is `Q` exactly for
any SPD `Q`. It conserves — and it is **a different estimand, not a bug fix**.

```{r whitened-estimand}
native_nodes <- line_energy(line_frame, dense)$total
whitened_nodes <- line_energy(line_frame, dense, "whitened")$total
node_gap <- max(abs(native_nodes - whitened_nodes)) / max(abs(native_nodes))
round(node_gap, 3)
stopifnot(node_gap > 0.01)
```

Under the native composition a node weights *features* and measures them in the
`Q` geometry. Under the whitened composition it weights *whitened coordinates*,
which are spatially delocalized whenever `Q` is dense — the node's support is no
longer its support. Here the two disagree by roughly a quarter of the largest
node value. Both are defensible; they answer different questions, and the
package makes you name which.

That naming is enforced through plan identity, so a whitened analysis is never
mistaken for a native one after the fact.

```{r plan-identity}
native_plan <- plan_geometry(line_relation, line_frame, line_pairing,
  metric = dense)
whitened_plan <- plan_geometry(line_relation, line_frame, line_pairing,
  metric = dense, composition = "whitened")
!identical(native_plan$scientific_plan_id, whitened_plan$scientific_plan_id)
stopifnot(
  !identical(native_plan$scientific_plan_id, whitened_plan$scientific_plan_id)
)
```

The composition and the root convention — `"whitened"` means the symmetric
positive semidefinite root, since conservation holds for *any* root `R R' = Q`
while the node values do not — both enter the plan's scientific identity. A
conservation certificate is therefore not sufficient evidence that two whitened
analyses computed the same thing. And the switch must never be applied silently
to repair a failed conservation check: that is a change of estimand, and it
belongs in the record.

## 7. What this article does not give you

**No inference.** Conservation is a point-estimate law. `sum_x theta_x =
theta_Omega` says nothing about uncertainty: node estimates of an overlapping
frame are strongly positively correlated, variances do not add, and per-node
standard errors must never be summed to put an error bar on a conserved budget
(§7.6a). The cross-node sampling covariance that would license one does not
exist yet; `sampling_covariance(scope = "cross_measurement")` refuses by
capability rather than returning per-node margins that could be mistaken for a
joint block. Until then a conservative attribution map is a point ledger
reported without inference.

**No cross-voxel content in a total.** Section 2 is a caveat as much as a
demonstration: under a feature-additive metric, nothing in a `total` column at
any radius is multivariate across space. The multivariate content of this
package lives in the coherent/configuration split and in the RDM and RSA views,
not in the ledger.

**Advanced tier.** This is an advanced-tier topic and uses advanced-tier
exports: `frame_conservation()`, `materialize_geometry()`, `latent_geometry()`,
`neural_metric()`, and `metric_capabilities()`. Everything else here is core:
`example_fmri_effects()`, `compile_frame()`, `searchlights()`, `voxelwise()`,
`whole_brain()`, `abstract_domain()`, `relation()`, `cross_partitions()`,
`plan_geometry()`, `contrast_energy()`, `contribution()`,
`coherence_spectrum()`, and `catch_refusal()`.

## See also

- `vignette("interpreting-results")` — reading a detection map, and the three
  interpretive traps that make a coherent share look like a finding when it is
  an artefact of the frame. Read it before reporting any share from this
  article.
- `vignette("introduction")` — the same fixture as a detection-first workflow,
  and where the planted blocks come from.
- [`design/conservative-geometry-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/conservative-geometry-contract.md)
  — the normative document: the two estimands, the conservation theorem, the
  smoothed-ledger claim and its metric precondition, the α-invariance argument,
  and the measured tolerances every assertion above is drawn from.
- [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001),
  script `07-conservative-geometry.R` and section 07 of its README — the same
  four instruments on one subject's ventral temporal cortex, with committed
  receipts.
- The population layer, which carries a conservative ledger from one
  participant onto a group frame, is experimental and under construction; it is
  what the transport-readiness section of the contract is written for.
