---
title: "Choosing distance: effect magnitude or pattern correlation"
description: >
  Choose between squared distance and Pearson correlation distance, understand
  their different scale behavior, and identify the views crossform supports.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing distance: effect magnitude or pattern correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

Should two condition patterns count as different when one is a larger version
of the other? **Squared distance retains that amplitude difference. Pearson
correlation distance removes it**, along with each pattern's mean across neural
features. Choose according to the scientific question before constructing an
RDM.

This guide assumes you know what an RDM represents; the
[introduction](introduction.html) shows how to compute one with `crossform`.
The distinction matters for both the point estimate and its uncertainty.

| Your question | Quantity | Current support |
|---|---|---|
| How far apart are the effects in a fixed feature metric? | Squared Euclidean or fixed-Mahalanobis distance | `rdm()`; crossnobis uses a declared noise metric |
| How different are the patterns after removing their means and positive scales? | `1 - Pearson correlation` | Illustrative base R calculation below; no exported `correlation_rdm()` |
| How does an RDM relate to a model? | RSA under a declared RDM and comparison rule | `rsa()` does not silently replace the input geometry with correlation distance |

## See what amplitude normalization removes

Each row below is a condition and each column is a neural feature. Pattern
`b` is twice pattern `a`, with five added to every feature. Pattern `c` changes
the arrangement of the middle two features.

```{r patterns}
a <- c(-2, -1, 1, 2)
patterns <- rbind(a = a, b = 2 * a + 5, c = c(-2, 1, -1, 2))
knitr::kable(patterns, col.names = paste0("feature ", 1:4))
```

Compute both distances on these observed patterns. `cor()` correlates columns,
so the transpose makes conditions its variables. These are ordinary self-data
calculations, with no crossvalidation or uncertainty estimate.

```{r distances}
squared <- as.matrix(stats::dist(patterns))^2
correlation_distance <- 1 - stats::cor(t(patterns))
pairs <- which(lower.tri(squared), arr.ind = TRUE)
comparison <- data.frame(
  pair = paste(rownames(patterns)[pairs[, 2]],
    rownames(patterns)[pairs[, 1]], sep = " versus "),
  squared_distance = squared[pairs],
  correlation_distance = correlation_distance[pairs]
)
knitr::kable(comparison, digits = 3)
```

The `a`–`b` pair has a much larger squared distance than `a`–`c`, yet its
correlation distance is zero.
That zero says that the centered patterns have the same direction. It does
not say that their effects have equal magnitude. The `a`–`c` pair changes
pattern direction, which both distances detect.

## When can a form be normalized as Pearson correlation?

Let `B` contain condition patterns in rows. First center each row across neural
features, giving `B_c`. Its self form is `G = B_c B_c'`. This construction makes
`G` positive semidefinite (PSD), and the correlation-distance formula is

\[
d_{ij}^{\mathrm{corr}}
= 1 - \frac{G_{ij}}{\sqrt{G_{ii}G_{jj}}}.
\]

**The formula requires a feature-centered PSD self form and strictly positive
relevant diagonal entries.** PSD alone does not supply feature centering. A
constant pattern has zero centered norm and its Pearson correlation is
undefined.

```{r form-normalization}
centered <- sweep(patterns, 1, rowMeans(patterns))
G <- tcrossprod(centered)
norms <- sqrt(diag(G))
from_form <- 1 - G / outer(norms, norms)
c(max_difference_from_cor = max(abs(from_form - correlation_distance)))
```

```{r normalization-check, include=FALSE}
stopifnot(
  all(diag(G) > 0),
  min(eigen(G, symmetric = TRUE, only.values = TRUE)$values) > -1e-12,
  max(abs(from_form - correlation_distance)) < 1e-12,
  abs(correlation_distance["a", "b"]) < 1e-12,
  squared["a", "b"] > squared["a", "c"]
)
```

Feature centering is a fixed linear projection and can be represented by a
neural metric. Dividing by the observed norms is nonlinear: its denominator
changes with the data. Consequently, correlation distance is not a fixed
linear query of the unnormalized form, even though it can be computed from a
suitable self form.

## Why does crossvalidation change the problem?

A cross-generalized form uses independently estimated patterns on its two
sides. It is signed: diagonal entries may be zero or negative, and the matrix
need not be PSD. Substituting such a form into the Pearson formula can produce
undefined denominators or values outside the usual correlation range. Clipping
diagonals changes the estimator and does not recover ordinary Pearson distance.

For independent partitions and a correctly specified fixed metric, squared
cross-distances estimate the corresponding signal distance without the self-noise
term. Their signed behavior around a zero effect is useful for inference.
Normalizing by a denominator estimated from the evaluation data introduces a
new estimator and requires its own sampling analysis. The covariance of a
linear RDM does not automatically become the covariance of a normalized one.

## What is available, and what would a future view need?

`rdm()` reads squared Euclidean or fixed-Mahalanobis distances from bilinear
geometry. It does not normalize condition patterns to unit length. The base R
example above illustrates a distinct estimand; it is not an alternate
`crossform` RDM route.

A proposed `correlation_rdm()` would need to declare centering and its metric,
require an appropriate PSD self form with positive diagonals, and refuse
undefined normalization. It would also need explicit spatial-frame and unit
semantics, independent numerical checks, and performance validation. Any
standard errors would require a separately validated delta-method, bootstrap,
or other sampling procedure. This named view is **not currently exported**.

Use the existing squared-distance workflow when effect magnitude in a fixed
metric is the target. Use the pattern-correlation estimand when invariance to
positive scaling and feature means is itself the question, while keeping its
construction and uncertainty separate from crossvalidated squared distances.
