---
title: "Extending crossform: the developer protocol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending crossform: the developer protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_neuroim2 <- requireNamespace("neuroim2", quietly = TRUE) &&
  utils::packageVersion("neuroim2") >= "0.19.0"
library(crossform)
set.seed(20260817)
```

This guide is for people writing a *package*, not an analysis: a storage
backend, a first-level fitting engine, a design compiler, a spatial provider,
or a result type. It states the surface crossform promises to such a package —
what you supply, what the compiler checks, and which refusal you get when the
declaration and the object disagree — and then builds a small adapter end to
end so the protocol is demonstrated rather than described.

If you are analysing data rather than writing a package, you want
`vignette("introduction")` or `vignette("from-observations")` instead. Nothing
here is required to use crossform.

## The protocol

An extension package meets crossform at two places.

**The extension-only surface** is six entry points. They exist so another
package can hand data, an estimator, an error channel, and a result type into
the typed core; an ordinary analysis never calls them.

| Entry point | Seam | Hands in |
|---|---|---|
| `file_matrix_source()` | data-in | a reopenable out-of-memory response store |
| `source_capabilities()` | data-in | what an opaque source can actually do |
| `relation_block()` | data-in | the bounded read the whole design rests on |
| `effect_extractor()` | design-in | the declared map `E` in `B = E Y` |
| `relation_fit()` | error-channel-in | the statistical envelope around a relation |
| `as_neurovol()` | result-out | an S3 generic your result class can register on |

**The ordinary core** is everything else, and adapters use much more of it than
of the list above. The four adapters shipped in this package call fourteen
crossform functions between them and only two — `effect_extractor()` and
`relation_fit()` — come from the extension-only list. The rest are the same
`study()`, `observation_events()`, `condition_space()`, `design_model()`,
`relation()`, `abstract_domain()` calls a user makes. That is deliberate: an
adapter is a *morphism into the typed core*, so it builds ordinary crossform
objects and hands them over. There is no privileged back door, and the absence
of one is the point.

The rest of this guide takes the six in turn, then builds an adapter.

### `file_matrix_source()` — an out-of-memory store

You supply a path, the two-element `dim` (observations by neural features),
and optionally an expected `sha256:` revision. You get an
`effect_source_descriptor`, which is a *description*: no handle is opened, and
the descriptor is serializable, so it survives being sent to a worker.

The constructor checks that the file is exactly `offset_bytes + prod(dim) * 8`
bytes, and computes the content revision itself. If you declare a
`stable_revision` that does not match the bytes on disk, it is an
`effect_contract_error` — two objects disagreeing, not a bad argument:

```{r file-source-refusal}
path <- tempfile(fileext = ".bin")
writeBin(as.vector(matrix(as.double(1:12), 4L, 3L)), path, size = 8L)

tryCatch(
  file_matrix_source(
    path, dim = c(4L, 3L),
    stable_revision = paste0("sha256:", strrep("0", 64))
  ),
  effect_contract_error = conditionMessage
)
```

The same check runs again every time the file is reopened for a block read
(`"Reopenable matrix source has a stale content revision."`), so a store that
is silently rewritten between planning and execution is caught rather than
used.

Your own store need not be a flat matrix file. Any function
`function(features) -> matrix` is accepted as a source by `relation()` when you
declare `source_dims`; `file_matrix_source()` is the one concrete descriptor
crossform ships, and the one whose capabilities it can derive for you.

### `source_capabilities()` — what the source can do

This is the declaration the compiler plans against: `block_read`,
`reopenable`, `thread_safe`, and a `stable_revision` that goes into the
execution receipt. A source descriptor implies its own capabilities, so
`relation()` derives them for you when every source is a descriptor. A source
that is an opaque closure implies nothing, and you must say.

Three ways to get it wrong, and what each raises:

```{r capability-refusals}
weak <- tryCatch(
  source_capabilities(TRUE, stable_revision = "2026-08-15"),
  effect_input_error = conditionMessage
)
flags <- tryCatch(
  source_capabilities(block_read = "yes", stable_revision = paste0(
    "sha256:", strrep("a", 64)
  )),
  effect_input_error = conditionMessage
)
c(weak_revision = weak, nonlogical_flag = flags)
```

```{r capability-refusal-reopenable}
# Claiming `reopenable` for a source that has no reopenable descriptor: the
# claim cannot be honoured, because there is nothing to reopen.
source_fn <- function(features) matrix(0, 4L, length(features))
tryCatch(
  relation(
    list(`run-1` = source_fn), effects = c("a", "b", "c", "d"),
    source_dims = list(c(4L, 3L)),
    capabilities = source_capabilities(
      block_read = TRUE, reopenable = TRUE,
      stable_revision = paste0("sha256:", strrep("a", 64))
    )
  ),
  effect_input_error = conditionMessage
)
```

Two more are worth knowing because they gate execution rather than
construction. Both come from `.relation_source_capabilities()` in
`R/capabilities.R`, which every route into the executor passes through:

- omitting capabilities for an opaque source —
  `"Opaque relation sources require explicit source_capabilities() before
  execution."`
- declaring `block_read = FALSE` —
  `"All relation sources must support bounded block reads."`

Note what class these are. A wrong capability declaration is an
`effect_input_error` or an `effect_contract_error`, not an
`effect_capability_refusal`: crossform reserves the refusal class for *"the
interpretation you asked for cannot be earned from the objects you supplied"*,
and a source that lies about itself is a bug in the adapter. Refusals appear
one layer up, at the error channel, and carry `$capability` and `$namespace`
(see `?crossform_conditions`). Both are demonstrated in the worked example
below.

### `relation_block()` — the bounded read

`relation_block(x, partition, features)` is where neural values are actually
read: only the requested feature columns are pulled from the source and mapped
through that partition's extractor. It is on the extension surface for two
reasons — an external executor needs it, and it is the observable that proves
your source is genuinely block-readable rather than a whole-array read wearing
a closure.

Feature indices must be unique and in range
(`"features must be unique valid neural feature indices."`), and the partition
must name one the relation declares
(`"partition must identify one relation partition."`). The returned matrix
always has one row per effect coordinate in effect-space order, whatever order
your source stored things in.

### `effect_extractor()` — the declared map

An extractor is the matrix `E` in `B = E Y`: effect-by-observation, holding no
neural data, and therefore reusable across every feature block. Supply `map`
plus an `estimator` identity string; the constructor requires a finite
nonempty matrix and names the rows from the effect space.

This is the seam for an engine that owns its own estimator. crossform's
`lm_extractor()` compiles one from a design and a target and is *not* part of
the developer protocol — maintainer decision 2 in `design/api-tiers.md`
re-tiered it as advanced, because it is an `extract =` argument value of
`relation()` and so reachable by an ordinary user. Use it if it fits; it
raises the capability refusals `full_rank_design` and `estimable_effects` in
namespace `relation_fit` when the design cannot identify what was asked for.
If your package does something `lm_extractor()` does not, compute `E` yourself
and hand it in through `effect_extractor()` — that is what the fmrireg adapter
does with `fmrireg::fmri_ols_fit()`.

### `relation_fit()` — the error channel

`relation_fit(relation, error_models, provenance)` wraps a mathematically
intact relation in a separate, capability-bearing statistical envelope. The
relation itself is untouched; what the wrapper adds is a per-partition record
of what uncertainty the fit can support, readable with
`relation_fit_capabilities()`.

**Be precise about what an external package can attach today.** The
`error_models` argument takes crossform's internal error-model values, and no
public constructor for one exists: the separable-GLM model is built inside
`lm_relation_fit()`. So from outside the package, `relation_fit()` attaches an
**explicitly absent** channel — `error_models = NULL` — which is a real and
useful statement, and different from an unstated one. It is exactly what
`fmrireg_relation()` does (`R/adapter-fmrireg.R`), recording
`analytic_error_channel = "withheld"` in its provenance. An engine that has
residuals and wants them honoured routes its raw responses through
`lm_relation_fit()` instead.

The consequence, once declared, is enforced. Every operation that needs the
channel refuses with a capability whose name is the missing column of
`relation_fit_capabilities()`, in namespace `relation_fit`:

| Capability | Refused operation |
|---|---|
| `error_model` | any statistical envelope at all |
| `residual_blocks` | `residual_block()`, learned metrics |
| `effect_covariance` | design-side factor of the separable error model |
| `residual_df` | `residual_df()`, the noise-variance divisor |
| `separable_error` | the separable second-moment law |
| `learned_metric_input` | metric learning from residuals |
| `within_participant_calibration` | analytic within-participant standard errors |

### `as_neurovol()` — writing a result type out

`as_neurovol()` is an S3 generic dispatching on `values`, so a package with its
own result class can teach crossform to write that class to a volume without
crossform importing it. Register the method the ordinary way —
`S3method(as_neurovol, my_result)` in your `NAMESPACE` — and it receives
`mask`, `domain`, `fill`, and `label` unchanged. The generic validates
nothing, so your method may take different arguments; it is expected to return
a `NeuroVol` on the space of `mask`. crossform ships the numeric-vector method
and a `default` that behaves identically.

## A worked adapter

Suppose you maintain `runstore`, a package that keeps one response matrix per
run in a binary file store and fits its own first-level model. This section
builds it against the protocol, with synthetic data so it runs here.

### The store

Three runs, twenty-four scans, forty features. Eight of the forty carry a real
`face` effect; the rest are noise. Each run is written to disk in R's
column-major order as consecutive doubles — the layout `file_matrix_source()`
requires.

```{r store}
n_scans <- 24L
n_features <- 40L
runs <- c("run-1", "run-2", "run-3")
signal_features <- 1:8

condition <- rep(c("face", "house", "tool"), times = 8L)
design <- stats::model.matrix(~ 0 + factor(condition))
colnames(design) <- levels(factor(condition))
design <- cbind(design, drift = as.numeric(scale(seq_len(n_scans))))

store <- file.path(tempdir(), "runstore")
dir.create(store, showWarnings = FALSE)
paths <- vapply(runs, function(run) {
  coefficients <- matrix(
    rnorm(ncol(design) * n_features, sd = 0.2), ncol(design), n_features,
    dimnames = list(colnames(design), NULL)
  )
  coefficients["face", signal_features] <-
    coefficients["face", signal_features] + 1.5
  responses <- design %*% coefficients +
    matrix(rnorm(n_scans * n_features, sd = 0.5), n_scans, n_features)
  path <- file.path(store, paste0(run, ".bin"))
  writeBin(as.vector(responses), path, size = 8L)
  path
}, character(1))
basename(paths)
```

### Data-in: sources and capabilities

`runstore_sources()` is the entire storage adapter. Because every source is a
descriptor, `relation()` will derive the capability record itself — including
the content hash that lands in the execution receipt.

```{r sources}
runstore_sources <- function(paths, dim) {
  sources <- lapply(paths, file_matrix_source, dim = dim)
  names(sources) <- names(paths)
  sources
}

sources <- runstore_sources(paths, dim = c(n_scans, n_features))
c(kind = sources[[1L]]$kind, access = sources[[1L]]$access)
```

### Design-in: your own estimator

`runstore` fits its own model, so it computes `E` and hands it in rather than
calling `lm_extractor()`. Two contrasts are requested on the design's
coefficient axis.

```{r extractor}
runstore_extractor <- function(design, targets) {
  # Your estimator, whatever it is. Here: ordinary least squares, composed
  # with the requested contrasts, giving the effect-by-observation map E.
  pseudoinverse <- solve(crossprod(design), t(design))
  effect_extractor(targets %*% pseudoinverse, estimator = "runstore::ols")
}

targets <- rbind(
  `face-house` = c(1, -1, 0, 0),
  `face-tool` = c(1, 0, -1, 0)
)
colnames(targets) <- colnames(design)

extractor <- runstore_extractor(design, targets)
c(effects = paste(extractor$effects, collapse = ", "),
  observations = extractor$n_observations)
```

### Assembling an ordinary relation

Nothing here is extension-specific: `abstract_domain()` and `relation()` are
core functions a user calls the same way. Recording the adapter and its
version in `provenance` is the convention every in-tree adapter follows.

```{r relation}
domain <- abstract_domain(n_features, id = "runstore-domain")

runstore_relation <- function(paths, design, targets, domain) {
  relation(
    runstore_sources(paths, dim = c(nrow(design), domain$n_features)),
    extract = runstore_extractor(design, targets),
    domain = domain,
    provenance = list(
      adapter = "runstore::runstore_relation",
      adapter_version = "0.3.1"
    )
  )
}

relation_value <- runstore_relation(paths, design, targets, domain)
relation_value$partitions
relation_value$capabilities[["run-1"]]
```

The capability record was derived from the descriptors: `block_read` and
`reopenable` are true because a file matrix supports both, `thread_safe` is
false because crossform will not assume it, and the revision is the content
hash of `run-1.bin`.

### The bounded read

```{r block-read}
round(relation_block(relation_value, "run-1", c(1L, 2L, 39L)), 3)
```

Three columns were read from a forty-column store. The first two features are
planted and the last is not, which the numbers show. Rows come back in
effect-space order regardless of how the store is laid out.

```{r block-refusals}
c(
  bad_feature = tryCatch(
    relation_block(relation_value, "run-1", c(1L, 99L)),
    effect_input_error = conditionMessage
  ),
  bad_partition = tryCatch(
    relation_block(relation_value, "run-9", 1L),
    effect_input_error = conditionMessage
  )
)
```

### Error-channel-in

`runstore` computed point estimates and kept no residuals, so it says so.

```{r fit}
fit <- relation_fit(
  relation_value,
  error_models = NULL,
  provenance = list(
    adapter = "runstore::runstore_relation",
    analytic_error_channel = "withheld"
  )
)
relation_fit_capabilities(fit)[, 1:5]
```

Every statistical capability is `FALSE`, and the refusals follow from that
rather than from a special case. Asking for the residual degrees of freedom
returns a refusal carrying the capability that was missing:

```{r fit-refusal}
refusal <- catch_refusal(residual_df(fit, "run-1"))
c(capability = refusal$capability, namespace = refusal$namespace)
refusal$remedies
```

`sampling_capabilities()` is the ask-before-you-provoke form of the same
question: it returns the unmet requirements as a value, so an adapter can
report them without provoking the refusal.

```{r sampling-capabilities}
frame <- compile_frame(
  regions(ifelse(seq_len(n_features) %in% signal_features,
    "planted", "background")),
  domain
)
plan <- plan_geometry(
  fit$relation,
  at = frame,
  over = cross_partitions(
    fit$relation, independence = "independent", generalizes_over = "run"
  )
)
sampling_capabilities(plan, fit)
```

### An opaque source, and why it needs a declaration

Not every store is a file. If `runstore` gained an in-memory or networked
backend exposed as a closure, the descriptor — and with it the derived
capability record — disappears, and the declaration becomes the adapter's job.

```{r opaque}
opaque <- function(features) {
  matrix(rnorm(n_scans * length(features)), n_scans, length(features))
}
undeclared <- relation(
  list(`run-1` = opaque), extract = extractor,
  source_dims = list(c(n_scans, n_features)), domain = domain
)

# Constructing the relation is fine; executing it is not.
tryCatch(relation_fit(undeclared), effect_input_error = conditionMessage)
```

```{r opaque-declared}
declared <- relation(
  list(`run-1` = opaque), extract = extractor,
  source_dims = list(c(n_scans, n_features)), domain = domain,
  capabilities = source_capabilities(
    block_read = TRUE,
    stable_revision = paste0("sha256:", strrep("a", 64))
  )
)
relation_fit_capabilities(relation_fit(declared))$error_model
```

In a real adapter the revision must actually identify the bytes your source
will return — it is what a receipt is checked against later. A literal
constant like the one above is only acceptable in a vignette.

### It is now an ordinary relation

The point of the protocol is that nothing downstream knows an adapter was
involved. The plan built above compiles and runs:

```{r downstream}
contrast_energy(plan, c(1, 0))
```

The planted region reproduces across runs and the background region does not,
which is the answer the data were built to give.

### Result-out: registering an `as_neurovol()` method

Finally, suppose `runstore` returns its maps in its own class. Registering one
method teaches crossform to write it out; the method delegates to the shipped
numeric method rather than reimplementing the index arithmetic.

```{r as-neurovol, eval=has_neuroim2}
# In a package this is `S3method(as_neurovol, runstore_map)` in NAMESPACE.
as_neurovol.runstore_map <- function(values, mask, ...) {
  as_neurovol(values$statistic, mask, ...)
}

voxels <- array(FALSE, c(4L, 4L, 3L))
voxels[2:3, 2:3, 1:2] <- TRUE
mask <- neuroim2::LogicalNeuroVol(
  voxels, neuroim2::NeuroSpace(c(4L, 4L, 3L), spacing = c(3, 3, 3))
)
volume_domain_value <- neuroim2_volume_domain(mask, id = "runstore-mask")

boxed <- structure(
  list(statistic = seq_len(volume_domain_value$n_features) /
    volume_domain_value$n_features),
  class = "runstore_map"
)
written <- as_neurovol(
  boxed, mask, volume_domain_value, label = "runstore statistic"
)
dim(written)
round(as.numeric(written[volume_domain_value$feature_ids]), 3)
```

```{r as-neurovol-skipped, eval=!has_neuroim2, echo=FALSE, results="asis"}
cat(
  "**neuroim2 (>= 0.19.0) is not installed, so the `as_neurovol()` section",
  "was not run.**\n"
)
```

```{r cleanup, include=FALSE}
unlink(path)
unlink(store, recursive = TRUE)
```

## What is not part of the protocol

The list above is exhaustive, and the exclusions are deliberate.

**The compiler and the executor are closed.** `plan_geometry()`,
`plan_relation()`, and `estimate_relation()` are user-facing verbs, not
extension points. Nothing in `R/compiler.R`, `R/execution-driver.R`,
`R/task.R`, or `R/storage.R` is reachable, and the compiled task IR, the memory
plan, and the reduction schedule are internal by construction. An adapter
supplies inputs and reads outputs; it does not schedule work.

**Kernels are closed.** The numerical kernels (`R/kernel.R`,
`R/measurement-kernel.R`, `R/evidence-sampling-kernel.R`, the crossnobis
execution driver, and the C++ in `src/`) are not an extension surface at any
version. `crossnobis()` is a user verb, not a hook into the kernel behind it. Their identity is recorded in the execution receipt precisely so
that a result can be attributed to one; a third-party kernel would make that
attribution meaningless.

**Everything the tier ledger demotes.** `design/api-tiers.md` moves seven
exports to internal in the subtraction release — `inner_product`,
`measurement_space`, `measurement_bridge`, `effect_covariance`,
`residual_pair_statistics`, `bids_events`, `bids_confounds` — and merges
`reverse_bridge` into `measurement_bridge`. Two of those were developer-tier:
`effect_covariance()` (a design-side factor with no call site anywhere in `R/`
outside its own file — `relation_fit_capabilities()` already answers the only
question anyone asks of it) and `residual_pair_statistics()` (a compiler-stage
accumulator that `plan_crossnobis()` runs internally). Do not build against
them.

**Anything reached with `:::`.** If you cannot get there from `NAMESPACE`, it
is not protocol, and the layering in `design/architecture.md` is not a public
contract either.

### The gaps, stated honestly

The four adapters shipped in this package are not written entirely within the
protocol they document. `tests/testthat/test-adapter-protocol.R` checks that
every crossform function an adapter calls is either an export or a layer-1
internal (argument checks, message formatting, refusal raising, hashing), and
carries an explicit register of the twelve calls that are neither. Those twelve fall into
four groups, and each is a place where an external package would today have to
use `:::` or hand-roll a replacement:

- **fact re-validation** (`.validate_observations()`,
  `.validate_partition_names()`, `.validate_study()`) — an adapter that accepts
  a study or an observations record cannot re-check it through a public call;
- **adapter version pinning** (`.require_adapter_version()`) — the helper that
  turns "the installed fmridesign is not the certified one" into an
  `installed_compiler_adapter` or `supported_compiler_version` refusal;
- **external execution of a plan** (`.validate_relation_plan()`,
  `.planned_observation_sources()`) — the entire seam an external *executor*
  needs, which is why `fmrireg_relation()` cannot be written outside this
  package;
- **spatial provider internals** (`.new_domain()`, `.validate_domain()`,
  `.same_domain_reference()`, `.normalize_frame()`,
  `.support_index_from_members()`, `.support_index_membership()`) — an
  external searchlight provider can reach `abstract_domain()` and
  `additive_frame()`, but not the support-index machinery that gives a frame
  its membership pattern.

The register is a ratchet, not a licence: the test fails on a new unregistered
internal call *and* on a registered one that has been removed, so the list can
only shrink. If you need one of these seams, that is the ticket to open.

## The compatibility promise

`design/ingestion-contract.md` §13 is the binding statement. In short: the
`effect_extractor()`, `lm_extractor()`, and `lm_relation_fit()` APIs remain
valid for explicit linear designs and keep their honest, narrower capabilities;
a future ingestion layer may reimplement them through shared internals but must
not silently grant symbolic provenance or change their numerical estimand.
`plan_geometry()` continues to consume the existing relation-fit output, and no
parallel downstream engine is introduced alongside it. The promise binds
regardless of tier — `lm_extractor()` moving from developer to advanced does
not weaken it.

§12 of the same document states the boundary the other way round: *the
conformance protocol, not a particular package, is the core boundary*. Any
compiler may participate if it emits the same portable receipt and passes
`compiler_conformance()`. The core is not BIDS-shaped, and no adapter's
dependency is mandatory — every one of them is in `Suggests:`.

Two obligations follow for you. Certify against *installed* package versions
and record them, as `fmridesign_design_model()` and `fmrireg_relation()` do
with `.require_adapter_version()`; source-checkout agreement is not evidence.
And declare capabilities you can honour: crossform's refusals are only as good
as the declarations they are computed from, and a source that overstates
`thread_safe`, or a fit that claims an error channel it does not have, converts
a principled refusal into a wrong number.

## Where to look next

- `design/api-tiers.md` — the tier ledger; every export's tier, disposition,
  and the reasoning behind it.
- `design/ingestion-contract.md` — §12 external compiler adapters, §13
  compatibility, §10 required refusals.
- `design/architecture.md` — the layering the internals obey. Read for
  orientation, not as a contract.
- `R/adapter-fmrireg.R`, `R/adapter-fmridesign.R`, `R/adapter-bids.R`,
  `R/neuroim2-adapter.R` — the four worked examples in the tree.
- `tests/testthat/test-adapter-protocol.R` — the executable statement of this
  guide.
