---
title: "Extending crossform: the developer protocol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending crossform: the developer protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_neuroim2 <- requireNamespace("neuroim2", quietly = TRUE) &&
  utils::packageVersion("neuroim2") >= "0.19.0"
library(crossform)
set.seed(20260817)
```

How can another R package supply data or fitted effects to `crossform` while
preserving bounded reads and honest uncertainty? This guide builds a small
adapter from binary response files to a cross-run contrast result. It is for
package authors; analysis users can start with
[From observations](from-observations.html).

You should be familiar with `relation()`, `plan_geometry()`, and the distinction
between a relation's point estimates and its error channel. The example uses
only public functions. Optional volume output requires `neuroim2` >= 0.19.0;
the storage and estimation workflow runs without it.

## Choose the boundary your package needs

| Your package supplies | Public entry points | What to demonstrate |
|---|---|---|
| A response store | `file_matrix_source()`, `source_capabilities()`, `relation_block()` | Stable source identity and bounded feature reads |
| A linear estimator | `effect_extractor()` | The declared effect-by-observation map `E` in `B = E Y` |
| Fitted effects and uncertainty provenance | `relation_fit()` | Which error-channel capabilities are actually available |
| A custom result class | `as_neurovol()` | Correct values and spatial indexing in the returned volume |
| A version-specific compiler adapter | `adapter_version_certificate()` | Installed dependency version matches a separately certified version |

Adapters also use ordinary constructors such as `relation()`, `study()`, and
`volume_domain()`. Build those objects through their public constructors so the
same checks apply to adapter and user inputs. A version certificate records a
tested version; it does not replace numerical or compiler-conformance tests.

The next sections explain the boundaries and their expected failures. Continue
to [the worked adapter](#a-worked-adapter) to see them assembled into one
workflow.

### `file_matrix_source()` — an out-of-memory store

You supply a path, the two-element `dim` (observations by neural features),
and optionally an expected `sha256:` revision. You get an
`effect_source_descriptor`, which is a *description*: no handle is opened, and
the descriptor is serializable, so it survives being sent to a worker.

The constructor checks that the file is exactly `offset_bytes + prod(dim) * 8`
bytes, and computes the content revision itself. If you declare a
`stable_revision` that does not match the bytes on disk, it is an
`effect_contract_error` — two objects disagreeing, not a bad argument:

```{r file-source-refusal}
path <- tempfile(fileext = ".bin")
writeBin(as.vector(matrix(as.double(1:12), 4L, 3L)), path, size = 8L)

revision_failure <- tryCatch(
  file_matrix_source(
    path, dim = c(4L, 3L),
    stable_revision = paste0("sha256:", strrep("0", 64))
  ),
  effect_contract_error = conditionMessage
)
knitr::kable(data.frame(problem = "Revision differs from file contents",
  message = revision_failure))
```

The same check runs again every time the file is reopened for a block read
(`"Reopenable matrix source has a stale content revision."`), so a store that
is silently rewritten between planning and execution is caught rather than
used.

Your own store need not be a flat matrix file. Any function
`function(features) -> matrix` is accepted as a source by `relation()` when you
declare `source_dims`; `file_matrix_source()` is the one concrete descriptor
crossform ships, and the one whose capabilities it can derive for you.

### `source_capabilities()` — what the source can do

This is the declaration the compiler plans against: `block_read`,
`reopenable`, `thread_safe`, and a `stable_revision` that goes into the
execution receipt. A source descriptor implies its own capabilities, so
`relation()` derives them for you when every source is a descriptor. A source
that is an opaque closure implies nothing, and you must say.

Three ways to get it wrong, and what each raises:

```{r capability-refusals}
weak <- tryCatch(
  source_capabilities(TRUE, stable_revision = "2026-08-15"),
  effect_input_error = conditionMessage
)
flags <- tryCatch(
  source_capabilities(block_read = "yes", stable_revision = paste0(
    "sha256:", strrep("a", 64)
  )),
  effect_input_error = conditionMessage
)
knitr::kable(data.frame(problem = c("Weak revision", "Nonlogical flag"),
  message = c(weak, flags)))
```

```{r capability-refusal-reopenable}
# Claiming `reopenable` for a source that has no reopenable descriptor: the
# claim cannot be honoured, because there is nothing to reopen.
source_fn <- function(features) matrix(0, 4L, length(features))
reopen_failure <- tryCatch(
  relation(
    list(`run-1` = source_fn), effects = c("a", "b", "c", "d"),
    source_dims = list(c(4L, 3L)),
    capabilities = source_capabilities(
      block_read = TRUE, reopenable = TRUE,
      stable_revision = paste0("sha256:", strrep("a", 64))
    )
  ),
  effect_input_error = conditionMessage
)
knitr::kable(data.frame(problem = "Opaque source declared reopenable",
  message = reopen_failure))
```

Two declarations are checked when a source reaches execution: opaque sources
must supply capabilities, and every source must support bounded block reads.
An adapter should exercise both failures as well as successful reads.

Note what class these are. A wrong capability declaration is an
`effect_input_error` or an `effect_contract_error`, not an
`effect_capability_refusal`: crossform reserves the refusal class for *"the
interpretation you asked for cannot be earned from the objects you supplied"*,
and a source that lies about itself is a bug in the adapter. Refusals appear
one layer up, at the error channel, and carry `$capability` and `$namespace`
(see `?crossform_conditions`). Both are demonstrated in the worked example
below.

### `relation_block()` — the bounded read

`relation_block(x, partition, features)` is where neural values are actually
read: only the requested feature columns are pulled from the source and mapped
through that partition's extractor. It is on the extension surface for two
reasons — an external executor needs it, and it is the observable that proves
your source is genuinely block-readable rather than a whole-array read wearing
a closure.

Feature indices must be unique and in range
(`"features must be unique valid neural feature indices."`), and the partition
must name one the relation declares
(`"partition must identify one relation partition."`). The returned matrix
always has one row per effect coordinate in effect-space order, whatever order
your source stored things in.

### `effect_extractor()` — the declared map

An extractor is the matrix `E` in `B = E Y`: effect-by-observation, holding no
neural data, and therefore reusable across every feature block. Supply `map`
plus an `estimator` identity string; the constructor requires a finite
nonempty matrix and names the rows from the effect space.

If your estimator is an ordinary linear model, `lm_extractor()` can compile
`E` from a design and target. It checks full design rank and target estimability.
An engine with its own estimator instead computes `E` and passes it to
`effect_extractor()`, as the worked example does.

### `relation_fit()` — the error channel

`relation_fit(relation, error_models, provenance)` wraps a mathematically
intact relation in a separate, capability-bearing statistical envelope. The
relation itself is untouched; what the wrapper adds is a per-partition record
of what uncertainty the fit can support, readable with
`relation_fit_capabilities()`.

**Be precise about what an external package can attach today.** The
`error_models` argument takes crossform's internal error-model values, and no
public constructor for one exists: the separable-GLM model is built inside
`lm_relation_fit()`. So from outside the package, `relation_fit()` attaches an
**explicitly absent** channel — `error_models = NULL` — which is a real and
useful statement, and different from an unstated one. It is exactly what
`fmrireg_relation()` does (`R/adapter-fmrireg.R`), recording
`analytic_error_channel = "withheld"` in its provenance. An engine that has
residuals and wants them honoured routes its raw responses through
`lm_relation_fit()` instead.

The consequence, once declared, is enforced. Every operation that needs the
channel refuses with a capability whose name is the missing column of
`relation_fit_capabilities()`, in namespace `relation_fit`:

| Capability | Refused operation |
|---|---|
| `error_model` | any statistical envelope at all |
| `residual_blocks` | `residual_block()`, learned metrics |
| `effect_covariance` | design-side factor of the separable error model |
| `residual_df` | `residual_df()`, the noise-variance divisor |
| `separable_error` | the separable second-moment law |
| `learned_metric_input` | metric learning from residuals |
| `within_participant_calibration` | analytic within-participant standard errors |

### `as_neurovol()` — writing a result type out

`as_neurovol()` is an S3 generic dispatching on `values`, so a package with its
own result class can teach crossform to write that class to a volume without
crossform importing it. Register the method the ordinary way —
`S3method(as_neurovol, my_result)` in your `NAMESPACE` — and it receives
`mask`, `domain`, `fill`, and `label` unchanged. The generic validates
nothing, so your method may take different arguments; it is expected to return
a `NeuroVol` on the space of `mask`. crossform ships the numeric-vector method
and a `default` that behaves identically.

## A worked adapter

Suppose you maintain `runstore`, a package that keeps one response matrix per
run in a binary file store and fits its own first-level model. This section
builds it against the protocol, with synthetic data so it runs here.

### The store

Three runs, twenty-four scans, forty features. Eight of the forty carry a real
`face` effect; the rest are noise. Each run is written to disk in R's
column-major order as consecutive doubles — the layout `file_matrix_source()`
requires.

```{r store}
n_scans <- 24L
n_features <- 40L
runs <- c("run-1", "run-2", "run-3")
signal_features <- 1:8

condition <- rep(c("face", "house", "tool"), times = 8L)
design <- stats::model.matrix(~ 0 + factor(condition))
colnames(design) <- levels(factor(condition))
design <- cbind(design, drift = as.numeric(scale(seq_len(n_scans))))

store <- tempfile("runstore-")
dir.create(store, showWarnings = FALSE)
paths <- vapply(runs, function(run) {
  coefficients <- matrix(
    rnorm(ncol(design) * n_features, sd = 0.2), ncol(design), n_features,
    dimnames = list(colnames(design), NULL)
  )
  coefficients["face", signal_features] <-
    coefficients["face", signal_features] + 1.5
  responses <- design %*% coefficients +
    matrix(rnorm(n_scans * n_features, sd = 0.5), n_scans, n_features)
  path <- file.path(store, paste0(run, ".bin"))
  writeBin(as.vector(responses), path, size = 8L)
  path
}, character(1))
basename(paths)
```

### Data-in: sources and capabilities

`runstore_sources()` is the entire storage adapter. Because every source is a
descriptor, `relation()` will derive the capability record itself — including
the content hash that lands in the execution receipt.

```{r sources}
runstore_sources <- function(paths, dim) {
  sources <- lapply(paths, file_matrix_source, dim = dim)
  names(sources) <- names(paths)
  sources
}

sources <- runstore_sources(paths, dim = c(n_scans, n_features))
c(kind = sources[[1L]]$kind, access = sources[[1L]]$access)
```

### Design-in: your own estimator

`runstore` fits its own model, so it computes `E` and hands it in rather than
calling `lm_extractor()`. Two contrasts are requested on the design's
coefficient axis.

```{r extractor}
runstore_extractor <- function(design, targets) {
  # Your estimator, whatever it is. Here: ordinary least squares, composed
  # with the requested contrasts, giving the effect-by-observation map E.
  pseudoinverse <- solve(crossprod(design), t(design))
  effect_extractor(targets %*% pseudoinverse, estimator = "runstore::ols")
}

targets <- rbind(
  `face-house` = c(1, -1, 0, 0),
  `face-tool` = c(1, 0, -1, 0)
)
colnames(targets) <- colnames(design)

extractor <- runstore_extractor(design, targets)
c(effects = paste(extractor$effects, collapse = ", "),
  observations = extractor$n_observations)
```

### Assembling an ordinary relation

Nothing here is extension-specific: `abstract_domain()` and `relation()` are
core functions a user calls the same way. Recording the adapter and its
version in `provenance` is the convention every in-tree adapter follows.

```{r relation}
domain <- abstract_domain(n_features, id = "runstore-domain")

runstore_relation <- function(paths, design, targets, domain) {
  relation(
    runstore_sources(paths, dim = c(nrow(design), domain$n_features)),
    extract = runstore_extractor(design, targets),
    domain = domain,
    provenance = list(
      adapter = "runstore::runstore_relation",
      adapter_version = "0.3.1"
    )
  )
}

relation_value <- runstore_relation(paths, design, targets, domain)
relation_value$partitions
relation_value$capabilities[["run-1"]]
```

The capability record was derived from the descriptors: `block_read` and
`reopenable` are true because a file matrix supports both, `thread_safe` is
false because crossform will not assume it, and the revision is the content
hash of `run-1.bin`.

### The bounded read

```{r block-read}
round(relation_block(relation_value, "run-1", c(1L, 2L, 39L)), 3)
```

Three columns were read from a forty-column store. The first two features are
planted and the last is not, which the numbers show. Rows come back in
effect-space order regardless of how the store is laid out.

```{r block-refusals}
block_failures <- c(
  bad_feature = tryCatch(
    relation_block(relation_value, "run-1", c(1L, 99L)),
    effect_input_error = conditionMessage
  ),
  bad_partition = tryCatch(
    relation_block(relation_value, "run-9", 1L),
    effect_input_error = conditionMessage
  )
)
knitr::kable(data.frame(problem = names(block_failures),
  message = unname(block_failures)))
```

### Error-channel-in

`runstore` computed point estimates and kept no residuals, so it says so.

```{r fit}
fit <- relation_fit(
  relation_value,
  error_models = NULL,
  provenance = list(
    adapter = "runstore::runstore_relation",
    analytic_error_channel = "withheld"
  )
)
relation_fit_capabilities(fit)[, 1:5]
```

Every statistical capability is `FALSE`, and the refusals follow from that
rather than from a special case. Asking for the residual degrees of freedom
returns a refusal carrying the capability that was missing:

```{r fit-refusal}
refusal <- catch_refusal(residual_df(fit, "run-1"))
c(capability = refusal$capability, namespace = refusal$namespace)
knitr::kable(data.frame(remedy = refusal$remedies))
```

`sampling_capabilities()` is the ask-before-you-provoke form of the same
question: it returns the unmet requirements as a value, so an adapter can
report them without provoking the refusal.

```{r sampling-capabilities}
frame <- compile_frame(
  regions(ifelse(seq_len(n_features) %in% signal_features,
    "planted", "background")),
  domain
)
plan <- plan_geometry(
  fit$relation,
  at = frame,
  over = cross_partitions(
    fit$relation, independence = "independent", generalizes_over = "run"
  )
)
sampling <- sampling_capabilities(plan, fit)
sampling$available
knitr::kable(sampling$reasons, col.names = c("Requirement", "Why unavailable", "Remedy"))
```

### An opaque source, and why it needs a declaration

Not every store is a file. If `runstore` gained an in-memory or networked
backend exposed as a closure, the descriptor — and with it the derived
capability record — disappears, and the declaration becomes the adapter's job.

```{r opaque}
opaque_values <- matrix(
  readBin(paths[[1L]], what = "double", n = n_scans * n_features),
  nrow = n_scans
)
opaque <- local({
  snapshot <- opaque_values
  function(features) snapshot[, features, drop = FALSE]
})
undeclared <- relation(
  list(`run-1` = opaque), extract = extractor,
  source_dims = list(c(n_scans, n_features)), domain = domain
)

# Constructing the relation is fine; executing it is not.
missing_declaration <- tryCatch(
  relation_fit(undeclared), effect_input_error = conditionMessage
)
knitr::kable(data.frame(message = missing_declaration))
```

The closure holds a fixed in-memory snapshot of the first file. Its revision
can therefore use the descriptor's content hash; the function returns selected
columns from that snapshot on every call.

```{r opaque-declared}
declared <- relation(
  list(`run-1` = opaque), extract = extractor,
  source_dims = list(c(n_scans, n_features)), domain = domain,
  capabilities = source_capabilities(
    block_read = TRUE,
    stable_revision = sources[[1L]]$stable_revision
  )
)
relation_fit_capabilities(relation_fit(declared))$error_model
```

The declaration admits bounded reads, but does not claim that the closure is
reopenable or thread-safe. A different backend must establish those properties
before advertising them.

### It is now an ordinary relation

The point of the protocol is that nothing downstream knows an adapter was
involved. The plan built above compiles and runs:

```{r downstream}
contrast_energy(plan, c(1, 0))
```

Compare the planted region with the background: the generated face effect
should produce stronger cross-run evidence in the planted region. This is a
point-estimate check of the adapter workflow; the fit has declared no analytic
error channel.

### Result-out: registering an `as_neurovol()` method

Finally, suppose `runstore` returns its maps in its own class. Registering one
method teaches crossform to write it out; the method delegates to the shipped
numeric method rather than reimplementing the index arithmetic.

```{r as-neurovol, eval=has_neuroim2}
# In a package this is `S3method(as_neurovol, runstore_map)` in NAMESPACE.
as_neurovol.runstore_map <- function(values, mask, ...) {
  as_neurovol(values$statistic, mask, ...)
}

voxels <- array(FALSE, c(4L, 4L, 3L))
voxels[2:3, 2:3, 1:2] <- TRUE
mask <- neuroim2::LogicalNeuroVol(
  voxels, neuroim2::NeuroSpace(c(4L, 4L, 3L), spacing = c(3, 3, 3))
)
volume_domain_value <- neuroim2_volume_domain(mask, id = "runstore-mask")

boxed <- structure(
  list(statistic = seq_len(volume_domain_value$n_features) /
    volume_domain_value$n_features),
  class = "runstore_map"
)
written <- as_neurovol(
  boxed, mask, volume_domain_value, label = "runstore statistic"
)
dim(written)
round(as.numeric(written[volume_domain_value$feature_ids]), 3)
```

```{r as-neurovol-skipped, eval=!has_neuroim2, echo=FALSE, results="asis"}
cat(
  "**neuroim2 (>= 0.19.0) is not installed, so the `as_neurovol()` section",
  "was not run.**\n"
)
```

```{r cleanup, include=FALSE}
unlink(path)
unlink(store, recursive = TRUE)
```

## Stay on the public extension boundary

An adapter supplies inputs and consumes results. The compiler's task
representation, execution schedule, storage internals, and numerical kernels
are not extension hooks. Call public planning and estimation functions when
needed; do not modify their internal representations or call private functions
through `:::`.

Several public constructors support specialized adapters without private
helpers:

| Adapter task | Public route |
|---|---|
| Validate an external compiler's plan and inspect its records | `relation_plan_receipts()` and `compiler_conformance()` |
| Describe a provider's spatial geometry and provenance | `volume_domain(metadata = ...)` |
| Supply precomputed neighbourhood membership | `additive_frame(members = ...)` |
| Inspect a study's admitted behavior | `study_capabilities()` |

A spatial provider supplies membership and metadata; the constructors apply
frame normalization and establish domain identity. A compiler adapter validates
its plan and builds readers over the retained observations. See the
[ingestion contract](https://github.com/bbuchsbaum/crossform/blob/main/design/ingestion-contract.md)
for the portable receipt schema, and the
[API tier ledger](https://github.com/bbuchsbaum/crossform/blob/main/design/api-tiers.md)
for the current public boundaries.

## The compatibility promise

The [ingestion contract](https://github.com/bbuchsbaum/crossform/blob/main/design/ingestion-contract.md)
§13 defines the compatibility promise: the
`effect_extractor()`, `lm_extractor()`, and `lm_relation_fit()` APIs remain
valid for explicit linear designs and keep their honest, narrower capabilities;
a future ingestion layer may reimplement them through shared internals but must
not silently grant symbolic provenance or change their numerical estimand.
`plan_geometry()` continues to consume the existing relation-fit output, and no
parallel downstream engine is introduced alongside it. This promise applies to the public functions regardless of their API tier.

§12 of the same document states the boundary the other way round: *the
conformance protocol, not a particular package, is the core boundary*. Any
compiler may participate if it emits the same portable receipt and passes
`compiler_conformance()`. The core is not BIDS-shaped, and no adapter's
dependency is mandatory — every one of them is in `Suggests:`.

Before releasing an adapter, test it against the installed dependency version
you will support. Record that version with `adapter_version_certificate()` so
unsupported installations produce the standard capability refusal. Then check
that successful bounded reads and failed capability requests match your
declarations. In particular, verify any claim of thread safety and any error
channel used to compute uncertainty.

## Where to look next

- `design/api-tiers.md` — the tier ledger; every export's tier, disposition,
  and the reasoning behind it.
- `design/ingestion-contract.md` — §12 external compiler adapters, §13
  compatibility, §10 required refusals.
- `design/architecture.md` — the layering the internals obey. Read for
  orientation, not as a contract.
- `R/adapter-fmrireg.R`, `R/adapter-fmridesign.R`, `R/adapter-bids.R`,
  `R/neuroim2-adapter.R` — the four worked examples in the tree.
- `tests/testthat/test-adapter-protocol.R` — the executable statement of this
  guide.
