---
title: "Measure effects within and between neural nodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measure effects within and between neural nodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

Use this guide when your result must relate two neural measurements to each
other, rather than report one number at each. Two regions, or two multivariate
populations, are the usual cases.

A **measurement** is the same spatial unit the other guides report at: a region,
a searchlight, a voxel, or the whole brain. This guide also calls a measurement
a **node** whenever it is one endpoint of a pair, because that is how the
functions name it (`node_ids`, `edge_frame`). The two words denote the same
object.

The inputs are repeated experimental observations, a fixed measurement operator
for each node, a list of requested node pairs, and an experimental query: a
fixed matrix that weights time points, conditions, or other experimental
coordinates. `measurement_form()` returns one matrix block for every requested
pair. A diagonal block describes one node; an off-diagonal block relates two
nodes.

The later sections apply several views to those blocks. `connectivity()` can
report an ordinary scalar correlation, `canonical_coupling()` reports a
multivariate correlation spectrum, and `geometry_alignment()` compares two
within-node geometries. Each view checks the assumptions it needs. Use
`effect_coupling()` when you need the matrix block itself without claiming that
it is covariance or connectivity.

## Can two regional signals recover ordinary correlation?

The hidden setup creates two sessions. In each matrix, eight repeated time
points are rows and four native neural features are columns. The first two
features are negatively related, which gives this example a known correlation
sign.

```{r scalar-data, include=FALSE}
sample_space <- effect_space(
  paste0("time", seq_len(8)), basis_id = "demo:time:v1"
)
native <- abstract_domain(4, id = "demo:native:v1")
trend <- seq(-1.75, 1.75, length.out = 8)
session1 <- cbind(
  trend,
  -0.85 * trend + c(0.2, -0.1, 0.1, 0, -0.1, 0.1, -0.2, 0.1),
  sin(seq(0, pi, length.out = 8)),
  cos(seq(0, pi, length.out = 8))
)
session2 <- cbind(
  trend + c(0.1, 0, -0.1, 0.1, 0, -0.1, 0, 0.1),
  -0.75 * trend + c(-0.1, 0.2, 0, -0.1, 0.1, 0, 0.1, -0.2),
  sin(seq(0.1, pi + 0.1, length.out = 8)),
  cos(seq(0.1, pi + 0.1, length.out = 8))
)
signals <- relation(
  list(session1 = session1, session2 = session2),
  effects = sample_space,
  domain = native
)
```

First declare two scalar measurements. Each one selects a single oriented
feature here; a regional mean would use one row of fixed weights over several
features.

```{r scalar-frame}
nodes <- measurement_frame(
  list(
    anterior = matrix(c(1, 0, 0, 0), 1),
    posterior = matrix(c(0, 1, 0, 0), 1)
  ),
  domain = native,
  id = "demo:regional-means:v1"
)

node_pairs <- expand.grid(
  from = c("anterior", "posterior"),
  to = c("anterior", "posterior"),
  stringsAsFactors = FALSE
)
between_nodes <- edge_frame(node_pairs$from, node_pairs$to, nodes)
```

`edge_frame()` computes only the pairs listed in `node_pairs`. This example
requests the two self-pairs and the two directed cross-pairs. Correlation needs
the self-pairs to obtain the two variances in its denominator. For a seed map
or selected network, list the scientific edges you need and the self-pairs
required by the chosen normalization.

The matrix `center / 7` centers the eight time points and divides their sum of
products by seven. `variation_query()` records that operation on the time
axis. The `joint_covariance` construction is valid here because both sides of
each product use the same session and the same fitted relation.

```{r scalar-form}
center <- diag(8) - matrix(1 / 8, 8, 8)
sample_covariance <- variation_query(
  center / 7,
  sample_space,
  sampling_axis = "time",
  construction = "joint_covariance",
  provenance = list(estimator = "centered within session")
)
self_products <- pairing(
  signals$partitions,
  signals$partitions,
  directed = TRUE,
  self_pairs = "allow_biased",
  independence = "not_independent"
)

measured <- measurement_form(
  left = signals,
  between = between_nodes,
  by = sample_covariance,
  over = self_products
)
```

`self_pairs = "allow_biased"` permits each session to multiply by itself; this
is ordinary within-session covariance, not a crossvalidated estimator.
`independence = "not_independent"` prevents the result from claiming that the
two sides were independently estimated. `measurement_form()` averages the two
session products and returns the four requested blocks.

Because both node measurements are scalar and oriented, the correlation view
returns ordinary signed Pearson correlations.

```{r scalar-correlation}
functional <- connectivity(measured, view = "correlation")
data.frame(
  from = measured$block_index$left,
  to = measured$block_index$right,
  correlation = functional$values$correlation
)
```

```{r scalar-check, include=FALSE}
expected_covariance <- (
  crossprod(session1, (center / 7) %*% session1) +
  crossprod(session2, (center / 7) %*% session2)
) / 2
expected_correlation <- expected_covariance[1, 2] /
  sqrt(expected_covariance[1, 1] * expected_covariance[2, 2])
cross_row <- measured$block_index$left == "anterior" &
  measured$block_index$right == "posterior"
observed_correlation <- functional$values$correlation[
  functional$values$edge_id == measured$block_index$edge_id[cross_row]
]
stopifnot(
  is.finite(observed_correlation),
  observed_correlation < 0,
  abs(observed_correlation - expected_correlation) < 1e-12
)
```

The negative cross-node value matches the generated data. Its sign is defined
because each scalar measurement has a fixed orientation. A multivariate node
can be rotated without changing its represented subspace, so individual
cross-block entries are not invariant there. The next section uses summaries
that do not depend on an arbitrary basis rotation.

## What changes when each node retains several modes?

Use two fixed feature selectors as multivariate nodes, then request the same
four blocks under the same experimental query.

```{r multivariate-form}
populations <- measurement_frame(
  list(
    anterior = diag(4)[1:2, , drop = FALSE],
    posterior = diag(4)[3:4, , drop = FALSE]
  ),
  domain = native,
  id = "demo:multivariate-populations:v1"
)
population_pairs <- edge_frame(
  node_pairs$from, node_pairs$to, populations
)
population_form <- measurement_form(
  signals, population_pairs, sample_covariance, self_products
)
```

```{r population-edge, include=FALSE}
cross_population_edge <- population_form$block_index$edge_id[
  population_form$block_index$left == "anterior" &
    population_form$block_index$right == "posterior"
]
```

Canonical coupling uses the two self-covariance blocks to scale the cross-node
block, then reports its singular values. These canonical correlations describe
the shared modes in descending order. `ridge = 0.05` stabilizes the two inverse
covariance calculations and is recorded because changing it changes the
reported values.

```{r canonical-view}
canonical <- canonical_coupling(population_form, ridge = 0.05)
canonical$values[
  canonical$values$edge_id == cross_population_edge,
  c("mode", "canonical_correlation")
]
```

`geometry_alignment()` answers a different question: do the two populations
induce similar pairwise geometry over the repeated observations? It computes a
linear CKA/RV-like alignment. Under the joint-covariance construction used
here, the result lies between zero and one.

```{r alignment-view}
alignment <- geometry_alignment(population_form)
alignment$values$geometry_alignment[
  alignment$values$edge_id == cross_population_edge
]
```

```{r multivariate-check, include=FALSE}
stopifnot(
  all(is.finite(canonical$values$canonical_correlation)),
  all(canonical$values$canonical_correlation >= 0),
  all(canonical$values$canonical_correlation <= 1),
  all(is.finite(alignment$values$geometry_alignment)),
  all(alignment$values$geometry_alignment >= 0),
  all(alignment$values$geometry_alignment <= 1)
)
```

Gaussian mutual information treats the normalized canonical modes as arising
from a joint Gaussian model. The call therefore requires an explicit model
declaration and a choice of information units.

```{r gaussian-view}
gaussian_model <- gaussian_covariance_model(
  list(assumption = "joint Gaussian time observations")
)
information <- connectivity(
  population_form,
  view = "gaussian_information",
  ridge = 0.05,
  model = gaussian_model,
  units = "bits"
)
information$values[
  information$values$edge_id == cross_population_edge,
  c("information", "units")
]
```

```{r gaussian-check, include=FALSE}
stopifnot(
  identical(information$units, "bits"),
  all(is.finite(information$values$information)),
  all(information$values$information >= 0)
)
```

## Why can one contrast not estimate connectivity?

A contrast vector $c$ defines the rank-one query $H=cc^\top$. This query can
measure whether two nodes express the same contrast, but it leaves only one
experimental direction. Its neural block is an outer product, so a normalized
correlation would equal one whenever both node effects are nonzero. The
repeated variation needed to estimate connectivity has been removed.

```{r rank-one-effect}
effect_direction <- c(-1, -1, -1, -1, 1, 1, 1, 1)
rank_one_query <- variation_query(
  tcrossprod(effect_direction),
  sample_space,
  sampling_axis = "time",
  construction = "joint_covariance"
)
rank_one_form <- measurement_form(
  signals, between_nodes, rank_one_query, self_products
)
rank_one_effect <- effect_coupling(rank_one_form)
c(
  effective_rank = rank_one_form$diagnostics$experimental_effective_rank,
  edge_blocks = length(rank_one_effect$values)
)
```

```{r rank-one-check, include=FALSE}
rank_one_failure <- try(
  connectivity(rank_one_form, view = "correlation"),
  silent = TRUE
)
stopifnot(
  rank_one_form$diagnostics$experimental_effective_rank == 1L,
  inherits(rank_one_failure, "try-error")
)
```

`effect_coupling()` returns the block because it makes no covariance claim.
`connectivity(..., view = "correlation")` stops with an error because the
experimental query has effective rank one. A matrix with two neural endpoints
is therefore not sufficient evidence for connectivity.

## How do coherent and configuration modes cross on an edge?

An additive measurement has a positively oriented weighted-mean direction,
called its coherent mode. The orthogonal remainder is its configuration
subspace. When both nodes use this decomposition, their cross-node block has
four components: coherent-to-coherent, coherent-to-configuration,
configuration-to-coherent, and configuration-to-configuration.

```{r decomposition-form}
additive <- additive_frame(
  matrix(c(
    1, 2, 1, 0,
    0, 1, 2, 1
  ), 2, 4, byrow = TRUE),
  domain = native
)
decomposed_nodes <- measurement_frame(
  additive, mode = "coherent_configuration"
)
decomposed_edge <- edge_frame(
  decomposed_nodes$node_ids[1],
  decomposed_nodes$node_ids[2],
  decomposed_nodes
)
decomposed_form <- measurement_form(
  signals, decomposed_edge, sample_covariance, self_products
)
components <- measurement_components(decomposed_form, edge = 1)
components[, c(
  "left_component", "right_component", "raw_entries_meaningful",
  "frobenius_strength"
)]
```

```{r decomposition-check, include=FALSE}
stopifnot(
  nrow(components) == 4L,
  identical(components$raw_entries_meaningful,
    c(TRUE, FALSE, FALSE, FALSE)),
  all(is.finite(components$frobenius_strength)),
  all(components$frobenius_strength >= 0)
)
```

The coherent direction has a fixed orientation because the measurement weights
define its positive direction. The configuration subspace is fixed, but any
orthonormal basis within it is interchangeable. `measurement_components()`
therefore reports rotation-invariant Frobenius strengths for components that
involve configuration. Frobenius strength is the square root of the sum of the
block's squared entries. The function does not present individual
configuration-to-configuration entries as uniquely defined quantities.

## How do encoding and retrieval use different axes?

`measurement_form()` accepts separate left and right relations. In this
example, the left relation has two encoding conditions over three neural
features. The right relation has three retrieval conditions over four neural
features. `pair_query()` therefore uses a rectangular 2 by 3 experimental
query, while `edge_frame()` relates a scalar encoding seed to a two-mode
retrieval target.

```{r cross-domain-setup, include=FALSE}
encoding_space <- effect_space(
  c("encode_a", "encode_b"), basis_id = "demo:encoding:v1"
)
retrieval_space <- effect_space(
  c("retrieve_a", "retrieve_b", "lure"),
  basis_id = "demo:retrieval:v1"
)
encoding_domain <- abstract_domain(3, id = "demo:encoding-neural:v1")
retrieval_domain <- abstract_domain(4, id = "demo:retrieval-neural:v1")
encoding_values <- matrix(c(1, 0, 2, 0, 1, -1), 2, 3, byrow = TRUE)
retrieval_values <- matrix(c(
  1, 0, 0, 1,
  0, 1, 1, 0,
  -1, 0, 1, 1
), 3, 4, byrow = TRUE)
encoding <- relation(
  list(encoding_run = encoding_values),
  effects = encoding_space,
  domain = encoding_domain
)
retrieval <- relation(
  list(retrieval_run = retrieval_values),
  effects = retrieval_space,
  domain = retrieval_domain
)
```

```{r cross-domain-form}
encoding_nodes <- measurement_frame(
  list(seed = matrix(c(1, 0, 1), 1)),
  encoding_domain,
  id = "demo:encoding-node:v1"
)
retrieval_nodes <- measurement_frame(
  list(target = matrix(c(1, 0, 0, 1, 0, 1, 1, 0), 2, 4)),
  retrieval_domain,
  id = "demo:retrieval-node:v1"
)
cross_node <- edge_frame(
  "seed", "target", encoding_nodes, to_frame = retrieval_nodes
)
cross_query <- pair_query(
  matrix(c(1, -0.5, 0, 0, 0.5, -1), 2, 3, byrow = TRUE),
  encoding_space,
  retrieval_space
)
cross_form <- measurement_form(
  left = encoding,
  right = retrieval,
  between = cross_node,
  by = cross_query,
  over = pairing("encoding_run", "retrieval_run", directed = TRUE)
)
effect_coupling(cross_form)$values[[1]]
```

```{r cross-domain-check, include=FALSE}
expected_cross <- encoding_nodes$legs$seed$operator %*%
  crossprod(encoding_values,
    as.matrix(cross_query$operator) %*% retrieval_values) %*%
  t(retrieval_nodes$legs$target$operator)
stopifnot(isTRUE(all.equal(
  effect_coupling(cross_form)$values[[1]],
  expected_cross,
  tolerance = 1e-12
)))
```

The returned matrix is 1 by 2. Its rows belong to the encoding seed and its
columns belong to the retrieval target. No square RDM is constructed because
the two experimental axes and the two neural measurements are different.

## Did the node and edge blocks retain all neural evidence?

`reconstruct_evidence()` tests whether the measured blocks retain the complete
global neural operator or only a projection of it. Lossless reconstruction
requires every directed node pair and a measurement frame with full column
rank. This example uses a Parseval frame: stacking the two node operators gives
the identity, so the local blocks can be added back without a correction
matrix.

```{r parseval-form}
parseval_nodes <- measurement_frame(
  list(
    first_half = diag(4)[1:2, , drop = FALSE],
    second_half = diag(4)[3:4, , drop = FALSE]
  ),
  native,
  id = "demo:parseval:v1"
)
parseval_pairs <- expand.grid(
  from = parseval_nodes$node_ids,
  to = parseval_nodes$node_ids,
  stringsAsFactors = FALSE
)
parseval_edges <- edge_frame(
  parseval_pairs$from, parseval_pairs$to, parseval_nodes
)
parseval_form <- measurement_form(
  signals, parseval_edges, sample_covariance, self_products
)
reconstructed <- reconstruct_evidence(parseval_form, parseval_edges)
c(
  method = reconstructed$method,
  status = reconstructed$status,
  lossless = reconstructed$lossless
)
```

```{r parseval-check, include=FALSE}
expected_global <- expected_covariance
stopifnot(
  reconstructed$lossless,
  identical(reconstructed$method, "parseval"),
  max(abs(reconstructed$operator - expected_global)) < 1e-12
)
```

A full-rank non-Parseval frame reconstructs through its canonical dual, a
correction that reverses overlapping or non-orthogonal frame weights. A
rank-deficient frame returns `status = "projected_reconstruction"` and
`lossless = FALSE`, which tells you that the measurements discarded neural
directions. The function refuses a lossless claim when edges are missing,
frame bases disagree, the reconstruction is too ill-conditioned, or the
declared workspace is too small. Diagonal node blocks alone are insufficient:
two global operators can share the same diagonal blocks and differ between
nodes.

## Which interpretation is permitted?

| View | Minimum interpretation contract |
|---|---|
| `effect_coupling()` | Any supported fixed experimental query; no covariance claim |
| `covariance_coupling()` | Repeated variation, effective rank above one, a joint-covariance construction, and valid self-blocks |
| `connectivity(..., "correlation")` | The covariance requirements plus scalar, oriented measurement axes |
| `canonical_coupling()` | The covariance requirements plus explicit ridge regularization |
| `geometry_alignment()` | Joint covariance and nonzero self-geometries; static normalization over form entries |
| Gaussian information | Canonical coupling plus an explicit joint Gaussian model and information units |
| `reconstruct_evidence()` | Every directed edge, matching frame bases, sufficient workspace, and acceptable frame rank and conditioning |

Each result records which axis was normalized. Pattern correlation normalizes
neural features before it produces an experimental form. Functional
correlation normalizes repeated experimental samples before it interprets a
neural form. The two operations both use the word correlation, but they act on
different axes and answer different questions.

The measurement workflow first averages the requested partition products and
then applies a connectivity normalization. Normalizing or Fisher-transforming
each partition pair before averaging would define a different estimator, so
the package does not substitute that order. The result identity records the
source order, view order, regularization, units, and sampling axis.

## Is geometry alignment informational connectivity?

No. `geometry_alignment()` compares static Gram matrices and is a
rotation-invariant multivariate generalization of squared scalar correlation.
Informational-connectivity analyses often correlate dynamic discriminability
series estimated across time or trials instead. Estimating those series
requires training and cross-fitting stages that this fixed-form package does
not implement.

Choose the function by the question:

- use `effect_coupling()` for an arbitrary fixed query;
- use `connectivity()` only when repeated variation and covariance capabilities
  are present;
- call static CKA/RV-like comparison `geometry_alignment()`;
- keep dynamic informational-connectivity estimators in an adjacent system
  until their training, cross-fitting, and inference contracts are explicit.

For contrasts, RDMs, RSA, and geometry spectra within one experimental space,
continue with `vignette("introduction", package = "crossform")`. Use the
measurement-form functions in this guide only when the returned object must
retain explicit neural node and edge blocks.
