---
title: "Measure node effects and coupling with one evidence form"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measure node effects and coupling with one evidence form}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(effectagram)
```

Suppose you have repeated time, trial, or subject observations over native
neural features. You want to measure selected voxels or populations, retain
their multivariate structure, and ask how evidence is organized within and
between them. The hard part is not multiplying matrices; it is keeping the
experimental sample axis, neural measurement axes, covariance assumptions,
and normalization order honest.

This workflow produces one complete measurement-space form. Its diagonal
blocks are within-node evidence and its off-diagonal blocks are coupling. A
view can then report signed scalar correlation, a canonical spectrum, static
geometry alignment, or Gaussian information only when the form has the
required capabilities.

## Can two regional signals recover ordinary correlation?

The rows below are eight repeated observations and the columns are four native
neural features. Two sessions remain separate and are combined as declared
self-products.

```{r scalar-data, include=FALSE}
sample_space <- effect_space(
  paste0("time", seq_len(8)), basis_id = "demo:time:v1"
)
native <- abstract_domain(4, id = "demo:native:v1")
trend <- seq(-1.75, 1.75, length.out = 8)
session1 <- cbind(
  trend,
  -0.85 * trend + c(0.2, -0.1, 0.1, 0, -0.1, 0.1, -0.2, 0.1),
  sin(seq(0, pi, length.out = 8)),
  cos(seq(0, pi, length.out = 8))
)
session2 <- cbind(
  trend + c(0.1, 0, -0.1, 0.1, 0, -0.1, 0, 0.1),
  -0.75 * trend + c(-0.1, 0.2, 0, -0.1, 0.1, 0, 0.1, -0.2),
  sin(seq(0.1, pi + 0.1, length.out = 8)),
  cos(seq(0.1, pi + 0.1, length.out = 8))
)
signals <- relation(
  list(session1 = session1, session2 = session2),
  effects = sample_space,
  domain = native
)
```

First declare two rank-one measurements. Their supports could contain many
features, but each retains one oriented output mode.

```{r scalar-frame}
nodes <- measurement_frame(
  list(
    anterior = matrix(c(1, 0, 0, 0), 1),
    posterior = matrix(c(0, 1, 0, 0), 1)
  ),
  domain = native,
  id = "demo:regional-means:v1"
)

node_pairs <- expand.grid(
  from = c("anterior", "posterior"),
  to = c("anterior", "posterior"),
  stringsAsFactors = FALSE
)
between_nodes <- edge_frame(node_pairs$from, node_pairs$to, nodes)
```

`edge_frame()` never invents an all-pairs graph. All four pairs are requested
here because correlation needs both self-blocks as denominators. For a seed
map, local-neighbour graph, or selected network, list only those scientifically
motivated edges plus the self-blocks needed by the intended normalization.

The centered sample operator supplies repeated variation. Calling it a joint
covariance is a construction claim: the same relation and the same partition
occur on both sides of each product.

```{r scalar-form}
center <- diag(8) - matrix(1 / 8, 8, 8)
sample_covariance <- variation_query(
  center / 7,
  sample_space,
  sampling_axis = "time",
  construction = "joint_covariance",
  provenance = list(estimator = "centered within session")
)
self_products <- pairing(
  signals$partitions,
  signals$partitions,
  directed = TRUE,
  self_pairs = "allow_biased",
  independence = "not_independent"
)

measured <- measurement_form(
  left = signals,
  between = between_nodes,
  by = sample_covariance,
  over = self_products
)
```

The quick result is ordinary signed Pearson correlation on rank-one
measurement axes.

```{r scalar-correlation}
functional <- connectivity(measured, view = "correlation")
functional$values[, c("edge_id", "correlation")]
```

```{r scalar-check, include=FALSE}
expected_covariance <- (
  crossprod(session1, (center / 7) %*% session1) +
  crossprod(session2, (center / 7) %*% session2)
) / 2
expected_correlation <- expected_covariance[1, 2] /
  sqrt(expected_covariance[1, 1] * expected_covariance[2, 2])
cross_row <- measured$block_index$left == "anterior" &
  measured$block_index$right == "posterior"
observed_correlation <- functional$values$correlation[
  functional$values$edge_id == measured$block_index$edge_id[cross_row]
]
stopifnot(
  is.finite(observed_correlation),
  observed_correlation < 0,
  abs(observed_correlation - expected_correlation) < 1e-12
)
```

The sign survives because each measurement has an oriented scalar mode. For
multivariate populations, arbitrary rotations change raw cross-block entries;
canonical correlations and geometry alignment are the appropriate invariant
summaries.

## What changes when each node retains several modes?

Use two fixed feature selectors as multivariate nodes, then request the same
four blocks under the same experimental query.

```{r multivariate-form}
populations <- measurement_frame(
  list(
    anterior = diag(4)[1:2, , drop = FALSE],
    posterior = diag(4)[3:4, , drop = FALSE]
  ),
  domain = native,
  id = "demo:multivariate-populations:v1"
)
population_pairs <- edge_frame(
  node_pairs$from, node_pairs$to, populations
)
population_form <- measurement_form(
  signals, population_pairs, sample_covariance, self_products
)
```

Canonical coupling whitens the two self-covariances and reports the singular
values of the normalized cross-form. Ridge values are explicit because the
regularization changes the estimand.

```{r canonical-view}
canonical <- canonical_coupling(population_form, ridge = 0.05)
canonical$values[canonical$values$edge_id == "edge_000003", ]
```

Static geometry alignment instead compares the observation geometries induced
by the two populations. It is linear CKA/RV-like alignment and lies between
zero and one for a coherent joint covariance.

```{r alignment-view}
alignment <- geometry_alignment(population_form)
alignment$values
```

```{r multivariate-check, include=FALSE}
stopifnot(
  all(is.finite(canonical$values$canonical_correlation)),
  all(canonical$values$canonical_correlation >= 0),
  all(canonical$values$canonical_correlation <= 1),
  all(is.finite(alignment$values$geometry_alignment)),
  all(alignment$values$geometry_alignment >= 0),
  all(alignment$values$geometry_alignment <= 1)
)
```

Gaussian mutual information adds another assumption. It is available only
after an explicit model declaration and reports its units.

```{r gaussian-view}
gaussian_model <- gaussian_covariance_model(
  list(assumption = "joint Gaussian time observations")
)
information <- connectivity(
  population_form,
  view = "gaussian_information",
  ridge = 0.05,
  model = gaussian_model,
  units = "bits"
)
information$values
```

```{r gaussian-check, include=FALSE}
stopifnot(
  identical(information$units, "bits"),
  all(is.finite(information$values$information)),
  all(information$values$information >= 0)
)
```

## Why is one contrast not connectivity?

A rank-one query such as (H = cc^T) is a valid effect direction. Its
measurement block is an outer product, so normalized coupling would be
perfect whenever both sides are nonzero. The axis whose variation should
estimate connectivity has already been contracted away.

```{r rank-one-effect}
effect_direction <- c(-1, -1, -1, -1, 1, 1, 1, 1)
rank_one_query <- variation_query(
  tcrossprod(effect_direction),
  sample_space,
  sampling_axis = "time",
  construction = "joint_covariance"
)
rank_one_form <- measurement_form(
  signals, between_nodes, rank_one_query, self_products
)
rank_one_effect <- effect_coupling(rank_one_form)
c(
  effective_rank = rank_one_form$diagnostics$experimental_effective_rank,
  edge_blocks = length(rank_one_effect$values)
)
```

```{r rank-one-check, include=FALSE}
rank_one_failure <- try(
  connectivity(rank_one_form, view = "correlation"),
  silent = TRUE
)
stopifnot(
  rank_one_form$diagnostics$experimental_effective_rank == 1L,
  inherits(rank_one_failure, "try-error")
)
```

Use `effect_coupling()` for this algebraic relation. Do not label it covariance
connectivity merely because it has two neural endpoints.

## How do coherent and configuration modes cross on an edge?

An additive measurement can be factored into its positively oriented common
mode and its orthogonal configuration subspace. That node decomposition
automatically induces four edge channels.

```{r decomposition-form}
additive <- additive_frame(
  matrix(c(
    1, 2, 1, 0,
    0, 1, 2, 1
  ), 2, 4, byrow = TRUE),
  domain = native
)
decomposed_nodes <- measurement_frame(
  additive, mode = "coherent_configuration"
)
decomposed_edge <- edge_frame(
  decomposed_nodes$node_ids[1],
  decomposed_nodes$node_ids[2],
  decomposed_nodes
)
decomposed_form <- measurement_form(
  signals, decomposed_edge, sample_covariance, self_products
)
components <- measurement_components(decomposed_form, edge = 1)
components[, c(
  "left_component", "right_component", "raw_entries_meaningful",
  "frobenius_strength"
)]
```

```{r decomposition-check, include=FALSE}
stopifnot(
  nrow(components) == 4L,
  identical(components$raw_entries_meaningful,
    c(TRUE, FALSE, FALSE, FALSE)),
  all(is.finite(components$frobenius_strength)),
  all(components$frobenius_strength >= 0)
)
```

The coherent direction has a fixed scientific orientation. Configuration is
canonical as a subspace, not as an arbitrary compact basis within that
subspace. `measurement_components()` therefore reports rotation-invariant
strengths and does not present configuration-to-configuration entries as
basis-free quantities.

## Can different experimental and neural spaces use the same engine?

Yes. The left and right relation arguments name experimental-neural sides;
`between` alone carries the neural measurement endpoints. Here encoding and
retrieval have different experimental coordinates and different native neural
domains.

```{r cross-domain-setup, include=FALSE}
encoding_space <- effect_space(
  c("encode_a", "encode_b"), basis_id = "demo:encoding:v1"
)
retrieval_space <- effect_space(
  c("retrieve_a", "retrieve_b", "lure"),
  basis_id = "demo:retrieval:v1"
)
encoding_domain <- abstract_domain(3, id = "demo:encoding-neural:v1")
retrieval_domain <- abstract_domain(4, id = "demo:retrieval-neural:v1")
encoding_values <- matrix(c(1, 0, 2, 0, 1, -1), 2, 3, byrow = TRUE)
retrieval_values <- matrix(c(
  1, 0, 0, 1,
  0, 1, 1, 0,
  -1, 0, 1, 1
), 3, 4, byrow = TRUE)
encoding <- relation(
  list(encoding_run = encoding_values),
  effects = encoding_space,
  domain = encoding_domain
)
retrieval <- relation(
  list(retrieval_run = retrieval_values),
  effects = retrieval_space,
  domain = retrieval_domain
)
```

```{r cross-domain-form}
encoding_nodes <- measurement_frame(
  list(seed = matrix(c(1, 0, 1), 1)),
  encoding_domain,
  id = "demo:encoding-node:v1"
)
retrieval_nodes <- measurement_frame(
  list(target = matrix(c(1, 0, 0, 1, 0, 1, 1, 0), 2, 4)),
  retrieval_domain,
  id = "demo:retrieval-node:v1"
)
cross_node <- edge_frame(
  "seed", "target", encoding_nodes, to_frame = retrieval_nodes
)
cross_query <- pair_query(
  matrix(c(1, -0.5, 0, 0, 0.5, -1), 2, 3, byrow = TRUE),
  encoding_space,
  retrieval_space
)
cross_form <- measurement_form(
  left = encoding,
  right = retrieval,
  between = cross_node,
  by = cross_query,
  over = pairing("encoding_run", "retrieval_run", directed = TRUE)
)
effect_coupling(cross_form)$values[[1]]
```

```{r cross-domain-check, include=FALSE}
expected_cross <- encoding_nodes$legs$seed$operator %*%
  crossprod(encoding_values,
    as.matrix(cross_query$operator) %*% retrieval_values) %*%
  t(retrieval_nodes$legs$target$operator)
stopifnot(isTRUE(all.equal(
  effect_coupling(cross_form)$values[[1]],
  expected_cross,
  tolerance = 1e-12
)))
```

This is the off-diagonal block on both axes: different experimental spaces and
different neural measurements. It is still one evidence-form contraction.

## When do node and edge blocks reconstruct the global operator?

Complete node plus edge blocks are coordinates for the global neural evidence
operator when the measurement frames are injective. This example uses a
Parseval frame: stacking its two operators gives the identity.

```{r parseval-form}
parseval_nodes <- measurement_frame(
  list(
    first_half = diag(4)[1:2, , drop = FALSE],
    second_half = diag(4)[3:4, , drop = FALSE]
  ),
  native,
  id = "demo:parseval:v1"
)
parseval_pairs <- expand.grid(
  from = parseval_nodes$node_ids,
  to = parseval_nodes$node_ids,
  stringsAsFactors = FALSE
)
parseval_edges <- edge_frame(
  parseval_pairs$from, parseval_pairs$to, parseval_nodes
)
parseval_form <- measurement_form(
  signals, parseval_edges, sample_covariance, self_products
)
reconstructed <- reconstruct_evidence(parseval_form, parseval_edges)
c(
  method = reconstructed$method,
  status = reconstructed$status,
  lossless = reconstructed$lossless
)
```

```{r parseval-check, include=FALSE}
expected_global <- expected_covariance
stopifnot(
  reconstructed$lossless,
  identical(reconstructed$method, "parseval"),
  max(abs(reconstructed$operator - expected_global)) < 1e-12
)
```

A full-rank non-Parseval frame uses its canonical dual. A rank-deficient frame
returns `status = "projected_reconstruction"` and `lossless = FALSE`.
Incomplete edges, incompatible frame bases, excessive condition numbers, and
insufficient workspace are rejected before a lossless claim. Diagonal node
blocks alone cannot distinguish two global operators with the same diagonal
and different off-diagonal structure.

## Which interpretation is permitted?

| View | Minimum interpretation contract |
|---|---|
| `effect_coupling()` | Any admitted fixed experimental query; no covariance claim |
| `covariance_coupling()` | Repeated variation, effective rank above one, coherent joint covariance, valid self-blocks |
| `connectivity(..., "correlation")` | The covariance contract plus rank-one oriented measurement axes |
| `canonical_coupling()` | The covariance contract plus explicit regularization |
| `geometry_alignment()` | Coherent joint covariance and nonzero self-geometries; static form-entry normalization |
| Gaussian information | Canonical contract plus an explicit joint Gaussian model and information units |
| `reconstruct_evidence()` | Complete edge frame, matching identified bases, resource preflight, and acceptable frame rank/conditioning |

The result records which axis was normalized. Pattern correlation normalizes
neural features before producing an experimental form. Functional correlation
normalizes repeated experimental samples before interpreting a neural form.
Calling both operations "correlation" does not make them interchangeable.

The public measurement workflow aggregates raw bilinear partition forms before
its connectivity views normalize them. It does not silently substitute the
different estimand obtained by normalizing or Fisher-transforming within each
partition pair before reduction. The source stage order, view stage order,
regularization, units, and sampling axis remain in result identity.

## Is geometry alignment informational connectivity?

Not by default. `geometry_alignment()` compares static Gram matrices and is a
rotation-tolerant multivariate generalization of squared scalar correlation.
Established informational-connectivity analyses may instead correlate dynamic
discriminability series estimated over time or trials. Those series require a
training and cross-fitting pipeline that this fixed-form core does not own.

Migration is therefore semantic rather than cosmetic:

- use `effect_coupling()` for an arbitrary fixed query;
- use `connectivity()` only when repeated variation and covariance capabilities
  are present;
- call static CKA/RV-like comparison `geometry_alignment()`;
- keep dynamic informational-connectivity estimators in an adjacent system
  until their training, cross-fitting, and inference contracts are explicit.

For ordinary self geometry, contrast, RDM, RSA, and spectrum views, continue
with `vignette("introduction", package = "effectagram")`. The measurement-form
workflow adds another partial materialization of the same evidence pairing; it
does not replace the established geometry API.
