---
title: "Failure gallery"
description: >
  Six realistic analysis errors a conventional pipeline can execute silently, and the classed refusals or absent transformations that stop them.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Failure gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this guide when a requested result is unavailable, or when a familiar
transformation would change the scientific question. Each case explains the
problem and a supported next step.

A capability refusal is an R condition of class `effect_capability_refusal`.
`catch_refusal()` captures it so you can inspect `$capability`, `$reasons`, and
`$remedies`. Ordinary errors still stop execution; a successful expression
returns `NULL` from `catch_refusal()`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

```{r example-plan}
library(crossform)
example <- example_fmri_effects()
plan <- plan_geometry(
  example$fit$relation, example$frame,
  cross_partitions(example$fit$relation, independence = "independent")
)
```

The examples below execute against this same fit and plan.

## 1. Correlation-style normalization of a signed cross-generalized form

Dividing crossvalidated similarities by crossvalidated "variances" looks like
`1 - Pearson` correlation distance. It is not: cross-partition diagonal
estimates can be zero or negative, so the normalization is undefined exactly
where crossvalidation is doing its job.

```{r refusal-1}
refusal <- catch_refusal({
  rdm(plan, normalize = "correlation")
})
refusal$capability
cat(paste(strwrap(refusal$message, width = 76), collapse = "\n"))
```

Refusal capability: `guaranteed_psd`. The full boundary is
[the correlation-distance policy](correlation-distance-policy.html)
(`vignette("correlation-distance-policy", package = "crossform")` offline).

## 2. "Remove the univariate signal"

Demeaning patterns before a multivariate analysis destroys information while
leaving voxelwise mean effects inside the residual subspace, and then invites
the claim that what remains is "purely multivariate."

```{r refusal-2}
refusal <- catch_refusal({
  contrast_energy(plan, example$contrast, remove_univariate = TRUE)
})
refusal$capability
cat(paste(strwrap(refusal$message, width = 76), collapse = "\n"))
```

Refusal capability: `nondestructive_decomposition`. The accounting it points
to is exact: `total = coherent + configuration` at every measurement, for
contrasts, RDMs, and RSA coefficients alike.

## 3. Clipping negative crossvalidated distances

Crossvalidated squared distances are unbiased around zero under the null, so
negative estimates are informative, not errors. Truncating them at zero
introduces positive bias.

There is nothing to refuse here because no clipping surface exists: `rdm()`
takes no truncation argument, its values are documented as signed
("Cross-generalized distances may be negative"), and `geometry_spectrum()`
reports eigenvalues that "are never truncated at zero." Any nonnegativity
transformation would be a new named view with its own estimand, not an option
on this one.

## 4. Changing the generalization question

What reproduces across runs within a session is a different scientific
quantity from what reproduces across sessions — even at identical fold
counts. The generalization axis is therefore part of estimand identity, not
an estimator knob:

```{r distinct-questions}
run_plan <- plan_geometry(
  example$fit$relation, example$frame,
  cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)
session_plan <- plan_geometry(
  example$fit$relation, example$frame,
  cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "session"
  )
)
identical(run_plan$scientific_plan_id, session_plan$scientific_plan_id)
```

These declarations give the plans distinct identities even though they pair
the same inputs and therefore compute the same numbers. Relabeling this
fixture does not create session-level evidence: a real session analysis also
needs partitions and pairings that represent the intended sessions.

## 5. Independence that was never declared

Different run labels and off-diagonal pairing do not prove statistical
independence. Point geometry remains available, but an analytic covariance law
that consumes endpoint independence refuses if the declaration is absent:

```{r refusal-3}
refusal <- catch_refusal({
  undeclared_plan <- plan_geometry(
    example$fit$relation, example$frame,
    cross_partitions(example$fit$relation)
  )
  rdm_sampling_covariance(
    undeclared_plan, example$fit, target = "null", at = 1L
  )
})
refusal$capability
cat(paste(strwrap(refusal$message, width = 76), collapse = "\n"))
```

Refusal capability: `sampling_covariance`. The remedy is to declare
`independence = "independent"` only when the acquisition and observation model
justify it; changing that declaration changes the estimand identity.

## 6. Uncertainty that was never earned

Analytic error bars need both an admitted metric and a fitted error channel.

**A learned metric wants an analytic law.** The metric was estimated from
data, so the fixed-metric sampling law does not cover the result:

```{r refusal-4}
refusal <- catch_refusal({
  learned_metric_plan <- plan_geometry(
    example$fit$relation, example$frame,
    cross_partitions(example$fit$relation, independence = "independent"),
    metric = neural_metric(
      diag(example$domain$n_features), example$domain,
      estimation = "learned_frozen",
      provenance = list(
        frozen = TRUE,
        training_signature = paste0("sha256:", strrep("ab", 32))
      )
    )
  )
  rdm_sampling_covariance(
    learned_metric_plan, example$fit, target = "null", at = 1L
  )
})
refusal$capability
cat(paste(strwrap(refusal$message, width = 76), collapse = "\n"))
```

Refusal capability: `fixed_metric_sampling_law`.

**Precomputed betas want residual uncertainty back.** Beta matrices alone
cannot recover the error channel that was discarded when they were exported —
and the refusal reports every unmet requirement, not only the first:

```{r refusal-5}
refusal <- catch_refusal({
  rdm_sampling_covariance(plan, example$fit$relation, target = "null", at = 1L)
})
refusal$capability
cat(paste(strwrap(refusal$message, width = 76), collapse = "\n"))
```

Both reasons are separately actionable. Requirements that only restate the
missing channel — such as whether the partitions share one error structure —
cannot be evaluated without a channel to inspect, so they are not listed as
extra failures.

Refusal capability: `sampling_covariance`. Check availability before requesting a covariance with
`sampling_capabilities(plan, example$fit)`. It reports the reasons and remedies
without computing the covariance.

## Choose a supported next step

Retain signed distances for cross-run geometry. Use `coherent`,
`configuration`, or `total` to select the contrast component you need. For
uncertainty, inspect `sampling_capabilities()` and address the reported
requirements; an independence declaration must be justified by the acquisition
and model. [Fit condition effects from observations](from-observations.html)
shows how to retain residuals, and [Reading results](interpreting-results.html)
explains the available uncertainty targets.

## Further checks

Every case is covered by executable tests:
[`tests/testthat/test-capability-refusals.R`](https://github.com/bbuchsbaum/crossform/blob/main/tests/testthat/test-capability-refusals.R),
[`tests/testthat/test-integrity-guards.R`](https://github.com/bbuchsbaum/crossform/blob/main/tests/testthat/test-integrity-guards.R),
and
[`tests/testthat/test-generalization-axis.R`](https://github.com/bbuchsbaum/crossform/blob/main/tests/testthat/test-generalization-axis.R).
The Haxby 2001 exemplar records three of these refusals firing against real
data in
[`exemplars/haxby2001/results/smoke-report.md`](https://github.com/bbuchsbaum/crossform/blob/main/exemplars/haxby2001/results/smoke-report.md).
