---
title: "Coming from rMVPA"
description: >
  A concept map from rMVPA's dataset/design/model vocabulary to crossform's relation/frame/plan vocabulary, with an honest list of what crossform does not do.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = TRUE
)
library(crossform)
```

Use this guide to translate an rMVPA analysis into `crossform`: identify the
corresponding objects, bring in a trial-level beta series, and understand which
results can be compared. You need trial condition/run labels and either a
masked beta matrix or neuroim2 images.

The `crossform` examples run on simulated data. The rMVPA example is an
unevaluated template requiring your own images and model RDM. A recorded
numerical comparison on a shared distance estimand appears at the end.

## From a model run to a reusable geometry plan

rMVPA is organized around a **model** that is *run over* a set of locations.
You build a dataset, a design, and a model specification, then hand the whole
thing to `run_searchlight()` or `run_regional()`, which returns a performance
map.

`crossform` is organized around a **plan** that is *queried*. You declare a
fitted relation, where to measure it, and what must generalize; that produces
a plan. Contrasts, RDMs, and RSA are then different reads of the same plan,
and they do not refit anything:

```r
# rMVPA: one run per question
run_searchlight(model_a, radius = 8)
run_searchlight(model_b, radius = 8)

# crossform: one plan, many questions
plan <- plan_geometry(rel, at = frame, over = pairing)
contrast_energy(plan, weights)
rdm(plan)
rsa(plan, models = list(category = model_rdm))
```

The practical consequence is where cross-validation lives. In rMVPA the
blocking variable is consumed by a cross-validation scheme that folds a
*model*. In `crossform` the generalization axis is bound into the *estimand's
identity* by `cross_partitions()`, when the geometry is planned, and it travels with the plan.

## Concept map

| rMVPA | crossform | Notes |
|---|---|---|
| `mvpa_dataset(train_data, mask =)` | `neuroim2_volume_domain(mask)` + `relation(list_of_run_matrices, ...)` | rMVPA holds the image; `crossform` splits *where* (domain) from *what was estimated* (relation). |
| `mvpa_design(df, y_train = ~y, block_var = ~block)` | `effect_space(condition_names)` for the conditions; run structure is the names of the list passed to `relation()` | The same table drives both routes in "Getting your own data in" below: the condition column becomes the effect space, the run column becomes the partitions. |
| `blocked_cross_validation(block_var)` | `cross_partitions(rel, independence = "independent", generalizes_over = "run")` | rMVPA folds train/test; `crossform` forms products between *different* partitions. Not the same object: see below. |
| `run_searchlight(spec, radius = 8)` | `compile_frame(searchlights(radius = 8), domain)` or `neuroim2_searchlights(mask, radius = 8)`, then `plan_geometry()` | Radius is millimeters in both. |
| `run_regional(spec, region_mask)` | `compile_frame(regions(labels), domain)` | `regions()` takes one label per **feature** (a vector), not a `NeuroVol`. |
| `rsa_design(~ model, data = list(model = D), block_var =)` | the `models =` list of `rsa()`, plus `cross_partitions()` for the blocking | `crossform` aligns model RDMs by condition **name**. |
| `rsa_model(dset, des, distmethod =, regtype =)` | `rsa(plan, models = list(...))` | Two separate arguments: `distmethod` picks the distance, `regtype` the second-order statistic. See "distance conventions" below. |
| `contrast_rsa_model(dset, msreve_des, output_metric = "beta_delta")` | `contrast_energy(plan, weights)` | The nearest analogue, not an equivalent. rMVPA regresses the crossvalidated second-moment matrix on several competing contrast RDMs at once and multiplies each signed coefficient by the *center voxel's* projection onto that contrast, giving one signed map per contrast. `contrast_energy()` scores one contrast on its own, returns that contrast's crossvalidated energy for the whole measurement, and splits it into coherent and configuration parts. |
| `vector_rsa_model(dset, des, distfun = cordist())` | no equivalent | Trial-level; `crossform` aggregates to condition level. |
| `compute_crossvalidated_means_sl(..., estimation_method = "crossnobis")` + `compute_crossnobis_distances_sl()` | `rdm(plan)` under `cross_partitions()`, or `plan_crossnobis()` / `crossnobis()` | This is the pairing where the two packages have been checked against each other. |
| `mvpa_model(model = "sda_notune", ...)`, classification accuracy | no equivalent | `crossform` has no classifier. |
| train/test folds | cross-partition pairing | A fold count is an execution detail in `crossform`; the *generalization axis* is part of plan identity. |
| performance metrics (`compute_performance`) | the views: `contrast_energy()`, `rdm()`, `rsa()` | Energies and signed squared distances, not accuracies. |
| `feature_selector("FTest", "top_k", 100)` | no equivalent | See absences below. |
| `run_permutation_searchlight(...)`, `permutation_control()` | no equivalent | See absences below. |
| `future::plan(multisession)` (rMVPA uses furrr/future.apply) | `compute_policy()`, sequential only | `workers > 1` is refused, not silently ignored. |
| `save_results(x, dir, ...)` | `as_neurovol()` + your own writer | `crossform` writes no files. |

### The cross-validation mapping deserves a second look

`blocked_cross_validation(block_var)` and
`cross_partitions(rel, generalizes_over = "run")` both use run structure, but neither run labels nor a fold scheme
establish statistical independence.

A leave-one-run-out fold scheme produces training means that **overlap across
folds**. Cross-partition pairing forms products between two *disjoint* run
estimates. When errors in those estimates have zero mean and are independent
across runs, their cross-product has zero expectation. This removes the
noise contribution that biases a within-run squared distance upward, while
allowing individual estimates to be negative. The Haxby exemplar records a concrete instance of the difference
mattering: an older rMVPA build derived crossnobis folds from leave-one-run-out
training means, whose overlap made the distances positively biased, and the
exemplar script asserts fold semantics rather than trusting a version number.

## Side-by-side minimal example

Both examples examine animacy-related geometry. They illustrate the two
workflows; their distance definitions and summary statistics differ, so their
output values are not interchangeable.

### rMVPA

```{r rmvpa-code, eval=FALSE}
library(rMVPA)

dset <- mvpa_dataset(train_data = beta_vec, mask = vt_mask)

des <- mvpa_design(
  data.frame(
    y = factor(rep(conditions, times = n_run), levels = conditions),
    block = run_of
  ),
  y_train = ~ y,
  block_var = ~ block
)

rsa_des <- rsa_design(~ animacy, data = list(animacy = model_rdm),
                      block_var = run_of)
mspec <- rsa_model(dset, rsa_des, distmethod = "pearson", regtype = "pearson")
res <- run_searchlight(mspec, radius = 6, method = "standard")

# res$results$animacy is a NeuroVol of second-order correlations.
```

### crossform

```{r crossform-code}
example <- example_fmri_effects()

plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)

effect <- contrast_energy(plan, example$contrast)
category <- rsa(plan, models = list(category = example$model_rdm))

c(
  measurements = length(effect$total),
  peak = which.max(effect$total),
  rsa_columns = ncol(category$coefficients)
)
```

Two things differ beyond syntax. `plan` is reusable: `rdm(plan)` and any
number of further contrasts cost no refit. And `rsa()` returns OLS
coefficients with an intercept column by default — `ncol()` above is 2 for one
model because the intercept is retained. rMVPA reaches the same family with
`rsa_model(..., regtype = "lm")`; its default `regtype = "pearson"` scores the
match as a correlation between the two vectorized RDMs instead.

```{r crossform-reuse}
distances <- rdm(plan)
dim(distances$values)
colnames(category$coefficients)
```

## Getting your own data in: from a trial-level beta series

The example above starts from a fitted relation. If what you have is what
rMVPA wants — a trial-level beta series plus a design table with a condition
column and a run column — this section is the bridge. Everything here runs.

```{r beta-series}
set.seed(2026)
conditions <- c("face", "body", "house", "tool")
runs <- paste0("run", 1:4)

# One row per trial, five trials per condition per run. This is the
# `mvpa_design()` table, and `beta_series` is the matrix behind the
# `mvpa_dataset()` NeuroVec once the mask has been applied.
trial_design <- expand.grid(
  trial = 1:5, condition = conditions, run = runs, stringsAsFactors = FALSE
)
trial_design$condition <- factor(trial_design$condition, levels = conditions)
n_trial <- nrow(trial_design)
n_voxel <- 40L

beta_series <- matrix(rnorm(n_trial * n_voxel), n_trial, n_voxel)
planted <- 1:8
animate <- trial_design$condition %in% c("face", "body")
beta_series[animate, planted] <- beta_series[animate, planted] +
  rep(rep(c(1.5, -1.5), length.out = length(planted)), each = sum(animate))

dim(beta_series)
head(trial_design, 3)
```

`relation()` wants one **condition-by-voxel matrix per run**, not a trial
list, because the second-order geometry is defined on condition estimates and
the runs are the partitions that must generalize. There are two ways to get
there. Both estimate condition means; fitting also retains residuals under
a declared trial-level error model.

### Route 1: average the betas within condition, per run

The direct translation. `split()` on the run column, then a column mean per
condition.

```{r beta-series-average}
domain <- abstract_domain(n_voxel, id = "beta-series-example")
by_run <- split(seq_len(n_trial), trial_design$run)

run_matrices <- lapply(by_run, function(rows) {
  t(vapply(conditions, function(condition) {
    colMeans(beta_series[rows[trial_design$condition[rows] == condition], ,
      drop = FALSE])
  }, numeric(n_voxel)))
})

averaged <- relation(run_matrices, effects = effect_space(conditions),
                     domain = domain)
plan_averaged <- plan_geometry(
  averaged,
  at = compile_frame(whole_brain(), domain),
  over = cross_partitions(averaged, independence = "independent",
                          generalizes_over = "run")
)
animacy <- c(face = 0.5, body = 0.5, house = -0.5, tool = -0.5)
energy_averaged <- contrast_energy(plan_averaged, animacy)
energy_averaged$total
```

Row names matter: `relation()` aligns partitions by effect **name**, so the
matrices must be named by condition. `t(vapply(...))` above produces that.

### Route 2: hand the trial betas over as observations

`lm_relation_fit()` takes the trial betas plus a
per-run design matrix and fits the condition estimates inside `crossform`,
which keeps the residual error channel that the analytic sampling law needs.
The design is the condition indicator; `effects` says each condition *is* its
own coefficient.

```{r beta-series-fit}
designs <- lapply(by_run, function(rows) {
  X <- stats::model.matrix(~ 0 + condition,
                           data = trial_design[rows, , drop = FALSE])
  colnames(X) <- conditions
  X
})
sources <- lapply(by_run, function(rows) beta_series[rows, , drop = FALSE])
targets <- diag(length(conditions))
dimnames(targets) <- list(conditions, conditions)

fit <- lm_relation_fit(
  sources, design = designs, effects = targets,
  effect_names = effect_space(conditions),
  sampling_unit = "trial", domain = domain
)
plan_fitted <- plan_geometry(
  fit$relation,
  at = compile_frame(whole_brain(), domain),
  over = cross_partitions(fit$relation, independence = "independent",
                          generalizes_over = "run")
)
energy_fitted <- contrast_energy(plan_fitted, animacy)
```

With this balanced indicator design the two routes estimate the same
condition means, so they report the same energy:

```{r beta-series-agree}
c(route_1 = energy_averaged$total,
  route_2 = energy_fitted$total,
  max_difference = max(abs(energy_averaged$total - energy_fitted$total)))
```

The fitted route retains an error channel for within-participant standard
errors. In this simulation trial errors are independent. For real beta series,
the first-level GLM can induce covariance among trial estimates; treating
those betas as independent observations requires justification. The following
standard errors use the null calibration target:

```{r beta-series-uncertainty}
sampling_capabilities(plan_fitted, fit)

round(sqrt(sampling_covariance(rdm_sampling_covariance(
  plan_fitted, fit, target = "null", at = 1L
))), 4)
```

The averaged relation has no residual channel:

```{r beta-series-refusal}
catch_refusal(
  rdm_sampling_covariance(plan_averaged, averaged, target = "null", at = 1L)
)
```

Two practical notes. Runs need not have equal trial counts — `design` accepts
a named list with one matrix per partition, which is why it is built inside
`lapply()` above. A fixed `observation_whitener` here must describe covariance among the
trial betas, not among the original scans. See
[Fit condition effects from observations](from-observations.html) for the
scan-level route and the fixed-whitener assumptions.

## Distance and second-order conventions are not the same

`rsa_model()` makes two choices: how to construct the neural RDM, and how
to compare it with a model RDM. Match both before comparing outputs.

- **The distance.** `crossform`'s `rdm()` reports signed **crossvalidated
  squared Euclidean** distances, `d_ij = G_ii + G_jj - 2 G_ij` (squared
  Mahalanobis under a fixed metric). They can be negative, and are not
  truncated. rMVPA's `rsa_model()` takes `distmethod` in
  `{"pearson", "spearman"}` and nothing else, so its neural RDM is always a
  **correlation distance**. This is the difference that cannot be bridged.
  (rMVPA's crossnobis distances live elsewhere, on
  `contrast_rsa_model(estimation_method = "crossnobis")` and the
  `compute_crossnobis_distances_sl()` helper — which is why the parity check
  below uses those and not `rsa_model()`.)
- **The second-order statistic.** This is a *separate* argument, `regtype`,
  taking `"pearson"`, `"spearman"`, `"lm"`, or `"rfit"`. `crossform`'s `rsa()`
  is OLS with an intercept, which is the same family as `regtype = "lm"`.
  rMVPA's default is `regtype = "pearson"`, a correlation between the two
  vectorized RDMs; set `regtype = "lm"` and this half of the comparison lines
  up.

`crossform` cannot express correlation distance at all: dividing each pattern
by its own norm is not a fixed linear operation on the patterns, so it is
outside the bilinear core. Per-sphere mean-centering *is* expressible as a
metric; the unit-norm denominator is not. The boundary and the conditions
under which it could be crossed are recorded in the
[correlation-distance policy](correlation-distance-policy.html)
(`vignette("correlation-distance-policy", package = "crossform")` offline).

If you want the two packages to compute the same thing, the estimand both can
express is the condition-level crossvalidated squared Euclidean (crossnobis)
distance — which is the pairing used for the parity check below.

## What remains outside crossform

Keep these parts of an existing analysis in other tools:

- **Classification and decoding.** There is no classifier, no accuracy, no AUC,
  no confusion matrix, no `MVPAModels` registry. `crossform` computes
  second-order geometry only.
- **Universal group inference.** The experimental population layer now has
  declared group models, classical and HC3 intervals, and a participant-level
  null-imposed wild bootstrap, calibrated only in its recorded synthetic
  regimes and conditional on realized transport. It does not provide random
  fields, simultaneous/maxT bands, general multiplicity correction, or a
  marginal interval under informative coverage or transport learning.
- **Feature selection.** No `FTest`/`catscore` ranking, no `top_k` cutoff. The
  frame decides which features each measurement reads, and it is declared in
  advance.
- **Label permutation and multiplicity correction.** There is no label
  shuffling, cluster/random-field procedure, or FDR correction. The exported
  `population_wild_bootstrap()` is a different object: it resamples participants
  under a declared group-model null and keeps its conditional target explicit.
- **Parallel backends.** Execution is sequential. This is refused rather than
  silently downgraded:

```{r no-workers}
worker_refusal <- catch_refusal(compute_policy(workers = 4))
cat(paste(strwrap(worker_refusal$message, width = 76), collapse = "\n"))
```

- **Preprocessing and I/O.** No registration, no smoothing, no file reading or
  writing. `as_neurovol()` hands you a `NeuroVol`; saving it is yours.

```{r absence-check, include=FALSE}
exports <- getNamespaceExports("crossform")
stopifnot(
  !any(grepl("classif|decod|accuracy|permut|feature_select|fdr",
             exports, ignore.case = TRUE)),
  "population_wild_bootstrap" %in% exports,
  inherits(try(compute_policy(workers = 2), silent = TRUE), "try-error")
)
```

Two things rMVPA users may expect that `crossform` *does* have, in a different
shape: a crossnobis path (`plan_crossnobis()`, `crossnobis()`, with
`shrinkage_precision()` and `metric_training_policy()` for the whitening), and
explicit refusals when an uncertainty claim is not supported by the fit
(`sampling_capabilities()`, `catch_refusal()`).

## The one measured comparison

The [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
runs both packages on eight stimulus categories estimated in twelve runs, over
577 ventral-temporal searchlights at an 11.25 mm radius. On the matched
estimand — `crossform`'s `rdm()` over a cross-run geometry against rMVPA's
`compute_crossvalidated_means_sl(estimation_method = "crossnobis")` followed by
`compute_crossnobis_distances_sl()` — the two agree to a **maximum absolute
difference of 8.88e-16** at every one of the 577 centers. Sphere membership
was verified identical rather than assumed.

This is a recorded result, not a benchmark rerun by this vignette. The
[exemplar scripts and receipts](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
document the inputs, sphere membership, and package behavior required to
reproduce it. Three limits matter when adapting it:

- It uses rMVPA's crossnobis helpers, including an unexported helper called
  with `:::`, rather than `rsa_model()`.
- It requires per-run condition means. The script checks this behavior because
  older fold implementations used overlapping training means.
- It substitutes a shared squared-distance estimand for the originally
  proposed correlation-distance RSA. The agreement supports that shared
  estimand only; it says nothing about runtime or correlation-distance RSA.

```{r beta-series-parity-check, include=FALSE}
stopifnot(isTRUE(all.equal(energy_averaged$total, energy_fitted$total)))
```

## Where to go next

- [Predict model-supported geometry on independent runs](predictive-geometry.html)
  (`vignette("predictive-geometry", package = "crossform")`) — a separate
  learning workflow: `model_basis()` declares the models, `fit_geometry()`
  learns on training runs, and `score_geometry()` evaluates signed gain on
  independent runs. This is regularized geometry prediction, not another
  `regtype` for the fixed RDM regression compared above.
- `vignette("introduction", package = "crossform")` — what the energies, RDMs,
  and RSA coefficients mean.
- `vignette("neuroim2-data", package = "crossform")` — the input and output
  adapters, if your data is already in neuroim2 objects.
- [Reading contrast energies, RDMs, and uncertainty](interpreting-results.html)
  (`vignette("interpreting-results", package = "crossform")`) — how to read
  the numbers, and the three interpretive traps.
- [What is novel in crossform?](novelty.html)
  (`vignette("novelty", package = "crossform")`) — what is architecture and
  what is established statistics.
