---
title: "How do observations become a representational analysis?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How do observations become a representational analysis?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(effectagram)
```

This guide starts before the beta matrices. Its question is: **how can raw
observation-by-feature responses become a cross-run geometry without letting
design coding, censoring, or solver choices silently redefine the requested
effects?**

The answer uses two plans. `plan_relation()` identifies the first-moment
request; `plan_geometry()` identifies the second-moment request. Between them,
portable receipts record what actually happened.

## Build a small study from facts

The example has three runs, three named conditions, twelve neural features,
and two censored scans per run. The event rows, observation clocks, confounds,
and partition hierarchy are facts. They do not decide which terms are nuisance,
which sampling unit matters, or what must generalize.

```{r facts}
set.seed(20260815)
partitions <- paste0("run-", 1:3)
n_scan <- 18L
conditions <- c("face", "body", "tool")
domain <- abstract_domain(12L, id = "from-observations:v1")

condition <- factor(
  rep(conditions, length.out = n_scan),
  levels = conditions
)
indicator <- stats::model.matrix(~ condition - 1)
colnames(indicator) <- conditions
motion <- sin(seq(0, 2 * pi, length.out = n_scan))
drift <- seq(-1, 1, length.out = n_scan)
design <- cbind(indicator, drift = drift, motion = motion)
rownames(design) <- as.character(seq_len(n_scan))

truth <- matrix(
  rnorm(length(conditions) * domain$n_features, sd = 0.35),
  nrow = length(conditions),
  dimnames = list(conditions, domain$feature_ids)
)
nuisance_truth <- matrix(
  rnorm(2L * domain$n_features, sd = 0.08),
  nrow = 2L
)
response_runs <- stats::setNames(lapply(partitions, function(partition) {
  design[, conditions, drop = FALSE] %*% truth +
    design[, c("drift", "motion"), drop = FALSE] %*% nuisance_truth +
    matrix(rnorm(n_scan * domain$n_features, sd = 0.4),
      nrow = n_scan)
}), partitions)

scan_indexes <- stats::setNames(lapply(partitions, function(partition) {
  observation_index(
    seq_len(n_scan), partition,
    time = seq(0, by = 2, length.out = n_scan), units = "seconds"
  )
}), partitions)

event_table <- do.call(rbind, lapply(partitions, function(partition) {
  data.frame(
    partition = partition,
    event_id = paste0(partition, "-", seq_len(n_scan)),
    onset = seq(0, by = 2, length.out = n_scan),
    duration = 0,
    condition = as.character(condition)
  )
}))
confound_table <- do.call(rbind, lapply(partitions, function(partition) {
  data.frame(
    partition = partition,
    observation_id = seq_len(n_scan),
    motion = motion,
    retained = seq_len(n_scan) <= 16L
  )
}))

facts <- study(
  observations(response_runs, scan_indexes, domain),
  events(event_table),
  observation_confounds(confound_table, censor = "retained"),
  partition_hierarchy(data.frame(
    partition = partitions,
    run = partitions,
    session = "session-1",
    subject = "subject-1"
  ))
)
facts
study_capabilities(facts)
```

`study()` joins by declared partition and observation identifiers, never by
positional luck. It also checks that event intervals and scan clocks use the
same units and coverage. The condition and motion columns still have no model
role.

## Declare effects before choosing coefficient coding

The scientific coordinates are the three condition amplitudes. The identity
effect map requests all three so the downstream relation has a named condition
axis. A contrast such as face minus body will be asked later without refitting.

```{r semantic-effects}
condition_coordinates <- condition_space(
  conditions,
  basis_id = "scan-level-condition-mean:v1",
  units = "arbitrary-BOLD"
)
effect_weights <- diag(length(conditions))
dimnames(effect_weights) <- list(conditions, conditions)
named_effects <- effect_map(effect_weights, condition_coordinates)
```

The compiled design uses cell means plus drift and motion. Its parameterization
declares how coefficient coordinates reconstruct the semantic condition means:

```{r design-model}
condition_from_coefficients <- cbind(
  diag(length(conditions)), drift = 0, motion = 0
)
dimnames(condition_from_coefficients) <- list(
  conditions, colnames(design)
)
parameterization <- coefficient_parameterization(
  condition_from_coefficients,
  condition_coordinates,
  coding_id = "cell-means-plus-nuisance"
)

designs <- stats::setNames(
  rep(list(design), length(partitions)), partitions
)
parameterizations <- stats::setNames(
  rep(list(parameterization), length(partitions)), partitions
)
declared_design <- design_model(
  specification = list(
    target = "scan-level condition means",
    nuisance = c("linear drift", "declared motion covariate"),
    censor_policy = "use the bound retained column"
  ),
  conditions = condition_coordinates,
  designs = designs,
  parameterizations = parameterizations,
  row_ids = lapply(scan_indexes, `[[`, "observation_id"),
  solver = "qr",
  protocol = "from-observations-example",
  protocol_version = "1",
  package_version = "1"
)
```

This explicit design constructor is the small self-contained route. A
conforming external compiler can own the event grammar and HRF construction;
effectagram requires it to emit the same semantic identity and portable
receipt fields.

## Compose the first-moment estimand

The observation model is an assumption declaration. Here OLS is fixed, scans
are the sampling unit, and run independence is asserted explicitly. None of
those claims is inferred from the event table or run labels.

```{r relation-plan}
noise_model <- observation_model(
  "ols",
  sampling_unit = "scan",
  independence = "runs independently acquired conditional on the model"
)
relation_request <- plan_relation(
  facts,
  declared_design,
  named_effects,
  noise_model
)
relation_request
compiler_conformance(relation_request)
```

`plan_relation()` compiles and validates design metadata without reading the
neural values. Non-estimable effects refuse here with the partition, rank,
aliased regressors, cause, and remedies.

The identity ladder separates the scientific request from its realized route:

```{r identity-ladder}
receipts <- relation_plan_receipts(relation_request)
fit <- estimate(relation_request)
data.frame(
  object = c("relation plan", "run-1 design receipt", "fitted relation"),
  identity = substr(c(
    relation_request$relation_plan_id,
    receipts$`run-1`$design_receipt_id,
    fit$signature
  ), 1, 32)
)
```

Changing QR to SVD or using a supported treatment coding changes the receipt,
not the condition-space request. Changing the named functional, units,
observation law, or independence declaration changes the plan. `fit$signature`
then binds that plan provenance, realized receipts, and observation revisions.

## Continue into geometry without an ingestion-specific analysis engine

`estimate()` emits the same `effect_relation_fit` consumed everywhere else in
the package. The second-moment API does not branch on how the relation was
obtained.

```{r geometry}
frame <- compile_frame(
  regions(rep(paste0("roi-", 1:3), each = 4L), normalization = "local"),
  domain
)
generalization <- cross_partitions(
  fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
geometry_plan <- plan_geometry(fit$relation, frame, generalization)

face_minus_body <- c(face = 1, body = -1, tool = 0)
effect <- contrast(geometry_plan, face_minus_body)
distances <- rdm(geometry_plan)

tool_category <- outer(
  c(face = 0, body = 0, tool = 1),
  c(face = 0, body = 0, tool = 1),
  function(left, right) as.numeric(left != right)
)
model_fit <- rsa(geometry_plan, models = list(tool = tool_category))

round(cbind(
  signed = effect$signed,
  coherent = effect$coherent,
  configuration = effect$configuration,
  total = effect$total
), 3)
round(distances$values, 3)
round(model_fit$coefficients, 3)
```

The point outputs inherit one exact spatial accounting:

```{r downstream-checks, include=FALSE}
stopifnot(
  isTRUE(all.equal(
    effect$total,
    effect$coherent + effect$configuration,
    tolerance = 1e-12
  )),
  identical(dim(distances$values), c(3L, 3L)),
  identical(dim(model_fit$coefficients), c(3L, 2L)),
  all(as.matrix(compiler_conformance(relation_request)[-1L]))
)
```

Because the fit retained residuals and the plan declared a fixed observation
model and endpoint independence, the admitted analytic RDM variance is also
available:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  geometry_plan, fit, target = "plugin", at = 1L
)
round(sqrt(sampling_covariance(distance_covariance)), 3)
```

That is within-participant uncertainty under the stated separable fixed-model
law. It is not population inference, and a learned frozen whitener would keep
the point relation while withholding this analytic channel.

## Use adapters without making the core BIDS-shaped

`bids_events()` maps one task `events.tsv` per declared partition into
`events()`. `bids_confounds()` maps row-aligned confound tables into
`observation_confounds()` and requires an explicit censor column if censoring
is intended. `bids_study()` returns the same generic `study()` used above.
Arbitrary event and confound columns are preserved; model roles are still
assigned by the design compiler.

```{r bids-adapter, eval=FALSE}
facts <- bids_study(
  declared_observations,
  event_files = named_event_tsvs,
  confound_files = named_confound_tsvs,
  partitions = names(named_event_tsvs),
  censor = "retained"
)
```

`fmridesign_design_model()` adapts the certified `fmridesign` version into a
semantic design model. `fmrireg_relation()` independently executes an OLS
relation plan for point-estimate parity; it deliberately withholds an analytic
error channel rather than translating unsupported standard errors.

If only already compiled matrices are available, use `raw_design_model()` with
`raw_effect_map()`. That path remains valid, but the concrete `(X, T)` values
become identity-bearing and the fit cannot claim symbolic coding invariance.
If only beta matrices remain, `relation()` supports point geometry while
honestly lacking a residual channel.

The distinction is the whole discipline: facts describe what was observed;
`plan_relation()` composes the scientific request and assumptions; receipts
record the realized compiler route; `estimate()` reads the data; and
`plan_geometry()` asks what must reproduce.
