---
title: "Coming from rMVPA"
description: >
  A concept map from rMVPA's dataset/design/model vocabulary to crossform's relation/frame/plan vocabulary, with an honest list of what crossform does not do.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from rMVPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

> **This is a mapping guide, not a benchmark.** Nothing here compares the two
> packages on speed, accuracy, or scientific quality. It answers one question:
> if you already think in [rMVPA](https://github.com/bbuchsbaum/rMVPA)'s
> vocabulary, what is the corresponding `crossform` object, and what is
> deliberately missing? The one place the two packages have been measured
> against each other is a numerical-parity check, reported at the end with its
> caveats.
>
> The rMVPA calls below are shown but not executed — rMVPA is not on CRAN, so
> this vignette cannot depend on it. The `crossform` calls are executed.

## The one structural difference

rMVPA is organized around a **model** that is *run over* a set of locations.
You build a dataset, a design, and a model specification, then hand the whole
thing to `run_searchlight()` or `run_regional()`, which returns a performance
map.

`crossform` is organized around a **plan** that is *queried*. You declare a
fitted relation, where to measure it, and what must generalize; that produces
a plan. Contrasts, RDMs, and RSA are then different reads of the same plan,
and they do not refit anything:

```r
# rMVPA: one run per question
run_searchlight(model_a, radius = 8)
run_searchlight(model_b, radius = 8)

# crossform: one plan, many questions
plan <- plan_geometry(rel, at = frame, over = pairing)
contrast_energy(plan, weights)
rdm(plan)
rsa(plan, models = list(category = model_rdm))
```

The practical consequence is where cross-validation lives. In rMVPA the
blocking variable is consumed by a cross-validation scheme that folds a
*model*. In `crossform` the generalization axis is bound into the *estimand's
identity* by `cross_partitions()`, before any data is read, and it travels
with the plan.

## Concept map

| rMVPA | crossform | Notes |
|---|---|---|
| `mvpa_dataset(train_data, mask =)` | `neuroim2_volume_domain(mask)` + `relation(list_of_run_matrices, ...)` | rMVPA holds the image; `crossform` splits *where* (domain) from *what was estimated* (relation). |
| `mvpa_design(df, y_train = ~y, block_var = ~block)` | `effect_space(condition_names)` for the conditions; run structure is the names of the list passed to `relation()` | The same table drives both routes in "Getting your own data in" below: the condition column becomes the effect space, the run column becomes the partitions. |
| `blocked_cross_validation(block_var)` | `cross_partitions(rel, independence = "independent", generalizes_over = "run")` | rMVPA folds train/test; `crossform` forms products between *different* partitions. Not the same object: see below. |
| `run_searchlight(spec, radius = 8)` | `compile_frame(searchlights(radius = 8), domain)` or `neuroim2_searchlights(mask, radius = 8)`, then `plan_geometry()` | Radius is millimeters in both. |
| `run_regional(spec, region_mask)` | `compile_frame(regions(labels), domain)` | `regions()` takes one label per **feature** (a vector), not a `NeuroVol`. |
| `rsa_design(~ model, data = list(model = D), block_var =)` | the `models =` list of `rsa()`, plus `cross_partitions()` for the blocking | `crossform` aligns model RDMs by condition **name**. |
| `rsa_model(dset, des, distmethod =, regtype =)` | `rsa(plan, models = list(...))` | Two separate arguments: `distmethod` picks the distance, `regtype` the second-order statistic. See "distance conventions" below. |
| `contrast_rsa_model(dset, msreve_des, output_metric = "beta_delta")` | `contrast_energy(plan, weights)` | The nearest analogue, not an equivalent. rMVPA regresses the crossvalidated second-moment matrix on several competing contrast RDMs at once and multiplies each signed coefficient by the *center voxel's* projection onto that contrast, giving one signed map per contrast. `contrast_energy()` scores one contrast on its own, returns that contrast's crossvalidated energy for the whole measurement, and splits it into coherent and configuration parts. |
| `vector_rsa_model(dset, des, distfun = cordist())` | no equivalent | Trial-level; `crossform` aggregates to condition level. |
| `compute_crossvalidated_means_sl(..., estimation_method = "crossnobis")` + `compute_crossnobis_distances_sl()` | `rdm(plan)` under `cross_partitions()`, or `plan_crossnobis()` / `crossnobis()` | This is the pairing where the two packages have been checked against each other. |
| `mvpa_model(model = "sda_notune", ...)`, classification accuracy | no equivalent | `crossform` has no classifier. |
| train/test folds | cross-partition pairing | A fold count is an execution detail in `crossform`; the *generalization axis* is part of plan identity. |
| performance metrics (`compute_performance`) | the views: `contrast_energy()`, `rdm()`, `rsa()` | Energies and signed squared distances, not accuracies. |
| `feature_selector("FTest", "top_k", 100)` | no equivalent | See absences below. |
| `run_permutation_searchlight(...)`, `permutation_control()` | no equivalent | See absences below. |
| `future::plan(multisession)` (rMVPA uses furrr/future.apply) | `compute_policy()`, sequential only | `workers > 1` is refused, not silently ignored. |
| `save_results(x, dir, ...)` | `as_neurovol()` + your own writer | `crossform` writes no files. |

### The cross-validation mapping deserves a second look

`blocked_cross_validation(block_var)` and
`cross_partitions(rel, generalizes_over = "run")` both encode "runs are the
independent unit", but they are not interchangeable objects.

A leave-one-run-out fold scheme produces training means that **overlap across
folds**. Cross-partition pairing forms products between two *disjoint* run
estimates, so the noise in the two factors is independent and cancels in
expectation. That is exactly why `crossform`'s estimates are unbiased and may
be negative. The Haxby exemplar records a concrete instance of the difference
mattering: an older rMVPA build derived crossnobis folds from leave-one-run-out
training means, whose overlap made the distances positively biased, and the
exemplar script asserts fold semantics rather than trusting a version number.

## Side-by-side minimal example

The same question — where does an animate-versus-inanimate pattern reproduce
across runs? — written both ways.

### rMVPA

```{r rmvpa-code, eval=FALSE}
library(rMVPA)

dset <- mvpa_dataset(train_data = beta_vec, mask = vt_mask)

des <- mvpa_design(
  data.frame(
    y = factor(rep(conditions, times = n_run), levels = conditions),
    block = run_of
  ),
  y_train = ~ y,
  block_var = ~ block
)

rsa_des <- rsa_design(~ animacy, data = list(animacy = model_rdm),
                      block_var = run_of)
mspec <- rsa_model(dset, rsa_des, distmethod = "pearson", regtype = "pearson")
res <- run_searchlight(mspec, radius = 6, method = "standard")

# res$results$animacy is a NeuroVol of second-order correlations.
```

### crossform

```{r crossform-code}
example <- example_fmri_effects()

plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)

effect <- contrast_energy(plan, example$contrast)
category <- rsa(plan, models = list(category = example$model_rdm))

c(
  measurements = length(effect$total),
  peak = which.max(effect$total),
  rsa_columns = ncol(category$coefficients)
)
```

Two things differ beyond syntax. `plan` is reusable: `rdm(plan)` and any
number of further contrasts cost no refit. And `rsa()` returns OLS
coefficients with an intercept column by default — `ncol()` above is 2 for one
model because the intercept is retained. rMVPA reaches the same family with
`rsa_model(..., regtype = "lm")`; its default `regtype = "pearson"` scores the
match as a correlation between the two vectorized RDMs instead.

```{r crossform-reuse}
distances <- rdm(plan)
dim(distances$values)
colnames(category$coefficients)
```

## Getting your own data in: from a trial-level beta series

The example above starts from a fitted relation. If what you have is what
rMVPA wants — a trial-level beta series plus a design table with a condition
column and a run column — this section is the bridge. Everything here runs.

```{r beta-series}
set.seed(2026)
conditions <- c("face", "body", "house", "tool")
runs <- paste0("run", 1:4)

# One row per trial, five trials per condition per run. This is the
# `mvpa_design()` table, and `beta_series` is the matrix behind the
# `mvpa_dataset()` NeuroVec once the mask has been applied.
trial_design <- expand.grid(
  trial = 1:5, condition = conditions, run = runs, stringsAsFactors = FALSE
)
trial_design$condition <- factor(trial_design$condition, levels = conditions)
n_trial <- nrow(trial_design)
n_voxel <- 40L

beta_series <- matrix(rnorm(n_trial * n_voxel), n_trial, n_voxel)
planted <- 1:8
animate <- trial_design$condition %in% c("face", "body")
beta_series[animate, planted] <- beta_series[animate, planted] +
  rep(rep(c(1.5, -1.5), length.out = length(planted)), each = sum(animate))

dim(beta_series)
head(trial_design, 3)
```

`relation()` wants one **condition-by-voxel matrix per run**, not a trial
list, because the second-order geometry is defined on condition estimates and
the runs are the partitions that must generalize. There are two ways to get
there, and they are not equivalent in what they earn you.

### Route 1: average the betas within condition, per run

The direct translation. `split()` on the run column, then a column mean per
condition.

```{r beta-series-average}
domain <- abstract_domain(n_voxel, id = "beta-series-example")
by_run <- split(seq_len(n_trial), trial_design$run)

run_matrices <- lapply(by_run, function(rows) {
  t(vapply(conditions, function(condition) {
    colMeans(beta_series[rows[trial_design$condition[rows] == condition], ,
      drop = FALSE])
  }, numeric(n_voxel)))
})

averaged <- relation(run_matrices, effects = effect_space(conditions),
                     domain = domain)
plan_averaged <- plan_geometry(
  averaged,
  at = compile_frame(whole_brain(), domain),
  over = cross_partitions(averaged, independence = "independent",
                          generalizes_over = "run")
)
animacy <- c(face = 0.5, body = 0.5, house = -0.5, tool = -0.5)
energy_averaged <- contrast_energy(plan_averaged, animacy)
energy_averaged$total
```

Row names matter: `relation()` aligns partitions by effect **name**, so the
matrices must be named by condition. `t(vapply(...))` above produces that.

### Route 2: hand the trial betas over as observations

Better, when you can. `lm_relation_fit()` takes the trials themselves plus a
per-run design matrix and fits the condition estimates inside `crossform`,
which keeps the residual error channel that the analytic sampling law needs.
The design is the condition indicator; `effects` says each condition *is* its
own coefficient.

```{r beta-series-fit}
designs <- lapply(by_run, function(rows) {
  X <- stats::model.matrix(~ 0 + condition,
                           data = trial_design[rows, , drop = FALSE])
  colnames(X) <- conditions
  X
})
sources <- lapply(by_run, function(rows) beta_series[rows, , drop = FALSE])
targets <- diag(length(conditions))
dimnames(targets) <- list(conditions, conditions)

fit <- lm_relation_fit(
  sources, design = designs, effects = targets,
  effect_names = effect_space(conditions),
  sampling_unit = "trial", domain = domain
)
plan_fitted <- plan_geometry(
  fit$relation,
  at = compile_frame(whole_brain(), domain),
  over = cross_partitions(fit$relation, independence = "independent",
                          generalizes_over = "run")
)
energy_fitted <- contrast_energy(plan_fitted, animacy)
```

With this balanced indicator design the two routes estimate the same
condition means, so they report the same energy:

```{r beta-series-agree}
c(route_1 = energy_averaged$total,
  route_2 = energy_fitted$total,
  max_difference = max(abs(energy_averaged$total - energy_fitted$total)))
```

What route 2 buys is the error channel. Only the fitted relation can be asked
for a within-measurement standard error:

```{r beta-series-uncertainty}
sampling_capabilities(plan_fitted, fit)

round(sqrt(sampling_covariance(rdm_sampling_covariance(
  plan_fitted, fit, target = "null", at = 1L
))), 4)
```

Ask the averaged relation the same question and you get a refusal naming
exactly what is missing, rather than a number:

```{r beta-series-refusal}
catch_refusal(
  rdm_sampling_covariance(plan_averaged, averaged, target = "null", at = 1L)
)
```

Two practical notes. Runs need not have equal trial counts — `design` accepts
a named list with one matrix per partition, which is why it is built inside
`lapply()` above. And if your betas came from a GLM whose residuals are
temporally autocorrelated, pass `observation_whitener =` as well; see
`vignette("from-observations", package = "crossform")` for what that argument
is asserting.

## Distance and second-order conventions are not the same

This is the most likely source of surprise, and it is not a bug on either
side.

The two arguments of `rsa_model()` do separate jobs, and only one of them is
where the packages part company.

- **The distance.** `crossform`'s `rdm()` reports signed **crossvalidated
  squared Euclidean** distances, `d_ij = G_ii + G_jj - 2 G_ij` (squared
  Mahalanobis under a fixed metric). They can be negative, and are not
  truncated. rMVPA's `rsa_model()` takes `distmethod` in
  `{"pearson", "spearman"}` and nothing else, so its neural RDM is always a
  **correlation distance**. This is the difference that cannot be bridged.
  (rMVPA's crossnobis distances live elsewhere, on
  `contrast_rsa_model(estimation_method = "crossnobis")` and the
  `compute_crossnobis_distances_sl()` helper — which is why the parity check
  below uses those and not `rsa_model()`.)
- **The second-order statistic.** This is a *separate* argument, `regtype`,
  taking `"pearson"`, `"spearman"`, `"lm"`, or `"rfit"`. `crossform`'s `rsa()`
  is OLS with an intercept, which is the same family as `regtype = "lm"`.
  rMVPA's default is `regtype = "pearson"`, a correlation between the two
  vectorized RDMs; set `regtype = "lm"` and this half of the comparison lines
  up.

`crossform` cannot express correlation distance at all: dividing each pattern
by its own norm is not a fixed linear operation on the patterns, so it is
outside the bilinear core. Per-sphere mean-centering *is* expressible as a
metric; the unit-norm denominator is not. The boundary and the conditions
under which it could be crossed are recorded in the
[correlation-distance policy](correlation-distance-policy.html)
(`vignette("correlation-distance-policy", package = "crossform")` offline).

If you want the two packages to compute the same thing, the estimand both can
express is the condition-level crossvalidated squared Euclidean (crossnobis)
distance — which is the pairing used for the parity check below.

## What remains outside crossform

Each absence below was checked against the package's actual exports rather
than recalled.

- **Classification and decoding.** There is no classifier, no accuracy, no AUC,
  no confusion matrix, no `MVPAModels` registry. `crossform` computes
  second-order geometry only.
- **Universal group inference.** The experimental population layer now has
  declared group models, classical and HC3 intervals, and a participant-level
  null-imposed wild bootstrap, calibrated only in its recorded synthetic
  regimes and conditional on realized transport. It does not provide random
  fields, simultaneous/maxT bands, general multiplicity correction, or a
  marginal interval under informative coverage or transport learning.
- **Feature selection.** No `FTest`/`catscore` ranking, no `top_k` cutoff. The
  frame decides which features each measurement reads, and it is declared in
  advance.
- **Label permutation and multiplicity correction.** There is no label
  shuffling, cluster/random-field procedure, or FDR correction. The exported
  `population_wild_bootstrap()` is a different object: it resamples participants
  under a declared group-model null and keeps its conditional target explicit.
- **Parallel backends.** Execution is sequential. This is refused rather than
  silently downgraded:

```{r no-workers, error=TRUE}
compute_policy(workers = 4)
```

- **Preprocessing and I/O.** No registration, no smoothing, no file reading or
  writing. `as_neurovol()` hands you a `NeuroVol`; saving it is yours.

```{r absence-check, include=FALSE}
exports <- getNamespaceExports("crossform")
stopifnot(
  !any(grepl("classif|decod|accuracy|permut|feature_select|fdr",
             exports, ignore.case = TRUE)),
  "population_wild_bootstrap" %in% exports,
  inherits(try(compute_policy(workers = 2), silent = TRUE), "try-error")
)
```

Two things rMVPA users may expect that `crossform` *does* have, in a different
shape: a crossnobis path (`plan_crossnobis()`, `crossnobis()`, with
`shrinkage_precision()` and `metric_training_policy()` for the whitening), and
explicit refusals when an uncertainty claim is not supported by the fit
(`sampling_capabilities()`, `catch_refusal()`).

## The one measured comparison

The [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
runs both packages on eight stimulus categories estimated in twelve runs, over
577 ventral-temporal searchlights at an 11.25 mm radius. On the matched
estimand — `crossform`'s `rdm()` over a cross-run geometry against rMVPA's
`compute_crossvalidated_means_sl(estimation_method = "crossnobis")` followed by
`compute_crossnobis_distances_sl()` — the two agree to a **maximum absolute
difference of 8.88e-16** at every one of the 577 centers. Sphere membership
was verified identical rather than assumed.

Read that number with the exemplar's own caveats, which are recorded there in
full:

1. **It is not `rsa_model`.** The exemplar originally named `rsa_model` for
   the rMVPA arm and could not use it: its distances are correlation
   distances, which `crossform` does not compute. `vector_rsa_model` is
   trial-level. `contrast_rsa_model` does build the cross-validated
   second-moment matrix internally, but what it returns is a signed
   per-contrast contribution evaluated at each searchlight's center voxel,
   never the condition-by-condition distances themselves. The crossnobis
   helper is the only route, and `compute_crossnobis_distances_sl()` is not
   exported, so the script calls it with `:::`.
2. **It is version-sensitive.** The comparison requires an rMVPA build whose
   crossnobis fold estimates are per-run condition means. The script asserts
   that semantics and stops with an explanatory error otherwise, rather than
   checking a version string.
3. **The originally specified estimand was not achievable.** A
   correlation-distance RDM scored by Spearman cannot be expressed by
   `crossform` at all: `rsa_model()`'s `distmethod` offers only `"pearson"`
   and `"spearman"`, and `crossform` has no correlation distance to meet
   either one. The second-order statistic is not the obstacle — that is a
   separate argument, `regtype`, whose `"lm"` setting is the same OLS family
   as `crossform`'s `rsa()`. It is the distance that cannot be shared. The
   matched comparison is therefore a substitution, and the exemplar says so.

What the number demonstrates is numerical parity on a shared estimand. It
demonstrates nothing about speed, and nothing about correlation-distance RSA.

## Where to go next

- `vignette("introduction", package = "crossform")` — what the energies, RDMs,
  and RSA coefficients mean.
- `vignette("neuroim2-data", package = "crossform")` — the input and output
  adapters, if your data is already in neuroim2 objects.
- [Reading contrast energies, RDMs, and uncertainty](interpreting-results.html)
  (`vignette("interpreting-results", package = "crossform")`) — how to read
  the numbers, and the three interpretive traps.
- [What is novel in crossform?](novelty.html)
  (`vignette("novelty", package = "crossform")`) — what is architecture and
  what is established statistics.
