---
title: "Fit condition effects from observations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fit condition effects from observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

Use this guide when you have one scan-by-feature response matrix per run, event
and confound tables, and a design matrix or design compiler. The goal is to fit
named condition effects in every run and pass them to `contrast_energy()`,
`rdm()`, or `rsa()`. The returned `effect_relation_fit` can also retain
residuals for within-participant uncertainty calculations.

The workflow has five steps: bind observations to their clocks and metadata,
declare the condition effects, declare the observation model, call
`plan_relation()`, and fit it with `estimate_relation()`. A separate
`plan_geometry()` then states what those fitted effects must reproduce across.

## What data does the fit require?

Three objects, and nothing else:

1. **one scan-by-feature response matrix per run**, in a named list — scans in
   rows, neural features in columns;
2. **an events table** — one row per event, with `partition`, `onset`,
   `duration`, and the condition label, on the same clock as the scan index;
3. **a confounds table** — one row per scan, joined to the responses by
   `partition` and `observation_id`, carrying nuisance regressors and an
   optional Boolean censor column.

The example below generates all three so the vignette can check its own
answers. It has three runs, three named conditions, twelve neural features, and
two censored scans per run. Read the code as a description of the shape your own
data must have; substitute your matrices and tables for the generated ones.

The design and the planted signal come first. Each run's responses are the
condition means plus a linear drift, a motion covariate, and noise.

```{r simulate-responses}
set.seed(20260815)
partitions <- paste0("run-", 1:3)
n_scan <- 18L
conditions <- c("face", "body", "tool")
domain <- abstract_domain(12L, id = "from-observations:v1")

condition <- factor(
  rep(conditions, length.out = n_scan),
  levels = conditions
)
indicator <- stats::model.matrix(~ condition - 1)
colnames(indicator) <- conditions
motion <- sin(seq(0, 2 * pi, length.out = n_scan))
drift <- seq(-1, 1, length.out = n_scan)
design <- cbind(indicator, drift = drift, motion = motion)
rownames(design) <- as.character(seq_len(n_scan))

truth <- matrix(
  rnorm(length(conditions) * domain$n_features, sd = 0.35),
  nrow = length(conditions),
  dimnames = list(conditions, domain$feature_ids)
)
nuisance_truth <- matrix(
  rnorm(2L * domain$n_features, sd = 0.08),
  nrow = 2L
)

# (1) one scan-by-feature response matrix per run
response_runs <- stats::setNames(lapply(partitions, function(partition) {
  design[, conditions, drop = FALSE] %*% truth +
    design[, c("drift", "motion"), drop = FALSE] %*% nuisance_truth +
    matrix(rnorm(n_scan * domain$n_features, sd = 0.4),
      nrow = n_scan)
}), partitions)
```

Each response matrix needs a clock. `observation_index()` records the scan
identifiers, their partition, and their acquisition times; the events table then
speaks the same units.

```{r simulate-tables}
scan_indexes <- stats::setNames(lapply(partitions, function(partition) {
  observation_index(
    seq_len(n_scan), partition,
    time = seq(0, by = 2, length.out = n_scan), units = "seconds"
  )
}), partitions)

# (2) one events table for the whole study
event_table <- do.call(rbind, lapply(partitions, function(partition) {
  data.frame(
    partition = partition,
    event_id = paste0(partition, "-", seq_len(n_scan)),
    onset = seq(0, by = 2, length.out = n_scan),
    duration = 0,
    condition = as.character(condition)
  )
}))

# (3) one confounds table, one row per scan, with a censor column
confound_table <- do.call(rbind, lapply(partitions, function(partition) {
  data.frame(
    partition = partition,
    observation_id = seq_len(n_scan),
    motion = motion,
    retained = seq_len(n_scan) <= 16L
  )
}))
```

That is the whole input. Its shape:

```{r inspect-inputs}
data.frame(
  partition = names(response_runs),
  scans = vapply(response_runs, nrow, integer(1)),
  features = vapply(response_runs, ncol, integer(1))
)
utils::head(event_table, 3)
utils::head(confound_table, 3)
```

## Bind observations to events and confounds

The constructors below turn the matrices and tables into records with named
axes. `study()` then verifies their correspondence. It rejects missing or extra
confound rows, incompatible identifier types, mismatched clock units, and
events outside the observation window.

```{r bind-study}
neural_record <- observations(response_runs, scan_indexes, domain)
event_record <- observation_events(event_table)
confound_record <- observation_confounds(
  confound_table, censor = "retained"
)
hierarchy <- partition_hierarchy(data.frame(
  partition = partitions, run = partitions,
  session = "session-1", subject = "subject-1"
))

study_facts <- study(
  neural_record, event_record, confound_record, hierarchy
)
study_capabilities(study_facts)
```

The returned row reports aligned observations, resolved timing, a bound
partition hierarchy, and stable source revisions. The study's row lineage also
records which scans censoring retained. `study()` does not decide whether
`condition` or `motion` is a target, nuisance term, or unused column. Those
roles belong to the design model.

## Declare effects before choosing coefficient coding

The requested effects are the three condition amplitudes, in arbitrary BOLD
units. `condition_space()` names those coordinates. `effect_map()` requests all
three, so later calls can form contrasts without refitting the run models.

```{r semantic-effects}
condition_coordinates <- condition_space(
  conditions,
  basis_id = "scan-level-condition-mean:v1",
  units = "arbitrary-BOLD"
)
effect_weights <- diag(length(conditions))
dimnames(effect_weights) <- list(conditions, conditions)
named_effects <- effect_map(effect_weights, condition_coordinates)
```

The compiled design has one column per condition plus linear drift and motion.
`coefficient_parameterization()` declares which linear combinations of those
coefficients equal the named condition amplitudes. This declaration makes the
effect request independent of whether the compiler uses cell-means or treatment
coding.

```{r design-model}
condition_from_coefficients <- cbind(
  diag(length(conditions)), drift = 0, motion = 0
)
dimnames(condition_from_coefficients) <- list(
  conditions, colnames(design)
)
parameterization <- coefficient_parameterization(
  condition_from_coefficients,
  condition_coordinates,
  coding_id = "cell-means-plus-nuisance"
)

designs <- stats::setNames(
  rep(list(design), length(partitions)), partitions
)
parameterizations <- stats::setNames(
  rep(list(parameterization), length(partitions)), partitions
)
declared_design <- design_model(
  specification = list(
    target = "scan-level condition means",
    nuisance = c("linear drift", "declared motion covariate"),
    censor_policy = "use the bound retained column"
  ),
  conditions = condition_coordinates,
  designs = designs,
  parameterizations = parameterizations,
  row_ids = lapply(scan_indexes, `[[`, "observation_id"),
  solver = "qr",
  protocol = "from-observations-example",
  protocol_version = "1",
  package_version = "1"
)
```

`design_model()` returns the declared design plus the metadata needed to check
and reproduce its compilation. This example supplies matrices directly. An
external compiler may instead build hemodynamic response functions and event
grammars, but it must provide the same named condition map, row lineage, ranks,
aliasing information, and solver diagnostics.

## Declare the fit and its assumptions

The observation model states assumptions that the data tables cannot prove.
Here the estimator is ordinary least squares (OLS), scans are the sampling
unit, and runs are declared independently acquired conditional on the model.
Changing any of these declarations changes the requested relation.

```{r relation-plan}
noise_model <- observation_model(
  "ols",
  sampling_unit = "scan",
  independence = "runs independently acquired conditional on the model"
)
relation_request <- plan_relation(
  study_facts,
  declared_design,
  named_effects,
  noise_model
)
relation_request
compiler_conformance(relation_request)
```

`plan_relation()` compiles and validates the design metadata without reading
neural values. It returns an `effect_relation_plan`. If a requested effect is
not estimable in a run, the call stops and names the run, rank, aliased
regressors, cause, and possible remedies.

The plan, design receipt, and fitted relation have separate identities:

```{r identity-ladder}
receipts <- relation_plan_receipts(relation_request)
fit <- estimate_relation(relation_request)
data.frame(
  object = c("relation plan", "run-1 design receipt", "fitted relation"),
  identity = substr(c(
    relation_request$relation_plan_id,
    receipts$`run-1`$design_receipt_id,
    fit$signature
  ), 1, 32)
)
```

Changing the solver from QR to SVD changes the design receipt, not the named
condition-space request. A supported change from cell-means to treatment coding
does the same. Changing the requested effects, units, observation law, or
independence declaration changes the relation plan. `fit$signature` binds the
plan, the realized receipts, and the exact observation revisions used by
`estimate_relation()`.

## Ask what reproduces across runs

`estimate_relation()` emits the same `effect_relation_fit` consumed everywhere
else in the package. `cross_partitions()` requests products between different
runs; `generalizes_over = "run"` records that run is the axis across which the
effect must reproduce.

The frame decides where results are reported. This example uses three
anatomical regions, so every result below has three rows. Searchlights, single
voxels, and the whole brain are the other choices; the package calls any one of
them a *measurement*, and nothing downstream changes when you swap one for
another.

```{r geometry}
frame <- compile_frame(
  regions(rep(paste0("roi-", 1:3), each = 4L), normalization = "local"),
  domain
)
generalization <- cross_partitions(
  fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
geometry_plan <- plan_geometry(fit$relation, frame, generalization)

face_minus_body <- c(face = 1, body = -1, tool = 0)
effect <- contrast_energy(geometry_plan, face_minus_body)
distances <- rdm(geometry_plan)

tool_category <- outer(
  c(face = 0, body = 0, tool = 1),
  c(face = 0, body = 0, tool = 1),
  function(left, right) as.numeric(left != right)
)
model_fit <- rsa(geometry_plan, models = list(tool = tool_category))
```

`contrast_energy()` returns the signed regional contrast and its coherent,
configuration, and total cross-run energies. `rdm()` returns the three pairwise
condition distances for each region. `rsa()` returns an intercept and the
coefficient for the named tool-category model.

```{r summarize-geometry}
round(cbind(
  signed = effect$signed,
  coherent = effect$coherent,
  configuration = effect$configuration,
  total = effect$total
), 3)
round(distances$values, 3)
round(model_fit$coefficients, 3)
```

For every region, total energy must equal coherent plus configuration energy.
The hidden check below also verifies the expected result dimensions and every
compiler-conformance guarantee.

```{r downstream-checks, include=FALSE}
stopifnot(
  isTRUE(all.equal(
    effect$total,
    effect$coherent + effect$configuration,
    tolerance = 1e-12
  )),
  identical(dim(distances$values), c(3L, 3L)),
  identical(dim(model_fit$coefficients), c(3L, 2L)),
  all(as.matrix(compiler_conformance(relation_request)[-1L]))
)
```

Because the fit retained residuals and declared a fixed observation model, the
package can calculate analytic standard errors for the three RDM distances at
a selected region:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  geometry_plan, fit, target = "plugin", at = 1L
)
round(sqrt(sampling_covariance(distance_covariance)), 3)
```

These are within-participant standard errors under the declared fixed,
separable error model, which represents sampling and neural-feature covariance
as separate factors. They are not population-level uncertainty. A learned
whitener would still permit the point estimates but would not receive this
analytic uncertainty capability.

## How do BIDS and external compilers fit?

`bids_study()` adapts one task `events.tsv` and one row-aligned confound table
per partition to the same `study()` object used above. It preserves extra event
and confound columns. The design compiler, not the adapter, decides which
columns become target or nuisance terms. If censoring is intended, the call
must name the Boolean censor column. This keeps BIDS at the file boundary; the
core study object is not BIDS-shaped.

The signature mirrors the `study()` call above: an `observations()` record,
then one events file and one confounds file per partition, in partition order.

```{r bids-adapter, eval=FALSE}
runs <- paste0("run-", 1:3)

bids_facts <- bids_study(
  observations(response_runs, scan_indexes, domain),
  event_files = file.path(
    "sub-01", "func",
    sprintf("sub-01_task-objects_%s_events.tsv", runs)
  ),
  confound_files = file.path(
    "sub-01", "func",
    sprintf("sub-01_task-objects_%s_desc-confounds_timeseries.tsv", runs)
  ),
  partitions = runs,
  censor = "retained",
  units = "seconds"
)
```

`fmridesign_design_model()` adapts the supported `fmridesign` version into a
design that names condition effects independently of coefficient columns.
`fmrireg_relation()` can execute an OLS relation plan for point-estimate
parity. That adapter does not return an analytic error channel because
crossform cannot verify the required residual covariance contract from its
result.

If you have compiled design and target matrices but not their semantic
condition mapping, use `raw_design_model()` with `raw_effect_map()`. Their
numeric values then become part of the plan identity, and the fit cannot claim
that a different coefficient coding represents the same request. If only beta
matrices remain, use `relation()` for point geometry; analytic uncertainty is
unavailable because the residual channel is missing.

The next call depends on the result you need. Use `contrast_energy()` for a
named effect, `rdm()` for condition distances, `rsa()` for a fixed linear
model, and `rdm_sampling_covariance()` only after `sampling_capabilities()`
confirms that the fit and plan meet its assumptions.
