---
title: "Reading contrast energies, RDMs, and uncertainty"
description: >
  Interpret signed energies, coherent shares, RDMs, RSA coefficients and predictive gain, with worked examples of frame dependence and uncertainty.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading contrast energies, RDMs, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

Use this guide after the [introduction](introduction.html) to interpret the
numbers returned by a geometry plan. Start with the contrast columns, then
follow the section relevant to your result: coherent shares, an exploratory
map, RDMs, fixed RSA coefficients, predictive gain or uncertainty. Everything below runs on the generated fixture, which plants two
blocks of equal amplitude at opposite ends of the volume — one whose sign
alternates between neighboring voxels, one that shifts every voxel the same
way — so each half of the energy decomposition has a known home.

```{r example}
example <- example_fmri_effects()
pairing <- cross_partitions(
  example$fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
plan <- plan_geometry(example$fit$relation, at = example$frame, over = pairing)
effect <- contrast_energy(plan, example$contrast)
peak <- which.max(effect$total)

pattern <- example$truth$pattern_measurements
mean_block <- example$truth$mean_measurements
c(
  measurements = length(effect$total),
  pattern_searchlights = length(pattern),
  mean_searchlights = length(mean_block),
  overlap = length(intersect(pattern, mean_block))
)
```

The two blocks have equal planted amplitude and size, and their searchlights
do not overlap. This controlled comparison lets us examine spatial sign
structure while keeping those factors fixed.

## The four columns of a contrast view

```{r columns}
head(as.data.frame(effect), 4)
```

- **`signed`** is the ordinary signed contrast of the frame-weighted mean
  pattern. It is a *first moment*: it has the units and the sign of your
  effect, and it is the only column that answers "which way?". It is not
  crossvalidated in the reproducibility sense — it is a retained marginal.
- **`coherent`** is the crossvalidated energy carried by that
  frame-weighted mean pattern. It is a *second moment* — a squared quantity —
  so it is not the signed effect and not its square either.
- **`configuration`** is the crossvalidated energy in reproducible spatial
  *departures* from the weighted mean, orthogonal to `coherent` under the
  frame's weights.
- **`total`** is their exact sum.

```{r partition}
max(abs(effect$total - (effect$coherent + effect$configuration)))
```

The fixture separates the two energies by construction. Here is the strongest
searchlight in each block:

```{r two-blocks}
strongest <- function(measurements) {
  measurements[which.max(effect$total[measurements])]
}
at <- c(pattern = strongest(pattern), mean = strongest(mean_block))
block_rows <- as.data.frame(effect)[at, -1]
rownames(block_rows) <- names(at)
round(block_rows, 4)
```

```{r two-blocks-check, include=FALSE}
stopifnot(
  at[["pattern"]] == peak,
  effect$configuration[at[["pattern"]]] >
    50 * abs(effect$coherent[at[["pattern"]]]),
  effect$coherent[at[["mean"]]] >
    50 * abs(effect$configuration[at[["mean"]]]),
  abs(effect$signed[at[["pattern"]]]) < 0.1,
  abs(effect$signed[at[["mean"]]]) > 1.5,
  is.na(effect$coherence_fraction[at[["pattern"]]]),
  effect$coherence_fraction[at[["mean"]]] > 0.99
)
```

Equal amplitude, equal block size, near-equal `total` — and opposite
decompositions. The pattern block's energy is almost all `configuration`: a
shape that reproduces across runs while the sphere's average barely moves,
which is why its `signed` value is close to zero and its signed regional-mean contrast is small. The mean block's energy is almost all `coherent`,
and its `signed` value is large, because the whole neighborhood shifts
together.

Note the `NA` in the first row. That searchlight is the largest `total` in the
map and it reports no coherence fraction at all, because its `coherent`
estimate landed a hair below zero. That is the subject of the section after
next.

### Negatives are estimates, not errors

Each energy combines products of estimates from different runs. With
independent, unbiased partition estimates, its expectation is zero under the
null. Negative estimates are therefore legitimate. Zero expectation alone does
not imply that positive and negative values are equally likely.

```{r negatives}
c(
  measurements = length(effect$total),
  total_below_zero = sum(effect$total < 0),
  coherent_below_zero = sum(effect$coherent < 0),
  configuration_below_zero = sum(effect$configuration < 0),
  min_total = min(effect$total)
)
```

Keep the negative values in summaries. Clipping them at zero introduces
positive bias, including where the true effect is absent. An average over
prespecified measurements or participants should use the signed estimates;
selecting only positive values changes the quantity being summarized.

## The coherence fraction and its validity flag

`coherence_fraction` is `coherent / total`: the share of reproducible energy
carried by the local mean rather than by pattern shape. It is reported only
where the raw components actually form a nonnegative partition — that is,
where `total > 0`, `coherent >= 0`, and `configuration >= 0`. Elsewhere it is
`NA`, and a companion logical says why you got `NA`.

```{r validity}
c(
  measurements = length(effect$total),
  valid = sum(effect$coherence_fraction_valid),
  na = sum(is.na(effect$coherence_fraction))
)
```

```{r validity-reasons}
c(
  total_not_positive = sum(effect$total <= 0),
  coherent_negative = sum(effect$coherent < 0),
  configuration_negative = sum(effect$configuration < 0)
)
```

Outside this gate the ratio can leave [0, 1], and a denominator near zero
makes it unstable. The package therefore withholds it. `NA` here means "this measurement has no
interpretable coherent share", not "missing data".

The gate is not a test for signal. The strongest pattern-block measurement
above has an `NA` share because its coherent estimate is slightly negative,
even though its total energy is large.

The three sections below are the caveats that matter once you start reporting
those fractions.

## Trap (a): the coherent share shrinks as the sphere grows

Changing the radius changes the spatial quantity being measured, even when
the neural data stay fixed. In this example, a larger sphere dilutes a fixed
block of mean-shift signal with surrounding voxels.

A simple noise-free case explains the effect. Suppose a fraction $f$ of a
uniformly weighted, locally normalized support carries contrast $a$, and the
rest carries zero. Then

\[
\mathrm{total}=a^2 f,\qquad
\mathrm{coherent}=a^2 f^2,\qquad
\mathrm{configuration}=a^2 f(1-f).
\]

The coherent share is $f$. Enlarging the support lowers that share as the
signal occupies less of it. Configuration need not stay constant; it is the
remainder after the squared local mean is removed.

Extra noise dimensions can change variability and which noisy ratios pass
the validity gate. They do not, by themselves, add positive expected
crossvalidated energy. Read the radius sweep as a change in the declared
measurement, not as a count of neural dimensions.

```{r sphere-size}
radii <- c(2, 3, 4.5, 6)
sweep <- t(vapply(radii, function(r) {
  frame_r <- compile_frame(searchlights(radius = r), example$domain)
  effect_r <- contrast_energy(
    plan_geometry(example$fit$relation, at = frame_r, over = pairing),
    example$contrast
  )
  c(
    median_support = median(Matrix::rowSums(frame_r$weights != 0)),
    valid = sum(effect_r$coherence_fraction_valid),
    median_fraction = median(effect_r$coherence_fraction, na.rm = TRUE),
    pattern_block = median(effect_r$coherence_fraction[pattern], na.rm = TRUE),
    mean_block = median(effect_r$coherence_fraction[mean_block], na.rm = TRUE)
  )
}, numeric(5)))
data.frame(radius = radii, round(sweep, 3))
```

```{r sphere-size-check, include=FALSE}
stopifnot(
  sweep[1, "median_fraction"] == 1,
  sweep[1, "pattern_block"] == 1,
  sweep[1, "mean_block"] == 1,
  all(diff(sweep[, "median_fraction"]) <= 0),
  all(diff(sweep[, "mean_block"]) <= 0),
  all(sweep[-1, "mean_block"] > sweep[-1, "pattern_block"])
)
```

The mechanism is visible directly. Hold the radius at 4.5 mm and compare, over
the searchlights whose support touches the mean-shift block, the two energies
against the squared weighted share of planted voxels in each support:

```{r sphere-dilution}
weighted_share <- function(frame, features) {
  as.numeric(Matrix::rowSums(frame$weights[, features, drop = FALSE]) /
               Matrix::rowSums(frame$weights))
}

frame_45 <- compile_frame(searchlights(radius = 4.5), example$domain)
effect_45 <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_45, over = pairing),
  example$contrast
)
share <- weighted_share(frame_45, example$truth$mean_features)
touching <- which(share > 0)

c(
  searchlights = length(touching),
  coherent = round(cor(effect_45$coherent[touching], share[touching]^2), 3),
  configuration = round(
    cor(effect_45$configuration[touching], share[touching]^2), 3
  )
)
```

```{r sphere-dilution-check, include=FALSE}
stopifnot(
  cor(effect_45$coherent[touching], share[touching]^2) > 0.99,
  cor(effect_45$configuration[touching], share[touching]^2) < 0.6
)
```

The coherent energy is almost a deterministic function of the squared planted
share. The configuration energy is not. This is the dilution mechanism predicted by the simple mean-shift example.

At radius 2 every support is a single voxel, so the weighted mean *is* the
pattern and the fraction is exactly 1 by construction — in both blocks. It
then falls. The decomposition keeps the two blocks in the right order at every
larger radius, the mean block several times above the pattern block, so the
ordering is real. But the mean block's own fraction still slides from 0.347 to
0.289 while nothing about the planted signal changes. A sentence such as
"roughly 30% of the reproducible signal was carried by the mean response" is
therefore a statement about your radius as much as about the brain. Report the
radius and the support sizes alongside the fraction, and never compare
fractions across analyses with different frames.

## Trap (b): "coherent" means coherent under *your* weights

The decomposition is orthogonal in the frame-weighted inner product, so the
weights are part of the definition. Reweighting the *same* voxels moves energy
between the two components — and changes the total.

```{r weighting}
frame_uniform <- compile_frame(searchlights(radius = 4.5), example$domain)

# Same supports, Gaussian instead of uniform weights.
coords <- example$domain$coordinates
uniform_weights <- as.matrix(frame_uniform$weights)
centers <- match(frame_uniform$index$measurement, example$domain$feature_ids)
gaussian_weights <- uniform_weights
for (i in seq_len(nrow(uniform_weights))) {
  members <- which(uniform_weights[i, ] != 0)
  offsets <- coords[members, , drop = FALSE] -
    matrix(coords[centers[i], ], length(members), 3, byrow = TRUE)
  w <- exp(-rowSums(offsets^2) / (2 * 3^2))
  gaussian_weights[i, ] <- 0
  gaussian_weights[i, members] <- w / sum(w)
}
frame_gaussian <- additive_frame(
  gaussian_weights, normalization = "local", domain = example$domain
)

identical(which(uniform_weights != 0), which(gaussian_weights != 0))
```

Read the two frames at the searchlight sitting on each block's center. That
measurement is chosen from geometry alone, not from the results, so nothing
below is a selected extreme.

```{r weighting-effect}
effect_uniform <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_uniform, over = pairing),
  example$contrast
)
effect_gaussian <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_gaussian, over = pairing),
  example$contrast
)

centered_on <- function(features) {
  centroid <- colMeans(coords[features, , drop = FALSE])
  which.min(rowSums(
    (coords[centers, , drop = FALSE] -
      matrix(centroid, length(centers), 3, byrow = TRUE))^2
  ))
}
middle <- c(
  pattern = centered_on(example$truth$planted_features),
  mean = centered_on(example$truth$mean_features)
)

reweighted <- cbind(
  uniform_coherent = effect_uniform$coherent[middle],
  gaussian_coherent = effect_gaussian$coherent[middle],
  uniform_total = effect_uniform$total[middle],
  gaussian_total = effect_gaussian$total[middle]
)
rownames(reweighted) <- names(middle)
round(reweighted, 4)
```

```{r weighting-check, include=FALSE}
stopifnot(
  identical(which(uniform_weights != 0), which(gaussian_weights != 0)),
  identical(frame_uniform$index$measurement, example$frame$index$measurement),
  middle[["pattern"]] %in% pattern,
  middle[["mean"]] %in% mean_block,
  effect_gaussian$coherent[middle[["pattern"]]] <
    0.6 * effect_uniform$coherent[middle[["pattern"]]],
  abs(
    effect_gaussian$coherent[middle[["mean"]]] -
      effect_uniform$coherent[middle[["mean"]]]
  ) < 0.02 * effect_uniform$coherent[middle[["mean"]]],
  effect_uniform$total[middle[["pattern"]]] !=
    effect_gaussian$total[middle[["pattern"]]],
  effect_uniform$total[middle[["mean"]]] !=
    effect_gaussian$total[middle[["mean"]]]
)
```

Identical voxels, identical data, different numbers — but not equally
different. The mean block's coherent energy barely moves: a shift that is the
same in every voxel is the same shift under any nonnegative weights. The pattern block's coherent energy roughly halves because its alternating
signal cancels differently under the two weightings. Both are valid weighted
projections of the same spatial pattern. Both totals move.

This is not instability: each frame defines a different, well-specified
estimand, and the plan records which one you asked for. It does mean that "the
coherent component" is not a property of the brain region. It is a property of
the region *and* the weights you chose to average it with.

## Trap (c): reported fractions are a selected sample

Because the fraction exists only where the components form a nonnegative
partition, the set of measurements that report one is not a random subset of
the map. The gate depends on the same quantities that appear in the ratio, so
it is enriched for measurements with real signal — and, among the rest, for
whichever noise realizations happened to make both components positive.

```{r selection}
frame_small <- compile_frame(searchlights(radius = 3), example$domain)
effect_small <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_small, over = pairing),
  example$contrast
)
signal <- example$truth$signal_measurements
valid <- effect_small$coherence_fraction_valid

c(
  signal_measurements = length(signal),
  signal_reporting_a_fraction = sum(valid[signal]),
  other_measurements = length(valid) - length(signal),
  other_reporting_a_fraction = sum(valid[-signal])
)
```

```{r selection-blocks}
round(c(
  pattern_block = mean(valid[pattern]),
  mean_block = mean(valid[mean_block]),
  elsewhere = mean(valid[-signal])
), 3)
```

```{r selection-share}
c(
  share_of_map_reporting = mean(valid),
  share_of_reporters_that_are_signal = mean(seq_along(valid)[valid] %in% signal)
)
```

```{r selection-check, include=FALSE}
stopifnot(
  # Every frame here is one measurement per voxel in the same order, so the
  # truth indices carry across radii.
  identical(frame_small$index$measurement, example$frame$index$measurement),
  mean(valid[pattern]) > 0.9,
  mean(valid[mean_block]) > 0.9,
  mean(valid[-signal]) < 0.25
)
```

In this fixture, both blocks clear the gate at similar rates. That need not
hold for every signal shape or noise regime. Under a quarter
of the remaining measurements report a fraction, on noise alone. The result is
that the average fraction "over the map" is really an average over a set that
is 70% planted signal. That is fine if you say so, and misleading if you do
not. If you want a map-wide summary, define the denominator explicitly — for
instance the share of *all* measurements that report a coherent share at all,
the first number above — rather than averaging the reported values and calling
the result a property of the map.

The same warning applies to any post-hoc gate: thresholding on `total > 0`
and then summarizing `coherent` conditions the summary on the numerator's
sibling.

## Reading a map without ground truth

Every figure so far could mark the right answer, because the fixture carries
`truth$signal_measurements`. Your data will not. This section is about what the
same picture tells you when nothing is marked.

```{r blind-map, fig.width=8, fig.height=4, fig.cap="The same contrast view with no highlight supplied. Nothing here identifies the two planted blocks; the reading below uses only what is on the page."}
plot(effect)
```

### Zero is a reference, not a threshold

Independent, unbiased partition estimates give null energy an expectation of
zero. That provides a useful reference line. It does not tell you which
locations are null, how variable their estimates are, or what threshold
controls false positives across the map.

Here the generated truth lets us examine measurements without planted signal:

```{r floor-known}
quiet <- setdiff(seq_along(effect$total), example$truth$signal_measurements)
round(c(
  measurements = length(quiet),
  mean = mean(effect$total[quiet]),
  sd = stats::sd(effect$total[quiet]),
  fraction_below_zero = mean(effect$total[quiet] < 0),
  largest = max(effect$total[quiet])
), 4)
```

Without the truth, the median and MAD summarize the observed map. Their
meaning depends on its mixture of signal and noise; they are not automatically
estimates of a null distribution:

```{r floor-blind}
round(c(
  median_over_all = stats::median(effect$total),
  mad_over_all = stats::mad(effect$total),
  fraction_below_zero_over_all = mean(effect$total < 0)
), 4)
```

```{r floor-check, include=FALSE}
stopifnot(
  abs(stats::median(effect$total)) < 0.01,
  abs(mean(effect$total[quiet])) < 0.01,
  stats::mad(effect$total) > stats::sd(effect$total[quiet])
)
```

In this fixture the blind median lies near the known null center, while the
MAD is more than twice the standard deviation of the known null measurements.
The two signal blocks occupy 37% of the map and affect that summary. A different
signal prevalence or spatial noise pattern could change both comparisons.

### Candidates, not findings

Sorted, this map has a visible shoulder: rank 104 is `0.478` and rank 105 is
`0.030`, while rank 1 is `4.10`. The shoulder falls exactly at the edge of the
planted set — all 104 planted searchlights outrank every other measurement.
That is a fact about this fixture, not a method.

```{r shoulder}
round(sort(effect$total, decreasing = TRUE)[c(1, 10, 40, 104, 105, 150, 280)], 4)
```

```{r shoulder-check, include=FALSE}
stopifnot(
  identical(
    sort(order(effect$total, decreasing = TRUE)[seq_along(signal)]),
    signal
  )
)
```

A visible shoulder is useful for exploration, but this view supplies no
spatial threshold or selected-peak p-value. A population analysis has its own
declared target and assumptions; it does not retroactively calibrate this map.
Treat selected locations as candidates for further study:

- **Describe its estimated composition.** High configuration attributes more
  estimated energy to departures from the local mean; high coherent energy
  attributes more to that mean. Selection can exaggerate either component,
  so the description does not establish replication.
- **Attach a within-measurement standard error at a measurement you chose in
  advance**, using `rdm_sampling_covariance()` with `target = "null"` — and
  only if the relation kept its residual channel, which means fitting with
  `lm_relation_fit()` or `estimate_relation()` rather than importing betas.
- **Take the candidates somewhere else**: a held-out session, an independent
  dataset, or a group-level analysis performed outside this package.

### A pointwise error bar does not account for peak selection

If you find `peak <- which.max(effect$total)` and then report a standard error at
`peak`, the pointwise standard error does not account for having chosen an extreme
estimate. An interval formed from it need not retain its nominal coverage
after selection. The peak summaries here illustrate the fixture; they do not
perform inference for a selected location.

Specify the measurement in advance, or treat the peak as a candidate and estimate it
somewhere you did not select it.

## Reading the RDM

`peak` is the pattern block's strongest searchlight, so the two within-category
pairs should sit near zero and the four across-category pairs near the planted
separation.

```{r rdm}
distances <- rdm(plan)
distances$pairs
round(distances$values[peak, ], 3)
```

For conditions *i* and *j*, `rdm()` reports
`d_ij = G_ii + G_jj - 2 G_ij`. Under cross-partition pairing this is a
**signed crossvalidated squared Euclidean** distance; with a fixed neural
precision metric it is a squared Mahalanobis distance. Three consequences:

1. **They are squared.** Do not take square roots to "get distances back" —
   the square root of a negative estimate is undefined, and the square root of
   a noisy positive one is biased.
2. **They can be negative**, for the same reason the energies can, and for
   the same reason they are kept.
3. **They are not `1 - Pearson correlation`.** Correlation distance divides
   each pattern by its own norm, which is not a fixed linear operation, so it
   cannot be a query against this geometry. The package refuses to substitute
   it silently; the reasoning and the conditions under which the boundary
   could be crossed are in the correlation-distance policy
   ([online](correlation-distance-policy.html), or
   `vignette("correlation-distance-policy")` offline).

Because the RDM is a geometry query and not a decomposition, it does not tell
the two blocks apart. The strongest searchlight in each block returns much the
same six distances, even though their energies split in opposite directions:

```{r rdm-blocks}
round(rbind(
  pattern = distances$values[at[["pattern"]], ],
  mean = distances$values[at[["mean"]], ]
), 3)
```

```{r rdm-blocks-check, include=FALSE}
stopifnot(
  max(abs(
    distances$values[at[["pattern"]], ] - distances$values[at[["mean"]], ]
  )) < 0.5
)
```

The total RDM alone cannot distinguish those spatial organizations. Use
`contrast_energy()` for a contrast decomposition; `rdm()` also accepts a
`component` when you want distances from a particular spatial component.

## Reading RSA coefficients

`rsa()` regresses the measured distances on your model RDMs. It is ordinary
least squares, not a rank correlation.

```{r rsa}
category <- rsa(plan, models = list(category = example$model_rdm))
colnames(category$coefficients)
round(category$coefficients[peak, ], 4)
```

There are two columns because an **intercept is fitted by default**. It absorbs
the overall level of signed distances, letting the model slope describe their
relationship after accounting for that level. Without an intercept, the slope
must explain both level and shape: uniformly large distances can increase the
coefficient of a positive-valued model even when its shape matches poorly.

```{r rsa-nointercept}
no_intercept <- rsa(
  plan, models = list(category = example$model_rdm), intercept = FALSE
)
c(
  with_intercept = category$coefficients[peak, "category"],
  without_intercept = no_intercept$coefficients[peak, "category"]
)
```

Drop the intercept only when you have a reason to believe the level is
meaningful and shared. Coefficients are in units of distance per unit of model
RDM, so they are comparable across measurements within one analysis but not
across analyses with differently scaled model RDMs.

## Reading a learned geometry prediction

`fit_geometry()` learns a model-supported PSD form on training data.
`score_geometry()` evaluates that frozen form on independent runs using
$2\langle F,G_{\mathrm{test}}\rangle_F-\|F\|_F^2$. The result is a signed
gain in squared geometry units, relative to predicting zero. Positive gain
supports improved prediction in expectation under the declared independence
assumptions; negative gain means the prediction's amplitude costs more than
its independent evidence earns.

For each fitted mode, keep three quantities apart:

| Quantity | Meaning |
|---|---|
| `amplitude` | Strength learned on training data |
| `evidence` | Signed geometry in that frozen direction on evaluation data |
| `gain` | Twice amplitude times evidence, minus amplitude squared |

A positive training amplitude does not establish replication. Even positive
test evidence can produce negative gain if the training amplitude is too
large. Do not drop such a mode after inspecting final evaluation results:
that would select a different prediction using the test data. Rank, penalty,
model weights and preprocessing belong to training or inner validation.

Gain is neither a fraction explained nor a trace energy. Under zero signal,
its conditional expectation is minus the prediction's squared norm; the
test inner product alone is centered on zero. Likewise, the descriptive
projection from `latent_geometry(rank = ...)` and the trace split from
`model_geometry()` do not measure independent predictive improvement.

The [predictive geometry guide](predictive-geometry.html)
(`vignette("predictive-geometry", package = "crossform")` offline) executes a
case with mode gains 4 and -1, total gain 3, and an exact rank-zero baseline.
It also shows the observation-origin manifest that keeps training and
evaluation separate. These scores concern new runs on the same conditions;
they do not establish transfer to new conditions or supply generic standard
errors for learned ranks and eigenvalues.

## Uncertainty: two targets, and what "refused" means

`rdm_sampling_covariance()` requires an explicit `target`, because the two
choices answer different questions:

```{r targets}
null_se <- sqrt(sampling_covariance(
  rdm_sampling_covariance(plan, example$fit, target = "null", at = peak)
))
plugin_se <- sqrt(sampling_covariance(
  rdm_sampling_covariance(plan, example$fit, target = "plugin", at = peak)
))
round(rbind(distance = distances$values[peak, ],
            null = null_se, plugin = plugin_se), 4)
```

- **`target = "null"`** evaluates the signal-dependent variance term at a fixed
  zero effect. It is **exact on the variance scale** under the declared model,
  and it is what you want when calibrating a test of no effect. Notice it is
  the same for every pair: with no signal assumed, the variance depends only
  on the noise.
- **`target = "plugin"`** substitutes the estimated partition-mean pattern for
  the unknown signal. It tracks the actual distances — large where the
  distance is large, and back down toward the null value for the two
  within-category pairs — which is what you want when reporting an error bar
  around a nonzero estimate. It is **biased upward**, by a term that shrinks
  like 1/M² in the number of partitions and is largest when noise dominates
  the true distances, so read it as mildly conservative. See
  `?rdm_sampling_covariance` for the exact expression.

Both targets share one estimated ingredient: the residual covariance
\(\Sigma_w\). "Exact" above means exact *given* that covariance — and because
the noise term is quadratic in it, a raw plug-in would inflate every standard
error by \(\sqrt{1+(1+P_{\text{eff}})/\nu}\). crossform applies the
Wishart-unbiased correction and reports the two numbers that govern it: the
residual degrees of freedom \(\nu\), and \(P_{\text{eff}}\), the number of
residual directions this measurement's support actually spends variance on.

```{r residual-diagnostics}
covariance <- rdm_sampling_covariance(
  plan, example$fit, target = "null", at = peak
)
c(residual_df = covariance$source$residual_df,
  effective_dimension = round(
    covariance$source$residual_effective_dimension, 2
  ))
```

A large support bought with few runs runs out of residual information, and
when \(\nu\) falls below \(P_{\text{eff}}\) the call refuses rather than
reporting a confidently small number.

There is no default. Choosing one silently would mean the same function
reported two different quantities depending on context.

This is a **within-measurement, within-participant** covariance under an
equal-partition, fixed-metric, separable error model. It is not a confidence
interval, not a spatial random-field correction, and not group inference.

### Refusal is an answer

If the relation cannot support the analytic law, the package says so with a
classed condition rather than returning a plausible-looking number. Ask
before provoking:

```{r refusal-setup, include=FALSE}
set.seed(3)
conditions <- c("a", "b", "c")
domain <- volume_domain(array(1L, c(4, 4, 3)), spacing = c(3, 3, 3))
beta_runs <- lapply(1:3, function(i) {
  m <- matrix(rnorm(3 * 48), 3, 48)
  rownames(m) <- conditions
  m
})
names(beta_runs) <- paste0("run", 1:3)
beta_only <- relation(beta_runs, effects = effect_space(conditions),
                      domain = domain)
beta_plan <- plan_geometry(
  beta_only,
  at = compile_frame(searchlights(radius = 3), domain),
  over = cross_partitions(beta_only, independence = "independent",
                          generalizes_over = "run")
)

```

```{r inspect-capabilities}
capabilities <- sampling_capabilities(beta_plan, beta_only)
capabilities$available
```

```{r capability-remedies, echo=FALSE}
knitr::kable(capabilities$reasons[c("reason", "remedy")])
```

A relation built only from precomputed beta matrices has no residuals and no
residual degrees of freedom, so the analytic law is unavailable. Provoking it
anyway yields a refusal you can inspect as data:

```{r refusal}
refusal <- catch_refusal(
  rdm_sampling_covariance(beta_plan, beta_only, target = "null", at = 1)
)
refusal$capability
refusal$reasons
```

Here the refusal identifies missing statistical information. Other refusals
can identify unsupported operations; read `$reasons` and `$remedies` to tell
which requirement failed. The point estimates —
`contrast_energy()`, `rdm()`, `rsa()` — remain perfectly valid on such a
relation; only the error bar is withheld.

## Summary

| Reading | Do | Do not |
|---|---|---|
| Negative energies or distances | keep and average them | truncate at zero |
| `signed` | read direction and units | square it and call it energy |
| `coherence_fraction` | report with radius and support sizes | compare across frames |
| `NA` fraction | read as "no interpretable share" | impute or drop silently |
| RDM values | treat as signed squared distances | take square roots, or read as `1 - r` |
| RSA coefficient | keep the intercept | compare across differently scaled models |
| Predictive gain | retain its sign and the frozen prediction cost | call it a fraction explained or a null-centered energy |
| Mode evidence | distinguish training amplitude, test evidence and gain | remove harmful modes using final test outcomes |
| `target = "null"` | calibrate a test of no effect | use as an error bar on a nonzero distance |
| `target = "plugin"` | error bar on an estimate, read as conservative | treat as exact |
| a refusal | read `$reasons` and `$remedies` | retry until something returns a number |
| a measurement above the floor | call it a candidate and say what kind it is | threshold it and report it as a finding |
| an error bar at `which.max()` | name the measurement in advance instead | read it as uncertainty about the peak |

## See also

- `vignette("introduction", package = "crossform")` — the workflow these
  results come from.
- [Correlation-distance policy](correlation-distance-policy.html), offline
  `vignette("correlation-distance-policy")` — why `rdm()` will not give you
  `1 - r`.
- [Failure gallery](failure-gallery.html), offline
  `vignette("failure-gallery")` — six realistic errors and the refusals that
  stop them.
- `?example_fmri_effects` for the two planted blocks, and `?contrast_energy`,
  `?rdm`, `?rsa`, `?rdm_sampling_covariance` for the exact contracts.
