---
title: "Reading contrast energies, RDMs, and uncertainty"
description: >
  What each returned column means, how to read negatives and the coherence fraction, the three interpretive traps that make a coherent share look like a scientific finding when it is an artefact of the frame, and how to read a real map when no ground truth marks the answer.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading contrast energies, RDMs, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

This article introduces no new statistics. It is about reading the numbers the
package already returns: what each column means, three ways a reader can be
misled by the coherence fraction, and how to read a map when nothing marks the
right answer. Everything below runs on the generated fixture, which plants two
blocks of equal amplitude at opposite ends of the volume — one whose sign
alternates between neighboring voxels, one that shifts every voxel the same
way — so each half of the energy decomposition has a known home.

```{r example}
example <- example_fmri_effects()
pairing <- cross_partitions(
  example$fit$relation,
  independence = "independent",
  generalizes_over = "run"
)
plan <- plan_geometry(example$fit$relation, at = example$frame, over = pairing)
effect <- contrast_energy(plan, example$contrast)
peak <- which.max(effect$total)

pattern <- example$truth$pattern_measurements
mean_block <- example$truth$mean_measurements
c(
  measurements = length(effect$total),
  pattern_searchlights = length(pattern),
  mean_searchlights = length(mean_block),
  overlap = length(intersect(pattern, mean_block))
)
```

The two blocks are the same size and never touch the same searchlight, so
every comparison below between them is a comparison of sign structure alone.

## The four columns of a contrast view

```{r columns}
head(as.data.frame(effect), 4)
```

- **`signed`** is the ordinary signed contrast of the frame-weighted mean
  pattern. It is a *first moment*: it has the units and the sign of your
  effect, and it is the only column that answers "which way?". It is not
  crossvalidated in the reproducibility sense — it is a retained marginal.
- **`coherent`** is the crossvalidated energy carried by that
  frame-weighted mean pattern. It is a *second moment* — a squared quantity —
  so it is not the signed effect and not its square either.
- **`configuration`** is the crossvalidated energy in reproducible spatial
  *departures* from the weighted mean, orthogonal to `coherent` under the
  frame's weights.
- **`total`** is their exact sum.

```{r partition}
max(abs(effect$total - (effect$coherent + effect$configuration)))
```

The fixture separates the two energies by construction. Here is the strongest
searchlight in each block:

```{r two-blocks}
strongest <- function(measurements) {
  measurements[which.max(effect$total[measurements])]
}
at <- c(pattern = strongest(pattern), mean = strongest(mean_block))
block_rows <- as.data.frame(effect)[at, -1]
rownames(block_rows) <- names(at)
round(block_rows, 4)
```

```{r two-blocks-check, include=FALSE}
stopifnot(
  at[["pattern"]] == peak,
  effect$configuration[at[["pattern"]]] >
    50 * abs(effect$coherent[at[["pattern"]]]),
  effect$coherent[at[["mean"]]] >
    50 * abs(effect$configuration[at[["mean"]]]),
  abs(effect$signed[at[["pattern"]]]) < 0.1,
  abs(effect$signed[at[["mean"]]]) > 1.5,
  is.na(effect$coherence_fraction[at[["pattern"]]]),
  effect$coherence_fraction[at[["mean"]]] > 0.99
)
```

Equal amplitude, equal block size, near-equal `total` — and opposite
decompositions. The pattern block's energy is almost all `configuration`: a
shape that reproduces across runs while the sphere's average barely moves,
which is why its `signed` value is close to zero and a univariate contrast
would find nothing there. The mean block's energy is almost all `coherent`,
and its `signed` value is large, because the whole neighborhood shifts
together.

Note the `NA` in the first row. That searchlight is the largest `total` in the
map and it reports no coherence fraction at all, because its `coherent`
estimate landed a hair below zero. That is the subject of the section after
next.

### Negatives are estimates, not errors

Each energy is a product of estimates from two different runs. Under the null
its expectation is zero, so roughly half the sampling distribution lies below
zero.

```{r negatives}
c(
  measurements = length(effect$total),
  total_below_zero = sum(effect$total < 0),
  coherent_below_zero = sum(effect$coherent < 0),
  configuration_below_zero = sum(effect$configuration < 0),
  min_total = min(effect$total)
)
```

Do not truncate them. Clipping at zero replaces an unbiased estimator with a
positively biased one, and the bias is largest exactly where there is no
effect — which is where a thresholded map does its talking. If you average
energies over an ROI, over participants, or over searchlights, average the
signed values. The negatives are what makes the average come out at zero when
nothing is there.

## The coherence fraction and its validity flag

`coherence_fraction` is `coherent / total`: the share of reproducible energy
carried by the local mean rather than by pattern shape. It is reported only
where the raw components actually form a nonnegative partition — that is,
where `total > 0`, `coherent >= 0`, and `configuration >= 0`. Elsewhere it is
`NA`, and a companion logical says why you got `NA`.

```{r validity}
c(
  measurements = length(effect$total),
  valid = sum(effect$coherence_fraction_valid),
  na = sum(is.na(effect$coherence_fraction))
)
```

```{r validity-reasons}
c(
  total_not_positive = sum(effect$total <= 0),
  coherent_negative = sum(effect$coherent < 0),
  configuration_negative = sum(effect$configuration < 0)
)
```

The gate is not fussiness. A ratio of two quantities that can each be negative
is not a share of anything: it can exceed 1, or be negative, or flip sign
under an arbitrarily small change in the denominator. Rather than print such
a number, the package withholds it. `NA` here means "this measurement has no
interpretable coherent share", not "missing data".

Nor is the gate a signal detector. The peak of the pattern block above is the
strongest measurement in the entire map and it is one of these `NA`s: an
effect that is genuinely all configuration has nothing left over to form a
nonnegative pair with.

The three sections below are the caveats that matter once you start reporting
those fractions.

## Trap (a): the coherent share shrinks as the sphere grows

The coherent component is the energy of *one* spatial mode — the
frame-weighted mean. The configuration component spans everything orthogonal
to it. As a searchlight grows, the configuration subspace gains dimensions
while the coherent subspace stays one-dimensional. The coherent share
therefore falls for mechanical reasons, with no change in the underlying
neural data.

```{r sphere-size}
radii <- c(2, 3, 4.5, 6)
sweep <- t(vapply(radii, function(r) {
  frame_r <- compile_frame(searchlights(radius = r), example$domain)
  effect_r <- contrast_energy(
    plan_geometry(example$fit$relation, at = frame_r, over = pairing),
    example$contrast
  )
  c(
    median_support = median(Matrix::rowSums(frame_r$weights != 0)),
    valid = sum(effect_r$coherence_fraction_valid),
    median_fraction = median(effect_r$coherence_fraction, na.rm = TRUE),
    pattern_block = median(effect_r$coherence_fraction[pattern], na.rm = TRUE),
    mean_block = median(effect_r$coherence_fraction[mean_block], na.rm = TRUE)
  )
}, numeric(5)))
data.frame(radius = radii, round(sweep, 3))
```

```{r sphere-size-check, include=FALSE}
stopifnot(
  sweep[1, "median_fraction"] == 1,
  sweep[1, "pattern_block"] == 1,
  sweep[1, "mean_block"] == 1,
  all(diff(sweep[, "median_fraction"]) <= 0),
  all(diff(sweep[, "mean_block"]) <= 0),
  all(sweep[-1, "mean_block"] > sweep[-1, "pattern_block"])
)
```

At radius 2 every support is a single voxel, so the weighted mean *is* the
pattern and the fraction is exactly 1 by construction — in both blocks. It
then falls. The decomposition keeps the two blocks in the right order at every
larger radius, the mean block several times above the pattern block, so the
ordering is real. But the mean block's own fraction still slides from 0.347 to
0.289 while nothing about the planted signal changes. A sentence such as
"roughly 30% of the reproducible signal was carried by the mean response" is
therefore a statement about your radius as much as about the brain. Report the
radius and the support sizes alongside the fraction, and never compare
fractions across analyses with different frames.

## Trap (b): "coherent" means coherent under *your* weights

The decomposition is orthogonal in the frame-weighted inner product, so the
weights are part of the definition. Reweighting the *same* voxels moves energy
between the two components — and changes the total.

```{r weighting}
frame_uniform <- compile_frame(searchlights(radius = 4.5), example$domain)

# Same supports, Gaussian instead of uniform weights.
coords <- example$domain$coordinates
uniform_weights <- as.matrix(frame_uniform$weights)
centers <- match(frame_uniform$index$measurement, example$domain$feature_ids)
gaussian_weights <- uniform_weights
for (i in seq_len(nrow(uniform_weights))) {
  members <- which(uniform_weights[i, ] != 0)
  offsets <- coords[members, , drop = FALSE] -
    matrix(coords[centers[i], ], length(members), 3, byrow = TRUE)
  w <- exp(-rowSums(offsets^2) / (2 * 3^2))
  gaussian_weights[i, ] <- 0
  gaussian_weights[i, members] <- w / sum(w)
}
frame_gaussian <- additive_frame(
  gaussian_weights, normalization = "local", domain = example$domain
)

identical(which(uniform_weights != 0), which(gaussian_weights != 0))
```

Read the two frames at the searchlight sitting on each block's center. That
measurement is chosen from geometry alone, not from the results, so nothing
below is a selected extreme.

```{r weighting-effect}
effect_uniform <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_uniform, over = pairing),
  example$contrast
)
effect_gaussian <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_gaussian, over = pairing),
  example$contrast
)

centered_on <- function(features) {
  centroid <- colMeans(coords[features, , drop = FALSE])
  which.min(rowSums(
    (coords[centers, , drop = FALSE] -
      matrix(centroid, length(centers), 3, byrow = TRUE))^2
  ))
}
middle <- c(
  pattern = centered_on(example$truth$planted_features),
  mean = centered_on(example$truth$mean_features)
)

reweighted <- cbind(
  uniform_coherent = effect_uniform$coherent[middle],
  gaussian_coherent = effect_gaussian$coherent[middle],
  uniform_total = effect_uniform$total[middle],
  gaussian_total = effect_gaussian$total[middle]
)
rownames(reweighted) <- names(middle)
round(reweighted, 4)
```

```{r weighting-check, include=FALSE}
stopifnot(
  identical(which(uniform_weights != 0), which(gaussian_weights != 0)),
  identical(frame_uniform$index$measurement, example$frame$index$measurement),
  middle[["pattern"]] %in% pattern,
  middle[["mean"]] %in% mean_block,
  effect_gaussian$coherent[middle[["pattern"]]] <
    0.6 * effect_uniform$coherent[middle[["pattern"]]],
  abs(
    effect_gaussian$coherent[middle[["mean"]]] -
      effect_uniform$coherent[middle[["mean"]]]
  ) < 0.02 * effect_uniform$coherent[middle[["mean"]]],
  effect_uniform$total[middle[["pattern"]]] !=
    effect_gaussian$total[middle[["pattern"]]],
  effect_uniform$total[middle[["mean"]]] !=
    effect_gaussian$total[middle[["mean"]]]
)
```

Identical voxels, identical data, different numbers — but not equally
different. The mean block's coherent energy barely moves: a shift that is the
same in every voxel is the same shift under any nonnegative weights. The
pattern block's coherent energy halves, because there it is not a signal at
all. It is whatever failed to cancel, and how much fails to cancel is decided
entirely by the weights. Both totals move.

This is not instability: each frame defines a different, well-specified
estimand, and the plan records which one you asked for. It does mean that "the
coherent component" is not a property of the brain region. It is a property of
the region *and* the weights you chose to average it with — and the smaller
the coherent component, the more of it belongs to the weights.

## Trap (c): reported fractions are a selected sample

Because the fraction exists only where the components form a nonnegative
partition, the set of measurements that report one is not a random subset of
the map. The gate depends on the same quantities that appear in the ratio, so
it is enriched for measurements with real signal — and, among the rest, for
whichever noise realizations happened to make both components positive.

```{r selection}
frame_small <- compile_frame(searchlights(radius = 3), example$domain)
effect_small <- contrast_energy(
  plan_geometry(example$fit$relation, at = frame_small, over = pairing),
  example$contrast
)
signal <- example$truth$signal_measurements
valid <- effect_small$coherence_fraction_valid

c(
  signal_measurements = length(signal),
  signal_reporting_a_fraction = sum(valid[signal]),
  other_measurements = length(valid) - length(signal),
  other_reporting_a_fraction = sum(valid[-signal])
)
```

```{r selection-blocks}
round(c(
  pattern_block = mean(valid[pattern]),
  mean_block = mean(valid[mean_block]),
  elsewhere = mean(valid[-signal])
), 3)
```

```{r selection-share}
c(
  share_of_map_reporting = mean(valid),
  share_of_reporters_that_are_signal = mean(seq_along(valid)[valid] %in% signal)
)
```

```{r selection-check, include=FALSE}
stopifnot(
  # Every frame here is one measurement per voxel in the same order, so the
  # truth indices carry across radii.
  identical(frame_small$index$measurement, example$frame$index$measurement),
  mean(valid[pattern]) > 0.9,
  mean(valid[mean_block]) > 0.9,
  mean(valid[-signal]) < 0.25
)
```

Both blocks clear the gate at about the same rate, so the selection is not
about which kind of effect you have; it is about having one. Under a quarter
of the remaining measurements report a fraction, on noise alone. The result is
that the average fraction "over the map" is really an average over a set that
is 70% planted signal. That is fine if you say so, and misleading if you do
not. If you want a map-wide summary, define the denominator explicitly — for
instance the share of *all* measurements that report a coherent share at all,
the first number above — rather than averaging the reported values and calling
the result a property of the map.

The same warning applies to any post-hoc gate: thresholding on `total > 0`
and then summarizing `coherent` conditions the summary on the numerator's
sibling.

## Reading a map without ground truth

Every figure so far could mark the right answer, because the fixture carries
`truth$signal_measurements`. Your data will not. This section is about what the
same picture tells you when nothing is marked.

```{r blind-map, fig.width=8, fig.height=4, fig.cap="The same contrast view with no highlight supplied. Nothing here identifies the two planted blocks; the reading below uses only what is on the page."}
plot(effect)
```

### The band at zero is the noise floor, and it comes free

A crossvalidated energy is a product of estimates from two different
partitions, so its expectation is zero wherever nothing reproduces. You do not
have to model that floor or estimate it from a separate null run. It is already
in the map, underneath your candidates, and about half of it is below the line
because the negatives were kept.

```{r floor-known}
quiet <- setdiff(seq_along(effect$total), example$truth$signal_measurements)
round(c(
  measurements = length(quiet),
  mean = mean(effect$total[quiet]),
  sd = stats::sd(effect$total[quiet]),
  fraction_below_zero = mean(effect$total[quiet] < 0),
  largest = max(effect$total[quiet])
), 4)
```

That used the truth. You can get close without it, because in a typical map the
floor is the majority and a robust summary is dominated by it:

```{r floor-blind}
round(c(
  median_over_all = stats::median(effect$total),
  mad_over_all = stats::mad(effect$total),
  fraction_below_zero_over_all = mean(effect$total < 0)
), 4)
```

```{r floor-check, include=FALSE}
stopifnot(
  abs(stats::median(effect$total)) < 0.01,
  abs(mean(effect$total[quiet])) < 0.01,
  stats::mad(effect$total) > stats::sd(effect$total[quiet])
)
```

The blind median lands near the true floor's center, and both sit under 0.3% of
the peak. The blind MAD is more than twice the floor's true spread, because 37%
of these measurements carry planted signal — two blocks, not one — and inflate
it. That is the right direction for a scale you are going to eyeball against,
but it is a description, not an estimate of a null.

### Candidates, not findings

Sorted, this map has a visible shoulder: rank 104 is `0.478` and rank 105 is
`0.030`, while rank 1 is `4.10`. The shoulder falls exactly at the edge of the
planted set — all 104 planted searchlights outrank every other measurement.
That is a fact about this fixture, not a method.

```{r shoulder}
round(sort(effect$total, decreasing = TRUE)[c(1, 10, 40, 104, 105, 150, 280)], 4)
```

```{r shoulder-check, include=FALSE}
stopifnot(
  identical(
    sort(order(effect$total, decreasing = TRUE)[seq_along(signal)]),
    signal
  )
)
```

`crossform` deliberately gives you no way to turn that shoulder into a finding.
There is no threshold, no p-value, no permutation null, no spatial
random-field correction, and no group inference; the README's **Status and
scope** section says so, and the failure gallery
([online](failure-gallery.html), or `vignette("failure-gallery")` offline)
treats manufactured uncertainty as an error to guard against. Measurements
standing clear of the floor are **candidates**. What you may legitimately do
with them:

- **Say what kind of effect each one is.** The coherent/configuration split is
  descriptive, not inferential, so it costs you nothing to read: a candidate
  high in configuration is a reproducible pattern, one far along the coherent
  axis is a regional mean that a univariate contrast would also have found.
  This map has both, at opposite ends of the volume, with the same amplitude.
- **Attach a within-measurement standard error at a measurement you chose in
  advance**, using `rdm_sampling_covariance()` with `target = "null"` — and
  only if the relation kept its residual channel, which means fitting with
  `lm_relation_fit()` or `estimate_relation()` rather than importing betas.
- **Take the candidates somewhere else**: a held-out session, an independent
  dataset, or a group-level analysis performed outside this package.

### The error bar at the peak is not an error bar

One trap deserves naming, because the package makes it easy to walk into. If
you find `peak <- which.max(effect$total)` and then report a standard error at
`peak`, that interval describes a measurement you selected *because* it was
extreme. The estimate is conditioned on being the largest; the standard error
is not. Every number in this article that uses `peak` is illustrative for that
reason, not a result.

Specify the measurement in advance, or treat the peak as a candidate and estimate it
somewhere you did not select it.

## Reading the RDM

`peak` is the pattern block's strongest searchlight, so the two within-category
pairs should sit near zero and the four across-category pairs near the planted
separation.

```{r rdm}
distances <- rdm(plan)
distances$pairs
round(distances$values[peak, ], 3)
```

For conditions *i* and *j*, `rdm()` reports
`d_ij = G_ii + G_jj - 2 G_ij`. Under cross-partition pairing this is a
**signed crossvalidated squared Euclidean** distance; with a fixed neural
precision metric it is a squared Mahalanobis distance. Three consequences:

1. **They are squared.** Do not take square roots to "get distances back" —
   the square root of a negative estimate is undefined, and the square root of
   a noisy positive one is biased.
2. **They can be negative**, for the same reason the energies can, and for
   the same reason they are kept.
3. **They are not `1 - Pearson correlation`.** Correlation distance divides
   each pattern by its own norm, which is not a fixed linear operation, so it
   cannot be a query against this geometry. The package refuses to substitute
   it silently; the reasoning and the conditions under which the boundary
   could be crossed are in the correlation-distance policy
   ([online](correlation-distance-policy.html), or
   `vignette("correlation-distance-policy")` offline).

Because the RDM is a geometry query and not a decomposition, it does not tell
the two blocks apart. The strongest searchlight in each block returns much the
same six distances, even though their energies split in opposite directions:

```{r rdm-blocks}
round(rbind(
  pattern = distances$values[at[["pattern"]], ],
  mean = distances$values[at[["mean"]], ]
), 3)
```

```{r rdm-blocks-check, include=FALSE}
stopifnot(
  max(abs(
    distances$values[at[["pattern"]], ] - distances$values[at[["mean"]], ]
  )) < 0.5
)
```

Only `contrast_energy()` separates coherent from configuration. If the
distinction matters to your question, the RDM is not where you will find it.

## Reading RSA coefficients

`rsa()` regresses the measured distances on your model RDMs. It is ordinary
least squares, not a rank correlation.

```{r rsa}
category <- rsa(plan, models = list(category = example$model_rdm))
colnames(category$coefficients)
round(category$coefficients[peak, ], 4)
```

There are two columns for one model because an **intercept is fitted by
default**. Keep it. Crossvalidated distances have an arbitrary additive level
per measurement; without an intercept the model coefficient absorbs that
level, and a searchlight with uniformly large distances will look like it
matches your model when it only has a large mean distance.

```{r rsa-nointercept}
no_intercept <- rsa(
  plan, models = list(category = example$model_rdm), intercept = FALSE
)
c(
  with_intercept = category$coefficients[peak, "category"],
  without_intercept = no_intercept$coefficients[peak, "category"]
)
```

Drop the intercept only when you have a reason to believe the level is
meaningful and shared. Coefficients are in units of distance per unit of model
RDM, so they are comparable across measurements within one analysis but not
across analyses with differently scaled model RDMs.

## Uncertainty: two targets, and what "refused" means

`rdm_sampling_covariance()` requires an explicit `target`, because the two
choices answer different questions:

```{r targets}
null_se <- sqrt(sampling_covariance(
  rdm_sampling_covariance(plan, example$fit, target = "null", at = peak)
))
plugin_se <- sqrt(sampling_covariance(
  rdm_sampling_covariance(plan, example$fit, target = "plugin", at = peak)
))
round(rbind(distance = distances$values[peak, ],
            null = null_se, plugin = plugin_se), 4)
```

- **`target = "null"`** evaluates the signal-dependent variance term at a fixed
  zero effect. It is **exact** under the declared model, and it is what you
  want when calibrating a test of no effect. Notice it is the same for every
  pair: with no signal assumed, the variance depends only on the noise.
- **`target = "plugin"`** substitutes the estimated partition-mean pattern for
  the unknown signal. It tracks the actual distances — large where the
  distance is large, and back down toward the null value for the two
  within-category pairs — which is what you want when reporting an error bar
  around a nonzero estimate. It is **biased upward**, by a term that shrinks
  like 1/M² in the number of features and is largest when noise dominates the
  true distances, so read it as mildly conservative. See
  `?rdm_sampling_covariance` for the exact expression.

There is no default. Choosing one silently would mean the same function
reported two different quantities depending on context.

This is a **within-measurement, within-participant** covariance under an
equal-partition, fixed-metric, separable error model. It is not a confidence
interval, not a spatial random-field correction, and not group inference.

### Refusal is an answer

If the relation cannot support the analytic law, the package says so with a
classed condition rather than returning a plausible-looking number. Ask
before provoking:

```{r refusal-setup}
set.seed(3)
conditions <- c("a", "b", "c")
domain <- volume_domain(array(1L, c(4, 4, 3)), spacing = c(3, 3, 3))
beta_runs <- lapply(1:3, function(i) {
  m <- matrix(rnorm(3 * 48), 3, 48)
  rownames(m) <- conditions
  m
})
names(beta_runs) <- paste0("run", 1:3)
beta_only <- relation(beta_runs, effects = effect_space(conditions),
                      domain = domain)
beta_plan <- plan_geometry(
  beta_only,
  at = compile_frame(searchlights(radius = 3), domain),
  over = cross_partitions(beta_only, independence = "independent",
                          generalizes_over = "run")
)

sampling_capabilities(beta_plan, beta_only)
```

A relation built only from precomputed beta matrices has no residuals and no
residual degrees of freedom, so the analytic law is unavailable. Provoking it
anyway yields a refusal you can inspect as data:

```{r refusal}
refusal <- catch_refusal(
  rdm_sampling_covariance(beta_plan, beta_only, target = "null", at = 1)
)
class(refusal)
refusal$capability
refusal$reasons
refusal$remedies
```

"Refused" does not mean "not implemented". It means the specific claim you
asked for is not supported by the evidence you supplied, and the condition
carries every unmet requirement with a concrete remedy. The point estimates —
`contrast_energy()`, `rdm()`, `rsa()` — remain perfectly valid on such a
relation; only the error bar is withheld.

## Summary

| Reading | Do | Do not |
|---|---|---|
| Negative energies or distances | keep and average them | truncate at zero |
| `signed` | read direction and units | square it and call it energy |
| `coherence_fraction` | report with radius and support sizes | compare across frames |
| `NA` fraction | read as "no interpretable share" | impute or drop silently |
| RDM values | treat as signed squared distances | take square roots, or read as `1 - r` |
| RSA coefficient | keep the intercept | compare across differently scaled models |
| `target = "null"` | calibrate a test of no effect | use as an error bar on a nonzero distance |
| `target = "plugin"` | error bar on an estimate, read as conservative | treat as exact |
| a refusal | read `$reasons` and `$remedies` | retry until something returns a number |
| a measurement above the floor | call it a candidate and say what kind it is | threshold it and report it as a finding |
| an error bar at `which.max()` | name the measurement in advance instead | read it as uncertainty about the peak |

## See also

- `vignette("introduction", package = "crossform")` — the workflow these
  results come from.
- [Correlation-distance policy](correlation-distance-policy.html), offline
  `vignette("correlation-distance-policy")` — why `rdm()` will not give you
  `1 - r`.
- [Failure gallery](failure-gallery.html), offline
  `vignette("failure-gallery")` — six realistic errors and the refusals that
  stop them.
- `?example_fmri_effects` for the two planted blocks, and `?contrast_energy`,
  `?rdm`, `?rsa`, `?rdm_sampling_covariance` for the exact contracts.
