---
title: "Where does a representational effect generalize?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where does a representational effect generalize?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

## The question

You have condition estimates from repeated runs, and you want to know where an
effect reproduces across those runs. This guide answers that question, and then
three more from the same fit without recomputing anything: what *kind* of effect
it is, what the whole representational geometry looks like at the strongest
location, and how much of that you are entitled to put an error bar on.

One word to fix before starting: a **measurement** is one spatial unit of the
frame, and every result below has one value per measurement. Here the
measurements are searchlights; anatomical regions, single voxels, and the whole
brain are the other choices, and they all behave the same way downstream.

The worked question is: **where does an animate-versus-inanimate pattern
reproduce across independent runs?** The data are generated with known truth, so
the example runs after installation and can check its own answer. These
generated data are not empirical evidence; the separate Haxby 2001 exemplar is
the public-data comparison.

```{r example}
example <- example_fmri_effects()
c(
  conditions = length(example$fit$relation$effects),
  runs = length(example$fit$relation$partitions),
  features = example$fit$relation$n_features,
  searchlights = nrow(example$frame$index)
)
```

`example$fit` contains four run-specific estimates for four conditions in a
small 8 by 7 by 5 volume. `example$frame` defines the 280 searchlights at which
the analysis will report results, one centered on each voxel. The generator
returns the frame separately because a fitted relation records the neural
domain but does not own a mutable searchlight definition.

The fit provides two kinds of information:

- `example$fit$relation` is the condition-by-feature relation used for point
  estimates; it reads feature blocks only when a query needs them;
- its error channel stores residual blocks, residual degrees of freedom, and
  effect-coordinate covariance for uncertainty calculations whose assumptions
  the fit can support.

`example$truth` records what was planted: two blocks of 19 voxels, at opposite
ends of the longest axis, never overlapping, at the same per-voxel amplitude.
They differ only in sign structure. In the **pattern** block the sign
alternates between neighboring voxels, so the block's local average nearly
cancels. In the **mean** block every voxel moves the same way. Same effect
size, different *kind* of effect — the section on kinds below turns entirely
on that difference.

```{r truth}
c(
  pattern_voxels = length(example$truth$planted_features),
  mean_voxels = length(example$truth$mean_features),
  pattern_searchlights = length(example$truth$pattern_measurements),
  mean_searchlights = length(example$truth$mean_measurements),
  signal_searchlights = length(example$truth$signal_measurements)
)
```

```{r truth-check, include=FALSE}
stopifnot(
  length(example$truth$planted_features) == 19L,
  length(example$truth$mean_features) == 19L,
  length(example$truth$pattern_measurements) == 52L,
  length(example$truth$mean_measurements) == 52L,
  length(intersect(
    example$truth$pattern_measurements, example$truth$mean_measurements
  )) == 0L,
  identical(
    example$truth$signal_measurements,
    sort(union(
      example$truth$pattern_measurements, example$truth$mean_measurements
    ))
  )
)
```

A searchlight counts as planted if it gives any weight to a planted voxel, so
each 19-voxel block occupies 52 of the 280 searchlights and the two sets stay
disjoint. `signal_searchlights` is their union, and it is what the figures
below highlight.

## Declare what must generalize

Three declarations make the estimand: the relation you fitted, the frame you
want answers at, and the pairing that says what must reproduce.

```{r plan}
plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)
plan
```

`cross_partitions()` requests products between different runs. Every unordered
pair of runs contributes to the estimate, but the function does not treat the
six run pairs as six independent samples. `plan_geometry()` checks that the
relation, condition names, searchlight frame, metric, units, and neural domain
agree before it reads neural values.

Read the `execution` and `state` lines of that print. The plan is query-first:
it holds the identity of the quantity, and nothing has been materialized yet.
Save `plan` with the analysis record. It identifies the estimand: the
scientific quantity the analysis is meant to estimate. Changing block size or
storage changes the execution receipt, not this plan identity, and
`print(plan, detail = TRUE)` is where those execution internals are shown.

### Choosing what must generalize

Two arguments to `cross_partitions()` decide what you are estimating, so it is
worth being deliberate about both.

`generalizes_over` names the sampling axis your partitions *are* — `"run"`,
`"session"`, `"task"`. Cross-run and cross-session reproduction are different
scientific quantities even at identical fold counts, so the declared axis is
bound into every plan identity built from this pairing. If your partitions are
sessions, pass `"session"`.

`independence = "independent"` is your declaration that distinct partition
estimates have independent estimation errors, which separate runs or sessions
with separately estimated noise satisfy. It has to be stated explicitly:
leaving it `NULL` records the endpoints as undeclared, which still yields a
point estimand but earns neither cross-generalized nor analytic sampling-law
capabilities.

```{r pairing-identity}
declared <- cross_partitions(
  example$fit$relation,
  independence = "independent", generalizes_over = "run"
)
undeclared <- cross_partitions(example$fit$relation)
c(
  declared = attr(declared, "estimate"),
  axis = attr(declared, "generalizes_over"),
  undeclared = attr(undeclared, "estimate")
)
```

```{r pairing-identity-check, include=FALSE}
stopifnot(
  identical(attr(declared, "estimate"), "cross_generalized"),
  identical(attr(undeclared, "estimate"), "independence_undeclared"),
  nrow(declared) == 6L,
  isTRUE(all.equal(sum(declared$weight), 1))
)
```

The plan print's `generalizes:` line is where you check that you declared what
you meant. Left undeclared it reads
`6 partition pairs (axis undeclared), endpoints undeclared` instead.

One thing the pairing is *not*: its six rows are estimator contributions, not
six independent replicates. Pairs that share a run share that run's estimation
error, which is why the standard-error section below cannot simply take their
spread. `?cross_partitions` states this; `?pairing` builds explicit or directed
edge sets for designs that need one, such as nested or ordered axes.

## The answer

```{r contrast}
effect <- contrast_energy(plan, example$contrast)
peak <- which.max(effect$total)
c(
  peak_searchlight = peak,
  signed_effect = effect$signed[peak],
  total_energy = effect$total[peak],
  coherent_energy = effect$coherent[peak],
  configuration_energy = effect$configuration[peak],
  overlaps_planted_signal = peak %in% example$truth$signal_measurements
)
```

`contrast_energy()` returns four values at every searchlight. `signed` is the
usual signed contrast of the local weighted mean. The other three values
measure cross-run reproducibility:

- `coherent` is the crossvalidated energy of that local weighted mean;
- `configuration` is the energy in reproducible spatial departures from the
  weighted mean;
- `total` is their sum.

The peak sits inside the pattern block, and its energy is almost entirely
`configuration` — its signed contrast is close to zero. Hold that; it is the
subject of the next section.

Because the fixture planted its signal in known searchlights, the answer can be
drawn against the truth. `highlight` marks both planted blocks in the picture;
it plays no part in the estimate.

```{r contrast-plot, fig.width=8, fig.height=4, fig.cap="Left: every searchlight in the coherent/configuration plane, with the dashed guides at constant total energy. The two planted blocks separate along the two axes. Right: total energy along the searchlight index. The planted searchlights are filled."}
plot(
  effect,
  highlight = example$truth$signal_measurements,
  highlight_label = "planted signal"
)
```

```{r contrast-check, include=FALSE}
stopifnot(
  peak %in% example$truth$pattern_measurements,
  effect$configuration[peak] > 0.99 * effect$total[peak],
  isTRUE(all.equal(
    effect$total, effect$coherent + effect$configuration, tolerance = 1e-14
  ))
)
```

The right panel is the map. The planted searchlights are the spikes; everything
else lies on zero rather than above it. That is deliberate. Crossvalidated
energy is centered on zero when no reproducible effect is present, so sampling
noise produces negative estimates about half the time. Clipping them to zero
would change the estimator and introduce positive bias, so `contrast_energy()`
retains them and the plot draws them where they fall.

```{r planted-summary}
summarize_region <- function(index) c(
  total = mean(effect$total[index]),
  coherent = mean(effect$coherent[index]),
  configuration = mean(effect$configuration[index]),
  fraction_below_zero = mean(effect$total[index] < 0)
)
signal_searchlights <- example$truth$signal_measurements
region_summary <- rbind(
  pattern = summarize_region(example$truth$pattern_measurements),
  mean = summarize_region(example$truth$mean_measurements),
  outside = summarize_region(
    setdiff(seq_along(effect$total), signal_searchlights)
  )
)
round(region_summary, 3)
```

```{r planted-summary-check, include=FALSE}
stopifnot(
  abs(region_summary["pattern", "total"] - region_summary["mean", "total"]) <
    0.01,
  region_summary["pattern", "configuration"] >
    10 * region_summary["pattern", "coherent"],
  region_summary["mean", "coherent"] >
    region_summary["mean", "configuration"],
  abs(region_summary["outside", "total"]) < 0.01,
  region_summary["outside", "fraction_below_zero"] > 0.4,
  region_summary["outside", "fraction_below_zero"] < 0.6
)
```

The two blocks carry the same total energy to within `0.01` — they were
planted at the same amplitude — and divide it in opposite directions: the
pattern block puts more than ten times as much into configuration as into
coherent, the mean block puts more into coherent than into configuration.
Outside them the mean total energy is `0.003` and roughly half the estimates
fall below zero. That is what a correctly centered null looks like, and you can
see it rather than assume it.

```{r contrast-map, echo=FALSE, fig.cap="Cross-generalized contrast energy in the middle slice, on a scale centered at zero: white is the noise floor, red is reproducible energy. The two generated blocks sit at opposite ends of the long axis, which runs vertically here. The map checks the analysis and is not empirical evidence."}
volume <- array(NA_real_, dim = example$domain$metadata$dim)
volume[example$domain$metadata$mask] <- effect$total
middle <- ceiling(dim(volume)[3] / 2)
limit <- max(abs(effect$total))
image(
  t(volume[, , middle]), axes = FALSE, asp = 1,
  zlim = c(-limit, limit), col = hcl.colors(25, "Blue-Red 3"),
  main = sprintf("Middle slice (z = %d)", middle)
)
box()
```

## What kind of effect is it?

The left panel of the two-panel figure is the part that is hard to obtain
elsewhere. It places every searchlight by how its reproducible energy divides:
`coherent` on the horizontal axis is the energy carried by the searchlight's
frame-weighted mean pattern, which is the familiar univariate story, and
`configuration` on the vertical axis is the reproducible pattern beyond that
mean.

The two components sum to `total` exactly at every measurement, which is why
lines of constant total appear as the dashed guides of slope `-1`. The
partition is checked, not assumed:

```{r partition-check}
max(abs(effect$total - (effect$coherent + effect$configuration)))
```

The coherent and configuration modes are orthogonal under the weights declared
by the frame. They need not be orthogonal under an unweighted Euclidean inner
product.

The two planted blocks make the split concrete, because they are the same
effect size and a different kind of effect. Take the strongest searchlight in
each:

```{r two-kinds}
strongest_in <- function(index) index[[which.max(effect$total[index])]]
pattern_peak <- strongest_in(example$truth$pattern_measurements)
mean_peak <- strongest_in(example$truth$mean_measurements)
readout <- function(measurement) c(
  signed = effect$signed[measurement],
  coherent = effect$coherent[measurement],
  configuration = effect$configuration[measurement],
  total = effect$total[measurement]
)
kinds <- rbind(
  pattern_block = readout(pattern_peak),
  mean_block = readout(mean_peak)
)
round(kinds, 3)
```

```{r two-kinds-check, include=FALSE}
stopifnot(
  pattern_peak == peak,
  abs(kinds["pattern_block", "total"] - kinds["mean_block", "total"]) < 0.2,
  kinds["pattern_block", "configuration"] >
    0.99 * kinds["pattern_block", "total"],
  kinds["mean_block", "coherent"] > 0.99 * kinds["mean_block", "total"],
  abs(kinds["mean_block", "signed"]) >
    30 * abs(kinds["pattern_block", "signed"])
)
```

Both peaks carry about `4` units of reproducible energy. The pattern block puts
essentially all of it into `configuration`; the mean block puts essentially all
of it into `coherent`. That first row is the signature of a multivariate
pattern a univariate map would miss: the sphere reproduces a *shape*, while its
average response barely moves. Nothing was removed to see this, and nothing was
fitted twice; the split is a property of the plan.

"A univariate map would miss it" is checkable here. `signed` is a first
moment — the ordinary contrast of the frame-weighted mean pattern, carrying
your effect's units and sign — while the three energies are second moments,
squared and crossvalidated. `signed` can therefore only ever see the coherent
half — and at the pattern block there is almost no coherent half to see. Take
the largest signed value anywhere outside the two blocks as the signed noise
level, and ask how far each block stands above it:

```{r signed-versus-configuration}
outside <- setdiff(seq_along(effect$total), signal_searchlights)
pattern <- example$truth$pattern_measurements
largest_signed_outside <- max(abs(effect$signed[outside]))
largest_total_outside <- max(effect$total[outside])
clearing <- function(index) {
  sum(abs(effect$signed[index]) >= largest_signed_outside)
}
comparison <- c(
  signed_noise_level = largest_signed_outside,
  mean_block_clearing_it = clearing(example$truth$mean_measurements),
  pattern_block_clearing_it = clearing(pattern),
  pattern_signed_over_noise =
    stats::median(abs(effect$signed[pattern])) / largest_signed_outside,
  pattern_energy_over_noise =
    stats::median(effect$total[pattern]) / largest_total_outside,
  signed_at_the_energy_peak = abs(effect$signed[peak]),
  total_at_the_energy_peak = effect$total[peak],
  largest_total_outside = largest_total_outside
)
round(comparison, 3)
```

```{r signed-versus-configuration-check, include=FALSE}
stopifnot(
  comparison[["mean_block_clearing_it"]] ==
    length(example$truth$mean_measurements),
  comparison[["pattern_signed_over_noise"]] < 3,
  comparison[["pattern_energy_over_noise"]] > 20,
  abs(effect$signed[peak]) < largest_signed_outside,
  effect$total[peak] > 100 * largest_total_outside
)
```

Every one of the 52 mean-block searchlights clears that level: the first moment
sees that block perfectly well. The pattern block is a different story. Its
alternating signs do not cancel exactly, so most of its searchlights clear the
level too — but only by about twice the noise level, where their energies
clear it by more than fifty times. And the single strongest searchlight in the
whole volume, the one you would actually report, has a signed contrast of
`0.052`, *smaller* than the `0.154` noise level. A signed map ranks the best
result in the volume below pure noise, while its total energy of `4.10` stands
against a largest-elsewhere total of `0.03`. The energies see both blocks; the
first moment sees only one of them clearly.

`contrast_energy()` also reports `coherence_fraction`, the share of reproducible
energy carried by the local mean, but only where the components form a
nonnegative partition. Elsewhere it is `NA`, because a ratio whose denominator
can be negative is not a share of anything; printing a contrast view repeats
that caveat under the table. The
[interpreting-results](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html)
article (`vignette("interpreting-results")`) covers how to read that fraction
and the three ways it can mislead you.

## Change the question without refitting

The plan is unchanged, and nothing about it is recompiled below. For conditions
`i` and `j`, `rdm()` applies the fixed contrast `c = e_i - e_j` to the geometry:

\[
d_{ij} = c^T G c = G_{ii} + G_{jj} - 2G_{ij}.
\]

```{r rdm}
distances <- rdm(plan)
round(distances$values[peak, ], 3)
```

```{r rdm-plot, fig.width=5, fig.height=4.5, fig.cap="Signed squared distances among the four conditions at the strongest searchlight. Distances are signed and may be negative; the color scale is centered on zero."}
plot(distances, measurement = peak)
```

Nothing asked for that block structure. The two animate conditions sit close
together, the two inanimate conditions sit close together, and every
animate-versus-inanimate pair is far apart. It is what the geometry contains,
read out of the same numbers that produced the contrast map.

With the current identity metric and cross-run pairing, these are signed,
crossvalidated squared Euclidean distances. Supplying a fixed neural precision
metric changes them to squared Mahalanobis distances. Either estimate can be
negative near the null because the two factors come from different runs.

Linear RSA compiles the RDM transform and regression coefficient into a fixed
query against the same plan:

```{r rsa}
category_fit <- rsa(
  plan,
  models = list(category = example$model_rdm)
)
round(category_fit$coefficients[peak, ], 3)
```

```{r geometry-view-checks, include=FALSE}
peak_distances <- distances$values[peak, ]
within_category <- c("face - body", "house - tool")
stopifnot(
  identical(dim(distances$values), c(280L, 6L)),
  identical(dim(category_fit$coefficients), c(280L, 2L)),
  category_fit$coefficients[peak, "category"] > 0,
  # The animate/inanimate block structure the text describes.
  max(peak_distances[within_category]) <
    min(peak_distances[setdiff(names(peak_distances), within_category)]) / 10
)
```

`rsa()` aligns the model's named rows and columns to the relation conditions.
It rejects a model with missing names or a rank-deficient regression design
before it reads geometry rows.

If you only want some pairs, ask for them. The rest is never materialized:

```{r rdm-pairs}
dim(rdm(plan, pairs = rbind(c("face", "house")))$values)
```

```{r rdm-pairs-check, include=FALSE}
stopifnot(
  identical(dim(rdm(plan, pairs = rbind(c("face", "house")))$values), c(280L, 1L))
)
```

## Uncertainty, and what the package refuses to fake

The six run pairs share run estimates and are not six independent sampling
units. Their empirical spread divided by the square root of six is not a
standard error.

`example_fmri_effects()` used `lm_relation_fit()`, so the fit retains the
residual error channel required by the package's fixed-metric covariance
model:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  plan,
  example$fit,
  target = "plugin",
  at = peak
)
distance_se <- sqrt(sampling_covariance(distance_covariance))
round(distance_se, 3)
```

```{r uncertainty-check, include=FALSE}
stopifnot(
  length(distance_se) == 6L,
  all(is.finite(distance_se)),
  all(distance_se > 0)
)
```

`target` is required because the two available choices answer different
variance questions, and the package will not pick one for you.

- `target = "null"` evaluates the signal-dependent variance term under a fixed
  zero-effect null. It is exact there, so prefer it whenever the standard error
  is referenced to a test of no effect.
- `target = "plugin"`, used above, substitutes the partition mean of the
  *estimates* for the unknown signal. That substitution biases the signal term
  upward, by an amount that shrinks like one over the squared number of
  partitions and is largest when noise dominates the true distances. Use it
  when reporting uncertainty around an estimated nonzero distance, and read it
  as mildly conservative.

`?rdm_sampling_covariance` states the bias term exactly.

The result is a within-searchlight covariance under an equal-partition,
fixed-metric, separable error model. It is not a cross-location random-field
model, a confidence interval, or group inference. The object stores a
factorized covariance operator. `sampling_covariance()` can read its diagonal,
selected entries, matrix action, quadratic form, or fixed linear transport
without constructing the full distance-by-distance matrix.

Now the part worth knowing before you need it. Those standard errors are
available because this fit kept residuals. Build the same plan from beta
matrices alone and the request is refused, by name:

```{r refusal}
conditions <- c("face", "body", "house", "tool")
betas_only_domain <- abstract_domain(40L, id = "betas-only:v1")

betas_only_runs <- lapply(seq_len(4), function(run) matrix(
  rnorm(4 * 40), nrow = 4, dimnames = list(conditions, NULL)
))
names(betas_only_runs) <- paste0("run", seq_len(4))

betas_only <- relation(
  betas_only_runs,
  effects = effect_space(conditions),
  domain = betas_only_domain
)
betas_only_plan <- plan_geometry(
  betas_only,
  compile_frame(whole_brain(), betas_only_domain),
  cross_partitions(
    betas_only, independence = "independent", generalizes_over = "run"
  )
)

catch_refusal(
  rdm_sampling_covariance(betas_only_plan, betas_only, target = "null", at = 1L)
)
```

```{r refusal-check, include=FALSE}
refusal <- catch_refusal(
  rdm_sampling_covariance(betas_only_plan, betas_only, target = "null", at = 1L)
)
stopifnot(
  inherits(refusal, "effect_capability_refusal"),
  identical(refusal$capability, "sampling_covariance"),
  "missing_error_channel" %in% refusal$reasons
)
```

That is a value, not a message: `catch_refusal()` hands back the unmet
requirement and its remedy so a script can branch on them. The point estimates
are unaffected — `contrast_energy()`, `rdm()`, and `rsa()` all work on
`betas_only_plan`. What is withheld is the one quantity the inputs cannot
support. `sampling_capabilities(plan, fit)` answers the same admission question
before you provoke it.

## Use your own data

`example_fmri_effects()` handed you a fitted relation; yours has to be built.
Every route ends at a plan you can pass to `contrast_energy()`, `rdm()`, and
`rsa()`. They differ only in what you start from, and therefore in whether the
analytic standard errors above remain admitted.

The route below is the common one: beta matrices in a volume, which is also the
starting point that makes searchlights available. It runs as written; the
generated inputs stand in for yours. Two other starting points — raw
scan-by-feature responses with an already-compiled design, which keeps the
residual channel, and beta matrices over a bare list of feature columns, which
does not — are worked in `vignette("from-observations")`.

### From beta matrices in a volume

If independent runs already contain condition-by-voxel estimates, build a point
relation directly. It needs a named list of matrices with one row per
condition, a named effect space, and a neural domain. Give that domain a mask
and voxel spacing and searchlights become available. Nothing downstream
changes: the same relation, pairing, and views apply, and the result simply has
one row per searchlight.

So that there is something to find, twelve voxels below carry the same
face-minus-house pattern in every run, alternating in sign so that the regional
mean stays near zero — the fixture's pattern block trick, done by hand. Delete
`volume_planted`, `volume_pattern`, and the two lines that add them to `b` when
you substitute your own betas.

```{r own-betas-volume}
set.seed(20260816)
condition_names <- c("face", "body", "house", "tool")
mask <- array(TRUE, c(8L, 8L, 6L))
mask_domain <- volume_domain(mask, spacing = c(3, 3, 3))
volume_frame <- compile_frame(searchlights(radius = 4), mask_domain)

volume_planted <- array(FALSE, dim(mask))
volume_planted[3:4, 3:5, 3:4] <- TRUE
volume_pattern <- rep(c(3, -3), length.out = sum(volume_planted))

# Substitute your own: one condition-by-voxel matrix per run, rows named.
volume_runs <- lapply(seq_len(4), function(run) {
  b <- matrix(
    rnorm(length(condition_names) * sum(mask)),
    nrow = length(condition_names),
    dimnames = list(condition_names, NULL)
  )
  b["face", volume_planted[mask]] <-
    b["face", volume_planted[mask]] + volume_pattern
  b["house", volume_planted[mask]] <-
    b["house", volume_planted[mask]] - volume_pattern
  b
})
names(volume_runs) <- paste0("run", seq_len(4))

volume_relation <- relation(
  volume_runs,
  effects = effect_space(condition_names, units = "percent-signal"),
  domain = mask_domain
)
volume_plan <- plan_geometry(
  volume_relation,
  volume_frame,
  cross_partitions(
    volume_relation,
    independence = "independent", generalizes_over = "run"
  )
)
volume_effect <- contrast_energy(
  volume_plan, c(face = 1, house = -1, body = 0, tool = 0)
)
dim(rdm(volume_plan)$values)
```

The same picture works here, and this time nothing was told where to look.

```{r own-betas-plot, fig.width=8, fig.height=4, fig.cap="The planted pattern on generated 'own' data. No highlight was supplied: the candidates are simply the measurements standing clear of the band at zero."}
plot(volume_effect)
```

Whether that is the right answer is checkable, because we planted it. The
searchlights that give any weight to a planted voxel are the ones that should
have won:

```{r own-betas-truth}
touching <- which(
  Matrix::rowSums(volume_frame$weights[, volume_planted[mask], drop = FALSE]) > 0
)
ranked <- order(volume_effect$total, decreasing = TRUE)
c(
  searchlights = nrow(volume_frame$index),
  touching_signal = length(touching),
  top_energies_touching = sum(ranked[seq_along(touching)] %in% touching),
  peak_energy = max(volume_effect$total),
  largest_elsewhere = max(volume_effect$total[-touching]),
  fraction_below_zero_elsewhere = mean(volume_effect$total[-touching] < 0)
)
```

```{r own-betas-checks, include=FALSE}
stopifnot(
  identical(dim(rdm(volume_plan)$values), c(sum(mask), 6L)),
  sum(ranked[seq_along(touching)] %in% touching) == length(touching),
  max(volume_effect$total[-touching]) < 0.1 * max(volume_effect$total),
  all(volume_effect$configuration[touching] > volume_effect$coherent[touching])
)
```

The largest energies in the map are exactly the searchlights that touch the
planted voxels, every one of them is configuration-dominated, and about half of
the rest fall below zero.

This relation is enough for `contrast_energy()`, `rdm()`, and `rsa()`. It
cannot supply the analytic standard errors shown above, for the reason the
refusal gave: beta matrices contain no residuals and no residual degrees of
freedom. Keeping them means starting further back, which is what
`vignette("from-observations")` covers.

### Reading a map without ground truth

Your own data will not come with `volume_planted`, and the figure still works.
`highlight` is optional, as the plot above shows. Read it in two steps.

- **The right panel gives you candidates.** The band around zero *is* the noise
  floor: crossvalidated energy is centered there when nothing reproduces, which
  is why roughly half of it falls below the line. Candidates are the
  measurements standing clear of that band.
- **The left panel gives you their kind.** High on the vertical axis is a
  reproducible pattern; far along the horizontal axis is a regional mean that a
  univariate analysis would also have found.

What the figure does not give you is a threshold or a p-value. `crossform`
performs no spatial and no group inference, and it will not invent a null
distribution over measurements for you. A within-measurement standard error is
available, but only from a fit that kept residuals, which the beta matrices
above did not.
[Reading a map without ground truth](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html#reading-a-map-without-ground-truth)
(`vignette("interpreting-results")`) works through what a candidate is and is
not, including why an error bar taken at `which.max()` is not an error bar.

### The other starting points

Everything above began with betas, which is why the standard errors were
refused. If you still have the scan-by-feature responses — or you want the
first-level model declared rather than hidden in an analysis script — continue
with `vignette("from-observations", package = "crossform")`. That guide binds
scan times, events, confounds, named effects, and an observation model before
it fits the relation, and `plan_relation()` there accepts either a semantic
`design_model()` or the version-checked `fmridesign_design_model()` adapter. It
also works the short version of the same journey: `lm_relation_fit()` straight
from an already-compiled design matrix, which keeps the error channel and
readmits the analytic standard errors, and `relation()` straight from beta
matrices over a plain list of feature columns, which does not.

For `neuroim2` volumes, `neuroim2_volume_domain()` and `neuroim2_searchlights()`
replace `volume_domain()` and `searchlights()` while preserving stable
full-volume indices, and `as_neurovol()` maps compact results back to a volume
without interpolation or smoothing.

Crossform records the requested effects, design identity, fit receipt, and
supported uncertainty operations. It does not perform preprocessing,
registration, masking, or general hemodynamic response modeling.

## Why correlation distance is not an option to `rdm()`

Feature centering is a fixed linear projection, so a plan can declare it as a
metric. Pearson correlation also divides each observed pattern by its norm.
That denominator depends on the data and is therefore not a fixed linear query.
A conventional correlation distance can be defined from a positive-semidefinite
self geometry. Applying the same diagonal normalization to a signed
cross-generalized form does not automatically define a valid crossvalidated
estimator.

The package therefore does not hide correlation normalization inside `rdm()`.
The accepted extension requirements and refusal cases are recorded in the
[correlation-distance policy](https://bbuchsbaum.github.io/crossform/articles/correlation-distance-policy.html)
(`vignette("correlation-distance-policy")`).

## What you just did that you could not before

- **You asked four questions of one fit.** The contrast map, the RDM, the RSA
  coefficient, and the single-pair query all read the same compiled plan, so
  they agree by construction rather than by care. The
  [novelty ledger](https://bbuchsbaum.github.io/crossform/articles/novelty.html)
  (`vignette("novelty")`) states what that architecture does and does not yet
  demonstrate.
- **You learned what kind of effect you had found.** The exact, non-destructive
  split into regional-mean and pattern-beyond-the-mean energy came free with the
  contrast, and separated two blocks of equal effect size into a pattern and a
  regional mean.
  [Reading the results](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html)
  (`vignette("interpreting-results")`) covers the traps.
- **You saw the null.** Negative crossvalidated estimates were kept, so the
  noise floor sat on zero in the figure instead of being clipped into a
  positive-looking band.
- **You were told what could not be estimated, and why.** The refusal named the
  missing error channel and the remedy. The
  [failure gallery](https://bbuchsbaum.github.io/crossform/articles/failure-gallery.html)
  (`vignette("failure-gallery")`) collects six such guards, this one among
  them.

## Where to go next

- `?plan_geometry` describes rectangular cross-axis plans
  (`right = `), read with axis-bound `pair_query()`s, and `?coupling` takes
  the adjoint neural-side closure from the same plan vocabulary.
- `vignette("evidence-pairing")` covers measurement forms and coupling
  views, including one bounded cross-domain contraction and the Parseval
  reconstruction law.
- `?rdm_sampling_covariance` states the admitted analytic covariance
  contract, and `?sampling_capabilities` answers the admission question
  before it is provoked.
- The Haxby 2001 exemplar supplies public-data parity and refusal evidence.
- [Coming from rMVPA](https://bbuchsbaum.github.io/crossform/articles/from-rmvpa.html)
  (`vignette("from-rmvpa")`) maps the familiar objects onto this vocabulary.

For the usual workflow, construct one plan from the fitted relation, spatial
frame, and cross-run pairing. Then pass that plan to `contrast_energy()`,
`rdm()`, or `rsa()` according to the result you need.
