---
title: "From partitioned effects to interpretable geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From partitioned effects to interpretable geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(effectagram)
```

`effectagram` computes one durable object: experimental geometry that
reproduces across declared partitions under a declared spatial measurement.
Changing a searchlight to an ROI changes the measurement frame. Changing a
contrast to RSA changes the view. Neither choice introduces another fitting
engine.

This vignette starts from precomputed run effects. The final section shows where
a linear extractor enters when responses have not yet been reduced to effects.

## Declare the native relation

The rows below are experimental effects and the columns are native neural
features. Runs remain separate because agreement between independent estimates
defines the cross-generalized quantity.

```{r data}
effects <- effect_space(
  c("face", "house", "object"),
  basis_id = "condition-means:v1",
  units = "percent-signal"
)
domain <- abstract_domain(
  6,
  coordinates = cbind(x = 0:5, y = 0),
  feature_ids = paste0("feature", 1:6),
  id = "native:demo"
)

run1 <- rbind(
  face   = c(2, 1, 0.5, -0.5, 0, 1),
  house  = c(0, 1, 2, 1, 0.5, 0),
  object = c(0.5, 0.5, 1, 1.5, 1, 0.5)
)
run2 <- rbind(
  face   = c(2.2, 0.8, 0.4, -0.4, 0.1, 0.9),
  house  = c(0.1, 1.2, 1.8, 0.9, 0.6, 0.1),
  object = c(0.4, 0.6, 1.1, 1.4, 0.9, 0.6)
)

rel <- relation(list(run1 = run1, run2 = run2), effects = effects,
  domain = domain)
```

`relation()` records the ordered experimental basis, units and scaling,
partition identity, exact ordered neural domain, and source revisions. It does
not concatenate runs or resample features. A matching name and dimension alone
cannot substitute for either identity.

## Measure locally and require cross-run agreement

```{r geometry}
at <- compile_frame(searchlights(radius = 1.01, normalization = "local"), domain)
over <- cross_partitions(rel)
g <- geometry(rel, at = at, over = over)
```

The searchlights are rows of one sparse additive frame. `over` is an undirected
edge table containing each run pair once. The geometry engine reads each
feature block once per partition, forms cross-run feature atoms, and contracts
the whole frame.

`g` retains complete packed total and coherent geometry. Configuration is
derived as `total - coherent`, so the decomposition cannot drift across reruns.

## Read level and configuration together

```{r contrast}
face_house <- contrast(g, c(face = 1, house = -1, object = 0))
round(cbind(
  signed = face_house$signed,
  coherent = face_house$coherent,
  configuration = face_house$configuration,
  total = face_house$total
), 3)
stopifnot(isTRUE(all.equal(
  face_house$total,
  face_house$coherent + face_house$configuration,
  tolerance = 0
)))
```

The signed column is the locally averaged contrast. Coherent energy asks
whether that local average reproduces across runs. Configuration energy asks
whether spatial departures from the average reproduce. Total energy is their
exact sum. Configuration is not evidence that every single feature is
marginally silent, nor that a biological interaction exists.

Because cross-products are unbiased around zero, coherent, configuration, and
total estimates can be negative. The package preserves those values.

## Change the question, not the analysis

Squared experimental distances are a fixed linear transform of geometry:

```{r rdm}
distances <- rdm(g)
round(distances$values[1:2, ], 3)
distances$pairs
```

Multiple-regression RSA compiles that same distance transform together with a
fixed regression design:

```{r rsa}
category_rdm <- matrix(
  c(0, 0, 1,
    0, 0, 1,
    1, 1, 0),
  3, 3, byrow = TRUE,
  dimnames = list(rownames(run1), rownames(run1))
)
model_fit <- rsa(g, models = list(category = category_rdm))
round(model_fit$coefficients, 3)
model_fit$terms
```

Named RDM axes are aligned to the geometry's effect identities. Rank-deficient
RSA designs fail before any geometry rows are read.

The geometry spectrum is the nonlinear descriptive view in this workflow:

```{r spectrum}
roots <- geometry_spectrum(g)
round(roots$values[1:2, ], 3)
```

Negative roots are retained. A later positive-semidefinite latent model may be
useful for description, but truncating the unbiased estimate would change its
inferential meaning.

## Choose the frame interpretation explicitly

```{r frames}
point_frame <- compile_frame(voxels(), domain)
roi_frame <- compile_frame(
  regions(c("anterior", "anterior", "anterior",
            "posterior", "posterior", "posterior")),
  domain
)
global_frame <- compile_frame(whole_brain(), domain)
```

These values all enter the same `geometry()` call. With `normalization =
"local"`, each row is an average. With `normalization = "conservative"`, each
feature distributes unit mass across overlapping rows and local total geometry
sums to global geometry.

## Avoid materializing unused geometry

If only one fixed bilinear question is needed, compile it before execution:

```{r direct}
query <- bilinear_query(tcrossprod(c(1, -1, 0)), effects = effects)
direct <- evaluate_geometry(rel, at, over, query, component = "total")
late <- query_geometry(g, query, component = "total")
stopifnot(isTRUE(all.equal(direct$values, late$values, tolerance = 1e-12)))
direct$completeness
```

`evaluate_geometry()` returns a query-only `effect_view`. `geometry()` always
returns complete geometry. Storage or workspace policy never changes that
semantic distinction.

## Supply an extractor when effects are still implicit

For responses `Y`, a fixed extractor `E` defines the lazy relation `B = E Y`.
`lm_extractor()` compiles an already constructed design and target map using QR,
or an estimability-checked SVD when the design is rank deficient.

```{r extractor, eval=FALSE}
extractor <- lm_extractor(
  design = X,
  effects = C,
  whiten = L
)
rel <- relation(response_runs, extract = extractor, domain = domain)
```

The package does not construct event designs, HRFs, nuisance models, temporal
whitening, or registration. Those choices remain upstream and explicit.

Function-backed sources must also provide `source_capabilities()` with a strong
immutable SHA-256 revision. Matrix sources receive this identity automatically.
The compiler validates the scientific objects and conservative workspace plan
before reading source blocks. A `workspace_bytes` limit covers only
effectagram-owned live workspace. Baseline, incremental peak, and absolute peak
process RSS remain separate observations.

If your native mask is a neuroim2 `NeuroVol`, use
`neuroim2_volume_domain()` and `neuroim2_searchlights()` to preserve its stable
full-volume feature indices and spacing. This optional adapter compiles geometry
only; neuroim2 ROI iteration, data extraction, results, and parallel state do
not enter effectagram.

Version 0.1 executes sequentially. The current claim is local and precise:
additive-frame searchlights are sparse contraction rows rather than independent
worker jobs. Locally learned covariance, adaptive feature selection, nonlinear
normalization, and classifier accuracy remain outside this bilinear core.
