---
title: "Where does a representational effect generalize?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where does a representational effect generalize?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

## The question

You have condition estimates from repeated runs, and you want to know where an
effect reproduces across those runs. This guide answers that question, and then
three more from the same fit without recomputing anything: what *kind* of effect
it is, what the whole representational geometry looks like at the strongest
location, and how much of that you are entitled to put an error bar on.

One word to fix before starting: a **measurement** is one spatial unit of the
frame, and every result below has one value per measurement. Here the
measurements are searchlights; anatomical regions, single voxels, and the whole
brain are the other choices, and they all behave the same way downstream.

The worked question is: **where does an animate-versus-inanimate pattern
reproduce across independent runs?** The data are generated with known truth, so
the example runs after installation and can check its own answer. These
generated data are not empirical evidence; the separate Haxby 2001 exemplar is
the public-data comparison.

```{r example}
example <- example_fmri_effects()
c(
  conditions = length(example$fit$relation$effects),
  runs = length(example$fit$relation$partitions),
  features = example$fit$relation$n_features,
  searchlights = nrow(example$frame$index)
)
```

`example$fit` contains four run-specific estimates for four conditions in a
small 7 by 7 by 5 volume. `example$frame` defines the 245 searchlights at which
the analysis will report results. The generator returns the frame separately
because a fitted relation records the neural domain but does not own a mutable
searchlight definition.

The fit provides two kinds of information:

- `example$fit$relation` is the condition-by-feature relation used for point
  estimates; it reads feature blocks only when a query needs them;
- its error channel stores residual blocks, residual degrees of freedom, and
  effect-coordinate covariance for uncertainty calculations whose assumptions
  the fit can support.

## Declare what must generalize

Three declarations make the estimand: the relation you fitted, the frame you
want answers at, and the pairing that says what must reproduce.

```{r plan}
plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)
plan
```

`cross_partitions()` requests products between different runs. Every unordered
pair of runs contributes to the estimate, but the function does not treat the
six run pairs as six independent samples. `plan_geometry()` checks that the
relation, condition names, searchlight frame, metric, units, and neural domain
agree before it reads neural values.

Read the last line of that print. The plan is query-first: it holds the identity
of the quantity, and nothing has been materialized yet. Save `plan` with the
analysis record. It identifies the estimand: the scientific quantity the
analysis is meant to estimate. Changing block size or storage changes the
execution receipt, not this plan identity.

### Choosing what must generalize

Two arguments to `cross_partitions()` decide what you are estimating, so it is
worth being deliberate about both.

`generalizes_over` names the sampling axis your partitions *are* — `"run"`,
`"session"`, `"task"`. Cross-run and cross-session reproduction are different
scientific quantities even at identical fold counts, so the declared axis is
bound into every plan identity built from this pairing. If your partitions are
sessions, pass `"session"`.

`independence = "independent"` is your declaration that distinct partition
estimates have independent estimation errors, which separate runs or sessions
with separately estimated noise satisfy. It has to be stated explicitly:
leaving it `NULL` records the endpoints as undeclared, which still yields a
point estimand but earns neither cross-generalized nor analytic sampling-law
capabilities.

```{r pairing-identity}
declared <- cross_partitions(
  example$fit$relation,
  independence = "independent", generalizes_over = "run"
)
undeclared <- cross_partitions(example$fit$relation)
c(
  declared = attr(declared, "estimate"),
  axis = attr(declared, "generalizes_over"),
  undeclared = attr(undeclared, "estimate")
)
```

```{r pairing-identity-check, include=FALSE}
stopifnot(
  identical(attr(declared, "estimate"), "cross_generalized"),
  identical(attr(undeclared, "estimate"), "independence_undeclared"),
  nrow(declared) == 6L,
  isTRUE(all.equal(sum(declared$weight), 1))
)
```

The plan print's `generalizes:` line is where you check that you declared what
you meant. Left undeclared it reads
`6 partition pairs (axis undeclared), endpoints undeclared` instead.

One thing the pairing is *not*: its six rows are estimator contributions, not
six independent replicates. Pairs that share a run share that run's estimation
error, which is why the standard-error section below cannot simply take their
spread. `?cross_partitions` states this; `?pairing` builds explicit or directed
edge sets for designs that need one, such as nested or ordered axes.

## The answer

```{r contrast}
effect <- contrast_energy(plan, example$contrast)
peak <- which.max(effect$total)
c(
  peak_searchlight = peak,
  signed_effect = effect$signed[peak],
  total_energy = effect$total[peak],
  coherent_energy = effect$coherent[peak],
  configuration_energy = effect$configuration[peak],
  overlaps_planted_signal = peak %in% example$truth$signal_measurements
)
```

`contrast_energy()` returns four values at every searchlight. `signed` is the
usual signed contrast of the local weighted mean. The other three values
measure cross-run reproducibility:

- `coherent` is the crossvalidated energy of that local weighted mean;
- `configuration` is the energy in reproducible spatial departures from the
  weighted mean;
- `total` is their sum.

Because the fixture planted its signal in known searchlights, the answer can be
drawn against the truth. `highlight` marks those searchlights in the picture;
it plays no part in the estimate.

```{r contrast-plot, fig.width=8, fig.height=4, fig.cap="Left: every searchlight in the coherent/configuration plane, with the dashed guides at constant total energy. Right: total energy along the searchlight index. The planted searchlights are filled."}
plot(
  effect,
  highlight = example$truth$signal_measurements,
  highlight_label = "planted signal"
)
```

```{r contrast-check, include=FALSE}
stopifnot(
  peak %in% example$truth$signal_measurements,
  isTRUE(all.equal(
    effect$total, effect$coherent + effect$configuration, tolerance = 1e-14
  ))
)
```

The right panel is the map. The planted searchlights are the spikes; everything
else lies on zero rather than above it. That is deliberate. Crossvalidated
energy is centered on zero when no reproducible effect is present, so sampling
noise produces negative estimates about half the time. Clipping them to zero
would change the estimator and introduce positive bias, so `contrast_energy()`
retains them and the plot draws them where they fall.

```{r planted-summary}
summarize_region <- function(index) c(
  total = mean(effect$total[index]),
  coherent = mean(effect$coherent[index]),
  configuration = mean(effect$configuration[index]),
  fraction_below_zero = mean(effect$total[index] < 0)
)
signal_searchlights <- example$truth$signal_measurements
region_summary <- rbind(
  signal = summarize_region(signal_searchlights),
  outside = summarize_region(
    setdiff(seq_along(effect$total), signal_searchlights)
  )
)
round(region_summary, 3)
```

```{r planted-summary-check, include=FALSE}
stopifnot(
  region_summary["signal", "configuration"] >
    region_summary["signal", "coherent"],
  abs(region_summary["outside", "total"]) < 0.01,
  region_summary["outside", "fraction_below_zero"] > 0.4,
  region_summary["outside", "fraction_below_zero"] < 0.6
)
```

Outside the planted searchlights the mean total energy is within `0.01` of zero
and roughly half the estimates are below it. That is what a correctly centered
null looks like, and you can see it rather than assume it.

```{r contrast-map, echo=FALSE, fig.cap="Cross-generalized contrast energy in the middle slice. The generated signal is centered in the volume; the map checks the analysis and is not empirical evidence."}
volume <- array(NA_real_, dim = example$domain$metadata$dim)
volume[example$domain$metadata$mask] <- effect$total
middle <- ceiling(dim(volume)[3] / 2)
image(
  t(volume[, , middle]), axes = FALSE, asp = 1,
  col = hcl.colors(24, "Blue-Red 3", rev = TRUE),
  main = sprintf("Middle slice (z = %d)", middle)
)
box()
```

## What kind of effect is it?

The left panel of the two-panel figure is the part that is hard to obtain
elsewhere. It places every searchlight by how its reproducible energy divides:
`coherent` on the horizontal axis is the energy carried by the searchlight's
frame-weighted mean pattern, which is the familiar univariate story, and
`configuration` on the vertical axis is the reproducible pattern beyond that
mean.

The two components sum to `total` exactly at every measurement, which is why
lines of constant total appear as the dashed guides of slope `-1`. The
partition is checked, not assumed:

```{r partition-check}
max(abs(effect$total - (effect$coherent + effect$configuration)))
```

The coherent and configuration modes are orthogonal under the weights declared
by the frame. They need not be orthogonal under an unweighted Euclidean inner
product.

In the region summary above, the planted searchlights carry more configuration
than coherent energy. That is the signature of a multivariate pattern a
univariate map would miss: the sphere reproduces a *shape*, while its average
response barely moves. Nothing was removed to see this, and nothing was fitted
twice; the split is a property of the plan.

"A univariate map would miss it" is checkable here. `signed` is a first
moment — the ordinary contrast of the frame-weighted mean pattern, carrying
your effect's units and sign — while the three energies are second moments,
squared and crossvalidated. `signed` can therefore only ever see the coherent
half.

```{r signed-versus-configuration}
outside <- setdiff(seq_along(effect$total), signal_searchlights)
largest_signed_outside <- max(abs(effect$signed[outside]))
quiet_but_reproducing <- signal_searchlights[
  abs(effect$signed[signal_searchlights]) < largest_signed_outside
]
c(
  largest_signed_outside_signal = largest_signed_outside,
  planted_below_that_signed_level = length(quiet_but_reproducing),
  planted_total = length(signal_searchlights),
  their_largest_total_energy = max(effect$total[quiet_but_reproducing])
)
```

```{r signed-versus-configuration-check, include=FALSE}
stopifnot(
  length(quiet_but_reproducing) > 10L,
  max(effect$total[quiet_but_reproducing]) >
    50 * max(effect$total[outside])
)
```

Seventeen of the 57 planted searchlights have a signed contrast smaller than
the largest signed value anywhere outside the signal, so a signed map cannot
distinguish them from noise. One of them still carries `1.99` of reproducible
energy, against a largest-elsewhere total of `0.03`. The energies see what the
first moment cannot.

`contrast_energy()` also reports `coherence_fraction`, the share of reproducible
energy carried by the local mean, but only where the components form a
nonnegative partition. Elsewhere it is `NA`, because a ratio whose denominator
can be negative is not a share of anything. The
[interpreting-results](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html)
article covers how to read that fraction and the three ways it can mislead you.

## Change the question without refitting

The plan is unchanged, and nothing about it is recompiled below. For conditions
`i` and `j`, `rdm()` applies the fixed contrast `c = e_i - e_j` to the geometry:

\[
d_{ij} = c^T G c = G_{ii} + G_{jj} - 2G_{ij}.
\]

```{r rdm}
distances <- rdm(plan)
round(distances$values[peak, ], 3)
```

```{r rdm-plot, fig.width=5, fig.height=4.5, fig.cap="Signed squared distances among the four conditions at the strongest searchlight. Distances are signed and may be negative; the color scale is centered on zero."}
plot(distances, measurement = peak)
```

Nothing asked for that block structure. The two animate conditions sit close
together, the two inanimate conditions sit close together, and every
animate-versus-inanimate pair is far apart. It is what the geometry contains,
read out of the same numbers that produced the contrast map.

With the current identity metric and cross-run pairing, these are signed,
crossvalidated squared Euclidean distances. Supplying a fixed neural precision
metric changes them to squared Mahalanobis distances. Either estimate can be
negative near the null because the two factors come from different runs.

Linear RSA compiles the RDM transform and regression coefficient into a fixed
query against the same plan:

```{r rsa}
category_fit <- rsa(
  plan,
  models = list(category = example$model_rdm)
)
round(category_fit$coefficients[peak, ], 3)
```

```{r geometry-view-checks, include=FALSE}
peak_distances <- distances$values[peak, ]
within_category <- c("face - body", "house - tool")
stopifnot(
  identical(dim(distances$values), c(245L, 6L)),
  identical(dim(category_fit$coefficients), c(245L, 2L)),
  category_fit$coefficients[peak, "category"] > 0,
  # The animate/inanimate block structure the text describes.
  max(peak_distances[within_category]) <
    min(peak_distances[setdiff(names(peak_distances), within_category)]) / 10
)
```

`rsa()` aligns the model's named rows and columns to the relation conditions.
It rejects a model with missing names or a rank-deficient regression design
before it reads geometry rows.

If you only want some pairs, ask for them. The rest is never materialized:

```{r rdm-pairs}
dim(rdm(plan, pairs = rbind(c("face", "house")))$values)
```

```{r rdm-pairs-check, include=FALSE}
stopifnot(
  identical(dim(rdm(plan, pairs = rbind(c("face", "house")))$values), c(245L, 1L))
)
```

## Uncertainty, and what the package refuses to fake

The six run pairs share run estimates and are not six independent sampling
units. Their empirical spread divided by the square root of six is not a
standard error.

`example_fmri_effects()` used `lm_relation_fit()`, so the fit retains the
residual error channel required by the package's fixed-metric covariance
model:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  plan,
  example$fit,
  target = "plugin",
  at = peak
)
distance_se <- sqrt(sampling_covariance(distance_covariance))
round(distance_se, 3)
```

```{r uncertainty-check, include=FALSE}
stopifnot(
  length(distance_se) == 6L,
  all(is.finite(distance_se)),
  all(distance_se > 0)
)
```

`target` is required because the two available choices answer different
variance questions, and the package will not pick one for you.

- `target = "null"` evaluates the signal-dependent variance term under a fixed
  zero-effect null. It is exact there, so prefer it whenever the standard error
  is referenced to a test of no effect.
- `target = "plugin"`, used above, substitutes the partition mean of the
  *estimates* for the unknown signal. That substitution biases the signal term
  upward, by an amount that shrinks like one over the squared number of
  partitions and is largest when noise dominates the true distances. Use it
  when reporting uncertainty around an estimated nonzero distance, and read it
  as mildly conservative.

`?rdm_sampling_covariance` states the bias term exactly.

The result is a within-searchlight covariance under an equal-partition,
fixed-metric, separable error model. It is not a cross-location random-field
model, a confidence interval, or group inference. The object stores a
factorized covariance operator. `sampling_covariance()` can read its diagonal,
selected entries, matrix action, quadratic form, or fixed linear transport
without constructing the full distance-by-distance matrix.

Now the part worth knowing before you need it. Those standard errors are
available because this fit kept residuals. Build the same plan from beta
matrices alone and the request is refused, by name:

```{r refusal}
conditions <- c("face", "body", "house", "tool")
betas_only_domain <- abstract_domain(40L, id = "betas-only:v1")

betas_only_runs <- lapply(seq_len(4), function(run) matrix(
  rnorm(4 * 40), nrow = 4, dimnames = list(conditions, NULL)
))
names(betas_only_runs) <- paste0("run", seq_len(4))

betas_only <- relation(
  betas_only_runs,
  effects = effect_space(conditions),
  domain = betas_only_domain
)
betas_only_plan <- plan_geometry(
  betas_only,
  compile_frame(whole_brain(), betas_only_domain),
  cross_partitions(
    betas_only, independence = "independent", generalizes_over = "run"
  )
)

catch_refusal(
  rdm_sampling_covariance(betas_only_plan, betas_only, target = "null", at = 1L)
)
```

```{r refusal-check, include=FALSE}
refusal <- catch_refusal(
  rdm_sampling_covariance(betas_only_plan, betas_only, target = "null", at = 1L)
)
stopifnot(
  inherits(refusal, "effect_capability_refusal"),
  identical(refusal$capability, "sampling_covariance"),
  "missing_error_channel" %in% refusal$reasons
)
```

That is a value, not a message: `catch_refusal()` hands back the unmet
requirement and its remedy so a script can branch on them. The point estimates
are unaffected — `contrast_energy()`, `rdm()`, and `rsa()` all work on
`betas_only_plan`. What is withheld is the one quantity the inputs cannot
support. `sampling_capabilities(plan, fit)` answers the same admission question
before you provoke it.

## Use your own data

The three routes below all end at a plan you can pass to `contrast_energy()`,
`rdm()`, and `rsa()`. They differ only in what you start from, and therefore in
whether analytic standard errors remain available. Each one runs as written; the
generated inputs stand in for yours.

### From beta matrices

If independent runs already contain condition-by-feature estimates, build a
point relation directly. The relation needs a named list of matrices with one
row per condition, a named effect space, and a neural domain.

```{r own-betas}
set.seed(20260815)
condition_names <- c("face", "body", "house", "tool")
native_domain <- abstract_domain(60L, id = "own-betas:v1")

# Substitute your own: one condition-by-feature matrix per run, rows named.
beta_runs <- lapply(seq_len(3), function(run) matrix(
  rnorm(length(condition_names) * native_domain$n_features),
  nrow = length(condition_names),
  dimnames = list(condition_names, native_domain$feature_ids)
))
names(beta_runs) <- paste0("run", seq_len(3))

rel <- relation(
  beta_runs,
  effects = effect_space(condition_names, units = "percent-signal"),
  domain = native_domain
)
own_plan <- plan_geometry(
  rel,
  compile_frame(whole_brain(), native_domain),
  cross_partitions(rel, independence = "independent", generalizes_over = "run")
)
own_plan
```

`abstract_domain()` is the right domain when the features are just a list of
columns. `whole_brain()` reports one measurement over all of them, which is the
simplest frame to check a pipeline against.

### From beta matrices in a volume

Give the domain a mask and voxel spacing, and searchlights become available.
Nothing downstream changes: the same relation, pairing, and views apply, and the
result simply has one row per searchlight instead of one row overall.

So that there is something to find, twelve voxels below carry the same
face-minus-house pattern in every run, alternating in sign so that the regional
mean stays near zero. Delete those three lines when you substitute your own
betas.

```{r own-betas-volume}
set.seed(20260816)
mask <- array(TRUE, c(8L, 8L, 6L))
mask_domain <- volume_domain(mask, spacing = c(3, 3, 3))
volume_frame <- compile_frame(searchlights(radius = 4), mask_domain)

volume_planted <- array(FALSE, dim(mask))
volume_planted[3:4, 3:5, 3:4] <- TRUE
volume_pattern <- rep(c(3, -3), length.out = sum(volume_planted))

# Substitute your own: one condition-by-voxel matrix per run, rows named.
volume_runs <- lapply(seq_len(4), function(run) {
  b <- matrix(
    rnorm(length(condition_names) * sum(mask)),
    nrow = length(condition_names),
    dimnames = list(condition_names, NULL)
  )
  b["face", volume_planted[mask]] <-
    b["face", volume_planted[mask]] + volume_pattern
  b["house", volume_planted[mask]] <-
    b["house", volume_planted[mask]] - volume_pattern
  b
})
names(volume_runs) <- paste0("run", seq_len(4))

volume_relation <- relation(
  volume_runs,
  effects = effect_space(condition_names, units = "percent-signal"),
  domain = mask_domain
)
volume_plan <- plan_geometry(
  volume_relation,
  volume_frame,
  cross_partitions(
    volume_relation,
    independence = "independent", generalizes_over = "run"
  )
)
volume_effect <- contrast_energy(
  volume_plan, c(face = 1, house = -1, body = 0, tool = 0)
)
dim(rdm(volume_plan)$values)
```

The same picture works here, and this time nothing was told where to look.

```{r own-betas-plot, fig.width=8, fig.height=4, fig.cap="The planted pattern on generated 'own' data. No highlight was supplied: the candidates are simply the measurements standing clear of the band at zero."}
plot(volume_effect)
```

Whether that is the right answer is checkable, because we planted it. The
searchlights that give any weight to a planted voxel are the ones that should
have won:

```{r own-betas-truth}
touching <- which(
  Matrix::rowSums(volume_frame$weights[, volume_planted[mask], drop = FALSE]) > 0
)
ranked <- order(volume_effect$total, decreasing = TRUE)
c(
  searchlights = nrow(volume_frame$index),
  touching_signal = length(touching),
  top_energies_touching = sum(ranked[seq_along(touching)] %in% touching),
  peak_energy = max(volume_effect$total),
  largest_elsewhere = max(volume_effect$total[-touching]),
  fraction_below_zero_elsewhere = mean(volume_effect$total[-touching] < 0)
)
```

```{r own-betas-checks, include=FALSE}
stopifnot(
  identical(dim(rdm(volume_plan)$values), c(sum(mask), 6L)),
  nrow(compile_frame(whole_brain(), native_domain)$index) == 1L,
  sum(ranked[seq_along(touching)] %in% touching) == length(touching),
  max(volume_effect$total[-touching]) < 0.1 * max(volume_effect$total),
  all(volume_effect$configuration[touching] > volume_effect$coherent[touching])
)
```

The largest energies in the map are exactly the searchlights that touch the
planted voxels, every one of them is configuration-dominated, and about half of
the rest fall below zero.

Either relation is enough for `contrast_energy()`, `rdm()`, and `rsa()`. Neither
can supply the analytic standard errors shown above, for the reason the refusal
gave: beta matrices contain no residuals and no residual degrees of freedom.

### Reading a map without ground truth

Your own data will not come with `volume_planted`, and the figure still works.
`highlight` is optional, as the plot above shows. Read it in two steps.

- **The right panel gives you candidates.** The band around zero *is* the noise
  floor: crossvalidated energy is centered there when nothing reproduces, which
  is why roughly half of it falls below the line. Candidates are the
  measurements standing clear of that band.
- **The left panel gives you their kind.** High on the vertical axis is a
  reproducible pattern; far along the horizontal axis is a regional mean that a
  univariate analysis would also have found.

What the figure does not give you is a threshold or a p-value. `crossform`
performs no spatial and no group inference, and it will not invent a null
distribution over measurements for you. A within-measurement standard error is
available, but only from a fit that kept residuals, which is the difference
between this section and the next one.
[Reading a map without ground truth](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html#reading-a-map-without-ground-truth)
works through what a candidate is and is not, including why an error bar taken
at `which.max()` is not an error bar.

### From raw responses and a compiled design

If you kept the scan-by-feature responses and already have a design matrix,
`lm_relation_fit()` is the compact route that keeps the error channel. The
`effects` matrix says which linear combination of design columns each named
condition is.

```{r own-raw}
n_scan <- 40L
scan_condition <- factor(
  rep(condition_names, length.out = n_scan), levels = condition_names
)
design_matrix <- cbind(
  stats::model.matrix(~ 0 + scan_condition),
  drift = seq(-1, 1, length.out = n_scan)
)
colnames(design_matrix) <- c(condition_names, "drift")

target_matrix <- cbind(diag(length(condition_names)), drift = 0)
dimnames(target_matrix) <- list(condition_names, colnames(design_matrix))

# Substitute your own: one scan-by-feature response matrix per run.
signal <- rbind(
  matrix(rnorm(length(condition_names) * 60L, sd = 0.5), length(condition_names)),
  rnorm(60L, sd = 0.1)
)
response_runs <- lapply(seq_len(3), function(run) {
  design_matrix %*% signal + matrix(rnorm(n_scan * 60L), n_scan, 60L)
})
names(response_runs) <- paste0("run", seq_len(3))

raw_fit <- lm_relation_fit(
  response_runs,
  design = design_matrix,
  effects = target_matrix,
  effect_names = effect_space(condition_names, units = "percent-signal"),
  sampling_unit = "trial",
  domain = native_domain
)
raw_plan <- plan_geometry(
  raw_fit$relation,
  compile_frame(whole_brain(), native_domain),
  cross_partitions(
    raw_fit$relation,
    independence = "independent", generalizes_over = "run"
  )
)
sampling_capabilities(raw_plan, raw_fit)
```

Because this fit retained residuals, the analytic law is admitted and the
standard errors from the earlier section are available here too:

```{r own-raw-uncertainty}
round(sqrt(sampling_covariance(
  rdm_sampling_covariance(raw_plan, raw_fit, target = "null", at = 1L)
)), 4)
```

```{r own-raw-checks, include=FALSE}
raw_se <- sqrt(sampling_covariance(
  rdm_sampling_covariance(raw_plan, raw_fit, target = "null", at = 1L)
))
stopifnot(length(raw_se) == 6L, all(is.finite(raw_se)), all(raw_se > 0))
```

### From events and confounds

If you have scan-by-feature responses but no compiled design — or you want the
first-level model declared rather than hidden in an analysis script — continue
with `vignette("from-observations", package = "crossform")`. That guide binds
scan times, events, confounds, named effects, and an observation model before it
fits the relation. `plan_relation()` there accepts either a semantic
`design_model()` or the version-checked `fmridesign_design_model()` adapter.

For `neuroim2` volumes, `neuroim2_volume_domain()` and `neuroim2_searchlights()`
replace `volume_domain()` and `searchlights()` while preserving stable
full-volume indices, and `as_neurovol()` maps compact results back to a volume
without interpolation or smoothing.

Crossform records the requested effects, design identity, fit receipt, and
supported uncertainty operations. It does not perform preprocessing,
registration, masking, or general hemodynamic response modeling.

## Why correlation distance is not an option to `rdm()`

Feature centering is a fixed linear projection, so a plan can declare it as a
metric. Pearson correlation also divides each observed pattern by its norm.
That denominator depends on the data and is therefore not a fixed linear query.
A conventional correlation distance can be defined from a positive-semidefinite
self geometry. Applying the same diagonal normalization to a signed
cross-generalized form does not automatically define a valid crossvalidated
estimator.

The package therefore does not hide correlation normalization inside `rdm()`.
The accepted extension requirements and refusal cases are recorded in the
[correlation-distance policy](https://bbuchsbaum.github.io/crossform/articles/correlation-distance-policy.html).

## What you just did that you could not before

- **You asked four questions of one fit.** The contrast map, the RDM, the RSA
  coefficient, and the single-pair query all read the same compiled plan, so
  they agree by construction rather than by care. The
  [novelty ledger](https://bbuchsbaum.github.io/crossform/articles/novelty.html)
  states what that architecture does and does not yet demonstrate.
- **You learned what kind of effect you had found.** The exact, non-destructive
  split into regional-mean and pattern-beyond-the-mean energy came free with the
  contrast, and said the planted effect was mostly pattern.
  [Reading the results](https://bbuchsbaum.github.io/crossform/articles/interpreting-results.html)
  covers the traps.
- **You saw the null.** Negative crossvalidated estimates were kept, so the
  noise floor sat on zero in the figure instead of being clipped into a
  positive-looking band.
- **You were told what could not be estimated, and why.** The refusal named the
  missing error channel and the remedy. The
  [failure gallery](https://bbuchsbaum.github.io/crossform/articles/failure-gallery.html)
  collects six such guards, this one among them.

## Where to go next

- `?plan_geometry` describes rectangular cross-axis plans
  (`right = `), read with axis-bound `pair_query()`s, and `?coupling` takes
  the adjoint neural-side closure from the same plan vocabulary.
- The `evidence-pairing` vignette covers measurement forms and coupling
  views, including one bounded cross-domain contraction and the Parseval
  reconstruction law.
- `?rdm_sampling_covariance` states the admitted analytic covariance
  contract, and `?sampling_capabilities` answers the admission question
  before it is provoked.
- The Haxby 2001 exemplar supplies public-data parity and refusal evidence.
- [Coming from rMVPA](https://bbuchsbaum.github.io/crossform/articles/from-rmvpa.html)
  maps the familiar objects onto this vocabulary.

For the usual workflow, construct one plan from the fitted relation, spatial
frame, and cross-run pairing. Then pass that plan to `contrast_energy()`,
`rdm()`, or `rsa()` according to the result you need.
