---
title: "Where does a representational effect generalize?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where does a representational effect generalize?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5)
library(effectagram)
```

This guide answers one question from beginning to end: **where does an
animate-versus-inanimate pattern reproduce across independent runs?** Along the
way, the same fitted geometry will supply a contrast map, a representational
dissimilarity matrix (RDM), a linear RSA coefficient, and an analytic standard
error where its assumptions are met.

The example is generated with known truth so that every line runs after
installation. It is not empirical evidence. The package's Haxby 2001 exemplar
provides the separate public-data parity workflow.

## Start from a ready relation

```{r example}
example <- example_fmri_effects()
example$fit
```

There are four conditions in four independent runs over a small 7 by 7 by 5
volume. The generator returns both the full domain and a compiled searchlight
frame because a fitted relation deliberately retains only the domain's exact
identity, not mutable spatial construction data.

`fit` contains two distinct things:

- `fit$relation` is the lazy condition-by-feature relation used for point
  evidence;
- its error channel retains residual blocks, residual degrees of freedom, and
  effect-coordinate covariance for admitted uncertainty calculations.

## Declare what must generalize

```{r plan}
plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(example$fit$relation, independence = "independent")
)
plan
```

`cross_partitions()` names the generalization requirement: every unordered pair
of different runs contributes to one estimator. It is not a request to treat
those run pairs as independent samples. `plan_geometry()` binds that pairing to
the condition axes, searchlight frame, metric, units, and exact neural domain.
It validates the request before reading neural values.

This is the central object to save in an audit trail. Compute policy may change
how the plan is executed, but it cannot change the plan's scientific identity.

## Find the reproducible contrast

```{r contrast}
effect <- contrast(plan, example$contrast)
peak <- which.max(effect$total)
c(
  peak_measurement = peak,
  signed_effect = effect$signed[peak],
  total_energy = effect$total[peak],
  coherent_energy = effect$coherent[peak],
  configuration_energy = effect$configuration[peak],
  overlaps_planted_signal = peak %in% example$truth$signal_measurements
)
```

One plan therefore reports the whole contrast family at once. `signed` is the
familiar signed contrast at each measurement — a retained first-moment
marginal, the number a GLM reader expects. The three energies are quadratic
quantities that decompose its reproducibility, so `coherent` is not the
signed univariate result: it is the crossvalidated energy carried by the
locally averaged contrast. `configuration` is the energy in reproducible
spatial departures from that weighted mean — orthogonal to the coherent mode
in the frame-weighted inner product, not the Euclidean one — and the package
preserves the exact identity `total = coherent + configuration`.

```{r contrast-check, include=FALSE}
stopifnot(
  peak %in% example$truth$signal_measurements,
  isTRUE(all.equal(
    effect$total, effect$coherent + effect$configuration, tolerance = 1e-14
  ))
)
```

Crossvalidated energy is unbiased around zero, so negative values are retained
rather than clipped. A negative estimate is not an invalid geometry.

The known truth makes that policy visible. Inside the affected searchlights the
planted multivariate pattern is carried mostly by configuration; outside them,
the estimate is centred near zero and therefore falls below zero about half the
time.

```{r planted-summary}
summarize_region <- function(index) c(
  total = mean(effect$total[index]),
  coherent = mean(effect$coherent[index]),
  configuration = mean(effect$configuration[index]),
  fraction_below_zero = mean(effect$total[index] < 0)
)
signal_nodes <- example$truth$signal_measurements
region_summary <- rbind(
  signal = summarize_region(signal_nodes),
  outside = summarize_region(setdiff(seq_along(effect$total), signal_nodes))
)
round(region_summary, 3)
```

```{r planted-summary-check, include=FALSE}
stopifnot(
  region_summary["signal", "configuration"] >
    region_summary["signal", "coherent"],
  abs(region_summary["outside", "total"]) < 0.01,
  region_summary["outside", "fraction_below_zero"] > 0.4,
  region_summary["outside", "fraction_below_zero"] < 0.6
)
```

```{r contrast-map, fig.cap="Cross-generalized contrast energy in the middle slice. The generated signal is centred in the volume; the map is an analysis check, not empirical evidence."}
volume <- array(NA_real_, dim = example$domain$metadata$dim)
volume[example$domain$metadata$mask] <- effect$total
middle <- ceiling(dim(volume)[3] / 2)
image(
  t(volume[, , middle]), axes = FALSE, asp = 1,
  col = hcl.colors(24, "Blue-Red 3", rev = TRUE),
  main = sprintf("Middle slice (z = %d)", middle)
)
box()
```

## Read distances and RSA from the same plan

For conditions `i` and `j`, `rdm()` applies the fixed contrast
`c = e_i - e_j` to the geometry:

\[
d_{ij} = c^T G c = G_{ii} + G_{jj} - 2G_{ij}.
\]

```{r rdm}
distances <- rdm(plan)
round(distances$values[peak, ], 3)
distances$pairs
```

Under cross-partition pairing these are signed crossvalidated squared
Euclidean distances. A plan with a fixed neural precision metric gives squared
Mahalanobis distances. Either can be negative near the null.

Linear RSA compiles the RDM transform and regression coefficient into a fixed
query against the same plan:

```{r rsa}
category_fit <- rsa(
  plan,
  models = list(category = example$model_rdm)
)
round(category_fit$coefficients[peak, ], 3)
```

```{r geometry-view-checks, include=FALSE}
stopifnot(
  identical(dim(distances$values), c(245L, 6L)),
  identical(dim(category_fit$coefficients), c(245L, 2L)),
  category_fit$coefficients[peak, "category"] > 0
)
```

Named RDM axes are aligned to the relation's exact condition identities, and a
rank-deficient design is rejected before geometry rows are read.

## Ask for uncertainty without inventing replicates

The six run pairs share run estimates and are not six independent sampling
units. Their empirical spread divided by the square root of six is not a valid
standard error.

Because `example_fmri_effects()` used `lm_relation_fit()`, the residual error
channel required by the admitted fixed-metric covariance model remains
available:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  plan,
  example$fit,
  target = "plugin",
  at = peak
)
distance_se <- sqrt(sampling_covariance(distance_covariance))
round(distance_se, 3)
```

```{r uncertainty-check, include=FALSE}
stopifnot(
  length(distance_se) == 6L,
  all(is.finite(distance_se)),
  all(distance_se > 0)
)
```

`target = "plugin"` evaluates the signal-dependent term at the partition-mean
pattern estimate. `target = "null"` evaluates it at the fixed zero null. The
package refuses to infer this choice.

The result is a within-searchlight covariance under an equal-partition,
fixed-metric, separable model. It is not a cross-location random-field model,
an automatic confidence interval, or group inference. The covariance is stored
as an exact factorized operator: its diagonal, selected entries, actions,
quadratic forms, or a fixed linear transport can be read without materializing
the full distance-by-distance matrix.

## Use your own data

The full raw-observation route is executable in
`vignette("from-observations", package = "effectagram")`. It binds observation
clocks, event and confound facts, a semantic condition-space effect map, an
observation model, portable design receipts, and the resulting fit before
entering the geometry layer.

If independent runs already contain condition-by-feature estimates, construct a
point relation directly:

```{r own-betas, eval=FALSE}
rel <- relation(
  list(run1 = beta_run1, run2 = beta_run2),
  effects = effect_space(condition_names, units = "percent-signal"),
  domain = native_domain
)
frame <- compile_frame(searchlights(radius = 6), native_domain)
plan <- plan_geometry(
  rel, frame, cross_partitions(rel, independence = "independent")
)
```

This is enough for `contrast()`, `rdm()`, and `rsa()`. Beta matrices do not
contain the residual information needed for within-participant analytic
uncertainty.

When raw observation-by-feature responses and an already compiled design are
available, the older direct fit remains the compact degenerate route:

```{r own-raw, eval=FALSE}
fit <- lm_relation_fit(
  response_runs,
  design = design_matrix,
  effects = target_matrix,
  effect_names = effect_space(condition_names, units = "percent-signal"),
  sampling_unit = "trial",
  domain = native_domain
)
```

The typed `plan_relation()` route can instead consume a semantic
`design_model()` or the exact-version `fmridesign_design_model()` adapter.
`effectagram` owns the identities, receipts, capabilities, and refusals; it
does not replace HRF grammars, general GLM engines, preprocessing, masks, or
registration. Those upstream choices remain explicit and identity-bearing.

## Why correlation distance is not an option to `rdm()`

Feature centring is a fixed linear projection and can be declared as a metric.
The unit-norm denominator in `1 - Pearson correlation` depends on the observed
patterns, so it is not a fixed linear query. Conventional correlation distance
can be a named nonlinear view of a positive-semidefinite self geometry, but
diagonal-normalizing a signed cross-generalized form is a different and not
automatically valid estimand.

The package therefore does not hide correlation normalization inside `rdm()`.
The accepted extension requirements and refusal cases are recorded in the
repository's `correlation-distance-policy.md`.

## Where to go next

- `?plan_geometry` describes rectangular cross-axis plans
  (`right = `), read with axis-bound `pair_query()`s, and `?coupling` takes
  the adjoint neural-side closure from the same plan vocabulary.
- The `evidence-pairing` vignette covers measurement forms and coupling
  views, including one bounded cross-domain contraction and the Parseval
  reconstruction law.
- `?rdm_sampling_covariance` states the admitted analytic covariance
  contract, and `?sampling_capabilities` answers the admission question
  before it is provoked.
- The repository's `failure-gallery.md` shows six realistic errors the
  package guards against through classed refusals, deliberately absent
  transformations, or distinct estimand identities.
- The Haxby 2001 exemplar supplies public-data parity and refusal evidence.
- `novelty.md` separates the package's architectural contribution from the
  established statistics it implements.

The key habit is simple: declare the relation, measurement, and generalization
requirement once; then ask fixed, named questions of the resulting plan.
