---
title: "Where does a representational effect generalize?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where does a representational effect generalize?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

Use this guide when you have condition estimates from repeated runs and want to
find where an effect reproduces across runs. The input is one
condition-by-feature matrix per run plus a spatial measurement frame. The
outputs are a contrast map, a representational dissimilarity matrix (RDM), and
a linear representational similarity analysis (RSA) coefficient at each
measurement. If the fitted relation retains the required residual information,
the same analysis can also return within-participant standard errors.

The worked question is: **where does an animate-versus-inanimate pattern
reproduce across independent runs?** The data are generated with known truth,
so the example runs after installation and can check its own answer. These
generated data are not empirical evidence; the separate Haxby 2001 exemplar is
the public-data comparison.

## Start from a ready relation

```{r example}
example <- example_fmri_effects()
c(
  conditions = length(example$fit$relation$effects),
  runs = length(example$fit$relation$partitions),
  features = example$fit$relation$n_features,
  searchlights = nrow(example$frame$index)
)
```

`example$fit` contains four run-specific estimates for four conditions in a
small 7 by 7 by 5 volume. `example$frame` defines the 245 searchlights at which
the analysis will report results. The generator returns the frame separately
because a fitted relation records the neural domain but does not own a mutable
searchlight definition.

The fit provides two kinds of information:

- `example$fit$relation` is the condition-by-feature relation used for point
  estimates; it reads feature blocks only when a query needs them;
- its error channel stores residual blocks, residual degrees of freedom, and
  effect-coordinate covariance for uncertainty calculations whose assumptions
  the fit can support.

## Declare what must generalize

```{r plan}
plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)
plan
```

`cross_partitions()` requests products between different runs. Every unordered
pair of runs contributes to the estimate, but the function does not treat the
six run pairs as six independent samples. `plan_geometry()` checks that the
relation, condition names, searchlight frame, metric, units, and neural domain
agree before it reads neural values.

Save `plan` with the analysis record. It identifies the estimand: the scientific
quantity the analysis is meant to estimate. Changing block size or storage
changes the execution receipt, not this plan identity.

## Find the reproducible contrast

```{r contrast}
effect <- contrast_energy(plan, example$contrast)
peak <- which.max(effect$total)
c(
  peak_measurement = peak,
  signed_effect = effect$signed[peak],
  total_energy = effect$total[peak],
  coherent_energy = effect$coherent[peak],
  configuration_energy = effect$configuration[peak],
  overlaps_planted_signal = peak %in% example$truth$signal_measurements
)
```

`contrast_energy()` returns four values at every searchlight. `signed` is the
usual signed contrast of the local weighted mean. The other three values
measure cross-run reproducibility:

- `coherent` is the crossvalidated energy of that local weighted mean;
- `configuration` is the energy in reproducible spatial departures from the
  weighted mean;
- `total` is their sum.

The coherent and configuration modes are orthogonal under the weights declared
by the frame. They need not be orthogonal under an unweighted Euclidean inner
product. The package checks `total = coherent + configuration` for every
measurement.

```{r contrast-check, include=FALSE}
stopifnot(
  peak %in% example$truth$signal_measurements,
  isTRUE(all.equal(
    effect$total, effect$coherent + effect$configuration, tolerance = 1e-14
  ))
)
```

Crossvalidated energy is centered on zero when no reproducible effect is
present. Sampling noise can therefore produce negative estimates. Clipping
them to zero would change the estimator and introduce positive bias, so
`contrast_energy()` retains them.

The known truth makes that policy visible. Inside the affected searchlights the
planted multivariate pattern is carried mostly by configuration; outside them,
the estimate is centered near zero and therefore falls below zero about half the
time.

```{r planted-summary}
summarize_region <- function(index) c(
  total = mean(effect$total[index]),
  coherent = mean(effect$coherent[index]),
  configuration = mean(effect$configuration[index]),
  fraction_below_zero = mean(effect$total[index] < 0)
)
signal_nodes <- example$truth$signal_measurements
region_summary <- rbind(
  signal = summarize_region(signal_nodes),
  outside = summarize_region(setdiff(seq_along(effect$total), signal_nodes))
)
round(region_summary, 3)
```

```{r planted-summary-check, include=FALSE}
stopifnot(
  region_summary["signal", "configuration"] >
    region_summary["signal", "coherent"],
  abs(region_summary["outside", "total"]) < 0.01,
  region_summary["outside", "fraction_below_zero"] > 0.4,
  region_summary["outside", "fraction_below_zero"] < 0.6
)
```

```{r contrast-map, echo=FALSE, fig.cap="Cross-generalized contrast energy in the middle slice. The generated signal is centered in the volume; the map checks the analysis and is not empirical evidence."}
volume <- array(NA_real_, dim = example$domain$metadata$dim)
volume[example$domain$metadata$mask] <- effect$total
middle <- ceiling(dim(volume)[3] / 2)
image(
  t(volume[, , middle]), axes = FALSE, asp = 1,
  col = hcl.colors(24, "Blue-Red 3", rev = TRUE),
  main = sprintf("Middle slice (z = %d)", middle)
)
box()
```

## Read distances and RSA from the same plan

For conditions `i` and `j`, `rdm()` applies the fixed contrast
`c = e_i - e_j` to the geometry:

\[
d_{ij} = c^T G c = G_{ii} + G_{jj} - 2G_{ij}.
\]

```{r rdm}
distances <- rdm(plan)
round(distances$values[peak, ], 3)
distances$pairs
```

With the current identity metric and cross-run pairing, these are signed,
crossvalidated squared Euclidean distances. Supplying a fixed neural precision
metric changes them to squared Mahalanobis distances. Either estimate can be
negative near the null because the two factors come from different runs.

Linear RSA compiles the RDM transform and regression coefficient into a fixed
query against the same plan:

```{r rsa}
category_fit <- rsa(
  plan,
  models = list(category = example$model_rdm)
)
round(category_fit$coefficients[peak, ], 3)
```

```{r geometry-view-checks, include=FALSE}
stopifnot(
  identical(dim(distances$values), c(245L, 6L)),
  identical(dim(category_fit$coefficients), c(245L, 2L)),
  category_fit$coefficients[peak, "category"] > 0
)
```

`rsa()` aligns the model's named rows and columns to the relation conditions.
It rejects a model with missing names or a rank-deficient regression design
before it reads geometry rows.

## Ask for uncertainty without inventing replicates

The six run pairs share run estimates and are not six independent sampling
units. Their empirical spread divided by the square root of six is not a valid
standard error.

`example_fmri_effects()` used `lm_relation_fit()`, so the fit retains the
residual error channel required by the package's fixed-metric covariance
model:

```{r uncertainty}
distance_covariance <- rdm_sampling_covariance(
  plan,
  example$fit,
  target = "plugin",
  at = peak
)
distance_se <- sqrt(sampling_covariance(distance_covariance))
round(distance_se, 3)
```

```{r uncertainty-check, include=FALSE}
stopifnot(
  length(distance_se) == 6L,
  all(is.finite(distance_se)),
  all(distance_se > 0)
)
```

`target = "plugin"` evaluates the signal-dependent variance term at the
estimated mean pattern. Use `target = "null"` to evaluate that term under a
fixed zero-effect null. The package requires this choice because the two calls
answer different variance questions.

The result is a within-searchlight covariance under an equal-partition,
fixed-metric, separable error model. It is not a cross-location random-field
model, a confidence interval, or group inference. The object stores a
factorized covariance operator. `sampling_covariance()` can read its diagonal,
selected entries, matrix action, quadratic form, or fixed linear transport
without constructing the full distance-by-distance matrix.

## Use your own data

If you have scan-by-feature responses rather than beta matrices, continue with
`vignette("from-observations", package = "crossform")`. That guide binds scan
times, events, confounds, named effects, and an observation model before it
fits the relation used here.

If independent runs already contain condition-by-feature estimates, construct a
point relation directly:

```{r own-betas, eval=FALSE}
rel <- relation(
  list(run1 = beta_run1, run2 = beta_run2),
  effects = effect_space(condition_names, units = "percent-signal"),
  domain = native_domain
)
frame <- compile_frame(searchlights(radius = 6), native_domain)
plan <- plan_geometry(
  rel, frame, cross_partitions(rel, independence = "independent")
)
```

This relation is enough for `contrast_energy()`, `rdm()`, and `rsa()`. It
cannot supply the analytic standard errors shown above because beta matrices
alone do not contain residuals or residual degrees of freedom.

When raw observation-by-feature responses and an already compiled design are
available, the older direct fit remains the compact degenerate route:

```{r own-raw, eval=FALSE}
fit <- lm_relation_fit(
  response_runs,
  design = design_matrix,
  effects = target_matrix,
  effect_names = effect_space(condition_names, units = "percent-signal"),
  sampling_unit = "trial",
  domain = native_domain
)
```

For a fully declared first-moment fit, `plan_relation()` accepts either a
semantic `design_model()` or the version-checked
`fmridesign_design_model()` adapter. Crossform records the requested effects,
design identity, fit receipt, and supported uncertainty operations. It does
not perform preprocessing, registration, masking, or general hemodynamic
response modeling.

## Why correlation distance is not an option to `rdm()`

Feature centering is a fixed linear projection, so a plan can declare it as a
metric. Pearson correlation also divides each observed pattern by its norm.
That denominator depends on the data and is therefore not a fixed linear query.
A conventional correlation distance can be defined from a positive-semidefinite
self geometry. Applying the same diagonal normalization to a signed
cross-generalized form does not automatically define a valid crossvalidated
estimator.

The package therefore does not hide correlation normalization inside `rdm()`.
The accepted extension requirements and refusal cases are recorded in the
[correlation-distance policy](https://bbuchsbaum.github.io/crossform/articles/correlation-distance-policy.html).

## Where to go next

- `?plan_geometry` describes rectangular cross-axis plans
  (`right = `), read with axis-bound `pair_query()`s, and `?coupling` takes
  the adjoint neural-side closure from the same plan vocabulary.
- The `evidence-pairing` vignette covers measurement forms and coupling
  views, including one bounded cross-domain contraction and the Parseval
  reconstruction law.
- `?rdm_sampling_covariance` states the admitted analytic covariance
  contract, and `?sampling_capabilities` answers the admission question
  before it is provoked.
- The [failure gallery](https://bbuchsbaum.github.io/crossform/articles/failure-gallery.html)
  shows six realistic errors the package guards against through classed
  refusals, deliberately absent transformations, or distinct estimand
  identities.
- The Haxby 2001 exemplar supplies public-data parity and refusal evidence.
- [What is novel in crossform?](https://bbuchsbaum.github.io/crossform/articles/novelty.html)
  separates the package's architectural contribution from the established
  statistics it implements.

For the usual workflow, construct one plan from the fitted relation, spatial
frame, and cross-run pairing. Then pass that plan to `contrast_energy()`,
`rdm()`, or `rsa()` according to the result you need.
