---
title: "From partitioned effects to interpretable geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From partitioned effects to interpretable geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(effectagram)
```

`effectagram` asks whether experimental effects reproduce across declared
partitions under a declared spatial measurement. In the common case, the same
effect coordinates appear on both sides and the durable result is a symmetric
geometry. More generally, an effect form can place different named coordinates
on its left and right axes, as in encoding-retrieval analyses.

Changing a searchlight to an ROI changes the measurement frame. Changing a
contrast to RSA changes the view. Neither choice introduces another fitting
engine. This vignette starts with the public self-geometry workflow, then shows
how unequal-axis pair hypotheses and cross-space bridges fit the same algebra.

This vignette starts from precomputed run effects. The final section uses a
small raw-response fit to show where residual uncertainty enters when responses
have not yet been reduced to effects.

## Declare the native relation

The rows below are experimental effects and the columns are native neural
features. Runs remain separate because agreement between independent estimates
defines the cross-generalized quantity.

```{r data}
effects <- effect_space(
  c("face", "house", "object"),
  basis_id = "condition-means:v1",
  units = "percent-signal"
)
domain <- abstract_domain(
  6,
  coordinates = cbind(x = 0:5, y = 0),
  feature_ids = paste0("feature", 1:6),
  id = "native:demo"
)

run1 <- rbind(
  face   = c(2, 1, 0.5, -0.5, 0, 1),
  house  = c(0, 1, 2, 1, 0.5, 0),
  object = c(0.5, 0.5, 1, 1.5, 1, 0.5)
)
run2 <- rbind(
  face   = c(2.2, 0.8, 0.4, -0.4, 0.1, 0.9),
  house  = c(0.1, 1.2, 1.8, 0.9, 0.6, 0.1),
  object = c(0.4, 0.6, 1.1, 1.4, 0.9, 0.6)
)

rel <- relation(list(run1 = run1, run2 = run2), effects = effects,
  domain = domain)
```

`relation()` records the ordered experimental basis, units and scaling,
partition identity, exact ordered neural domain, and source revisions. It does
not concatenate runs or resample features. A matching name and dimension alone
cannot substitute for either identity.

## What is an effect form?

An effect form is a matrix whose rows belong to one named experimental space
and whose columns belong to another. A geometry is the self-form specialization:
its two spaces are identical and the estimator is explicitly symmetric. That
specialization permits packed storage and the familiar RDM, RSA, and spectrum
views.

These are declared capabilities, not properties guessed from rounded output.
A rectangular form is not symmetric. A symmetric cross-generalized estimate is
not automatically positive semidefinite, so negative energies, distances, and
spectrum roots remain meaningful estimates rather than values to truncate.

## Measure locally and require cross-run agreement

```{r geometry}
at <- compile_frame(searchlights(radius = 1.01, normalization = "local"), domain)
over <- cross_partitions(rel)
plan <- plan_geometry(rel, at = at, over = over)
g <- geometry(plan)
```

The searchlights are rows of one sparse additive frame. `over` is an undirected
edge table containing each run pair once. `plan` validates this scientific
request without reading relation blocks. This section materializes it because
several later views need the complete form; the geometry engine then reads each
feature block once per partition, forms cross-run feature atoms, and contracts
the whole frame.

`g` retains complete packed total and coherent geometry. Configuration is
derived as `total - coherent`, so the decomposition cannot drift across reruns.

## Read level and configuration together

```{r contrast}
face_house <- contrast(g, c(face = 1, house = -1, object = 0))
round(cbind(
  signed = face_house$signed,
  coherent = face_house$coherent,
  configuration = face_house$configuration,
  total = face_house$total
), 3)
```

```{r contrast-check, include=FALSE}
stopifnot(isTRUE(all.equal(
  face_house$total,
  face_house$coherent + face_house$configuration,
  tolerance = 0
)))
```

The signed column is the locally averaged contrast. Coherent energy asks
whether that local average reproduces across runs. Configuration energy asks
whether spatial departures from the average reproduce. Total energy is their
exact sum. Configuration is not evidence that every single feature is
marginally silent, nor that a biological interaction exists.

Because cross-products are unbiased around zero, coherent, configuration, and
total estimates can be negative. The package preserves those values.

## Change the question, not the analysis

Squared experimental distances are a fixed linear transform of geometry:

```{r rdm}
distances <- rdm(g)
round(distances$values[1:2, ], 3)
distances$pairs
```

Multiple-regression RSA compiles that same distance transform together with a
fixed regression design:

```{r rsa}
category_rdm <- matrix(
  c(0, 0, 1,
    0, 0, 1,
    1, 1, 0),
  3, 3, byrow = TRUE,
  dimnames = list(rownames(run1), rownames(run1))
)
model_fit <- rsa(g, models = list(category = category_rdm))
round(model_fit$coefficients, 3)
model_fit$terms
```

Named RDM axes are aligned to the geometry's effect identities. Rank-deficient
RSA designs fail before any geometry rows are read.

The geometry spectrum is the nonlinear descriptive view in this workflow:

```{r spectrum}
roots <- geometry_spectrum(g)
round(roots$values[1:2, ], 3)
```

Negative roots are retained. A later positive-semidefinite latent model may be
useful for description, but truncating the unbiased estimate would change its
inferential meaning.

## Ask a question across unequal axes

Encoding-retrieval analyses have different row and column meanings. Here three
encoding conditions are compared with four retrieval conditions, including a
lure with no matched encoding item.

```{r unequal-axis-spaces}
encoding_space <- effect_space(
  c("word", "face", "place"), basis_id = "encoding:v1"
)
retrieval_space <- effect_space(
  c("word_old", "face_old", "place_old", "lure"),
  basis_id = "retrieval:v1"
)
```

The small matrices below stand in for two effect families measured in one
common five-feature space. Their ordered cross-form has shape `3 x 4`.

```{r unequal-axis-form}
encoding_profiles <- matrix(c(
  2, 1, 0, -1, 1,  0, 2, 1, 0, -1,  1, 0, 2, 1, 0
), 3, byrow = TRUE)
retrieval_profiles <- matrix(c(
  1.8, 1.1, 0.2, -0.8, 0.9,  0.1, 1.9, 1.2, 0.1, -0.8,
  0.9, 0.2, 2.1, 0.8, 0.1,  -0.5, 0.4, 0.2, 1.2, 1.1
), 4, byrow = TRUE)
cross_form <- encoding_profiles %*% t(retrieval_profiles)
```

`match_control()` compiles a matched-versus-control coefficient as one
axis-bound query matrix. Encoding-item and retrieval-item nuisance effects are
ordinary columns in that pair-space linear model.

```{r unequal-axis-query}
matches <- match_coupling(
  c("word", "face", "place"),
  c("word_old", "face_old", "place_old"),
  encoding_space, retrieval_space
)
er_query <- match_control(matches)
er_score <- sum(cross_form * as.matrix(er_query$operator))
round(er_score, 3)
```

```{r unequal-axis-check, include=FALSE}
balance <- er_query$metadata$diagnostics$balance
stopifnot(
  is.finite(er_score),
  balance$zero_row_marginals,
  balance$zero_column_marginals,
  isTRUE(all.equal(er_score, 4.3, tolerance = 1e-12))
)
```

Zero row and column marginals make this final query invariant to arbitrary
additive encoding and retrieval baselines. That guarantee belongs to the final
operator. Later unequal weighting can destroy the balance, so the constructor
reports observed diagnostics rather than promising downstream invariance.

## Keep operation order explicit

The contraction plan records the scientific order:

```text
ordered partition-pair product
  -> spatial normalization
  -> edge transform
  -> partition reduction
  -> pair query
```

Version 0.1 executes fixed inner-product forms and exposes reduction choices
only where a public workflow can consume them. `reduce_partitions()` and
`aggregate_first()` name different estimands; neither is selected as a hidden
optimization for the other. Generic correlation, ranking, and Fisher edge
pipelines are intentionally not exported until an executable,
scale-qualified path exists.

## Require a bridge between neural spaces

Relations with different neural feature identities cannot be multiplied just
because their dimensions happen to agree. A fixed factorized bridge maps both
sides into one named measurement space and records the exact source domains,
legs, and portable provenance.

```{r bridge}
left_neural <- abstract_domain(5, id = "subject-A:native")
right_neural <- abstract_domain(6, id = "subject-B:native")
common_measurement <- measurement_space(3, "atlas-measurement:v1")
bridge <- measurement_bridge(
  left_leg = matrix(seq_len(15) / 15, 3, 5),
  right_leg = matrix(seq_len(18) / 18, 3, 6),
  left_domain = left_neural,
  right_domain = right_neural,
  common_space = common_measurement,
  provenance = list(method = "fixed-external-map", revision = "v1")
)
```

```{r bridge-check, include=FALSE}
stopifnot(
  identical(dim(bridge$left_leg), c(3L, 5L)),
  identical(dim(bridge$right_leg), c(3L, 6L)),
  identical(reverse_bridge(reverse_bridge(bridge)), bridge)
)
```

The induced operator is the factorization
`t(left_leg) %*% right_leg`; an arbitrary dense operator is not a public
substitute. No explicit bridge is needed only when both relations carry the
same exact neural-domain identity. Bridges are fixed inputs here: learning an
alignment, hyperalignment, and group inference remain outside this package.

## Choose the frame interpretation explicitly

```{r frames}
point_frame <- compile_frame(voxels(), domain)
roi_frame <- compile_frame(
  regions(c("anterior", "anterior", "anterior",
            "posterior", "posterior", "posterior")),
  domain
)
global_frame <- compile_frame(whole_brain(), domain)
```

These values all enter the same `geometry()` call. With `normalization =
"local"`, each row is an average. On the feature-additive metric path, including
the current Euclidean default `K = I`, `normalization = "conservative"` makes
each feature distribute unit mass across overlapping rows and local total
geometry sums to global geometry. That conservation is a capability of the
composed frame and metric; it is not claimed for arbitrary non-diagonal local
metrics.

The exact composition is

\[
K_{x,\mathrm{effective}}
=
D(\sqrt{w_x})K_xD(\sqrt{w_x}).
\]

It recovers the additive frame exactly at `K_x = I` and preserves positive
semidefiniteness. A native-diagonal metric can therefore use the shared sparse
additive contraction. A non-diagonal metric remains part of the same evidence
pairing, but executes as a support-streamed pair contraction with an explicitly
larger cost.

## Avoid materializing unused geometry

Compile the reusable scientific plan before choosing whether to query or
materialize it. Plan construction validates sources, axes, the frame, pairing,
and compute policy without reading a neural block:

```{r direct}
plan <- plan_geometry(rel, at, over)
query <- bilinear_query(tcrossprod(c(1, -1, 0)), effects = effects)
direct <- evaluate_geometry(plan, query = query, component = "total")
late <- query_geometry(g, query, component = "total")
direct$completeness
```

```{r direct-check, include=FALSE}
stopifnot(isTRUE(all.equal(direct$values, late$values, tolerance = 1e-12)))
```

`evaluate_geometry()` fuses the fixed query before spatial contraction and
returns a query-only `effect_view`. `geometry(plan)` explicitly materializes
complete geometry. The compatibility call `geometry(rel, at, over)` compiles
the same plan; storage or workspace policy never changes the semantic
distinction.

## Supply an extractor when effects are still implicit

For responses `Y`, a fixed extractor `E` defines the lazy relation `B = E Y`.
`lm_extractor()` compiles an already constructed design and target map using QR,
or an estimability-checked SVD when the design is rank deficient.

```{r extractor, eval=FALSE}
extractor <- lm_extractor(
  design = X,
  effects = C,
  observation_whitener = L
)
rel <- relation(response_runs, extract = extractor, domain = domain)
```

If a later query will estimate a neural precision metric or calibrate
within-participant uncertainty, use `lm_relation_fit()` instead. It preserves
the same pure relation while attaching a distinct, capability-bearing error
channel with lazy whitened residual blocks, residual degrees of freedom, and
an unscaled effect-coordinate covariance. A relation built only from
precomputed effects remains valid but cannot recreate information discarded
upstream.

That distinction matters as soon as you ask for an error bar. The unordered
rows made by `cross_partitions()` are contributions to one crossvalidated
estimator, not independent replicates: pairs sharing a run share estimation
noise. Their empirical spread divided by the square root of the number of pairs
is therefore not a valid standard error.

For a plan using one common fixed neural metric, equal partition error
structures, and an `lm_relation_fit()` error channel, construct the exact local
RDM sampling covariance separately from the point estimate:

```{r uncertainty-setup, include=FALSE}
set.seed(19)
uncertainty_domain <- abstract_domain(
  6, coordinates = cbind(x = 0:5, y = 0), id = "uncertainty:demo"
)
uncertainty_design <- model.matrix(
  ~ 0 + factor(rep(c("face", "house", "object"), each = 8))
)
colnames(uncertainty_design) <- c("face", "house", "object")
uncertainty_effects <- diag(3)
rownames(uncertainty_effects) <- colnames(uncertainty_design)
true_patterns <- rbind(
  face = c(0.6, 0.3, 0.1, -0.1, 0.0, 0.2),
  house = c(0.0, 0.2, 0.6, 0.3, 0.1, 0.0),
  object = c(0.1, 0.1, 0.3, 0.5, 0.3, 0.1)
)
response_runs <- setNames(lapply(seq_len(4), function(run) {
  uncertainty_design %*% true_patterns +
    matrix(rnorm(24 * 6, sd = 0.25), 24, 6)
}), paste0("run", seq_len(4)))
fit <- lm_relation_fit(
  response_runs, uncertainty_design, uncertainty_effects,
  effect_names = rownames(true_patterns), sampling_unit = "trial",
  domain = uncertainty_domain
)
known_precision <- noise_precision(
  diag(6), uncertainty_domain, covariance = diag(6),
  provenance = list(source = "fixed-demo-metric")
)
```

```{r fixed-metric-sampling-covariance}
fixed_plan <- plan_geometry(
  fit$relation,
  at = compile_frame(whole_brain(), uncertainty_domain),
  over = cross_partitions(fit$relation),
  metric = known_precision
)

distance_estimate <- rdm(fixed_plan)
distance_covariance <- rdm_sampling_covariance(
  fixed_plan, fit, target = "plugin"
)
distance_se <- sqrt(sampling_covariance(distance_covariance))
round(distance_se, 4)
```

```{r fixed-metric-sampling-covariance-check, include=FALSE}
stopifnot(
  inherits(distance_covariance, "effect_rdm_sampling_covariance"),
  length(distance_se) == 3L,
  all(is.finite(distance_se)),
  all(distance_se > 0),
  identical(names(distance_se), colnames(distance_estimate$values))
)
```

`target = "plugin"` evaluates the signal-dependent term using the partition-
mean pattern estimate. `target = "null"` evaluates it at the fixed zero null.
Neither is selected silently. The covariance object is exact and factorized;
`sampling_covariance()` can read its diagonal, selected entries, actions,
quadratic forms, or linear transports without materializing the complete
distance-by-distance matrix.

This is a within-measurement covariance law under the fixed-metric,
equal-partition separable model. It is not a cross-location random-field model,
and it is not automatically a confidence interval or test. Learned local
precision metrics also remain outside this calibration claim because their
estimation uncertainty has not yet been propagated. These boundaries are why
the precomputed-effect example at the start of this vignette supports point
geometry but cannot be upgraded to analytic first-level uncertainty after the
residual channel has been discarded.

The package does not construct event designs, HRFs, nuisance models, temporal
whitening, or registration. Those choices remain upstream and explicit.

Function-backed sources must also provide `source_capabilities()` with a strong
immutable SHA-256 revision. Matrix sources receive this identity automatically.
The compiler validates the scientific objects and conservative workspace plan
before reading source blocks. A `workspace_bytes` limit covers only
effectagram-owned live workspace. Baseline, incremental peak, and absolute peak
process RSS remain separate observations.

If your native mask is a neuroim2 `NeuroVol`, use
`neuroim2_volume_domain()` and `neuroim2_searchlights()` to preserve its stable
full-volume feature indices and spacing. This optional adapter compiles geometry
only; neuroim2 ROI iteration, data extraction, results, and parallel state do
not enter effectagram.

Version 0.1 executes sequentially. The current performance claim is local and
precise: compiler-proven feature-additive searchlights are sparse contraction
rows rather than independent worker jobs. Non-diagonal local metrics require a
support-streamed lowering and do not receive that claim. Locally learned
covariance, adaptive feature selection, nonlinear normalization, classifier
accuracy, automatic pair-edge independence, and group-level inference remain
outside this bilinear core.
