---
title: "Equal total effects can have different spatial organization"
description: >
  A matched simulation comparing conventional summaries with coherent and configuration spectra across spatial scale.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equal total effects can have different spatial organization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE)
library(crossform)
artifact <- function(name) {
  candidates <- c(
    system.file("extdata", "certification", name, package = "crossform"),
    file.path("..", "inst", "extdata", "certification", name),
    file.path("inst", "extdata", "certification", name)
  )
  candidates <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(candidates)) stop("Missing certification artifact: ", name)
  candidates[[1L]]
}
figure_data <- utils::read.csv(
  artifact("matched-interpretability-figure-data.csv"),
  stringsAsFactors = FALSE
)
```

Read this example after [Reading contrast energies and RDMs](interpreting-results.html),
which introduces `total`, `coherent`, and `configuration`. Here *scale* means
the searchlight radius: widening a neighborhood changes how much of a
pattern looks spatially constant within it.

This simulation asks one narrow question: if total crossvalidated contrast
magnitude is fixed, can its spatial organization still differ? Three planted
patterns have equal total magnitude. One is a constant regional mean, one is a
50/50 mean-plus-alternating mixture, and one is a pure alternating
configuration. Every fitted comparison uses the same balanced design and the
same Gaussian noise draw within a replication.

The conventional activation comparator is the mean signed regional contrast.
The aggregate multivariate comparator is the average squared-Euclidean
cross-partition contrast magnitude over every unordered run pair. A comparator
is called ambiguous here only when its truth differs by at most `1e-12` while
the configuration-share spectra differ by at least `0.2` at some scale.

```{r canonical-figure, echo=FALSE, fig.cap="A: planted patterns. B: regional activation. C: aggregate crossvalidated magnitude. D-E: coherent and configuration magnitude at the widest radius, on a shared y-axis. F: coherent share across scale. Crosses mark analytic truth; points and lines show simulation means; whiskers and bands span the empirical 2.5th-97.5th percentiles.", fig.alt="Six panels compare broad coherent, mixed, and fine configuration effects. The total multivariate magnitude is equal, but coherent magnitude falls and configuration magnitude rises across scenarios; their coherence spectra separate with radius."}
figure_path <- knitr::fig_path("png")
dir.create(dirname(figure_path), recursive = TRUE, showWarnings = FALSE)
stopifnot(file.copy(
  artifact("matched-interpretability-figure.png"), figure_path,
  overwrite = TRUE
))
knitr::include_graphics(figure_path)
```

Read the panels in three steps. **A-C** establish that the planted patterns
differ while total magnitude is held fixed. **D-E** separate the common
spatial mode from its orthogonal remainder at the widest neighborhood.
**F** follows their balance as the neighborhood grows: a voxelwise pattern
is entirely coherent within its one-voxel neighborhood, whereas alternating
signs increasingly contribute to configuration in wider neighborhoods.

**Why do D and E sum to less than C?** Panel C reports the unscaled aggregate
magnitude. The four scales in D-F form a conservative frame family with
weight \(\alpha_s = 1/4\) per scale. Panels D and E therefore sum to one
quarter of C: analytic truth is 2.72 at the widest scale, compared with 10.88
in C. Summing `coherent + configuration` over all four scales recovers the
aggregate total. The coherent share in F divides within a scale, so this
common weight cancels.

Panel F also has a different availability rule: a signed crossvalidated
estimate need not admit a fraction between zero and one. Its mean and band
use the available fractions only, so their replication count can be below
24. They describe the admitted subset, rather than uncertainty over every
simulated dataset.

The regional activation panel is a negative control, not a failed comparator:
it correctly distinguishes a pure mean effect from a mixed effect. The
aggregate magnitude panel is the ambiguous one. Its truth is equal by
construction even though the two component panels and the spectrum differ.

```{r numeric-check, echo=FALSE}
aggregate <- figure_data[
  figure_data$metric == "aggregate_crossvalidated_magnitude",
  c("scenario_label", "truth", "estimate_mean", "interval_lower",
    "interval_upper", "n_replications")
]
knitr::kable(aggregate, digits = 4,
             col.names = c("Scenario", "Truth", "Mean", "2.5%", "97.5%",
                           "Replications"))
```

```{r normalization-check, include=FALSE}
components <- subset(figure_data, metric %in% c("coherent", "configuration"))
by_scale <- stats::aggregate(truth ~ scenario + scale, components, sum)
stopifnot(
  max(aggregate$truth) - min(aggregate$truth) < 1e-12,
  all(abs(aggregate$truth - 10.88) < 1e-12),
  all(abs(by_scale$truth - 10.88 / 4) < 1e-12)
)
```

This is matched-simulation evidence, not a real-data result. The figure does
not establish that a particular brain region is coherent or configurational,
and its intervals show Monte Carlo variation rather than population coverage.
The generator, thresholds, and evidence boundary are specified in
[`design/matched-interpretability-simulation-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/matched-interpretability-simulation-contract.md).


To apply this decomposition to your own fitted effects, continue with
[Conservative frames](conservative-frames.html). It shows how to
choose radii, declare scale weights, and read the components without counting
overlapping support repeatedly. The [neuroim2 workflow](neuroim2-data.html)
shows how to bring image estimates into a geometry plan.
