---
title: "Bring your own neuroim2 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring your own neuroim2 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = TRUE
)
has_neuroim2 <- requireNamespace("neuroim2", quietly = TRUE)
knitr::opts_chunk$set(eval = has_neuroim2)
library(crossform)
```

```{r no-neuroim2, eval=!has_neuroim2, echo=FALSE, results="asis"}
cat(
  "**neuroim2 is not installed, so the code below was not run.**",
  "Install it to reproduce this vignette:",
  "`remotes::install_github(\"bbuchsbaum/neuroim2\")`.\n"
)
```

You already have per-run condition estimates as
[neuroim2](https://github.com/bbuchsbaum/neuroim2) images and a brain mask.
This guide converts those objects into a cross-run contrast map and returns
it as a `NeuroVol`. The main steps are the orientation of the matrices `relation()` expects, which searchlight
provider to use, the difference between a *measurement* and a *feature*, and
how to expand region results back to voxels.

The `crossform` side is the same as in
`vignette("introduction", package = "crossform")`; only the input and output
adapters are new.

```{r libs}
library(neuroim2)
```

## A small example volume

Real data would be read with `neuroim2::read_vec()` and
`neuroim2::read_vol()`. So that this vignette is self-contained and fast, the
same objects are generated here: a 10-by-10-by-8 volume at 3 mm spacing with a
boxy mask, and one 4-D image holding 3 runs by 4 conditions with a
multivariate animacy pattern planted in a small central blob.

```{r data}
set.seed(11)
dims <- c(10L, 10L, 8L)
spacing <- c(3, 3, 3)
conditions <- c("face", "house", "cat", "chair")
n_run <- 3L
n_volume <- n_run * length(conditions)

volume_space <- NeuroSpace(dims, spacing = spacing)
mask_array <- array(0L, dims)
mask_array[2:9, 2:9, 2:7] <- 1L
mask <- NeuroVol(mask_array, volume_space)

# One planted multivariate pattern, present in every run.
blob <- array(0, dims)
blob[5:6, 5:6, 4:5] <- 1
pattern <- array(rnorm(prod(dims)), dims)
animacy <- c(face = 0.5, house = -0.5, cat = 0.5, chair = -0.5)

estimates <- array(rnorm(prod(dims) * n_volume), c(dims, n_volume))
k <- 0L
for (run in seq_len(n_run)) {
  for (condition in conditions) {
    k <- k + 1L
    estimates[, , , k] <- estimates[, , , k] +
      3 * animacy[[condition]] * blob * pattern
  }
}

runs_vec <- DenseNeuroVec(
  estimates,
  NeuroSpace(c(dims, n_volume), spacing = spacing)
)
runs_vec
```

The volumes are ordered run-major: volumes 1 to 4 are run 1's four conditions,
volumes 5 to 8 are run 2's, and so on. Keep that bookkeeping explicit —
`crossform` will check the condition names but cannot check that you sliced
the runs correctly.

## The domain fixes the voxel order

```{r domain}
domain <- neuroim2_volume_domain(mask)
c(
  features = domain$n_features,
  first_voxel_index = domain$feature_ids[1],
  last_voxel_index = domain$feature_ids[domain$n_features]
)
```

`neuroim2_volume_domain()` records the mask's **stable full-volume indices**
(`domain$feature_ids`), the physical coordinates, the spacing, and a hash of
the full `NeuroSpace`. Those indices are the contract between every later
step: they define which column of a response matrix is which voxel, and they
let `as_neurovol()` put results back without interpolation, smoothing, or any
reinterpretation of the coordinate frame.

```{r domain-ids}
identical(domain$feature_ids, which(mask_array != 0))
```

## Convert images to one condition-by-feature matrix per run

`relation()` wants a named list with one matrix per run, each
**condition-by-feature**: rows are the named conditions, columns are the
domain's features in `domain$feature_ids` order.

`neuroim2::series(x, i)` extracts the time series at a set of voxel indices
and returns a **volume-by-voxel** matrix. That is already the orientation
`crossform` wants, so no transpose is needed — just index the rows for one
run's volumes and index the columns with the domain's feature ids.

```{r convert}
volume_by_voxel <- series(runs_vec, domain$feature_ids)
run_matrices <- lapply(seq_len(n_run), function(run) {
  volumes <- ((run - 1L) * length(conditions) + 1L):(run * length(conditions))
  m <- volume_by_voxel[volumes, , drop = FALSE]
  rownames(m) <- conditions
  m
})
names(run_matrices) <- paste0("run", seq_len(n_run))

dim(run_matrices$run1)
rownames(run_matrices$run1)
```

If your estimates are a *list of `NeuroVol`s* rather than one `NeuroVec`, index
each volume at the same feature ids and stack the rows. `vol[ids]` returns the
values at those full-volume indices:

```{r convert-vols}
condition_vols <- lapply(seq_len(length(conditions)), function(k) {
  NeuroVol(estimates[, , , k], volume_space)
})

from_vols <- t(vapply(
  condition_vols,
  function(vol) as.numeric(vol[domain$feature_ids]),
  numeric(domain$n_features)
))
rownames(from_vols) <- conditions

isTRUE(all.equal(from_vols, run_matrices$run1))
```

Both routes go through `domain$feature_ids`, which is what makes them agree.
Do not use `as.array(vol)[mask_array != 0]` unless you are certain the mask
you are subsetting with is the same one that built the domain.

With the matrices in hand the relation is ordinary `crossform`:

```{r relation}
rel <- relation(
  run_matrices,
  effects = effect_space(conditions, units = "beta"),
  domain = domain
)
```

## Two searchlight providers, one support index

Use `neuroim2_searchlights()` when you need the same neighborhoods as other
neuroim2 tools. A built-in grid provider is also available:

```{r frames}
frame_neuroim2 <- neuroim2_searchlights(mask, radius = 6, domain = domain)
frame_grid <- compile_frame(searchlights(radius = 6), domain)

c(
  neuroim2_measurements = nrow(frame_neuroim2$index),
  grid_measurements = nrow(frame_grid$index)
)
```

- `neuroim2_searchlights(mask, radius)` delegates to
  `neuroim2::searchlight_indices()`. Use it when the same neighborhoods must
  be shared with other neuroim2-based tools: the sphere membership is
  literally the one that provider returns, and the frame records the upstream
  provider and commit in its specification.
- `compile_frame(searchlights(radius), domain)` uses `crossform`'s own grid
  stencil. It needs no neuroim2 at all, so it also works on a `volume_domain()`
  built from a plain array — which is what the rest of the documentation uses.

Both produce the same kind of object: an additive `effect_frame` carrying a
**support index** (which features each measurement reads) and a sparse
measurement-by-feature weight matrix. On a regular grid with a Euclidean ball
the providers agree for this example:

```{r frame-agreement}
identical(
  as.matrix(frame_neuroim2$weights),
  as.matrix(frame_grid$weights)
)
```

```{r frame-spec}
range(Matrix::rowSums(frame_neuroim2$weights != 0))
```

The radius is in millimeters for both, and both clip spheres at the mask
edge, so the sphere sizes vary. Use whichever provider matches the rest of
your pipeline; the plan below does not care which one produced the frame.

## Plan and read the contrast

```{r plan}
plan <- plan_geometry(
  rel,
  at = frame_neuroim2,
  over = cross_partitions(
    rel,
    independence = "independent",
    generalizes_over = "run"
  )
)

effect <- contrast_energy(plan, animacy)
peak <- which.max(effect$total)
c(
  peak_measurement = peak,
  peak_voxel_index = domain$feature_ids[peak],
  total_energy = effect$total[peak]
)
```

The planted blob is recovered. Searchlights whose support touches a planted
voxel carry most of the reproducible energy:

```{r recovery}
planted_features <- which(domain$feature_ids %in% which(blob != 0))
membership <- frame_neuroim2$weights != 0
touches_blob <- Matrix::rowSums(membership[, planted_features, drop = FALSE]) > 0

c(
  searchlights_touching_blob = sum(touches_blob),
  mean_energy_touching = mean(effect$total[touches_blob]),
  mean_energy_elsewhere = mean(effect$total[!touches_blob]),
  peak_touches_blob = unname(touches_blob[peak])
)
```

```{r recovery-check, include=FALSE}
stopifnot(
  touches_blob[peak],
  mean(effect$total[touches_blob]) > 10 * abs(mean(effect$total[!touches_blob]))
)
```

Away from the blob the mean energy is near zero. Individual crossvalidated
estimates can be negative; see `vignette("introduction", package = "crossform")` for why those
values are kept rather than truncated.

## Measurements are not features

This is the distinction that decides how you write results back.

- A **feature** is a voxel in the domain, addressed by `domain$feature_ids`.
  There are `domain$n_features` of them.
- A **measurement** is a row of the frame — one searchlight, one region, one
  whole-brain summary. Results (`effect$total`, `rdm()$values`,
  `rsa()$coefficients`) have one row per **measurement**.

`as_neurovol()` takes one finite value per **feature**. With searchlights the
two coincide: the frame has exactly one measurement per domain feature, in the
same order, because each sphere is labeled by its center voxel.

```{r measurement-vs-feature}
c(
  measurements = nrow(frame_neuroim2$index),
  features = domain$n_features,
  index_matches_features =
    identical(frame_neuroim2$index$measurement, domain$feature_ids)
)
```

So a searchlight map can be written straight back:

```{r write-back}
energy_vol <- as_neurovol(
  effect$total,
  mask,
  domain,
  label = "animacy total energy"
)
energy_vol
```

```{r write-back-check, include=FALSE}
# Values land at exactly the domain's voxels; everything else is `fill`.
stopifnot(
  isTRUE(all.equal(as.numeric(energy_vol[domain$feature_ids]), effect$total)),
  all(is.na(as.array(energy_vol)[mask_array == 0]))
)
```

`as_neurovol()` re-derives the domain from `mask` and refuses if the geometry
does not match the domain that produced the values, so a mask/result mismatch
is an error rather than a silently misplaced map. It also requires every value
to be finite: it will not guess what a missing measurement means.

## Regions from an atlas, and expanding them back

`regions()` takes one label per **feature** — a vector as long as
`domain$feature_ids`, not an image. Read an atlas `NeuroVol` at the domain's
voxel indices to get it:

```{r regions}
atlas_array <- array(0L, dims)
atlas_array[2:4, 2:9, 2:7] <- 1L
atlas_array[5:7, 2:9, 2:7] <- 2L
atlas_array[8:9, 2:9, 2:7] <- 3L
atlas_array[mask_array == 0L] <- 0L
atlas <- NeuroVol(atlas_array, volume_space)

region_labels <- as.integer(atlas[domain$feature_ids])
table(region_labels)

frame_regions <- compile_frame(regions(region_labels), domain)
frame_regions$index
```

```{r regions-run}
plan_regions <- plan_geometry(
  rel,
  at = frame_regions,
  over = cross_partitions(
    rel,
    independence = "independent",
    generalizes_over = "run"
  )
)
region_effect <- contrast_energy(plan_regions, animacy)
round(region_effect$total, 3)
```

Now there are only 3 measurements but still `domain$n_features` features, so
`as_neurovol(region_effect$total, ...)` would be an error. Expand the region
values to voxels first. For these disjoint regions, the frame's membership pattern assigns each
region value to its member voxels:

```{r regions-expand}
region_membership <- frame_regions$weights != 0
voxel_values <- as.numeric(
  Matrix::crossprod(region_membership, region_effect$total)
)

length(voxel_values)
region_vol <- as_neurovol(voxel_values, mask, domain, label = "region energy")
```

Use the membership pattern (`weights != 0`), not `weights` itself: under the
default `"local"` normalization the weights are `1/region_size`, so
`crossprod(weights, values)` would rescale the numbers you are trying to
display.

For disjoint labels the same thing is a lookup, which is easier to read:

```{r regions-expand-match}
by_lookup <- region_effect$total[
  match(region_labels, frame_regions$index$measurement)
]
isTRUE(all.equal(by_lookup, voxel_values))
```

For a partially labeled atlas, decide which voxels belong in the analysis
before building the domain. Numeric zero is a region label in `regions()`;
only missing or empty labels are excluded. If zero denotes background in your
atlas, intersect its nonzero coverage with the analysis mask and use that
mask throughout the workflow. Missing labels otherwise yield `NA` in the
lookup, which `as_neurovol()` does not accept.

The membership expansion above is for disjoint regions. For overlapping
measurements it sums their values at shared voxels, so it is not a general
method for reconstructing a voxel map.

## Tidy output

Every result view has an `as.data.frame()` method. The measurement index is
carried through, so for a searchlight frame the `measurement` column is the
center voxel's full-volume index:

```{r tidy}
head(as.data.frame(effect), 3)
```

```{r tidy-rdm}
distances <- rdm(plan)
head(as.data.frame(distances)[, 1:4], 3)
```

That frame joins directly to anything else keyed by voxel index, and
`as_neurovol()` remains available for the map itself.

## Where to go next

- `vignette("introduction", package = "crossform")` for what the energies,
  RDMs, and RSA coefficients mean.
- `vignette("from-observations", package = "crossform")` if you have
  scan-by-feature responses rather than per-run condition estimates, and want
  the residual channel that analytic standard errors require.
- `?neuroim2_volume_domain`, `?neuroim2_searchlights`, and `?as_neurovol` for
  the exact adapter contracts, including what each one refuses.
