---
title: "Bring your own neuroim2 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring your own neuroim2 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_neuroim2 <- requireNamespace("neuroim2", quietly = TRUE)
knitr::opts_chunk$set(eval = has_neuroim2)
library(crossform)
```

```{r no-neuroim2, eval=!has_neuroim2, echo=FALSE, results="asis"}
cat(
  "**neuroim2 is not installed, so the code below was not run.**",
  "Install it to reproduce this vignette:",
  "`remotes::install_github(\"bbuchsbaum/neuroim2\")`.\n"
)
```

You already have per-run condition estimates as
[neuroim2](https://github.com/bbuchsbaum/neuroim2) images and a brain mask.
This guide is the shortest correct path from those objects to a contrast map
written back as a `NeuroVol`. It covers four things that are easy to get
wrong: the orientation of the matrices `relation()` expects, which searchlight
provider to use, the difference between a *measurement* and a *feature*, and
how to expand region results back to voxels.

The `crossform` side is the same as in
`vignette("introduction", package = "crossform")`; only the input and output
adapters are new.

```{r libs}
library(neuroim2)
```

## A small example volume

Real data would be read with `neuroim2::read_vec()` and
`neuroim2::read_vol()`. So that this vignette is self-contained and fast, the
same objects are generated here: a 10-by-10-by-8 volume at 3 mm spacing with a
boxy mask, and one 4-D image holding 3 runs by 4 conditions with a
multivariate animacy pattern planted in a small central blob.

```{r data}
set.seed(11)
dims <- c(10L, 10L, 8L)
spacing <- c(3, 3, 3)
conditions <- c("face", "house", "cat", "chair")
n_run <- 3L
n_volume <- n_run * length(conditions)

volume_space <- NeuroSpace(dims, spacing = spacing)
mask_array <- array(0L, dims)
mask_array[2:9, 2:9, 2:7] <- 1L
mask <- NeuroVol(mask_array, volume_space)

# One planted multivariate pattern, present in every run.
blob <- array(0, dims)
blob[5:6, 5:6, 4:5] <- 1
pattern <- array(rnorm(prod(dims)), dims)
animacy <- c(face = 0.5, house = -0.5, cat = 0.5, chair = -0.5)

estimates <- array(rnorm(prod(dims) * n_volume), c(dims, n_volume))
k <- 0L
for (run in seq_len(n_run)) {
  for (condition in conditions) {
    k <- k + 1L
    estimates[, , , k] <- estimates[, , , k] +
      3 * animacy[[condition]] * blob * pattern
  }
}

runs_vec <- DenseNeuroVec(
  estimates,
  NeuroSpace(c(dims, n_volume), spacing = spacing)
)
runs_vec
```

The volumes are ordered run-major: volumes 1 to 4 are run 1's four conditions,
volumes 5 to 8 are run 2's, and so on. Keep that bookkeeping explicit —
`crossform` will check the condition names but cannot check that you sliced
the runs correctly.

## The domain fixes the voxel order

```{r domain}
domain <- neuroim2_volume_domain(mask)
c(
  features = domain$n_features,
  first_voxel_index = domain$feature_ids[1],
  last_voxel_index = domain$feature_ids[domain$n_features]
)
```

`neuroim2_volume_domain()` records the mask's **stable full-volume indices**
(`domain$feature_ids`), the physical coordinates, the spacing, and a hash of
the full `NeuroSpace`. Those indices are the contract between every later
step: they define which column of a response matrix is which voxel, and they
let `as_neurovol()` put results back without interpolation, smoothing, or any
reinterpretation of the coordinate frame.

```{r domain-ids}
identical(domain$feature_ids, which(mask_array != 0))
```

## Convert images to run-by-condition matrices

`relation()` wants a named list with one matrix per run, each
**condition-by-feature**: rows are the named conditions, columns are the
domain's features in `domain$feature_ids` order.

`neuroim2::series(x, i)` extracts the time series at a set of voxel indices
and returns a **volume-by-voxel** matrix. That is already the orientation
`crossform` wants, so no transpose is needed — just index the rows for one
run's volumes and index the columns with the domain's feature ids.

```{r convert}
run_matrices <- lapply(seq_len(n_run), function(run) {
  volumes <- ((run - 1L) * length(conditions) + 1L):(run * length(conditions))
  m <- series(runs_vec, domain$feature_ids)[volumes, , drop = FALSE]
  rownames(m) <- conditions
  m
})
names(run_matrices) <- paste0("run", seq_len(n_run))

dim(run_matrices$run1)
rownames(run_matrices$run1)
```

If your estimates are a *list of `NeuroVol`s* rather than one `NeuroVec`, index
each volume at the same feature ids and stack the rows. `vol[ids]` returns the
values at those full-volume indices:

```{r convert-vols}
condition_vols <- lapply(seq_len(length(conditions)), function(k) {
  NeuroVol(estimates[, , , k], volume_space)
})

from_vols <- t(vapply(
  condition_vols,
  function(vol) as.numeric(vol[domain$feature_ids]),
  numeric(domain$n_features)
))
rownames(from_vols) <- conditions

isTRUE(all.equal(from_vols, run_matrices$run1))
```

Both routes go through `domain$feature_ids`, which is what makes them agree.
Do not use `as.array(vol)[mask_array != 0]` unless you are certain the mask
you are subsetting with is the same one that built the domain.

With the matrices in hand the relation is ordinary `crossform`:

```{r relation}
rel <- relation(
  run_matrices,
  effects = effect_space(conditions, units = "beta"),
  domain = domain
)
```

## Two searchlight providers, one support index

There are two ways to get spheres over this domain, and they are not
redundant.

```{r frames}
frame_neuroim2 <- neuroim2_searchlights(mask, radius = 6, domain = domain)
frame_grid <- compile_frame(searchlights(radius = 6), domain)

c(
  neuroim2_measurements = nrow(frame_neuroim2$index),
  grid_measurements = nrow(frame_grid$index)
)
```

- `neuroim2_searchlights(mask, radius)` delegates to
  `neuroim2::searchlight_indices()`. Use it when the same neighborhoods must
  be shared with other neuroim2-based tools: the sphere membership is
  literally the one that provider returns, and the frame records the upstream
  provider and commit in its specification.
- `compile_frame(searchlights(radius), domain)` uses `crossform`'s own grid
  stencil. It needs no neuroim2 at all, so it also works on a `volume_domain()`
  built from a plain array — which is what the rest of the documentation uses.

Both produce the same kind of object: an additive `effect_frame` carrying a
**support index** (which features each measurement reads) and a sparse
measurement-by-feature weight matrix. On a regular grid with a Euclidean ball
they agree, and you can check it rather than trust it:

```{r frame-agreement}
identical(
  as.matrix(frame_neuroim2$weights),
  as.matrix(frame_grid$weights)
)
```

```{r frame-spec}
frame_neuroim2$specification
range(Matrix::rowSums(frame_neuroim2$weights != 0))
```

The radius is in millimeters for both, and both clip spheres at the mask
edge, so the sphere sizes vary. Use whichever provider matches the rest of
your pipeline; the plan below does not care which one produced the frame.

## Plan and read the contrast

```{r plan}
plan <- plan_geometry(
  rel,
  at = frame_neuroim2,
  over = cross_partitions(
    rel,
    independence = "independent",
    generalizes_over = "run"
  )
)

effect <- contrast_energy(plan, animacy)
peak <- which.max(effect$total)
c(
  peak_measurement = peak,
  peak_voxel_index = domain$feature_ids[peak],
  total_energy = effect$total[peak]
)
```

The planted blob is recovered. Searchlights whose support touches a planted
voxel carry most of the reproducible energy:

```{r recovery}
planted_features <- which(domain$feature_ids %in% which(blob != 0))
membership <- frame_neuroim2$weights != 0
touches_blob <- Matrix::rowSums(membership[, planted_features, drop = FALSE]) > 0

c(
  searchlights_touching_blob = sum(touches_blob),
  mean_energy_touching = mean(effect$total[touches_blob]),
  mean_energy_elsewhere = mean(effect$total[!touches_blob]),
  peak_touches_blob = unname(touches_blob[peak])
)
```

```{r recovery-check, include=FALSE}
stopifnot(
  touches_blob[peak],
  mean(effect$total[touches_blob]) > 10 * abs(mean(effect$total[!touches_blob]))
)
```

Away from the blob the estimate is centered near zero and is negative about as
often as not. That is the crossvalidated estimator behaving correctly, not a
bug; see `vignette("introduction", package = "crossform")` for why those
values are kept rather than truncated.

## Measurements are not features

This is the distinction that decides how you write results back.

- A **feature** is a voxel in the domain, addressed by `domain$feature_ids`.
  There are `domain$n_features` of them.
- A **measurement** is a row of the frame — one searchlight, one region, one
  whole-brain summary. Results (`effect$total`, `rdm()$values`,
  `rsa()$coefficients`) have one row per **measurement**.

`as_neurovol()` takes one finite value per **feature**. With searchlights the
two coincide: the frame has exactly one measurement per domain feature, in the
same order, because each sphere is labeled by its center voxel.

```{r measurement-vs-feature}
c(
  measurements = nrow(frame_neuroim2$index),
  features = domain$n_features,
  index_matches_features =
    identical(frame_neuroim2$index$measurement, domain$feature_ids)
)
```

So a searchlight map can be written straight back:

```{r write-back}
energy_vol <- as_neurovol(
  effect$total,
  mask,
  domain,
  label = "animacy total energy"
)
energy_vol

# Values land at exactly the domain's voxels; everything else is `fill`.
isTRUE(all.equal(as.numeric(energy_vol[domain$feature_ids]), effect$total))
all(is.na(as.array(energy_vol)[mask_array == 0]))
```

`as_neurovol()` re-derives the domain from `mask` and refuses if the geometry
does not match the domain that produced the values, so a mask/result mismatch
is an error rather than a silently misplaced map. It also requires every value
to be finite: it will not guess what a missing measurement means.

## Regions from an atlas, and expanding them back

`regions()` takes one label per **feature** — a vector as long as
`domain$feature_ids`, not an image. Read an atlas `NeuroVol` at the domain's
voxel indices to get it:

```{r regions}
atlas_array <- array(0L, dims)
atlas_array[2:4, 2:9, 2:7] <- 1L
atlas_array[5:7, 2:9, 2:7] <- 2L
atlas_array[8:9, 2:9, 2:7] <- 3L
atlas_array[mask_array == 0L] <- 0L
atlas <- NeuroVol(atlas_array, volume_space)

region_labels <- as.integer(atlas[domain$feature_ids])
table(region_labels)

frame_regions <- compile_frame(regions(region_labels), domain)
frame_regions$index
```

```{r regions-run}
plan_regions <- plan_geometry(
  rel,
  at = frame_regions,
  over = cross_partitions(
    rel,
    independence = "independent",
    generalizes_over = "run"
  )
)
region_effect <- contrast_energy(plan_regions, animacy)
round(region_effect$total, 3)
```

Now there are only 3 measurements but still `domain$n_features` features, so
`as_neurovol(region_effect$total, ...)` would be an error. Expand the region
values to voxels first. The general route uses the frame's own membership
pattern, which works for any additive frame:

```{r regions-expand}
region_membership <- frame_regions$weights != 0
voxel_values <- as.numeric(
  Matrix::crossprod(region_membership, region_effect$total)
)

length(voxel_values)
region_vol <- as_neurovol(voxel_values, mask, domain, label = "region energy")
```

Use the membership pattern (`weights != 0`), not `weights` itself: under the
default `"local"` normalization the weights are `1/region_size`, so
`crossprod(weights, values)` would rescale the numbers you are trying to
display.

For disjoint labels the same thing is a lookup, which is easier to read:

```{r regions-expand-match}
by_lookup <- region_effect$total[
  match(region_labels, frame_regions$index$measurement)
]
isTRUE(all.equal(by_lookup, voxel_values))
```

One caveat. If your atlas does not cover the whole mask, `regions()` drops the
unlabeled features, and the lookup above yields `NA` there — which
`as_neurovol()` refuses. Build the domain from the atlas coverage
(`neuroim2_volume_domain(NeuroVol((atlas_array != 0) * 1, volume_space))`) so
that domain and results describe the same voxels, rather than filling missing
regions with a number that will later be read as data.

## Tidy output

Every result view has an `as.data.frame()` method. The measurement index is
carried through, so for a searchlight frame the `measurement` column is the
center voxel's full-volume index:

```{r tidy}
head(as.data.frame(effect), 3)
```

```{r tidy-rdm}
distances <- rdm(plan)
head(as.data.frame(distances)[, 1:4], 3)
```

That frame joins directly to anything else keyed by voxel index, and
`as_neurovol()` remains available for the map itself.

## Where to go next

- `vignette("introduction", package = "crossform")` for what the energies,
  RDMs, and RSA coefficients mean.
- `vignette("from-observations", package = "crossform")` if you have
  scan-by-feature responses rather than per-run condition estimates, and want
  the residual channel that analytic standard errors require.
- `?neuroim2_volume_domain`, `?neuroim2_searchlights`, and `?as_neurovol` for
  the exact adapter contracts, including what each one refuses.
