---
title: "Population form: one participant's ledger, carried to a group"
description: >
  Experimental. A group analysis of conservative geometry is three typed objects — a per-participant plan, a transport you supply, and a group model — and one theorem: the query, the transport and the group fit act on different axes, so modelling then querying is the same number as querying then modelling. This article builds the objects on a generated six-participant fixture, asserts the commutation and the conservation identity, finds a planted odd participant with the cross-fitted subject Gram, and reports the two error bars the layer keeps apart — separately, and labelled uncalibrated.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population form: one participant's ledger, carried to a group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

**This layer is experimental.** The API and the returned kinds may change, and
nothing in it is calibrated group inference. What it *is* is a typed way to
carry each participant's conservative attribution ledger onto a shared set of
group nodes, fit one group model there, and read the result without any of the
three steps quietly changing what was estimated.

Read `vignette("conservative-frames")` first. Everything here operates on the
attribution instrument from that article: a map whose values are shares of one
fixed budget, and which therefore *adds*. Adding is what makes a group-level
ledger definable at all, and it is why this layer refuses a detection map.

Every identity below is asserted in code you can see. If one stopped holding,
this article would stop knitting. The governing document is
[`design/population-form-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/population-form-contract.md)
(`population-form-v1`), and section numbers refer to it.

## 1. The three objects

### 1.1 Six participants, six conservative geometry plans

The fixture is generated, tiny, and deterministic. Six participants on
**different native frame sizes** — that is the case §3 says forces transport to
precede the fit, and a fixture where everyone shared a frame would never
exercise it. Four runs each, because the cross-fitted subject Gram of section 4
needs two disjoint halves and each half needs a cross-generalized estimate of
its own.

Every participant carries the same planted **consensus** direction in effect
space (face above house, present in every run, so it survives the
cross-partition product). One participant, `s06`, additionally carries a
direction nobody else has. Section 4 is about finding it without being told.

```{r fixture}
pop_effects <- effect_space(c("face", "house", "tool"),
  basis_id = "population-vignette:v1")

pop_subject <- function(id, features, gain = 1, tilt = 0, runs = 4L,
                        normalization = "conservative") {
  domain <- abstract_domain(features,
    coordinates = cbind(x = seq_len(features) - 1),
    feature_ids = paste0("f", seq_len(features)), id = id)
  consensus <- outer(c(0.9, -0.9, 0), rep(1, features))
  odd <- outer(c(0, 0.8, -0.8), rep(1, features))
  block <- function(k) {
    set.seed(1000L * k + sum(as.integer(charToRaw(id))) + features)
    values <- matrix(gain * stats::rnorm(3L * features), 3L, features,
      dimnames = list(c("face", "house", "tool"), NULL))
    values + consensus + tilt * odd
  }
  relation <- relation(
    stats::setNames(lapply(seq_len(runs), block), paste0("run", seq_len(runs))),
    effects = pop_effects, domain = domain
  )
  plan_geometry(relation,
    compile_frame(voxelwise(normalization = normalization), domain),
    cross_partitions(relation))
}

sizes <- c(s01 = 9L, s02 = 11L, s03 = 13L, s04 = 10L, s05 = 12L, s06 = 14L)
gains <- c(s01 = 1, s02 = 1.3, s03 = 0.8, s04 = 1.1, s05 = 0.9, s06 = 1.2)
tilts <- c(s01 = 0, s02 = 0, s03 = 0, s04 = 0, s05 = 0, s06 = 1.6)

subjects <- stats::setNames(lapply(names(sizes), function(id)
  pop_subject(id, sizes[[id]], gains[[id]], tilts[[id]])), names(sizes))
sizes
```

These are ordinary single-participant objects: `plan_geometry()` sealed six
estimands and read no data. `voxelwise()` is conservative by default, which is
the gate section 1.3 enforces.

### 1.2 A transport is an input, and it is typed

A **transport** is the operator that carries one participant's native nodes onto
the shared group nodes. `crossform` accepts one and refuses to learn one:
image registration and functional-transport fitting stay outside the package
(§9.2). What it does is check the operator's shape, record its provenance, and
put both into the estimand's identity — two transports are two estimands (§1.5).

The cheapest typed transport is built from coordinates. Group centres at
`x = 0, 5, 11`, a radius of 2, and each native node assigned to its nearest
centre within that radius:

```{r anatomical}
pop_carrier <- function(features, semantics = "budget", radius = 2) {
  anatomical_transport(
    native_coords = cbind(seq_len(features) - 1),
    group_coords = cbind(c(0, 5, 11)),
    semantics = semantics, radius = radius,
    native_index = paste0("f", seq_len(features))
  )
}
transports <- stats::setNames(lapply(names(sizes), function(id)
  pop_carrier(sizes[[id]])), names(sizes))
transports$s01
```

Three things in that print are load-bearing.

**The sink is a column, always.** The operator is `n × (m + 1)`, not `n × m`:
group nodes plus one accounting column for native mass that reached no group
node. It is materialized even when it is empty, so partial coverage is a number
you read rather than budget that quietly went missing.

**`semantics` has no default.** `"budget"` and `"density"` are two different
estimands, and the constructor makes you name which.

**Provenance is required.** `method` is a closed set, and `details` must say how
the operator was built.

The same laws hold for an operator you computed elsewhere and are declaring.
Here is a four-node example small enough to check by hand: `f2` splits 0.6/0.4,
`f3` leaves 30 % of its territory unmapped, `f4` is unmapped entirely.

```{r external}
P <- rbind(
  f1 = c(anterior = 1,   posterior = 0),
  f2 = c(anterior = 0.6, posterior = 0.4),
  f3 = c(anterior = 0,   posterior = 0.7),
  f4 = c(anterior = 0,   posterior = 0)
)
carrier <- external_transport(P, semantics = "budget",
  provenance = list(details = "partial-volume warp, atlas-tool 2.1"))
c(native = nrow(carrier$matrix), columns = ncol(carrier$matrix))
stopifnot(
  identical(dim(carrier$matrix), c(4L, 3L)),
  identical(nrow(carrier$group_index), 2L),
  max(abs(Matrix::rowSums(carrier$matrix) - 1)) < 1e-12
)
```

Now carry a **signed** ledger through it — signed because a crossvalidated
conservative ledger is signed by construction, and a fixture of nonnegative
numbers would let a mass-law bug pass.

```{r budget-semantics}
ledger <- c(f1 = 1.5, f2 = -0.5, f3 = 2, f4 = 0.25)
carried <- transport_values(carrier, ledger)
round(carried, 4)

c(
  closes = sum(carried) - sum(ledger),
  without_sink = sum(carried[c("anterior", "posterior")]) - sum(ledger)
)
stopifnot(
  abs(sum(carried) - sum(ledger)) <= 1e-12 * sum(abs(ledger)),
  abs(sum(carried[c("anterior", "posterior")]) - sum(ledger)) > 0.8
)
```

Budget semantics preserve the total **exactly, including the sink** — and lose
0.85 units of signed mass the moment the sink is deleted. That is why it is
required rather than optional.

The same operator read as a **density** is a different estimand: each group
column is divided by the row mass that reached it, so the answer is a value per
unit of declared territory and satisfies no conservation law at all.

```{r density-semantics}
dense <- external_transport(P, semantics = "density",
  provenance = list(details = "the same operator, read as a density"))
carried_density <- transport_values(dense, ledger)
round(carried_density, 4)
c(density_gap = sum(carried_density) - sum(ledger))
stopifnot(abs(sum(carried_density) - sum(ledger)) > 1e-6)
```

That gap is arithmetic, not a discovery. Density trades conservation away; what
it keeps is section 2's commutation, because a declared row-mass ratio is still
a fixed linear map.

**A transport fitted to the response data is refused unless it says which
partitions built it.** An operator that saw the same runs the geometry is
estimated from is circular, and the contract measures what that circularity buys
(§7): the refusal is a gate, not a formality.

```{r cross-fit-refusal}
refusal <- catch_refusal(external_transport(P, semantics = "budget",
  provenance = list(method = "functional",
    details = "hyperalignment on the analysis runs")))
c(capability = refusal$capability, reason = refusal$reasons)
stopifnot(
  inherits(refusal, "effect_capability_refusal"),
  identical(refusal$capability, "cross_fit_provenance"),
  identical(refusal$namespace, "location_transport"),
  identical(refusal$reasons, "cross_fit_partitions_not_declared")
)

honest <- external_transport(P, semantics = "budget",
  provenance = list(method = "functional", details = "hyperalignment",
    cross_fit = c("run1", "run2")))
honest$provenance$cross_fit
stopifnot(identical(honest$provenance$cross_fit, c("run1", "run2")))
```

Naming the partitions is what the refusal asks for, and the names travel with
the operator into the estimand's identity. An anatomical or external transport
never saw the responses, so it owes no such record — which is why the six
carriers above passed without one.

### 1.3 `plan_population()`: the group estimand

```{r plan}
plan <- plan_population(subjects, transports)
plan
```

```{r plan-fields}
c(
  semantics = plan$semantics,
  normalization = plan$normalization,
  order = plan$fit$evaluation_order
)
plan$subject_index[, c("subject", "measurements", "declared_normalization",
  "conserved", "sink_territory")]
stopifnot(
  all(plan$subject_index$conserved),
  identical(plan$semantics, "budget"),
  identical(plan$normalization, "none"),
  identical(plan$fit$evaluation_order, "transport_then_fit"),
  identical(plan$fit$commuting, TRUE),
  identical(nrow(plan$group_index), 3L)
)
```

The evaluation order is **recorded, not chosen**. Participants have different
native frames, so there is no common node axis to fit at before transporting:
fit-then-transport is not definable here at all. Section 2 is about the orders
that *are* definable.

The conservative gate is the plan's headline refusal. A participant whose frame
reports a density has no budget to partition, and the plan says so by name:

```{r conservative-gate}
loose <- subjects
loose$s01 <- pop_subject("s01", sizes[["s01"]], gains[["s01"]],
  normalization = "local")
gate <- catch_refusal(plan_population(loose, transports))
c(capability = gate$capability, reason = gate$reasons[[1L]])
stopifnot(
  identical(gate$capability, "conservative_subject_geometry"),
  "normalization_not_conservative:s01:local" %in% gate$reasons,
  any(grepl("allow_nonconservative", gate$remedies))
)
```

The escape hatch exists, is named in the remedy, and is *recorded* — it enters
the plan's scientific identity, because a non-conservative population estimand
is a different estimand and not a relaxed setting on the same one.

The group model is the last piece. It is one-sided, its rows bind to
participants **by name** rather than by position, and it is factorized once on
the plan:

```{r group-model}
covariates <- data.frame(
  age = c(24, 31, 27, 44, 38, 22),
  row.names = c("s02", "s01", "s04", "s03", "s06", "s05")
)
aged <- plan_population(subjects, transports, model = ~ age, data = covariates)
aged$model$matrix
stopifnot(
  identical(aged$model$columns, c("(Intercept)", "age")),
  identical(aged$model$rank, 2L),
  identical(rownames(aged$model$matrix), names(sizes)),
  # The shuffled `data` rows followed their participant, not their position.
  identical(unname(aged$model$matrix[, "age"]), c(31, 24, 44, 27, 22, 38)),
  # A different group model is a different estimand.
  !identical(plan$scientific_plan_id, aged$scientific_plan_id)
)
```

The rest of this article uses the intercept-only `plan`, whose single term is
the group mean.

## 2. Model then query is query then model

This is the claim the layer rests on (§3). The query mixes **experimental**
coordinates, the transport mixes **spatial** nodes, and the group fit mixes
**participants**. Three different axes of the same array, so under the OLS
default with subject-constant weights they commute — and the answer does not
depend on the order you ran them in.

`estimate_population()` runs `query → transport → fit`: it contracts each
participant's geometry against the bank first and carries **one number per node
per query**.

```{r estimate}
bank <- rbind(`face-house` = c(1, -1, 0), `house-tool` = c(0, 1, -1))
fit <- estimate_population(plan, bank)
dim(fit$coefficients)
dimnames(fit$coefficients)[c("query", "term")]
fit$index
```

### The check that is not a check

Before the real one, the trap. Comparing `transport_values()` applied to a
participant's own `contrast_energy()` against `fit$values` looks like a
commutation test and is not: `contrast_energy()` lowers to exactly the packed
operator the bank carries, so both sides are the *same call on the same
numbers*. It is worth recording as what it is — evidence that the executor's
plumbing does not rescale or reorder rows — and it is exactly zero, which is the
tell.

```{r plumbing}
one <- estimate_population(plan, rbind(`face-house` = c(1, -1, 0)))
plumbing <- max(vapply(names(plan$subjects), function(id) {
  native <- contrast_energy(plan$subjects[[id]], c(1, -1, 0))$total
  max(abs(transport_values(plan$transport[[id]], native) -
    one$values[, "face-house", id]))
}, numeric(1)))
c(executor_plumbing = plumbing)
stopifnot(plumbing == 0)
```

### The real order reversal

`materialize_population()` runs the *other* order. It carries all six packed
coordinates of the complete `3 × 3` form to the group nodes, fits the group
model at every one of them, and hands back a coefficient **form** per node. The
`rdm()` view then contracts the (face, house) edge out of those coefficients —
**afterwards**. That is `transport → fit → query`.

The face − house contrast *is* that RDM edge. Two genuinely different sequences
of arithmetic, one number.

```{r commutation}
form <- materialize_population(plan)
edges <- as.data.frame(rdm(form))
edge <- edges[edges$left == "face" & edges$right == "house", ]
node_ids <- as.character(fit$index$node)

# One row per node: with a richer group model `rdm()` emits one row per node
# per term, and `match()` would silently take the first.
stopifnot(nrow(edge) == length(node_ids), !anyDuplicated(edge$node))

commutation <- max(abs(
  edge$estimate[match(node_ids, edge$node)] -
    fit$coefficients[node_ids, "face-house", "(Intercept)"]
))
c(query_first = fit$basis, transport_first = form$basis)
c(commutation = commutation)
stopifnot(
  identical(fit$basis, "query_bank"),
  identical(form$basis, "complete_form"),
  commutation < 1e-12
)
```

`1e-12` is the tolerance §11 states for the commutation of query, transport and
OLS under subject-constant weights. Unlike the plumbing check this one lands at
a rounding scale rather than at exactly zero, because the two routes really are
different arithmetic: one contraction before transport against six after it.

This is what licenses the reader verbs. A population view is a fixed linear
combination of estimated query columns, not a second execution — so
`contrast_energy()`, `rdm()`, `rsa()` and `contribution()` on a result reopen no
participant's geometry, and the combination is exact rather than approximate.

Reading the query first is also what makes the route cheap: complete geometry at
a native node is `q(q+1)/2` packed coordinates, a bank of `K` queries is `K`
numbers, and complete geometry is never allocated.

## 3. Conservation survives the transport, sink included

Section 2 of the contract is the certificate: under budget semantics each
participant's transported total, sink included, equals its native total exactly.
`estimate_population()` asserts it per participant and per query at fit time,
with a tolerance of `1e-12` times the ledger's **L1 norm** — not its total,
which a signed ledger can drive near zero.

```{r conservation}
native_total <- fit$receipt$native_total
carried_total <- apply(fit$values, c("subject", "query"), sum)
l1 <- apply(abs(fit$values), c("subject", "query"), sum)
budget_deviation <- max(abs(carried_total - native_total) / l1)

fit$receipt$budget$asserted
c(
  recorded = fit$receipt$budget$max_relative_deviation,
  recomputed = budget_deviation
)
stopifnot(
  isTRUE(fit$receipt$budget$asserted),
  identical(fit$receipt$budget$scale, "relative_to_ledger_l1_norm"),
  fit$receipt$budget$max_relative_deviation < 1e-12,
  budget_deviation < 1e-12
)
```

The sink is not a discarded remainder. Drop it and the identity breaks, by an
amount that is a real number about a real failure of coverage:

```{r sink}
group_only <- apply(fit$values[!fit$index$sink, , , drop = FALSE],
  c("subject", "query"), sum)
without_sink <- max(abs(group_only - native_total) / l1)
c(
  with_sink = budget_deviation,
  without_sink = without_sink,
  sink_territory = max(plan$subject_index$sink_territory)
)
stopifnot(without_sink > 1e-3, all(plan$subject_index$sink_territory > 0))
```

Between 7 % and 11 % of each participant's native territory here falls outside
every group node's radius. The sink is also **fitted like any other row** — its
coefficients are estimated rather than dropped — because covariate-linked sink
mass is how a differentially failing transport becomes visible (§3.3). A
transport that works worse in older participants shows up as an age effect in
the sink, and only if you left the sink in the model.

Note the units discipline in `fit$index`: the sink is reported in **budget**
units under either semantics, so a density population still has one honest
column of unmapped mass.

```{r units}
fit$index$units
stopifnot(all(fit$index$units == "budget"))
```

## 4. Heterogeneity: what the participants disagree about

The group fit gives the consensus. The other half of the section 6 decomposition
is the scatter: the `N × N` Gram of participants' deviations from the group fit,
in the packed geometry coordinates. Geometry-space covariance across `N`
participants has rank at most `N − 1`, so the whole spectrum lives in that small
matrix and the `D × D` covariance never has to be formed.

```{r heterogeneity}
het <- heterogeneity(plan, estimator = "cross_fit")
c(estimator = het$estimator, space = het$space)
c(residual_df = het$residual_df)
round(het$spectrum, 3)
stopifnot(
  identical(het$estimator, "cross_fit"),
  identical(het$space, "packed_form"),
  identical(het$residual_df, 5L)
)
```

**The Gram is indefinite, and that is reported rather than repaired.** The
cross-fitted estimator subtracts a cross-participant term to remove the
consensus-squared part, and nothing constrains the difference of two estimates
to be positive semidefinite. The contract measures it in 100 % of 2 000
replications (§6.2); here it shows up as two genuinely negative modes.

```{r indefinite}
c(
  negative_modes = het$latent$negative_modes,
  moved_share = round(het$latent$moved_share, 4),
  n_eff = round(het$latent$n_eff, 3)
)
stopifnot(
  sum(het$spectrum < -1e-8) == 2L,
  identical(het$latent$negative_modes, 2L),
  het$latent$moved_share > 0
)
```

The effective mode count and the cumulative curve are legal only on the declared
PSD projection, which records the mass it moved — the same latent-layer
discipline `latent_geometry()` applies to a single participant's map.

### The planted participant, found

The loadings are the eigenvectors of that Gram: one number per participant per
mode. Nothing above told the estimator that `s06` is different.

```{r loadings}
round(het$loadings[, 1:2], 3)
leading <- het$loadings[, 1L]
others <- leading[names(leading) != "s06"]
names(which.max(abs(leading)))
c(
  separation = round(max(abs(leading)) / max(abs(others)), 2),
  mode1_share = round(het$latent$cumulative[[1L]], 3)
)
stopifnot(
  identical(names(which.max(abs(leading))), "s06"),
  max(abs(leading)) > 4 * max(abs(others)),
  het$latent$cumulative[[1L]] > 0.85
)
```

Mode 1 is one participant against the rest — `s06` at 0.91 with everybody else
between −0.15 and −0.22, the signature of a single odd participant rather than a
gradient — and it holds over 90 % of the projected heterogeneity mass. Pass
`nodes =` to reconstruct the *geometry* of a mode at named group nodes and see
which effect pair it lives in.

### Why the estimator is not a free choice

The plug-in Gram books every participant's within-subject sampling noise as
between-participant heterogeneity. The contract measures the consequence: over
2 000 Monte Carlo replications at `N = 12`, the plug-in inflates the
heterogeneity trace by **+62.7 %** (mean `tr Q^H` of `+17.61` against a true
`+10.82`, predicted bias `+6.72` versus measured `+6.79`), while the
cross-fitted estimator's bias is within one Monte Carlo standard error of zero
(§6.2). That number is a property of the estimator, not of this fixture; one
draw of a trace estimates nothing. What *this* fixture can show is the
direction, and that the two Grams disagree about size while agreeing about
direction.

```{r plug-in}
plug <- heterogeneity(plan, estimator = "plug_in")
c(
  cross_fit_trace = round(sum(diag(het$gram)), 3),
  plug_in_trace = round(sum(diag(plug$gram)), 3),
  mode1_agreement = round(abs(sum(het$loadings[, 1L] * plug$loadings[, 1L])), 4)
)
stopifnot(
  sum(diag(plug$gram)) > sum(diag(het$gram)),
  abs(sum(het$loadings[, 1L] * plug$loadings[, 1L])) > 0.99,
  identical(names(which.max(abs(plug$loadings[, 1L]))), "s06")
)
```

Section 6.3 makes the split normative: **loading directions may be read from
either Gram provided the source is named; any eigenvalue, spectrum,
variance-explained figure or effective mode count must come from the
cross-fitted Gram.** The package enforces it rather than advising it — the
plug-in record carries no latent layer, and says why.

```{r plug-in-refusal}
plug_refusal <- plug$receipt$latent_refusal
c(capability = plug_refusal$capability, reasons = plug_refusal$reasons)
stopifnot(
  is.null(plug$latent),
  identical(plug_refusal$capability, "plug_in_spectrum_functionals"),
  "plug_in_trace_inflated_62_7_percent_on_the_contract_fixture" %in%
    plug_refusal$reasons
)
```

## 5. Two error bars, kept apart — and a count that is neither

### 5.1 The between-subject layer

The scatter of the participants about the group fit. For the intercept-only
model this has a closed form nobody needs a package for, which is exactly why it
is worth checking against one:

```{r between}
uncertainty <- population_uncertainty(fit)
between <- uncertainty$between

hand_estimate <- apply(fit$values, c("node", "query"), mean)
hand_se <- apply(fit$values, c("node", "query"),
  function(y) stats::sd(y) / sqrt(length(y)))
c(
  residual_df = between$residual_df,
  estimate_gap = max(abs(between$estimate[, , "(Intercept)"] - hand_estimate)),
  se_gap = max(abs(between$se[, , "(Intercept)"] - hand_se))
)
stopifnot(
  identical(between$residual_df, 5L),
  max(abs(between$estimate[, , "(Intercept)"] - hand_estimate)) < 1e-12,
  max(abs(between$se[, , "(Intercept)"] - hand_se)) < 1e-12
)
```

**The `t` is labelled uncalibrated, and the label does not move.** The
arithmetic has been checked against a 2 000-replication null simulation and it
is right: under a *correctly specified* group model the nominal 95 % interval
covered the null term in 0.9485 of replications at `N = 6` and 0.9520 at
`N = 12`. The arithmetic was never the part in doubt. The second arm of the same
simulation is why the label stays: when each participant's transported value
carries a variance that depends on the group covariates — which is what a
transport whose quality varies with age or motion produces — coverage falls to
0.9230 at `N = 6` and **0.8850 at `N = 24`**. It gets *worse* with more
participants, because the bias is in the standard error and not in the sample
size.

```{r calibration}
between$calibration
c(level = between$level, t_max = round(max(abs(between$t), na.rm = TRUE), 3))
stopifnot(identical(between$calibration, "uncalibrated"))
```

Report the statistic; do not report a p-value derived from it without an
argument that the section 7.5 transport diagnostics are benign.

### 5.2 The within-subject layer, and where it is refused

The second bar is the sampling variance of one participant's transported value.
It exists **only where it is exact**. A transported value is a fixed linear
functional `w'z` of that participant's native query values, so its variance
needs the covariance *between* native nodes — an object the package refuses
(capability `cross_node_sampling_covariance`). The layer is therefore admitted
exactly where `w` has one nonzero entry: there the cross terms carry weight zero
and `Var = w²Var(z)` is exact, with **no independence assumption anywhere**.

That needs participants with an error channel, so this is a separate small
fixture arriving through the ingestion route. Group nodes sit on the native grid
itself, so a four-node participant maps one-to-one; the five-node participant
sends two native rows into the last group node.

```{r within-fixture}
wu_subject <- function(id, features, seed) {
  set.seed(seed)
  domain <- abstract_domain(features,
    coordinates = cbind(x = seq_len(features) - 1),
    feature_ids = paste0("f", seq_len(features)), id = id)
  runs <- c("run-1", "run-2", "run-3")
  scans <- function(run) paste0(run, "-scan-", 1:6)
  design <- cbind(face = c(1, 0, 0, 1, 0, 0), house = c(0, 1, 0, 0, 1, 0),
    tool = c(0, 0, 1, 0, 0, 1))
  target <- diag(3)
  dimnames(target) <- list(c("face", "house", "tool"), colnames(design))
  relation_fit <- estimate_relation(plan_relation(
    study(observations(
      stats::setNames(lapply(runs, function(run)
        matrix(stats::rnorm(6 * features), 6L, features)), runs),
      stats::setNames(lapply(runs, function(run)
        observation_index(scans(run), run)), runs),
      domain)),
    raw_design_model(stats::setNames(lapply(runs, function(run) {
      value <- design; rownames(value) <- scans(run); value
    }), runs)),
    raw_effect_map(target),
    observation_model("ols", sampling_unit = "scan")
  ))
  list(fit = relation_fit, features = features,
    plan = plan_geometry(relation_fit$relation,
      compile_frame(voxelwise(), domain),
      cross_partitions(relation_fit$relation, independence = "independent")))
}

built <- list(u01 = wu_subject("u01", 4L, 41L),
  u02 = wu_subject("u02", 4L, 52L), u03 = wu_subject("u03", 5L, 63L))
grid_carrier <- function(features) anatomical_transport(
  native_coords = cbind(seq_len(features) - 1), group_coords = cbind(0:3),
  semantics = "budget", native_index = paste0("f", seq_len(features)))

wplan <- plan_population(lapply(built, `[[`, "plan"),
  lapply(built, function(value) grid_carrier(value$features)))
wbank <- rbind(`face-house` = c(1, -1, 0), `face-tool` = c(1, 0, -1))
wfit <- estimate_population(wplan, wbank,
  uncertainty = lapply(built, function(value)
    rdm_sampling_covariance(value$plan, value$fit, target = "null",
      at = seq_len(value$features))))
within <- population_uncertainty(wfit)$within
```

```{r within}
within$admitted
c(scope = within$scope, assumption = within$assumption)
c(admitted_columns = within$admitted_columns)
stopifnot(
  identical(within$scope, "transported_single_source_column"),
  identical(within$assumption, "none: the cross-node terms carry weight zero"),
  # One-to-one for the four-node participants at every group node; `u03`'s
  # last group node collects two native rows and is therefore absent.
  identical(unname(within$admitted[, "u01"]), c(rep(TRUE, 4L), FALSE)),
  identical(unname(within$admitted[, "u03"]), c(rep(TRUE, 3L), FALSE, FALSE)),
  identical(within$admitted_columns, 11L),
  all(is.na(within$variance["<sink>", , ])),
  all(is.na(within$variance["group4", , "u03"])),
  "transported_value_mixes_native_nodes" %in% within$refusal$reasons
)
```

Where the column mixes native rows there is an **absence**, never a diagonal
approximation standing in for the missing cross terms. Overlapping supports
under spatially correlated noise are positively correlated, so a diagonal sum
would under-estimate the variance in a known direction — worse than nothing.

The six-participant population of sections 1 to 4 admits nothing at all: every
group node there collects several native nodes. **That is the refusal being
visible, not the layer being broken**, and it is the ordinary situation for a
hard anatomical parcellation.

### 5.3 The two are never pooled

```{r separation}
uncertainty$layers
uncertainty$separation
stopifnot(
  identical(uncertainty$layers, c("between_subject", "within_subject")),
  # There is no field holding their sum, and `as.data.frame()` emits one
  # layer at a time.
  is.null(uncertainty$total),
  !any(c("total", "pooled", "combined") %in% names(uncertainty)),
  identical(
    sort(unique(as.data.frame(uncertainty, layer = "between")$layer)),
    "between_subject")
)
```

They answer different questions and are not summands. The between-subject
residual already contains whatever measurement error survived into each
participant's transported value, so adding the within-subject variance would
double-count the shared part *and* still miss the covariance a
variance-components model would need.

### 5.4 Prevalence is descriptive, and says so

`population_prevalence()` counts how many participants stand behind a group
value. It is not the third error bar; it is on the **latent layer**, the same
place `latent_geometry()` confines fractions to, and it carries no interval.

```{r prevalence}
before <- fit$uncertainty
prevalence <- population_prevalence(fit, coverage_floor = length(subjects))
round(prevalence$sign$fraction, 3)
c(reference = prevalence$reference)
c(layer = prevalence$layer, reading = prevalence$reading)
# Nothing on the record may look like an error bar, at any depth.
named <- local({
  flatten <- function(x) {
    if (!is.list(x)) return(character())
    c(names(x), unlist(lapply(x, flatten), use.names = FALSE))
  }
  flatten(prevalence[c("sign", "alignment", "coverage")])
})
stopifnot(
  identical(prevalence$layer, "latent_descriptive"),
  identical(prevalence$reading, "latent descriptive layer; not for inference"),
  identical(prevalence$reference, 0.5),
  identical(intersect(named, c("se", "t", "lower", "upper", "p_value", "p",
    "statistic", "level", "confidence")), character(0)),
  identical(prevalence$receipt$prevalence$inference, "none_derivable"),
  # And constructing one does not reach into the inferential layer.
  identical(before, fit$uncertainty)
)
```

**A group node and query at which nothing reproduces reports a fraction near
`0.5`, not near `0`.** Thresholding a signed crossvalidated estimate discards
its magnitude and keeps its sign, and the sign is the noisy part; every
participant contributes an independent coin flip. `$reference` carries that
number so the comparison is not made against zero out of habit.

Coverage is reported beside it, because a high prevalence at a node only a few
participants reached is a different object from a high prevalence at a node they
all reached:

```{r coverage}
prevalence$coverage$minimum
prevalence$coverage$below_floor
stopifnot(
  identical(prevalence$coverage$floor, 6L),
  identical(prevalence$coverage$below_floor, "group3"),
  prevalence$coverage$minimum[["group3"]] < length(subjects)
)
```

`group3` sits at `x = 11`, out of radius of every native node the smallest
participant has. Five participants stand behind it, not six — a fact about the
transport, available before any of its numbers are interpreted.

## 6. What this article does not give you

**No calibrated group inference.** Section 5.1 is the whole story: the standard
error is the group OLS's own and the `t` is arithmetically correct, but the
recorded null simulation shows the interval losing coverage in exactly the
regime a real population fit occupies — transport quality varying with the group
covariates. Nothing in this package converts that `t` into a defensible
p-value, and `population_prevalence()` is descriptive rather than an alternative
route to one.

**No transport learning, and no registration.** §9.2 is a list of four things
the package refuses and requires as typed input: image registration of any kind;
functional-transport learning (a `P^F` bearing cross-fit provenance is
*accepted and evaluated*, never fitted); resampling or interpolation of subject
images — transport acts on nodes, not voxels; and any group frame over group
features, and therefore any group-node coherent component. That last one is why
the transported components take their own names — `native_coherent_ledger` is
native-node coherent evidence carried to a group node, not a group-node common
mode, and reading it as one is the error the name exists to prevent.

```{r ledger-names}
c(total = fit$ledger,
  coherent = estimate_population(plan, bank, component = "coherent")$ledger)
stopifnot(identical(fit$ledger, "transported_total"))
```

**No cross-node sampling covariance**, and therefore no error bar on a conserved
budget. Node estimates of an overlapping frame are strongly positively
correlated; variances do not add. Section 5.2's carve-out is the whole of what
exists.

**No precision weighting.** `normalization = "precision_weighted"` is in the
closed set and refused at plan construction, because the per-subject budget
variance it would need is exactly the missing object above.

```{r precision-refusal}
weighted <- catch_refusal(plan_population(subjects, transports,
  normalization = "precision_weighted"))
weighted$capability
stopifnot(
  identical(weighted$capability, "precision_weighted_normalization"),
  "per_subject_budget_variance_unavailable" %in% weighted$reasons
)
```

## See also

- [`design/population-form-contract.md`](https://github.com/bbuchsbaum/crossform/blob/main/design/population-form-contract.md)
  — the normative document. §1 the transport object, §2 the budget certificate,
  §3 the commutation claim and the pinned evaluation order, §4 the
  normalizations, §5–6 the subject Gram and the measured plug-in bias, §6.5 the
  latent layer, §7 transport diagnostics and `η_transport`, §9.2 the four
  refusals of section 6 above, §11 every tolerance asserted here.
- [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001),
  script `10-population-slice1.R` and **§10 of its README** — this article's
  identities on six real participants at region-level nodes, with committed
  receipts. Read its note on what an exact zero would mean: the
  executor-plumbing check of section 2 is copied from there.
- [`exemplars/population-slice2/DECISION.md`](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/population-slice2)
  — the hard half, and the record of why OpenNeuro `ds003745` was chosen for it:
  searchlight-level transport in a common normalized space, with real sink mass
  from partial coverage and a functionally-informed transport cross-fitted from
  independent data. Slice 1 is the case where transport is trivial; slice 2 is
  the case this contract was written for.
- `vignette("conservative-frames")` — the attribution instrument every
  participant's ledger comes from, and why summing a detection map is the error
  this layer is built to avoid.
- `vignette("interpreting-results")` — the interpretive traps that apply to a
  single participant's map and apply again, unchanged, to a group one.
