---
title: "Predict model-supported geometry on independent runs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict model-supported geometry on independent runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(crossform)
```

You have condition-by-feature effects from several runs and a model of the
conditions' representational relationships. You want to learn which
model-supported dimensions predict neural geometry, then evaluate that
prediction on independent runs. `fit_geometry()` learns a low-rank PSD form;
`score_geometry()` measures its signed predictive gain over a zero-geometry
prediction. A fitted dimension can help or harm prediction.

The [introduction](introduction.html) explains relations, frames and geometry
plans. Here a **form** is a condition-by-condition inner-product matrix. A
positive-semidefinite (PSD) form assigns nonnegative energy to every contrast;
it describes the fitted prediction. The signed crossvalidated geometry used
to fit and score it can have negative eigenvalues.

This example uses four conditions, two neural features and four runs. Its
values are deterministic so that the score can be checked exactly. In an
analysis, the matrices would contain your independently estimated effects.
The first pair of runs trains the prediction; the second pair evaluates it.

## Declare the model and the observations

The model distinguishes face from body and house from tool. These two
contrasts form centered, orthonormal directions: each has length one and
they are mutually orthogonal. Their model strengths are 0.6 and 0.4.
The kernel `K` records inner products among the four conditions.

```{r model-input}
condition_names <- c("face", "body", "house", "tool")
V <- cbind(c(1, -1, 0, 0), c(0, 0, 1, -1)) / sqrt(2)
rownames(V) <- condition_names
model_values <- c(.6, .4)
K <- V %*% diag(model_values) %*% t(V)
```

Choose penalty 0.1 before looking at the evaluation runs. The training
effects below give fitted amplitudes 2 and 1 after this penalty. Only the
first dimension is present in the evaluation effects.

```{r observations}
penalty <- .1
B_train <- V %*% diag(sqrt(c(2, 1) + penalty / model_values))
B_test <- V %*% diag(sqrt(c(2, 0)))
domain <- abstract_domain(2, id = "predictive-guide")
```

An origin manifest identifies the underlying observations. It records your
sampling assumption; equal numerical values do not determine whether
observations are independent. Use a shared manifest for training and
evaluation so that copied or renamed observations retain their identity.

```{r declared-origins}
origins <- list(id = "four-independent-runs",
  partitions = list(train1 = "raw-run1", train2 = "raw-run2",
    test1 = "raw-run3", test2 = "raw-run4"),
  independence = "independent",
  assumption = "Independent runs; preprocessing and effect extraction were fixed externally.")
rel <- relation(list(train1 = B_train, train2 = B_train,
  test1 = B_test, test2 = B_test), domain = domain,
  provenance = list(observation_origins = origins))
```

The frame sums over the two features. Its normalization sets the geometry
scale and therefore matters when choosing the penalty.

```{r independent-plans}
at <- compile_frame(whole_brain(normalization = "none"), domain)
train_plan <- plan_geometry(rel, at, pairing("train1", "train2",
  independence = "independent", generalizes_over = "run"))
test_plan <- plan_geometry(rel, at, pairing("test1", "test2",
  independence = "independent", generalizes_over = "run"))
```

## Fit once, then score independently

`model_basis()` retains the model's directions and their relative strengths.
Trace normalization is applied after any declared model-rank truncation.
With several models, supply named nonnegative weights summing to one when
fitting. Overlapping model directions are allowed.

```{r fit-and-score}
basis <- model_basis(kernels = list(model = K),
  conditions = rel$effect_space, normalize = "trace")
fit <- fit_geometry(train_plan, basis, rank = 2, penalty = penalty)
evidence <- score_geometry(fit, test_plan, modes = TRUE)
as.data.frame(evidence)
```

The gain is **3 squared geometry units**. For a frozen prediction $F$ and
independent signed test geometry $G_{\mathrm{test}}$, it is

\[
\Delta = 2\langle F,G_{\mathrm{test}}\rangle_F-\|F\|_F^2.
\]

Conditional on training and under the declared independence assumptions,
its expectation is the reduction in squared geometry error relative to
zero. It can be negative. It is not an explained fraction, and the score
does not estimate generic standard errors for fitted eigenvalues or ranks.

Under a zero signal, the expected gain of a nonzero frozen prediction is
$-\|F\|_F^2$: predicting structure that is absent incurs a cost. The signed
test inner product is centered on zero in that case; predictive gain is not.
This distinction matters when comparing gain with a cross-fitted energy.

```{r gain-check, include=FALSE}
stopifnot(is.finite(evidence$table$gain), abs(evidence$table$gain - 3) < 1e-10)
```

## Read the evidence for each fitted dimension

```{r mode-evidence}
modes <- as.data.frame(evidence, view = "modes")
knitr::kable(modes, digits = 6)
```

The first mode has fitted amplitude 2 and signed test evidence 2. Its
predictive gain is $2(2)(2)-2^2=4$. The second mode has amplitude 1 and test
evidence zero, so its gain is $-1$. Together they give 3.

Training amplitude, test evidence and predictive benefit answer different
questions. Scoring keeps the frozen prediction, including modes that harm
prediction. Discarding a mode after inspecting this table would use the
evaluation data to select a new model.

```{r mode-check, include=FALSE}
stopifnot(max(abs(modes$amplitude - c(2, 1))) < 1e-10,
  max(abs(modes$evidence - c(2, 0))) < 1e-10,
  max(abs(modes$gain - c(4, -1))) < 1e-10,
  abs(sum(modes$gain) - evidence$table$gain) < 1e-10)
```

Use `view = "groups"` for invariant evidence across a fully retained tied
eigenspace. Individual eigenvectors within such a group have arbitrary
orientation. A rank that cuts through a positive tie refuses rather than
choosing an arbitrary predictor.

The rank-zero predictor is a useful exact baseline:

```{r zero-prediction}
zero_fit <- fit_geometry(train_plan, basis, rank = 0, penalty = penalty)
as.data.frame(score_geometry(zero_fit, test_plan))
```

```{r zero-check, include=FALSE}
stopifnot(score_geometry(zero_fit, test_plan)$table$gain == 0)
```

## Choose what the prediction is allowed to learn

There are two rank choices and one penalty:

| Choice | What it controls |
|---|---|
| `model_basis(rank = ...)` | Dimensions retained from each supplied model |
| `fit_geometry(rank = ...)` | Maximum dimensions in the learned neural prediction |
| `fit_geometry(penalty = ...)` | How strongly weak model directions are discouraged |

At zero penalty, only the model span matters. If that span covers the whole
centered condition space, the model imposes no directional preference and the
fit is labelled a generic baseline. A positive penalty preserves preferences
from the model eigenvalues, even when the span is full.

Choose model ranks, weights, response rank and penalty before final
evaluation, or select them using inner validation contained entirely within
the training data. Four partitions support this fixed two-run/two-run
example; they do not provide an arbitrary nested tuning design. The public
verbs shown here fit and score a declared choice, without automatic tuning.

The estimator minimizes Frobenius error on effect geometry with an
inverse-kernel trace penalty. It compresses the signed training form, applies
the penalty, then retains the largest positive eigenvalues within the rank
budget. It never first clips the neural geometry. This objective differs
from `rsa()`'s fixed linear regression on RDM entries and from the descriptive
trace energies returned by `model_geometry()`.

### How the model controls the fit

More precisely, let $K_\alpha = \sum_i \alpha_i K_{i,r_i}$ be the weighted
sum of the retained, normalized model kernels. In its positive eigenspace,
$K_\alpha = Q D Q^\top$, the fitted form is

\[
\widehat F = Q\left[Q^\top G_{\mathrm{train}}Q
  - \lambda D^{-1}\right]_{+,s}Q^\top.
\]

Here $[\cdot]_{+,s}$ retains at most the $s$ largest positive eigenvalues.
The response rank is $s$ and `penalty` is $\lambda$.
The model directions can overlap: their strengths combine in $K_\alpha$.
The package transforms each partition's effects into reduced coordinates
before accumulating geometry, so fitting does not require a complete
condition-by-condition neural form. Compressed residual diagnostics describe
that reduced problem; they do not identify a whole-geometry explained fraction.

### Starting with squared distances

If you start with an RDM, declare that its entries are **squared** Euclidean
distances. The same model in this example can be supplied as follows:

```{r squared-distance-input}
squared_distances <- outer(diag(K), diag(K), "+") - 2 * K
distance_basis <- model_basis(models = list(model = squared_distances),
  distance = "squared_euclidean", conditions = rel$effect_space,
  normalize = "trace")
distance_fit <- fit_geometry(train_plan, distance_basis, rank = 2,
  penalty = penalty)
as.data.frame(score_geometry(distance_fit, test_plan))
```

```{r input-parity-check, include=FALSE}
stopifnot(abs(score_geometry(distance_fit, test_plan)$table$gain -
  evidence$table$gain) < 1e-10)
```

This recovers the same gain. Ordinary Euclidean distances must be squared
before this conversion; arbitrary dissimilarities need not define an admitted
PSD kernel.

## Keep the evaluation target explicit

Training and test plans must describe the same conditions, effect meanings,
spatial measurements, normalization, fixed neural metric and generalization
axis. Stable measurement IDs can appear in a different row order. Used
observation origins and upstream training dependencies must be disjoint.
There are two independence requirements: the effects within each
cross-partition product must be independent, and the evaluation observations
must be independent of everything used to learn or select the prediction.
Scoring on the training observations refuses:

```{r overlap-refusal}
overlap <- catch_refusal(score_geometry(fit, train_plan))
overlap$reasons
```

```{r overlap-check, include=FALSE}
stopifnot("overlapping_training_evaluation_origins" %in% overlap$reasons)
```

These scores evaluate independent runs on the **same conditions**.
Prediction on new conditions needs a training-defined model transformation
and a cross-kernel; it is outside this interface. Generic RDM/GLS losses and
learned neural metric schedules are also outside the current spectral path.

## Read the spatial components with the same prediction

For a total-geometry prediction, `components = TRUE` retains coherent and
configuration inner products read with the same frozen form. Their evidence
adds before subtracting the prediction cost once. Separately fitting each
component does not conserve the original signed decomposition.

```{r component-evidence}
component_score <- score_geometry(fit, test_plan, components = TRUE)
component_score$components
```

```{r component-check, include=FALSE}
parts <- component_score$components
stopifnot(abs(parts$coherent_inner_product - 2) < 1e-10,
  abs(parts$configuration_inner_product - 2) < 1e-10,
  abs(component_score$table$prediction_norm_sq - 5) < 1e-10,
  abs(2 * (parts$coherent_inner_product +
  parts$configuration_inner_product) -
  component_score$table$prediction_norm_sq - component_score$table$gain) < 1e-10,
  abs(component_score$table$gain - evidence$table$gain) < 1e-10)
```

In this two-feature example, the coherent and configuration inner products
are each 2. Twice their sum is 8; subtracting the fitted squared norm of 5
once gives the same gain of 3. These component inner products describe how
the frozen prediction reads the two spatial forms. They are not two
independently fitted predictive scores.

## Larger analyses and next steps

For many measurements, use a declared workspace budget and, with the
implicit identity neural metric, `storage = "block"`. Supply a new directory
through `storage_path`; completed records reopen with `readRDS()` while that
directory stays in place. A fixed SPD neural metric currently uses memory
storage. See `fit_geometry()` and `score_geometry()` for retained fields and
storage controls. The workspace budget covers planned numerical buffers,
not a hard limit on the R process's resident memory. This path currently
uses one worker.

For the signed energies and fixed RSA coefficients, see
[Reading results](interpreting-results.html)
(`vignette("interpreting-results")` offline). For a descriptive rank budget
without independent prediction, see the latent-layer example in
[Conservative frames](conservative-frames.html)
(`vignette("conservative-frames")`). The repository's
[standalone example](https://github.com/bbuchsbaum/crossform/blob/main/exemplars/predictive-geometry/fixed-split.R)
also checks equivalent kernel, feature and squared-Euclidean-RDM inputs.
