---
title: "Coming from rMVPA"
description: >
  A concept map from rMVPA's dataset/design/model vocabulary to crossform's relation/frame/plan vocabulary, with an honest list of what crossform does not do.
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(crossform)
```

> **This is a mapping guide, not a benchmark.** Nothing here compares the two
> packages on speed, accuracy, or scientific quality. It answers one question:
> if you already think in [rMVPA](https://github.com/bbuchsbaum/rMVPA)'s
> vocabulary, what is the corresponding `crossform` object, and what is
> deliberately missing? The one place the two packages have been measured
> against each other is a numerical-parity check, reported at the end with its
> caveats.
>
> The rMVPA calls below are shown but not executed — rMVPA is not on CRAN, so
> this article cannot depend on it. The `crossform` calls are executed.

## The one structural difference

rMVPA is organized around a **model** that is *run over* a set of locations.
You build a dataset, a design, and a model specification, then hand the whole
thing to `run_searchlight()` or `run_regional()`, which returns a performance
map.

`crossform` is organized around a **plan** that is *queried*. You declare a
fitted relation, where to measure it, and what must generalize; that produces
a plan. Contrasts, RDMs, and RSA are then different reads of the same plan,
and they do not refit anything:

```r
# rMVPA: one run per question
run_searchlight(model_a, radius = 8)
run_searchlight(model_b, radius = 8)

# crossform: one plan, many questions
plan <- plan_geometry(rel, at = frame, over = pairing)
contrast_energy(plan, weights)
rdm(plan)
rsa(plan, models = list(category = model_rdm))
```

The practical consequence is where cross-validation lives. In rMVPA the
blocking variable is consumed by a cross-validation scheme that folds a
*model*. In `crossform` the generalization axis is bound into the *estimand's
identity* by `cross_partitions()`, before any data is read, and it travels
with the plan.

## Concept map

| rMVPA | crossform | Notes |
|---|---|---|
| `mvpa_dataset(train_data, mask =)` | `neuroim2_volume_domain(mask)` + `relation(list_of_run_matrices, ...)` | rMVPA holds the image; `crossform` splits *where* (domain) from *what was estimated* (relation). |
| `mvpa_design(df, y_train = ~y, block_var = ~block)` | `effect_space(condition_names)` for the conditions; run structure is the names of the list passed to `relation()` | `crossform` has no trial-level design object at this stage; conditions are already estimated. |
| `blocked_cross_validation(block_var)` | `cross_partitions(rel, independence = "independent", generalizes_over = "run")` | rMVPA folds train/test; `crossform` forms products between *different* partitions. Not the same object: see below. |
| `run_searchlight(spec, radius = 8)` | `compile_frame(searchlights(radius = 8), domain)` or `neuroim2_searchlights(mask, radius = 8)`, then `plan_geometry()` | Radius is millimeters in both. |
| `run_regional(spec, region_mask)` | `compile_frame(regions(labels), domain)` | `regions()` takes one label per **feature** (a vector), not a `NeuroVol`. |
| `rsa_design(~ model, data = list(model = D), block_var =)` | the `models =` list of `rsa()`, plus `cross_partitions()` for the blocking | `crossform` aligns model RDMs by condition **name**. |
| `rsa_model(dset, des, distmethod =, regtype =)` | `rsa(plan, models = list(...))` | See "distance conventions" below — these are not the same estimator. |
| `vector_rsa_model(dset, des, distfun = cordist())` | no equivalent | Trial-level; `crossform` aggregates to condition level. |
| `compute_crossvalidated_means_sl(..., estimation_method = "crossnobis")` + `compute_crossnobis_distances_sl()` | `rdm(plan)` under `cross_partitions()`, or `plan_crossnobis()` / `crossnobis()` | This is the pairing where the two packages have been checked against each other. |
| `mvpa_model(model = "sda_notune", ...)`, classification accuracy | no equivalent | `crossform` has no classifier. |
| train/test folds | cross-partition pairing | A fold count is an execution detail in `crossform`; the *generalization axis* is part of plan identity. |
| performance metrics (`compute_performance`) | the views: `contrast_energy()`, `rdm()`, `rsa()` | Energies and signed squared distances, not accuracies. |
| `feature_selector("FTest", "top_k", 100)` | no equivalent | See absences below. |
| `run_permutation_searchlight(...)`, `permutation_control()` | no equivalent | See absences below. |
| `future::plan(multisession)` (rMVPA uses furrr/future.apply) | `compute_policy()`, sequential only | `workers > 1` is refused, not silently ignored. |
| `save_results(x, dir, ...)` | `as_neurovol()` + your own writer | `crossform` writes no files. |

### The cross-validation mapping deserves a second look

`blocked_cross_validation(block_var)` and
`cross_partitions(rel, generalizes_over = "run")` both encode "runs are the
independent unit", but they are not interchangeable objects.

A leave-one-run-out fold scheme produces training means that **overlap across
folds**. Cross-partition pairing forms products between two *disjoint* run
estimates, so the noise in the two factors is independent and cancels in
expectation. That is exactly why `crossform`'s estimates are unbiased and may
be negative. The Haxby exemplar records a concrete instance of the difference
mattering: an older rMVPA build derived crossnobis folds from leave-one-run-out
training means, whose overlap made the distances positively biased, and the
exemplar script asserts fold semantics rather than trusting a version number.

## Side-by-side minimal example

The same question — where does an animate-versus-inanimate pattern reproduce
across runs? — written both ways.

### rMVPA

```{r rmvpa-code, eval=FALSE}
library(rMVPA)

dset <- mvpa_dataset(train_data = beta_vec, mask = vt_mask)

des <- mvpa_design(
  data.frame(
    y = factor(rep(conditions, times = n_run), levels = conditions),
    block = run_of
  ),
  y_train = ~ y,
  block_var = ~ block
)

rsa_des <- rsa_design(~ animacy, data = list(animacy = model_rdm),
                      block_var = run_of)
mspec <- rsa_model(dset, rsa_des, distmethod = "pearson", regtype = "pearson")
res <- run_searchlight(mspec, radius = 6, method = "standard")

# res$results$animacy is a NeuroVol of second-order correlations.
```

### crossform

```{r crossform-code}
example <- example_fmri_effects()

plan <- plan_geometry(
  example$fit$relation,
  at = example$frame,
  over = cross_partitions(
    example$fit$relation,
    independence = "independent",
    generalizes_over = "run"
  )
)

effect <- contrast_energy(plan, example$contrast)
category <- rsa(plan, models = list(category = example$model_rdm))

c(
  measurements = length(effect$total),
  peak = which.max(effect$total),
  rsa_columns = ncol(category$coefficients)
)
```

Two things differ beyond syntax. `plan` is reusable: `rdm(plan)` and any
number of further contrasts cost no refit. And `rsa()` returns an OLS
coefficient with an intercept column by default, not a rank correlation —
`ncol()` above is 2 for one model because the intercept is retained.

```{r crossform-reuse}
distances <- rdm(plan)
dim(distances$values)
colnames(category$coefficients)
```

## Distance and second-order conventions are not the same

This is the most likely source of surprise, and it is not a bug on either
side.

- `crossform`'s `rdm()` reports signed **crossvalidated squared Euclidean**
  distances, `d_ij = G_ii + G_jj - 2 G_ij` (squared Mahalanobis under a fixed
  metric). They can be negative, and are not truncated.
- rMVPA's `rsa_model()` takes `distmethod` in `{"pearson", "spearman"}`, i.e.
  a **correlation distance**, and scores it with a correlation.
- `crossform`'s `rsa()` is **OLS**, not a rank correlation.

`crossform` cannot express correlation distance at all: dividing each pattern
by its own norm is not a fixed linear operation on the patterns, so it is
outside the bilinear core. Per-sphere mean-centering *is* expressible as a
metric; the unit-norm denominator is not. The boundary and the conditions
under which it could be crossed are recorded in the
[correlation-distance policy](correlation-distance-policy.html).

If you want the two packages to compute the same thing, the estimand both can
express is the condition-level crossvalidated squared Euclidean (crossnobis)
distance — which is the pairing used for the parity check below.

## What crossform does not do (yet)

Each absence below was checked against the package's actual exports rather
than recalled.

- **Classification and decoding.** There is no classifier, no accuracy, no AUC,
  no confusion matrix, no `MVPAModels` registry. `crossform` computes
  second-order geometry only.
- **Group inference.** No second-level model, no random-effects map, no
  across-subject test. `rdm_sampling_covariance()` is a *within*-measurement,
  within-participant covariance law and says so.
- **Feature selection.** No `FTest`/`catscore` ranking, no `top_k` cutoff. The
  frame decides which features each measurement reads, and it is declared in
  advance.
- **Permutation and bootstrap.** No label shuffling, no null distributions, no
  FDR correction.
- **Parallel backends.** Execution is sequential. This is refused rather than
  silently downgraded:

```{r no-workers, error=TRUE}
compute_policy(workers = 4)
```

- **Preprocessing and I/O.** No registration, no smoothing, no file reading or
  writing. `as_neurovol()` hands you a `NeuroVol`; saving it is yours.

```{r absence-check, include=FALSE}
exports <- getNamespaceExports("crossform")
stopifnot(
  !any(grepl("classif|decod|accuracy|permut|bootstrap|feature_select",
             exports, ignore.case = TRUE)),
  inherits(try(compute_policy(workers = 2), silent = TRUE), "try-error")
)
```

Two things rMVPA users may expect that `crossform` *does* have, in a different
shape: a crossnobis path (`plan_crossnobis()`, `crossnobis()`, with
`shrinkage_precision()` and `metric_training_policy()` for the whitening), and
explicit refusals when an uncertainty claim is not supported by the fit
(`sampling_capabilities()`, `catch_refusal()`).

## The one measured comparison

The [Haxby 2001 exemplar](https://github.com/bbuchsbaum/crossform/tree/main/exemplars/haxby2001)
runs both packages on 12 conditions and 577 ventral-temporal searchlights at
an 11.25 mm radius. On the matched estimand — `crossform`'s `rdm()` over a
cross-run geometry against rMVPA's
`compute_crossvalidated_means_sl(estimation_method = "crossnobis")` followed by
`compute_crossnobis_distances_sl()` — the two agree to a **maximum absolute
difference of 8.88e-16** at every one of the 577 centers. Sphere membership
was verified identical rather than assumed.

Read that number with the exemplar's own caveats, which are recorded there in
full:

1. **It is not `rsa_model`.** The exemplar originally named `rsa_model` for
   the rMVPA arm and could not use it: it never aggregates to condition level.
   `vector_rsa_model` is trial-level, and `contrast_rsa_model` computes the
   right cross-validated Gram matrix internally but exposes only contrast
   betas. The crossnobis helper is the only route, and
   `compute_crossnobis_distances_sl()` is not exported, so the script calls it
   with `:::`.
2. **It is version-sensitive.** The comparison requires an rMVPA build whose
   crossnobis fold estimates are per-run condition means. The script asserts
   that semantics and stops with an explanatory error otherwise, rather than
   checking a version string.
3. **The originally specified estimand was not achievable.** A
   correlation-distance RDM scored by Spearman cannot be expressed by
   `crossform` at all, and rMVPA's `rsa_model` couples the distance and the
   second-order statistic through one `distmethod` argument. The matched
   comparison is a substitution, and the exemplar says so.

What the number demonstrates is numerical parity on a shared estimand. It
demonstrates nothing about speed, and nothing about correlation-distance RSA.

## Where to go next

- `vignette("introduction", package = "crossform")` — what the energies, RDMs,
  and RSA coefficients mean.
- `vignette("neuroim2-data", package = "crossform")` — the input and output
  adapters, if your data is already in neuroim2 objects.
- [Reading contrast energies, RDMs, and uncertainty](interpreting-results.html)
  — how to read the numbers, and the three interpretive traps.
- [What is novel in crossform?](novelty.html) — what is architecture and what
  is established statistics.
